### 2008-08-07 doc/papers/cgo08/paper.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@243 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

\documentclass[11pt]{article}

\usepackage{url}
\usepackage{html}
\usepackage{setspace}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{cite}
\usepackage[margin=1in]{geometry}

\begin{document}

\title{Generating Empirically Optimized Numerical Software from Matlab Prototypes}

\abstract{}

\section{Introduction}
\label{sec:intro}

%Words here in Intro and Back are lifted from a 2004 proposal by Liz, Bill, and Jack, written almost completely by Liz.   Perhaps dated but
% at least a jumping off point.

The growing demand for higher levels of detail and accuracy in results means that the size and complexity of scientific computations is increasing at least as fast as the improvements in processor technology. Programming scientific applications is hard, and optimizing them for high performance is even harder.  The development of optimized codes requires extensive knowledge, not only of the costs of floating-point arithmetic but also of  memory access issues and of the compiler being used. Hardware platforms are continually evolving. In addition, experiments show that the complexity of this  hardware-software system means that performance is difficult to predict fully.  Therefore, computational scientists are often forced to choose between investing too much time in tuning code or  accepting performance that is lower (and sometimes significantly lower) than peak.

Historically, the research community has pursued two separate paths toward the goal of making software run at near-peak levels.  The first of these builds on research into compilers and their associated technologies.  The holy grail of compilation research is to take an arbitrary code as input and produce optimal code as output for a given language and hardware platform.  Despite the immense effort that has been poured into this approach, its success has been limited both by practical constraints (users will not tolerate compile times that extend into several days) and by the amount of detailed information the compiler can obtain about the software to be compiled and the hardware on which it is to execute~\cite{Ken99,Wol89,Wol96}.

The second path has been to identify kernel routines that constitute the dominant costs of a wide variety of applications.  When such kernels can be identified and agreed upon by the members of the community, programmers with the required level of technical knowledge can concentrate on producing these optimized kernel libraries for architectures of interest. A prime example is the high-performance BLAS libraries produced by a combination of hardware vendors, independent software vendors, and researchers.  Developers who write their codes calling these routines can achieve high performance across all supported architectures. The efforts in high-performance library development aptly demonstrate, however, that the research into compiler technologies has not been fully successful.  If compilers could take the relatively simple code for basic tasks such as a dense matrix-matrix product and produce optimal code, this second approach would be unnecessary.

Like compiler optimization, the kernel library-oriented approach has significant limitations.  Support for a variety of architectures is the most serious problem because the optimizations necessary to achieve near-peak performance are by nature not portable.   This lack of portability is particularly problematic because processor designs are constantly changing,  and constant updates are nearly impossible to maintain by hand.  It is a sad fact of the computing industry that by the time highly-optimized code is available for a given architecture, that architecture is generally well on its way to obsolescence.

%This doesn't match our paper but I leave it here anyway.
We believe that addressing these barriers to performance optimization requires a paradigm shift from standard approaches, a shift that augments traditional techniques with an automated process that experimentally tests software on target architectures and modifies the code accordingly.

To begin, we are using matrix algebra operations.

\section{Background}
\label{sec:background}

To date, efforts toward automatically tuned software have focused largely on linear algebra.  These systems all have the same structure.  A collection of implementations is generated and timed, and the best choice is then identified.  One such early effort that combined automated code generation and empirical timings for automatic tuning was the PHiPAC (Portable High Performance Ansi C) \cite{lawn111} project.  PHiPAC generates parameterized code that can be tuned to machine specifications.  Basic optimizations in the code include manual loop unrolling, explicit removal of unnecessary dependencies in code blocks, and use of machine sympathetic C constructs. Search scripts are provided that, for a given code generator, find the best set of parameters for a given architecture and compiler. A follow-on to PHiPAC, the Sparsity system \cite{SPARSITY}, allows users to build sparse matrix kernels that are tuned to their matrices and machines automatically. Sparsity combines traditional techniques such as loop transformations with data structure transformations and optimization heuristics that are specific to sparse  matrices. It is based on a framework for selecting optimization parameters, such as block sizes, using a combination of performance models and search.

Should update to include OSKI.

ATLAS produces dense linear algebra kernels that have portable high efficiency across a variety of platforms. Like PHiPAC, ATLAS employs a test and search strategy. The automatic output of ATLAS is highly efficient implementations of the BLAS and of a limited set of LAPACK routines. Even in this narrow incarnation, ATLAS produces valuable results, since these subprograms are the essential computational building blocks of many scientific computing codes and of many general-purpose libraries such as LAPACK or PETSc~\cite{BGC+98}.  Moreover, ATLAS-based BLAS are immediately and directly reusable in software designed and written for distributed-memory concurrent computers, since optimized kernels are also heavily used locally by each processor in many scientific computing applications.

Another project that lies outside of the domain of linear algebra, is FFTW (Fastest FFT in the West)~\cite{frigo98, frigo99} for tuning discrete fast Fourier transforms.  FFTW uses the Cooley-Tukey algorithm and relies on explicit recursion in order to improve locality.   It also follows the test-and-search procedure of the linear algebra systems. FFTW creates sequences of instructions for the evaluation of the FFT called {\em plans}. Using a dynamic programming algorithm, it then efficiently searches through the plans to identify an optimal one.

\section{Composed Operations}
\label{sec:composed}

%These words swiped from the recent DOE proposal.

To begin, we will focus on the matrix algebra that is a foundational part of many scientific applications. Codes based on matrix algebra are typically constructed as a sequence of calls to the Basic Linear Algebra Subprograms (BLAS)~\cite{Dongarra:1988uq,Lawson:1979kx,Dongarra:1990fk} and similar sparse matrix libraries~\cite{Saad:fr,George:1981uq}. Writing programs in this way promotes readability and maintainability but can be costly in terms of memory efficiency. While highly tuned implementations of the BLAS are available~\cite{Bilmes:1997ye,Whaley:1998fk,IntelMath:oq,ESSL:kl,Goto:2006fk}, the retrieval of a large order matrix at each routine call can have a profound effect on performance. Indeed, the PIs and other researchers have observed that a sequence of BLAS calls can be substantially less efficient than a call to a single specialized routine that performs the same operations (i.e, a {\it composed} BLAS routine)~\cite{Ashby:uq,Blackford:2002vn,Stanley:CSD-98-992,baker03blgmres,baker03lgmres,Dennis:2005tg,Howell:2008,gropp01,Vuduc:2003kl}. There are several ways to construct composed BLAS. As an example, consider the pair of matrix-vector products
$q  = Ap$, $s  = A^Tr$, where $A$ is a matrix and $p$, $q$, $r$, and $s$ are vectors, that comprise the computational
bottleneck of the BiConjugate Gradient Method~\cite{Barrett:1994kx,Saad:2003fk}.  These two operations can be implemented as a pair of calls to the BLAS routine {\tt GEMV} or they can be rewritten as a single doubly nested loop encompassing both matrix-vector multiplies. In the former case, the matrix $A$ is accessed twice while, in the latter, is it accessed only a single time. Our preliminary results indicate that this loop fusion leads to a routine that delivers 90\% more megaflops than does a pair of calls to the highly optimized Goto BLAS~\cite{Goto:2006fk} {\tt GEMV} routines for large matrix orders.

Composed routines are the focus of the work presented here, but much of what we discuss generalizes to a much broader array of computations.   Extending our approach will be the subject of future work.

% While large speedups are probable, composition can  also force memory accesses to be consecutive, and it can

%lead to the need to bring more data into cache. These bad

% effects can be remedied by combining composition with other tuning techniques like cache blocking and software pipelining.

\section{Our Approach}
\label{sec:approach}

To generate highly optimized code from a MATLAB prototype, we will follow a three-step approach.    To begin, we will use a compiler that converts a MATLAB script into simple C code.

Matlab compiler -> Pluto ->  Orio

\subsection{A Matlab Compiler}
\label{sec:matlab}

Figure~\ref{fig:compiler} gives an overview of the compilation process. The MATLAB kernel specification is parsed into a high-level intermediate representation in the form of a dataflow graph.  This dataflow graph is then iteratively processed until all of the implementation choices have been made.  The compilation process consists of three phases: analysis, lowering, and optimization, which are together iterated until all of the implementation decisions have been made.  The graph is then translated into C code.

\begin{figure}[htbp]
  \centering

  \caption{Overview of the Matlab compiler.}
\label{fig:compiler}
\end{figure}

\subsubsection{The Dataflow Graph}

An example dataflow graph for the GEMVER kernel, defined below, is given in figure~\ref{fig:gemver-dataflow}.
\begin{eqnarray*}
  A &\gets A + u_1 v_1^T + u_2 v_2^T \\[-0.5ex]
  x &\gets \beta A^T y + z \\[-0.5ex]
  w &\gets \alpha A x
\end{eqnarray*}
Each node represents a parameter of the kernel or an operation.  The arrows indicate the flow of data. At first the graph specifies what operations are to be performed but does not contain any implementation details. The symbol $\times$ in the depicted graph stands variously for outer product, matrix-vector multiplication, and scalar-vector multiplication, and it does not yet specify, for example, whether the outer products compute a row or a column at a time of the result matrix.

\begin{figure}[htbp]
  \centering

\caption{Dataflow graph for the GEMVER kernel.}
  \label{fig:gemver-dataflow}
\end{figure}

\subsubsection{Analyze the Dataflow Graph}

At the beginning of the analysis phase, each input and output node of the dataflow graph is labeled with its kind (scalar, vector, matrix, or unknown) and storage format (dense row-major, dense column-major, or unknown) according to the annotation given by the user.  During analysis, the kind and storage format are computed for the intermediate nodes and algorithms are assigned to operation nodes. The choice of algorithm and the determination of kinds of storage formats are interrelated so they must be computed together.  For example, the $\times$ symbol in the graph in Figure~\ref{fig:gemver-dataflow} having the inputs $u_1$ and $v_1^T$ can be implemented by iterating over rows first, or over columns first. The choice depends on how the result is used downstream in the dataflow graph.  In this case, the result is added to the outer product of $u_2$ and $v_2^T$, so we still could choose either option as long as we make the same choice for both outer products. Going one more step downstream, there is an addition with $A$, which was annotated in Figure~\ref{fig:gemver} to be column major. At this point it is clear that the outer-products should be computed in column-major order.

We do not hard code this knowledge of basic linear algebra in the compiler algorithm itself, but instead use a data-driven approach and store information regarding how to implement basic linear algebra in a separate database, which we refer to as the \emph{linear algebra database}.  This separation allows us to add new matrix formats, operations, and basic linear algebra algorithms without changes to the
compiler algorithm.

The analysis algorithm makes implementation choices using the most-constrained first strategy (also called minimum remaining values)
from the artificial intelligence literature~\cite{Russell:2003mz}.This may need to be updated depending on how we decide to talk about the empirical search. -Jeremy 8/6/08 10:15 AM 
The compiler chooses the node with the fewest matching implementations (in the linear algebra database) and assigns a matching implementation to the node.  If there is more than one match, the prototype compiler picks the first.  Once we add backtracking we will push each alternative onto the stack to be explored later. Once the choice is made, the kind and storage format labels for the operation node itself and for the input nodes are updated with the labels specified in the linear algebra database for the chosen algorithm. The algorithm then goes on to find the next node with the fewest matching implementations and repeats.  In this phase, only the name of the algorithm is assigned to the operation node.  The details of the implementation are not specified until the refinement phase.

\subsection{Refine the Dataflow Graph}
\label{sec:refine}

The refinement phase resolves the implementation for each operation node in the graph into a subgraph defining the details of the chosen algorithm.  Each subgraph is an abstract representation of the loop that implements the given operation.  In each case, the subgraph has an iteration strategy that specifies how to traverse the elements of the matrix or vector.

Figure~\ref{fig:refine-mv-dot} shows two steps of refinement for a matrix-vector multiplication. The first step expands the matrix-vector multiplication according to the mv-dot algorithm from figure~\ref{fig:algos}.  The refinement step replaces the $\times$ node with a subgraph that computes the inner product of a row of the matrix---the node labeled $(i,:)$---with the vector, storing the result in the $i$th element of the result vector---the node labeled $(i)$. The subgraph is labeled with $i=1\!:\!m$ indicating that the iteration strategy is to traverse the rows of the dense matrix.
%
The second pass of refinement introduces another subgraph to implement the inner products (the dot algorithm from figure~\ref{fig:algos}). The new subgraph iterates through each dense row, as indicated by the $j=1:n$ annotation. The sign $\Sigma$ indicates that the sum of all of the iterations is computed.

\begin{figure*}[btp]
  \centering

  \caption{The refinement of a matrix-vector product to a set of inner products and the refinement of inner products to scalar multiplications and additions.}
  \label{fig:refine-mv-dot}
\end{figure*}

\subsubsection{Optimize the Dataflow Graph}

In the optimization step, the dataflow graph is optimized by applying conditional graph rewrite rules.  One example rewrite rule is \emph{Merge Operand Sharing Subgraphs}: if two subgraphs share a common operand but do not depend on one another, and if the iteration strategies of the two subgraphs are compatible, then merge the subgraphs into a single subgraph.  This rule is responsible for fusing the loops of the two matrix-vector products in the GEMVER kernel.  Another example rule is \emph{Merge Dependent Subgraphs}: if one subgraph depends on another subgraph, and the iteration strategies of the two subgraphs are compatible, then merge them into a single subgraph.  This rule, for example, is responsible for fusing the loops of the two outer products in the GEMVER kernel.

The choice of which optimizations to apply to various parts of the graph depends on the characteristics of the kernel, the algorithm, the matrix order and kind, and architectural features.  The following is an instructive example: $y \gets A^T A (x + z + w)$.  Assume that the compiler has already merged the two vector additions into one subgraph, call it $S_1$, and the two matrix products into another subgraph, call it $S_2$. The question then is whether it is profitable to apply the Merge Dependent Subgraphs rule one more time to merge $S_1$ and $S_2$. If the three vectors plus a row of $A$ do not fit in cache, then the merger will not be profitable because it will cause each row of $A$ to fall out of cache before the row gets used a second time.  On the other hand, if all three vectors plus a row of $A$ do it in cache, then $S_1$ and $S_2$ should be merged.  In general, we need a way to search through the many different combinations of optimization choices and choose a combination that minimizes memory traffic.

\subsubsection{Search Space Exploration and Evaluation}
\label{sec:search}

This will need updating depending on how we decide to address this topic. -Jeremy 8/6/08 10:19 AM 
The set of all possible combinations of implementation and optimization choices is called the search space~\cite{Kisuki:2000uq,Triantafyllis:2003uq, Cooper:2005kx}.  Although the dataflow graphs are quite small in our setting (usually less than 100 nodes), it is still important to prune the search space to avoid an exponential compilation time. Our prototype already performs some pruning: the optimization rewrite rules only apply under certain circumstances.  However, as mentioned above we are sometimes overly aggressive in applying optimizations. To solve this problem we plan to use a combination of analytic methods and empirical
methods~\cite{Chen:2007vn}.
%Yotov:2005kx
We plan to use a fast analytic cost function to prune out regions of the search space that will not produce competitive implementations. The cost function is based on a model of the computer architecture, but does not necessarily account for all the details.  Once the search space is narrowed, we will use empirical testing with representative data sets to determine the exact performance of the alternatives and choose the best one.

\subsubsection{Translate the Dataflow Graph to C}

A graph that cannot be further refined has been reduced to a collection of subgraphs representing loops over scalar operations. The generator outputs a C loop for each subgraph.  The order in which the loops, and scalar operations within the loops, appear in the output is determined by performing a topological sort.

\subsection{PLuTo}
\label{sec:plutonorio}
% http://www.cse.ohio-state.edu/~bondhugu/pluto/

PLuTo~\cite{uday07tr70,uday08cc} is a source-to-source transformation tool for optimizing sequences of (possibly nested) loops.

\subsection{Orio}
\label{sec:orio}

Orio~\cite{Norris:2007} is an empirical tuning tool that takes as input annotated C code and generates multiple tuned versions of the annotated code. Orio does not parse the application code; rather, it relies on annotations expressed as structured comments, which contain a simplified expression of the computation, as well as directives on what transformations to apply.

% TODO (Boyana): redo this figure

\subsection{Putting it All Together}
\label{sec:combining}

\section{Experimental Results}
\label{sec:experiments}

\section{Multicore?}
\label{sec:multicore}

\section{Conclusions}

\bibliography{paper}
\bibliographystyle{siam}

\end{document}

### 2008-08-07 testsuite/matlab/bicgkernel__3.c
Jeremy Siek's Matlab compiler-generated C code from composed operations implemented in Matlab.

@@ -1,0 +1,23 @@
+void
+BICG (float *A, int A_nrows, int A_ncols, float *p, int p_nrows, float *r,
+	  int r_nrows, float *q, int q_nrows, float *s, int s_nrows)
+{
+	for (int ii = 0; ii < q_nrows; ++ii)
+		q[ii] = 0.0;
+	for (int i = 0; i < A_ncols; i += 1) {
+		float t9 = p[i];
+		float *t8 = A + i * A_nrows;
+		float &t13 = s[i];
+		float t19 = 0;
+
+		for (int j = 0; j < A_nrows; j += 1) {
+			float t17 = r[j];
+			float t14 = t8[j];
+			float &t16 = q[j];
+
+			t19 += (t17 * t14);
+			t16 += (t9 * t14);
+		}
+		t13 = t19;
+	}
+}

### 2008-08-07 testsuite/matlab/gemver__108.c

@@ -1,0 +1,77 @@
+void
+GEMVER (float *A, int A_nrows, int A_ncols, float a, float b, float *u1,
+		int u1_nrows, float *u2, int u2_nrows, float *v1, int v1_nrows,
+		float *v2, int v2_nrows, float *y, int y_nrows, float *z, int z_nrows,
+		float *B, int B_nrows, int B_ncols, float *w, int w_nrows, float *x,
+		int x_nrows)
+{
+	float *t5 = new float[A_ncols * A_nrows];
+	float *t15 = new float[A_ncols];
+	float *t19 = new float[A_nrows];
+
+	for (int i = 0; i < A_ncols; i += 1) {
+		float t24 = v1[i];
+		float *t27 = A + i * A_nrows;
+		float *t30 = t5 + i * A_nrows;
+
+		for (int j = 0; j < A_nrows; j += 1) {
+			float t54 = u1[j];
+			float t57 = t27[j];
+			float &t60 = t30[j];
+
+			t60 = (t57 + (t54 * t24));
+		}
+	}
+	for (int i = 0; i < A_ncols; i += 1) {
+		float t31 = v2[i];
+		float *t34 = t5 + i * A_nrows;
+		float *t37 = B + i * A_nrows;
+
+		for (int j = 0; j < A_nrows; j += 1) {
+			float t61 = u2[j];
+			float t64 = t34[j];
+			float &t67 = t37[j];
+
+			t67 = (t64 + (t61 * t31));
+		}
+	}
+	for (int i = 0; i < A_ncols; i += 1) {
+		float *t38 = B + i * A_nrows;
+		float &t43 = t15[i];
+		float t70 = 0;
+
+		for (int j = 0; j < A_nrows; j += 1) {
+			float t69 = t38[j];
+			float t68 = y[j];
+
+			t70 += (t68 * t69);
+		}
+		t43 = (b * t70);
+	}
+	for (int i = 0; i < A_ncols; i += 1) {
+		float t45 = z[i];
+		float t44 = t15[i];
+		float &t47 = x[i];
+
+		t47 = (t44 + t45);
+	}
+	for (int ii = 0; ii < A_nrows; ++ii)
+		t19[ii] = 0.0;
+	for (int i = 0; i < A_ncols; i += 1) {
+		float t49 = x[i];
+		float *t48 = B + i * A_nrows;
+
+		for (int j = 0; j < A_nrows; j += 1) {
[…]

### 2008-08-07 testsuite/matlab/new/bicgkernel_row__3.c

@@ -1,0 +1,18 @@
+void BICG(float* A, int A_nrows, int A_ncols, float* p, int p_nrows, float* r, int r_nrows, float* q, int q_nrows, float* s, int s_nrows) {
+for (int ii = 0; ii < s_nrows; ++ii)
+s[ii] = 0.0;
+for (int i = 0; i < A_nrows; i+=1) {
+float t12 = r[i];
+float* t8 = A + i * A_ncols;
+float& t10 = q[i];
+float t16 = 0;
+for (int j = 0; j < A_ncols; j+=1) {
+float t15 = p[j];
+float t14 = t8[j];
+float& t19 = s[j];
+t19 += (t12*t14);
+t16 += (t14*t15);
+}
+t10 = t16;
+}
+}

### 2008-08-07 testsuite/matlab/AxATy__3.c
Jeremy Siek's Matlab compiler-generated C code from composed operations implemented in Matlab.
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@244 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@ -1,0 +1,23 @@
+void
+AxATy (float *A, int A_nrows, int A_ncols, float *x, int x_nrows, float *y,
+	   int y_nrows, float *w, int w_nrows, float *z, int z_nrows)
+{
+	for (int ii = 0; ii < z_nrows; ++ii)
+		z[ii] = 0.0;
+	for (int i = 0; i < A_nrows; i += 1) {
+		float t12 = y[i];
+		float *t8 = A + i * A_ncols;
+		float &t10 = w[i];
+		float t16 = 0;
+
+		for (int j = 0; j < A_ncols; j += 1) {
+			float t15 = x[j];
+			float t14 = t8[j];
+			float &t19 = z[j];
+
+			t19 += (t12 * t14);
+			t16 += (t14 * t15);
+		}
+		t10 = t16;
+	}
+}

### 2008-08-07 testsuite/matlab/atax__2.c

@@ -1,0 +1,24 @@
+void
+ATAX (float *A, int A_nrows, int A_ncols, float *x, int x_nrows, float *y,
+	  int y_nrows)
+{
+	for (int ii = 0; ii < y_nrows; ++ii)
+		y[ii] = 0.0;
+	for (int i = 0; i < A_nrows; i += 1) {
+		float *t6 = A + i * A_ncols;
+		float t14 = 0;
+
+		for (int j = 0; j < A_ncols; j += 1) {
+			float t13 = x[j];
+			float t12 = t6[j];
+
+			t14 += (t12 * t13);
+		}
+		for (int j = 0; j < A_ncols; j += 1) {
+			float t15 = t6[j];
+			float &t17 = y[j];
+
+			t17 += (t15 * t14);
+		}
+	}
+}

### 2008-08-07 testsuite/matlab/waxpby__4.c

@@ -1,0 +1,12 @@
+void
+WAXPBY (float a, float b, float *x, int x_nrows, float *y, int y_nrows,
+		float *w, int w_nrows)
+{
+	for (int i = 0; i < x_nrows; i += 1) {
+		float t11 = y[i];
+		float t8 = x[i];
+		float &t17 = w[i];
+
+		t17 = ((a * t8) + (b * t11));
+	}
+}

### 2008-08-07 testsuite/matlab/AxATy.m
Jeremy Siek's Matlab compiler-generated C code from composed operations implemented in Matlab.
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@244 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@ -1,0 +1,9 @@
+AxATy
+in
+A : row matrix, x : vector, y : vector
+out
+  w : vector, z : vector
+{
+  w = A * x
+  z = A' * y
+}

### 2008-08-07 testsuite/matlab/axpydot__4.c

@@ -1,0 +1,17 @@
+void
+AXPYDOT (float a, float *u, int u_nrows, float *v, int v_nrows, float *w,
+		 int w_nrows, float &r, float *z, int z_nrows)
+{
+	float t19 = 0;
+
+	for (int i = 0; i < w_nrows; i += 1) {
+		float t10 = v[i];
+		float t13 = w[i];
+		float &t16 = z[i];
+		float t18 = u[i];
+
+		t16 = (t13 - (a * t10));
+		t19 += (t18 * t16);
+	}
+	r = t19;
+}

### 2008-08-07 testsuite/matlab/bicgkernel.m

@@ -1,0 +1,9 @@
+BICG
+in
+  A : column matrix, p : vector, r : vector
+out
+  q : vector, s : vector
+{
+  q = A * p
+  s = A' * r
+}

### 2008-08-07 testsuite/matlab/dgemv.m

@@ -1,0 +1,8 @@
+DGEMV
+in
+  A : column matrix, x : vector, y : vector, alpha : scalar, beta : scalar
+out
+  z : vector
+{
+  z = alpha*(A*x) + beta*y
+}

### 2008-08-07 testsuite/matlab/dscal.m

@@ -1,0 +1,8 @@
+DSCAL
+in
+  A : column matrix, a : scalar
+out
+  B : column matrix
+{
+  B = a * A
+}

### 2008-08-07 testsuite/matlab/dscal__1.c

@@ -1,0 +1,16 @@
+void
+DSCAL (float *A, int A_nrows, int A_ncols, float a, float *B, int B_nrows,
+	   int B_ncols)
+{
+	for (int i = 0; i < A_ncols; i += 1) {
+		float *t4 = A + i * A_nrows;
+		float *t6 = B + i * A_nrows;
+
+		for (int j = 0; j < A_nrows; j += 1) {
+			float t7 = t4[j];
+			float &t9 = t6[j];
+
+			t9 = (a * t7);
+		}
+	}
+}

### 2008-08-07 testsuite/matlab/gesummv_row__21.c

@@ -1,0 +1,38 @@
+void GESUMMV(float* A, int A_nrows, int A_ncols, float* B, int B_nrows, int B_ncols, float a, float b, float* x, int x_nrows, float* y, int y_nrows) {
+float* t7 = new float[B_nrows];
+float* t8 = new float[B_nrows];
+float* t3 = new float[A_nrows];
+for (int i = 0; i < B_nrows; i+=1) {
+float* t17 = B + i * B_ncols;
+float& t19 = t7[i];
+float t32 = 0;
+for (int j = 0; j < B_ncols; j+=1) {
+float t31 = x[j];
+float t30 = t17[j];
+t32 += (t30*t31);
+}
+t19 = t32;
+}
+for (int i = 0; i < B_nrows; i+=1) {
+float t20 = t7[i];
+float& t22 = t8[i];
+t22 = (b*t20);
+}
+for (int i = 0; i < A_nrows; i+=1) {
+float* t11 = A + i * A_ncols;
+float& t13 = t3[i];
+float t29 = 0;
+for (int j = 0; j < A_ncols; j+=1) {
+float t28 = x[j];
+float t27 = t11[j];
+t29 += (t27*t28);
+}
+t13 = t29;
+}
+for (int i = 0; i < A_nrows; i+=1) {
+float t14 = t3[i];
+float t24 = t8[i];
+float& t26 = y[i];
+t26 = (t24+(a*t14));
+}
+}

### 2008-08-07 testsuite/matlab/madd.m

@@ -1,0 +1,9 @@
+madd
+in
+  A : matrix, B : matrix
+out
+  C : matrix
+{
+  C = A + B
+}
+

### 2008-08-07 testsuite/matlab/madd__1.c

@@ -1,0 +1,18 @@
+void
+madd (float *A, int A_nrows, int A_ncols, float *B, int B_nrows, int B_ncols,
+	  float *C, int C_nrows, int C_ncols)
+{
+	for (int i = 0; i < A_ncols; i += 1) {
+		float *t5 = B + i * B_nrows;
+		float *t4 = A + i * A_nrows;
+		float *t7 = C + i * A_nrows;
+
+		for (int j = 0; j < A_nrows; j += 1) {
+			float t9 = t5[j];
+			float t8 = t4[j];
+			float &t11 = t7[j];
+
+			t11 = (t8 + t9);
+		}
+	}
+}

### 2008-08-07 testsuite/matlab/new/bicgkernel_row.m

@@ -1,0 +1,9 @@
+BICG
+in
+  A : row matrix, p : vector, r : vector
+out
+  q : vector, s : vector
+{
+  q = A * p
+  s = A' * r
+}

### 2008-08-08 testsuite/spmv.c
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@246 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@ -15,10 +15,10 @@
  inode_rows = inode_rows;       # the accumulated version of inode_sizes (e.g. [0,2,5,9])
 
  # transformation parameters
- out_unroll_factor = 3;
- in_unroll_factor = 4;
- num_threads = 1;
- simd = 'xlc';    # 'none' (default), 'gcc', 'sse', 'xlc'
+ out_unroll_factor = 4;
+ in_unroll_factor = 6;
+ num_threads = 4;
+ simd = 'none';    # 'none' (default), 'gcc', 'sse', 'xlc'
  block_structure = 'none';    # 'none' (default), 'inode', 'bcsr' (still unsupported)
 
 ) @*/

### 2008-08-08 testsuite/petsc/matmult/bgp/basic/spmv.c
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@246 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@ -1,9 +1,9 @@
 /*@ begin PerfTuning (  
  def build { 
    arg build_command = 'mpixlc_r -O3 -qsmp=omp:noauto -qstrict';
-   arg batch_command = 'qsub -n 4 -t 5 -q short --env "BG_MAXALIGNEXP=0"';
+   arg batch_command = 'qsub -n 64 -t 5 -q short --env "BG_MAXALIGNEXP=0"';
    arg status_command = 'qstat';
-   arg num_procs = 4;
+   arg num_procs = 64;
  }
 
  def performance_counter {
@@ -12,16 +12,16 @@
  }
 
  def performance_params {
-   param UNROLL_FAC_OUT[] = [1];
-   param UNROLL_FAC_IN[] = [2];
+   param UNROLL_FAC_OUT[] = [1,2,3,4];
+   param UNROLL_FAC_IN[] = [1,2,3,4,5,6,7,8];
    param N_THREADS[] = [1];
-   param SIMD_TYPE[] = ['xlc'];
+   param SIMD_TYPE[] = ['none','xlc'];
 
    constraint simd_unroll_factor = (SIMD_TYPE=='none' or UNROLL_FAC_IN%2==0);
  }
 
  def input_params {
-   param NROWS[] = [1000];
+   param NROWS[] = [700,1200,10000];
    param NCOLS[] = [25];
  }
  

### 2008-08-17 testsuite/fdtd/cookie/exp/fdtd-2d.orio.seq.large.c

@@ -67,19 +67,19 @@ int main()
 #   [4,8,16,32,64,128];
 #   [1,4,8,16];
 
-    param T1_1[] = [16];
-    param T1_2[] = [32];
-    param T1_3[] = [32];
+    param T1_1[] = [64];
+    param T1_2[] = [64];
+    param T1_3[] = [64];
     param T2_1[] = [1];
     param T2_2[] = [1];
     param T2_3[] = [1];
 
 #    constraint c1 = (T1_1*T2_1<=1024 and T1_2*T2_2<=1024 and T1_3*T2_3<=1024);
-#    constraint c2 = (T1_1==T1_3 and T2_1==T2_3);
+    constraint c2 = (T1_1==T1_3 and T2_1==T2_3);
     
-    param U1[] = [8];
+    param U1[] = [7];
     param U2[] = [2];
-    param U3[] = [4];
+    param U3[] = [3];
 
     constraint c3 = (U1*U2*U3<=256);
 
@@ -93,7 +93,7 @@ int main()
 
     ];
 
-    param PAR[] = [True];
+    param PAR[] = [False];
     param SCREP[] = [False];
     param IVEC[] = [True];
   }
@@ -125,25 +125,25 @@ int main()
 
 /**-- (Generated by Orio) 
 Best performance cost: 
-  1.681530 
+  7.484850 
 Tuned for specific problem sizes: 
   nx = 2000 
   ny = 2000 
   tmax = 500 
 Best performance parameters: 
   IVEC = True 
-  PAR = True 
+  PAR = False 
   PERM = [0, 2, 1] 
   SCREP = False 
-  T1_1 = 16 
-  T1_2 = 32 
-  T1_3 = 32 
+  T1_1 = 64 
+  T1_2 = 64 
+  T1_3 = 64 
   T2_1 = 1 
   T2_2 = 1 
   T2_3 = 1 
-  U1 = 8 
+  U1 = 7 
   U2 = 2 
-  U3 = 4 
+  U3 = 3 
 --**/
 
    
[…]

### 2008-08-17 testsuite/fdtd/cookie/exp/fdtd-2d.pluto.par.c

@@ -54,459 +54,509 @@ int main()
   {
     annot_t_start = rtclock();
     
-
-
 #include <math.h>
 #include <assert.h>
-
+#include <omp.h>
 #define ceild(n,d)  ceil(((double)(n))/((double)(d)))
 #define floord(n,d) floor(((double)(n))/((double)(d)))
 #define max(x,y)    ((x) > (y)? (x) : (y))
 #define min(x,y)    ((x) < (y)? (x) : (y))
 
-	#define S1(zT0,zT1,t,j)	{ey[0][j]=t;}
-	#define S2(zT0,zT1,zT2,t,i,j)	{ey[i][j]=ey[i][j]-((double)(1))/2*(hz[i][j]-hz[i-1][j]);}
-	#define S3(zT0,zT1,zT2,t,i,j)	{ex[i][j]=ex[i][j]-((double)(1))/2*(hz[i][j]-hz[i][j-1]);}
-	#define S4(zT0,zT1,zT2,t,i,j)	{hz[i][j]=hz[i][j]-((double)(7))/10*(ey[1+i][j]+ex[i][1+j]-ex[i][j]-ey[i][j]);}
 
-	int c1, c2, c3, c4, c5, c6;
 
-	register int lb, ub, lb1, ub1, lb2, ub2;
-	register int lbv, ubv;
 
-/* Generated from PLuTo-produced CLooG file by CLooG v0.14.1 64 bits in 0.26s. */
+
+ int c1, c2, c3, c4, c5, c6;
+ register int lb, ub, lb1, ub1, lb2, ub2;
+ register int lbv, ubv;
+
 for (c1=-1;c1<=floord(2*tmax+ny-2,32);c1++) {
-	lb1=max(max(ceild(32*c1-tmax+1,32),ceild(32*c1-31,64)),0);
-	ub1=min(min(floord(32*c1+ny+31,64),floord(tmax+ny-1,32)),floord(32*c1+31,32));
+ lb1=max(max(ceild(32*c1-tmax+1,32),0),ceild(32*c1-31,64));
+ ub1=min(min(floord(32*c1+31,32),floord(32*c1+ny+31,64)),floord(tmax+ny-1,32));
[…]

### 2008-08-17 testsuite/fdtd/cookie/exp/fdtd-2d.pluto.seq.c

@@ -54,453 +54,502 @@ int main()
   {
     annot_t_start = rtclock();
     
-
-
 #include <math.h>
 #include <assert.h>
-
 #define ceild(n,d)  ceil(((double)(n))/((double)(d)))
 #define floord(n,d) floor(((double)(n))/((double)(d)))
 #define max(x,y)    ((x) > (y)? (x) : (y))
 #define min(x,y)    ((x) < (y)? (x) : (y))
 
-	#define S1(zT0,zT1,t,j)	{ey[0][j]=t;}
-	#define S2(zT0,zT1,zT2,t,i,j)	{ey[i][j]=ey[i][j]-((double)(1))/2*(hz[i][j]-hz[i-1][j]);}
-	#define S3(zT0,zT1,zT2,t,i,j)	{ex[i][j]=ex[i][j]-((double)(1))/2*(hz[i][j]-hz[i][j-1]);}
-	#define S4(zT0,zT1,zT2,t,i,j)	{hz[i][j]=hz[i][j]-((double)(7))/10*(ey[1+i][j]+ex[i][1+j]-ex[i][j]-ey[i][j]);}
 
-	int c1, c2, c3, c4, c5, c6;
 
-	register int lbv, ubv;
 
-/* Generated from PLuTo-produced CLooG file by CLooG v0.14.1 64 bits in 0.25s. */
+
+ int c1, c2, c3, c4, c5, c6;
+ register int lbv, ubv;
+
 for (c1=0;c1<=floord(tmax-1,32);c1++) {
-  for (c2=max(ceild(32*c1-31,32),0);c2<=min(floord(tmax+ny-1,32),floord(32*c1+ny+31,32));c2++) {
-    for (c3=max(max(max(max(ceild(32*c2-ny-30,32),0),ceild(64*c1-32*c2-61,32)),ceild(32*c1-31,32)),ceild(32*c1-992*c2-1891,992));c3<=min(min(floord(32*c2+nx+30,32),floord(tmax+nx-1,32)),floord(32*c1+nx+31,32));c3++) {
+  for (c2=max(0,ceild(32*c1-31,32));c2<=min(floord(tmax+ny-1,32),floord(32*c1+ny+31,32));c2++) {
[…]

### 2008-08-17 testsuite/fdtd/cookie/exp/runscript.py

@@ -101,7 +101,7 @@ def checkCorrectness(optflag = '-O0'):
     fnames = [
         #'fdtd-2d.pluto.seq.c', 
         #'fdtd-2d.pluto.par.c', 
-        'fdtd-2d.orio.seq.small.c',
+        #'fdtd-2d.orio.seq.small.c',
         #'fdtd-2d.orio.seq.large.c',
         #'fdtd-2d.orio.par.c',
         ]
@@ -142,7 +142,7 @@ checkCorrectness()
 checkCorrectness(OPTFLAG)
 
 # parallel case
-if 0:
+if 1:
     reps = 1
     T = 500
     N = 2000
@@ -152,17 +152,17 @@ if 0:
     mflopss_pluto = []
     mflopss_orio = []
 
-    rtimes_base = runExp([1,2,3,4,5,6,7,8], 'icc %s -parallel' % OPTFLAG, 
-                         'fdtd-2d.base.c', 'base_par', flags, '-lm')
-    mflopss_base = countFlops(T,N,rtimes_base)
+    #rtimes_base = runExp([1,2,3,4,5,6,7,8], 'icc %s -parallel' % OPTFLAG, 
+    #                     'fdtd-2d.base.c', 'base_par', flags, '-lm')
+    #mflopss_base = countFlops(T,N,rtimes_base)
     
     rtimes_pluto = runExp([1,2,3,4,5,6,7,8], 'icc %s -openmp' % OPTFLAG, 
                           'fdtd-2d.pluto.par.c', 'pluto_par', flags, '-lm')
     mflopss_pluto = countFlops(T,N,rtimes_pluto)
     
-    rtimes_orio = runExp([1,2,3,4,5,6,7,8], 'icc %s -openmp' % OPTFLAG, 
-                         'fdtd-2d.orio.par.c', 'orio_par', flags, '-lm')
-    mflopss_orio = countFlops(T,N,rtimes_orio)
+    #rtimes_orio = runExp([1,2,3,4,5,6,7,8], 'icc %s -openmp' % OPTFLAG, 
+    #                     'fdtd-2d.orio.par.c', 'orio_par', flags, '-lm')
+    #mflopss_orio = countFlops(T,N,rtimes_orio)
     
     printFloats(mflopss_base)
     printFloats(mflopss_pluto)
@@ -186,15 +186,15 @@ if 1:
         #p = countFlops(T,N,rtimes_base)
         #mflopss_base.append(p[0])
[…]

### 2008-08-17 testsuite/fdtd/cookie/fdtd-2d.tiled.c
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@292 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@ -1,447 +1,0 @@
-
-
-#include <math.h>
-#include <assert.h>
-
-#define ceild(n,d)  ceil(((double)(n))/((double)(d)))
-#define floord(n,d) floor(((double)(n))/((double)(d)))
-#define max(x,y)    ((x) > (y)? (x) : (y))
-#define min(x,y)    ((x) < (y)? (x) : (y))
-
-	#define S1(zT0,zT1,t,j)	{ey[0][j]=t;}
-	#define S2(zT0,zT1,zT2,t,i,j)	{ey[i][j]=ey[i][j]-((double)(1))/2*(hz[i][j]-hz[i-1][j]);}
-	#define S3(zT0,zT1,zT2,t,i,j)	{ex[i][j]=ex[i][j]-((double)(1))/2*(hz[i][j]-hz[i][j-1]);}
-	#define S4(zT0,zT1,zT2,t,i,j)	{hz[i][j]=hz[i][j]-((double)(7))/10*(ey[1+i][j]+ex[i][1+j]-ex[i][j]-ey[i][j]);}
-
-	int c1, c2, c3, c4, c5, c6;
-
-	register int lbv, ubv;
-
-/* Generated from PLuTo-produced CLooG file by CLooG v0.14.1 64 bits in 0.25s. */
-for (c1=0;c1<=floord(tmax-1,32);c1++) {
-  for (c2=max(ceild(32*c1-31,32),0);c2<=min(floord(tmax+ny-1,32),floord(32*c1+ny+31,32));c2++) {
-    for (c3=max(max(max(max(ceild(32*c2-ny-30,32),0),ceild(64*c1-32*c2-61,32)),ceild(32*c1-31,32)),ceild(32*c1-992*c2-1891,992));c3<=min(min(floord(32*c2+nx+30,32),floord(tmax+nx-1,32)),floord(32*c1+nx+31,32));c3++) {
-      if ((c1 <= floord(32*c3-nx,32)) && (c2 <= floord(32*c3-nx+ny,32)) && (c3 >= ceild(nx,32))) {
-        for (c5=max(32*c3-nx+1,32*c2);c5<=min(32*c2+31,32*c3-nx+ny);c5++) {
-          S4(c1,-c1+c3,-c1+c2,32*c3-nx,nx-1,-32*c3+c5+nx-1) ;
-        }
[…]

### 2008-08-17 testsuite/fdtd/cookie/fdtd-2d.par.c
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@292 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@ -1,453 +1,0 @@
-
-
-#include <math.h>
-#include <assert.h>
-
-#define ceild(n,d)  ceil(((double)(n))/((double)(d)))
-#define floord(n,d) floor(((double)(n))/((double)(d)))
-#define max(x,y)    ((x) > (y)? (x) : (y))
-#define min(x,y)    ((x) < (y)? (x) : (y))
-
-	#define S1(zT0,zT1,t,j)	{ey[0][j]=t;}
-	#define S2(zT0,zT1,zT2,t,i,j)	{ey[i][j]=ey[i][j]-((double)(1))/2*(hz[i][j]-hz[i-1][j]);}
-	#define S3(zT0,zT1,zT2,t,i,j)	{ex[i][j]=ex[i][j]-((double)(1))/2*(hz[i][j]-hz[i][j-1]);}
-	#define S4(zT0,zT1,zT2,t,i,j)	{hz[i][j]=hz[i][j]-((double)(7))/10*(ey[1+i][j]+ex[i][1+j]-ex[i][j]-ey[i][j]);}
-
-	int c1, c2, c3, c4, c5, c6;
-
-	register int lb, ub, lb1, ub1, lb2, ub2;
-	register int lbv, ubv;
-
-/* Generated from PLuTo-produced CLooG file by CLooG v0.14.1 64 bits in 0.26s. */
-for (c1=-1;c1<=floord(2*tmax+ny-2,32);c1++) {
-	lb1=max(max(ceild(32*c1-tmax+1,32),ceild(32*c1-31,64)),0);
-	ub1=min(min(floord(32*c1+ny+31,64),floord(tmax+ny-1,32)),floord(32*c1+31,32));
-#pragma omp parallel for shared(c1,lb1,ub1) private(c2,c3,c4,c5,c6)
-	for (c2=lb1; c2<=ub1; c2++) {
-    for (c3=max(max(max(max(ceild(32*c2-ny-30,32),0),ceild(64*c1-96*c2-61,32)),ceild(32*c1-32*c2-31,32)),ceild(32*c1-1024*c2-1891,992));c3<=min(min(floord(32*c2+nx+30,32),floord(tmax+nx-1,32)),floord(32*c1-32*c2+nx+31,32));c3++) {
[…]

### 2008-08-17 testsuite/fdtd/cookie/fdtd-2d.src1.c

@@ -1,31 +1,31 @@
 /*@ begin PerfTuning (
-  def build
-  {
+  def build {
     arg build_command = 'icc -O3 -openmp -lm';
   }
 
-  def performance_counter
-  {
+  def performance_counter {
     arg repetitions = 1;
   }  
   
-  def performance_params
-  {  
-#   [1,32,64,128,256,512];
-#   [1,4,8,16,32];
+  def performance_params {  
+#   [4,8,16,32,64,128];
+#   [1,4,8,16];
 
-    param T1_1[] = [8,16,64];
-    param T1_2[] = [16,32,64];
-    param T1_3[] = [16,32,64];
+    param T1_1[] = [16];
+    param T1_2[] = [32];
+    param T1_3[] = [32];
     param T2_1[] = [1];
     param T2_2[] = [1];
     param T2_3[] = [1];
 
-    param U1[] = [1];
-    param U2[] = [1];
-    param U3[] = [1];
+#    constraint c1 = (T1_1*T2_1<=1024 and T1_2*T2_2<=1024 and T1_3*T2_3<=1024);
+#    constraint c2 = (T1_1==T1_3 and T2_1==T2_3);
+    
+    param U1[] = [8];
+    param U2[] = [2];
+    param U3[] = [4];
 
-    constraint c1 = (U1*U2*U3<=512);
+    constraint c3 = (U1*U2*U3<=256);
 
     param PERM[] = [
 #      [0,1,2],

### 2008-08-17 testsuite/fdtd/cookie/fdtd-2d.src1.c
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@293 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@ -40,6 +40,7 @@
     param PAR[] = [True];
     param SCREP[] = [False];
     param IVEC[] = [True];
+    param RECTILE[] = [False];
   }
   
   def search
@@ -79,6 +80,7 @@ register int newub_c1, newub_c2, newub_c3, newub_c4, newub_c5, newub_c6,
   tiles = [T1_1,T1_2,T1_3,T2_1,T2_2,T2_3];
   permut = PERM;
   unroll_factors = [U1,U2,U3]; 
+  rect_regtile = RECTILE;
   scalar_replace = SCREP; 
   vectorize = IVEC; 
  

### 2008-08-17 testsuite/seidel/cookie/exp/seidel.orio.par.c

@@ -1,0 +1,284 @@
+#include <stdio.h>
+#include <stdlib.h>
+#include <sys/time.h>
+#include <math.h>
+
+double A[N][N+13];
+
+void init_arrays()
+{
+  int i, j;
+  for (i=0; i<N; i++) 
+    for (j=0; j<N; j++) 
+      A[i][j] = i*i+j*j;
+}
+
+double rtclock()
+{
+  struct timezone tzp;
+  struct timeval tp;
+  int stat;
+  gettimeofday (&tp, &tzp);
+  return (tp.tv_sec + tp.tv_usec*1.0e-6);
+}
+
+int main()
+{
+  init_arrays();
+
+  double annot_t_start=0, annot_t_end=0, annot_t_total=0;
+  int annot_i;
+
+  for (annot_i=0; annot_i<REPS; annot_i++)
+  {
+    annot_t_start = rtclock();
+    
+
+/*@ begin PerfTuning (          
+  def build  
+  {  
+    arg build_command = 'icc -O3 -openmp -lm';  
+  }
+     
+  def performance_counter           
+  {  
+    arg repetitions = 1;  
+  } 
+ 
+  def performance_params  
+  { 
+# [1,8,16,32,64,128,256]; 
+# [1,4,8,16,32];
+
+    param T1_1[] = [32];
+    param T1_2[] = [32];
+    param T1_3[] = [32];
+    param T2_1[] = [1];
+    param T2_2[] = [1];
+    param T2_3[] = [1];
+ 
+    constraint c1 = (T1_1*T2_1<=1024 and T1_1*T2_1<=1024 and T1_1*T2_1<=1024);
+
+    param U1[] = [1];
+    param U2[] = [4,8];
+    param U3[] = [4,8];
+
+    constraint c2 = (U1*U2*U3<=256);
+
+    param PERM[] = [
+     [0,1,2],
[…]

### 2008-08-17 testsuite/seidel/cookie/seidel.src.c

@@ -1,0 +1,98 @@
+/*@ begin PerfTuning (          
+  def build  
+  {  
+    arg build_command = 'icc -O3 -openmp -lm';  
+  }
+     
+  def performance_counter           
+  {  
+    arg repetitions = 1;  
+  } 
+ 
+  def performance_params  
+  { 
+# [1,8,16,32,64,128,256]; 
+# [1,4,8,16,32];
+
+    param T1_1[] = [32];
+    param T1_2[] = [32];
+    param T1_3[] = [32];
+    param T2_1[] = [1];
+    param T2_2[] = [1];
+    param T2_3[] = [1];
+ 
+    constraint c1 = (T1_1*T2_1<=1024 and T1_1*T2_1<=1024 and T1_1*T2_1<=1024);
+
+    param U1[] = [1];
+    param U2[] = [4];
+    param U3[] = [4];
+
+    constraint c2 = (U1*U2*U3<=256);
+
+    param PERM[] = [
+     [0,1,2],
+#     [0,2,1],
+#     [1,0,2],
+#     [1,2,0],
+#     [2,0,1],
+#     [2,1,0],
+    ];
+
+    param PAR[] = [True];
+    param SCREP[] = [False];
+    param IVEC[] = [False];
+    param RECTILE[] = [True,False];
+  } 
+ 
+  def search  
+  {
+    arg algorithm = 'Exhaustive';  
+#    arg algorithm = 'Simplex';  
+#    arg total_runs = 1; 
+  }
+
+  def input_params  
+  {
+    param T[] = [500];
+    param N[] = [4000];
+  } 
+
+  def input_vars
+  {
+    decl static double A[N][N+17] = random;
+  }
+) @*/  
+
+register int i,j,k,t; 
+register int c1t, c2t, c3t, c4t, c5t, c6t, c7t, c8t, c9t, c10t, c11t, c12t; 
+register int newlb_c1, newlb_c2, newlb_c3, newlb_c4, newlb_c5, newlb_c6, 
+  newlb_c7, newlb_c8, newlb_c9, newlb_c10, newlb_c11, newlb_c12; 
+register int newub_c1, newub_c2, newub_c3, newub_c4, newub_c5, newub_c6, 
+  newub_c7, newub_c8, newub_c9, newub_c10, newub_c11, newub_c12; 
+
[…]

### 2008-08-17 src/module/polysyn/syn_transformator.py

@@ -9,7 +9,7 @@ import os, re, shlex, sys
 class SynTransformator:
     '''The syntactic transformator'''
 
-    def __init__(self, verbose, permut, unroll_factors, scalar_replace, vectorize):
+    def __init__(self, verbose, permut, unroll_factors, scalar_replace, vectorize, rect_regtile):
         '''To instantiate a syntactic transformator instance'''
 
         self.verbose = verbose
@@ -17,6 +17,7 @@ class SynTransformator:
         self.unroll_factors = unroll_factors
         self.scalar_replace = scalar_replace
         self.vectorize = vectorize
+        self.rect_regtile = rect_regtile
 
     #---------------------------------------------------------
 
@@ -165,7 +166,10 @@ class SynTransformator:
             else:
                 lnest = lnest[-mlength:]
                 ufactors = self.unroll_factors[:mlength]
-                regtile_code = 'regtile = (%s,%s)' % (lnest, ufactors)
+                if self.rect_regtile:
+                    regtile_code = 'unrolljam = (%s,%s)' % (lnest, ufactors)
+                else:
+                    regtile_code = 'regtile = (%s,%s)' % (lnest, ufactors)
 
             # get the scalar-replacement code
             srep_code = "scalarreplace = (%s, '%s')" % self.scalar_replace

### 2008-08-17 src/module/polysyn/transf_info.py

@@ -10,13 +10,14 @@ class TransfInfo:
     '''The transformation information'''
 
     def __init__(self, parallel, tiles, permut, unroll_factors, scalar_replace, vectorize,
-                 profiling_code, compile_cmd, compile_opts):
+                 profiling_code, compile_cmd, compile_opts, rect_regtile):
         '''To instantiate a transformation information instance'''
 
         self.parallel = parallel
         self.tiles = tiles
         self.permut = permut
         self.unroll_factors = unroll_factors
+        self.rect_regtile = rect_regtile
         self.scalar_replace = scalar_replace
         self.vectorize = vectorize
 
@@ -51,6 +52,7 @@ class TransfInfoGen:
         TILES = 'tiles'
         PERMUT = 'permut'
         UFACTORS = 'unroll_factors'
+        RECTRTILE = 'rect_regtile'
         SREPLACE = 'scalar_replace'
         VECTOR = 'vectorize'
         PCODE = 'profiling_code'
@@ -62,6 +64,7 @@ class TransfInfoGen:
         tiles = None
         permut = None
         unroll_factors = None
+        rect_regtile = None
         scalar_replace = None
         vectorize = None
         profiling_code = None
@@ -114,6 +117,14 @@ class TransfInfoGen:
                                (rhs_line_no, t))
                         sys.exit(1)
                 unroll_factors = rhs_val
+
+            # rectangular register tiling
+            elif vname == RECTRTILE:
+                if not isinstance(rhs_val, bool):
+                    print ('error:%s: rectangular register-tiling value must be a boolean' % 
+                           rhs_line_no)
+                    sys.exit(1)
+                rect_regtile = rhs_val
 
             # parallelization
             elif vname == PARALLEL:
@@ -177,6 +188,8 @@ class TransfInfoGen:
             permut = []
         if unroll_factors == None:
             unroll_factors = []
+        if rect_regtile == None:
+            rect_regtile = False
         if scalar_replace == None:
             scalar_replace = (False, 'double')
         if vectorize == None:
@@ -187,6 +200,6 @@ class TransfInfoGen:
             
         # generate and return the transformation information
         return TransfInfo(parallel, tiles, permut, unroll_factors, scalar_replace, vectorize,
-                          profiling_code, compile_cmd, compile_opts)
+                          profiling_code, compile_cmd, compile_opts, rect_regtile)
 
 

### 2008-08-17 testsuite/seidel/cookie/exp/seidel.pluto.par.c

@@ -1,0 +1,162 @@
+#include <stdio.h>
+#include <stdlib.h>
+#include <sys/time.h>
+#include <math.h>
+
+double A[N][N+13];
+
+void init_arrays()
+{
+  int i, j;
+  for (i=0; i<N; i++) 
+    for (j=0; j<N; j++) 
+      A[i][j] = i*i+j*j;
+}
+
+double rtclock()
+{
+  struct timezone tzp;
+  struct timeval tp;
+  int stat;
+  gettimeofday (&tp, &tzp);
+  return (tp.tv_sec + tp.tv_usec*1.0e-6);
+}
+
+int main()
+{
+  init_arrays();
+
+  double annot_t_start=0, annot_t_end=0, annot_t_total=0;
+  int annot_i;
+
+  for (annot_i=0; annot_i<REPS; annot_i++)
+  {
+    annot_t_start = rtclock();
+    
+
+#include <math.h>
+#include <assert.h>
+#include <omp.h>
+#define ceild(n,d)  ceil(((double)(n))/((double)(d)))
+#define floord(n,d) floor(((double)(n))/((double)(d)))
+#define max(x,y)    ((x) > (y)? (x) : (y))
+#define min(x,y)    ((x) < (y)? (x) : (y))
+
+
+
+ int c1, c2, c3, c4, c5, c6;
+ register int lb, ub, lb1, ub1, lb2, ub2;
+ register int lbv, ubv;
+
+if (N >= 3) {
+  for (c1=-1;c1<=floord(2*T+N-4,32);c1++) {
+ lb1=max(max(ceild(16*c1-15,32),0),ceild(32*c1-T+1,32));
+ ub1=min(min(floord(32*c1+31,32),floord(T+N-3,32)),floord(32*c1+N+29,64));
[…]

### 2008-08-17 src/module/polysyn/polysyn.py
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@293 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@ -87,7 +87,8 @@ class PolySyn(module.module.Module):
                                                     tinfo.permut,
                                                     tinfo.unroll_factors,
                                                     tinfo.scalar_replace,
-                                                    tinfo.vectorize)
+                                                    tinfo.vectorize,
+                                                    tinfo.rect_regtile)
         transformed_code = strans.transform(pluto_code)
 
         # return the transformed code

### 2008-08-17 testsuite/seidel/cookie/exp/runscript.py

@@ -1,0 +1,200 @@
+#!/usr/bin/env python
+
+import os, sys, re
+
+def printFloats(ls):
+    s=''
+    s+='['
+    for i,f in enumerate(ls):
+        if i: s+=', '
+        s+='%2.3f' % f
+    s+=']'
+    print s
+
+def runExp(nthreadss, cc, src_file, out_file, flags, libs):
+    compile_cmd = '%s %s -o %s %s %s' % (cc, flags, out_file, src_file, libs)
+    print '*************************************'
+    print compile_cmd
+    print '*************************************'
+    os.system(compile_cmd)
+
+    rtimes = []
+    for nt in nthreadss:
+        run_cmd = 'export OMP_NUM_THREADS=%s; ./%s' % (nt, out_file)
+        print '*************************************'
+        print run_cmd
+        print '*************************************'
+        f = os.popen(run_cmd)
+        output = f.read()
+        f.close()
+        rtime = eval(output)
+        rtimes.append(rtime)
+    return rtimes
+
+def countFlops(T, N, rtimes):
+    ops = 9*T*(N-2)*(N-2)
+    mflopss = []
+    for rtime in rtimes:
+        mflops = 1.0*ops/(rtime*1000000)
+        mflopss.append(mflops)
+    return mflopss
+
+def myDiff(fname1, fname2):
+    f1 = open(fname1)
+    c1 = f1.read()
+    f1.close()
+    f2 = open(fname2)
+    c2 = f2.read()
+    f2.close()
+    cls1 = c1.split('\n')
+    cls2 = c2.split('\n')
+    maxlength=100
+    diffs = []
+    total_diffs = 0
+    total_nums = 0
+    for l1,l2 in zip(cls1, cls2):
+        l1 = l1.split()
+        l2 = l2.split()
+        for n1,n2 in zip(l1,l2):
+            total_nums+=1
+            try:
+                n1=eval(n1)
+                n2=eval(n2)
+            except:
+                print 'error: cannot evaluate'
+                print 'n1=%s' % n1
+                print 'n2=%s' % n2
+                sys.exit(1)
+            d=n1-n2
+            if not (-1<d<1):
+                total_diffs+=1
+                if len(diffs) == 0:
+                    diffs.append(d)
+                    continue
+                maxdiff=diffs[-1]
+                if len(diffs) < maxlength:
[…]

### 2008-08-18 testsuite/seidel/cookie/exp/runscript.py
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@294 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@ -99,9 +99,9 @@ def checkCorrectness(optflag = '-O0'):
     f.close()
     
     fnames = [
-        #'seidel.pluto.seq.c', 
-        #'seidel.pluto.par.c', 
-        #'seidel.orio.seq.c',
+        'seidel.pluto.seq.c', 
+        'seidel.pluto.par.c', 
+        'seidel.orio.seq.c',
         'seidel.orio.par.c',
         ]
     for fname in fnames:
@@ -151,15 +151,15 @@ if 1:
     mflopss_pluto = []
     mflopss_orio = []
 
-    #rtimes_base = runExp([1,2,3,4,5,6,7,8], 'icc %s -parallel' % OPTFLAG, 
-    #                     'seidel.base.c', 'base_par', flags, '-lm')
-    #mflopss_base = countFlops(T,N,rtimes_base)
+    rtimes_base = runExp([1,2,3,4,5,6,7,8], 'icc %s -parallel' % OPTFLAG, 
+                         'seidel.base.c', 'base_par', flags, '-lm')
+    mflopss_base = countFlops(T,N,rtimes_base)
     
-    #rtimes_pluto = runExp([1,2,3,4,5,6,7,8], 'icc %s -openmp' % OPTFLAG, 
-    #                      'seidel.pluto.par.c', 'pluto_par', flags, '-lm')
-    #mflopss_pluto = countFlops(T,N,rtimes_pluto)
+    rtimes_pluto = runExp([1,2,3,4,5,6,7,8], 'icc %s -openmp' % OPTFLAG, 
+                          'seidel.pluto.par.c', 'pluto_par', flags, '-lm')
+    mflopss_pluto = countFlops(T,N,rtimes_pluto)
     
-    rtimes_orio = runExp([2,4,6,8], 'icc %s -openmp' % OPTFLAG, 
+    rtimes_orio = runExp([1,2,3,4,5,6,7,8], 'icc %s -openmp' % OPTFLAG, 
                          'seidel.orio.par.c', 'orio_par', flags, '-lm')
     mflopss_orio = countFlops(T,N,rtimes_orio)
     
@@ -168,7 +168,7 @@ if 1:
     printFloats(mflopss_orio)
     
 # sequential case
-if 0:
+if 1:
     reps = 1
     T = 500
 
@@ -179,15 +179,15 @@ if 0:
     for N in [125,250,500,1000,2000,4000,8000]:
         flags = '-DREPS=%s -DT=%s -DN=%s' % (reps, T, N)
         
[…]

### 2008-08-18 testsuite/seidel/cookie/exp/seidel.orio.par.c

@@ -50,18 +50,18 @@ int main()
 # [1,8,16,32,64,128,256]; 
 # [1,4,8,16,32];
 
-    param T1_1[] = [32];
-    param T1_2[] = [32];
-    param T1_3[] = [32];
+    param T1_1[] = [16,32];
+    param T1_2[] = [8];
+    param T1_3[] = [512];
     param T2_1[] = [1];
     param T2_2[] = [1];
     param T2_3[] = [1];
  
     constraint c1 = (T1_1*T2_1<=1024 and T1_1*T2_1<=1024 and T1_1*T2_1<=1024);
 
-    param U1[] = [1];
-    param U2[] = [4,8];
-    param U3[] = [4,8];
+    param U1[] = [3];
+    param U2[] = [1];
+    param U3[] = [4];
 
     constraint c2 = (U1*U2*U3<=256);
 
@@ -75,16 +75,16 @@ int main()
     ];
 
     param PAR[] = [True];
-    param SCREP[] = [False];
-    param IVEC[] = [False];
-    param RECTILE[] = [True,False];
+    param SCREP[] = [True];
+    param IVEC[] = [True];
+    param RECTILE[] = [True];
   } 
  
   def search  
   {
     arg algorithm = 'Exhaustive';  
 #    arg algorithm = 'Simplex';  
-#    arg total_runs = 1; 
+#    arg total_runs = 5; 
   }
 
   def input_params  
@@ -101,24 +101,24 @@ int main()
 
 /**-- (Generated by Orio) 
 Best performance cost: 
-  7.561850 
+  28.572200 
 Tuned for specific problem sizes: 
   N = 4000 
   T = 500 
 Best performance parameters: 
-  IVEC = False 
+  IVEC = True 
   PAR = True 
   PERM = [0, 1, 2] 
   RECTILE = True 
-  SCREP = False 
+  SCREP = True 
   T1_1 = 32 
-  T1_2 = 32 
[…]

### 2008-08-18 testsuite/seidel/cookie/seidel.src.c

@@ -14,17 +14,17 @@
 # [1,8,16,32,64,128,256]; 
 # [1,4,8,16,32];
 
-    param T1_1[] = [32];
-    param T1_2[] = [32];
-    param T1_3[] = [32];
+    param T1_1[] = [16];
+    param T1_2[] = [8];
+    param T1_3[] = [512];
     param T2_1[] = [1];
     param T2_2[] = [1];
     param T2_3[] = [1];
  
     constraint c1 = (T1_1*T2_1<=1024 and T1_1*T2_1<=1024 and T1_1*T2_1<=1024);
 
-    param U1[] = [1];
-    param U2[] = [4];
+    param U1[] = [3];
+    param U2[] = [1];
     param U3[] = [4];
 
     constraint c2 = (U1*U2*U3<=256);
@@ -38,17 +38,17 @@
 #     [2,1,0],
     ];
 
-    param PAR[] = [True];
-    param SCREP[] = [False];
-    param IVEC[] = [False];
-    param RECTILE[] = [True,False];
+    param PAR[] = [False];
+    param SCREP[] = [True];
+    param IVEC[] = [True];
+    param RECTILE[] = [True];
   } 
  
   def search  
   {
     arg algorithm = 'Exhaustive';  
 #    arg algorithm = 'Simplex';  
-#    arg total_runs = 1; 
+#    arg total_runs = 5; 
   }
 
   def input_params  

### 2008-08-18 doc/papers/cgo08/paper.tex
added authors, updated figures, formatting
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@299 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@
+% CGO 09: http://www.cgo.org/
+% svn co https://svn.mcs.anl.gov/repos/performance/orio/doc/papers/cgo08
+% Max 22 pages, double-spaced, 11pt or larger font
+% Abstract: Sep. 12, 2008
+% Full paper: Sep. 19, 2008
+
 \documentclass[11pt]{article}
@@
 \usepackage{html}
-\usepackage{setspace}
 \usepackage{graphicx}
 \usepackage{subfigure}
+\usepackage{amsmath}
 \usepackage{cite}
+\usepackage{times}
+\usepackage{fullpage}
+\usepackage{setspace}
 \usepackage[margin=1in]{geometry}
+
+\title{Generating Empirically Optimized Numerical Software from MATLAB Prototypes}
+
+\author{
+Elizabeth Jessup,$^1$ Boyana Norris,$^2$ Albert Hartono,$^3$ and Jeremy Siek$^4$\\
+\\
+\vspace{-.05in}
+{\small $^1$University of Colorado at Boulder Dept. of Computer Science, Boulder, CO~80302}\\
+\vspace{-.05in}
+{\small \url{jessup@cs.colorado.edu}}\\
+\vspace{-.05in}
+{\small $^2$Argonne National Laboratory Mathematics and Computer Science Division, Argonne, IL~60439}\\
+\vspace{-.05in}
+{\small \url{norris@mcs.anl.gov}}\\
+\vspace{-.05in}
+{\small $^3$Ohio State University Dept. of Computer Science and Engineering, Columbus, OH 43210}\\
+\vspace{-.05in}
+{\small \url{hartonoa@cse.ohio-state.edu}}\\
+\vspace{-.05in}
+{\small $^4$University of Colorado at Boulder Dept. of Electrical and Computer Engineering, Boulder, CO~80302}\\
+\vspace{-.05in}
+{\small \url{jeremy.siek@colorado.edu}}
+}
 
@@
 
-\title{Generating Empirically Optimized Numerical Software from Matlab Prototypes}
+\maketitle
 
-\abstract{}
-
+\abstract{TODO}
+\doublespacing
 \section{Introduction}
@@
 The holy grail of compilation research is to take an arbitrary code as input and produce optimal code as output for a given language and hardware platform.
-Despite the immense effort that has been poured into this approach, its success has been limited both by practical constraints (users will not tolerate compile times that extend into several days) and by the amount of detailed information the compiler can obtain about the software to be compiled and the hardware on which it is to execute~\cite{Ken99,Wol89,Wol96}.
-
+Despite the immense effort that has been poured into that approach, its success has been limited both by practical constraints (users will not tolerate compile times that extend into several days) and by the amount of detailed information the compiler can obtain about the software to be compiled and the hardware on which it is to execute~\cite{Ken99,Wol89,Wol96}.
 The second path has been to identify kernel routines that constitute the dominant costs of a wide variety of applications.
@@
 The efforts in high-performance library development aptly demonstrate, however, that the research into compiler technologies has not been fully successful.
-If compilers could take the relatively simple code for basic tasks such as a dense matrix-matrix product and produce optimal code, this second approach would be unnecessary.
-
+If compilers could take the relatively simple code even for basic tasks such as a dense matrix-matrix product and produce optimal code, this second approach would be unnecessary.
 Like compiler optimization, the kernel library-oriented approach has significant limitations.
 Support for a variety of architectures is the most serious problem because the optimizations necessary to achieve near-peak performance are by nature not portable.
-This lack of portability is particularly problematic because processor designs are constantly changing,  and constant updates are nearly impossible to maintain by hand.
+Lack of portability is particularly problematic because processor designs are constantly changing,  and constant updates are nearly impossible to maintain by hand.
 It is a sad fact of the computing industry that by the time highly-optimized code is available for a given architecture, that architecture is generally well on its way to obsolescence.
 
-%This doesn't match our paper but I leave it here anyway.
-We believe that addressing these barriers to performance optimization requires a paradigm shift from standard approaches, a shift that augments traditional techniques with an automated process that experimentally tests software on target architectures and modifies the code accordingly.
-
-To begin, we are using matrix algebra operations.
+In this paper, we describe the first steps toward a fully automated system for the optimization of the matrix algebra kernels that are a foundational part of many scientific applications.
+[Ugly] Beginning with a MATLAB prototype, we first generate C or Fortran then run it through Pluto~\cite{} to optimize the sequence of loops and then Pluto~\cite{} to generate a collection of organizations of code and to identify the best choice.
+Comment on manual parts of this.
 
@@
 
-To date, efforts toward automatically tuned software have focused largely on linear algebra.
-These systems all have the same structure.
-A collection of implementations is generated and timed, and the best choice is then identified.
-One such early effort that combined automated code generation and empirical timings for automatic tuning was the PHiPAC (Portable High Performance Ansi C) \cite{lawn111} project.
-PHiPAC generates parameterized code that can be tuned to machine specifications.
-Basic optimizations in the code include manual loop unrolling, explicit removal of unnecessary dependencies in code blocks, and use of machine sympathetic C constructs.
-Search scripts are provided that, for a given code generator, find the best set of parameters for a given architecture and compiler.
-A follow-on to PHiPAC, the Sparsity system \cite{SPARSITY}, allows users to build sparse matrix kernels that are tuned to their matrices and machines automatically.
-Sparsity combines traditional techniques such as loop transformations with data structure transformations and optimization heuristics that are specific to sparse  matrices.
-It is based on a framework for selecting optimization parameters, such as block sizes, using a combination of performance models and search.
+%Modified from recent DOE proposal
 
-Should update to include OSKI.
+The most common approach  to tuning numerical codes is for an expert to transform the source manually, unrolling loops, blocking for multiple levels of cache, and inserting prefetch instructions.
+The pitfalls of this approach are well-understood~\cite{Goedecker01}.
+It requires a significant amount of time and effort.
+Optimizing code for one particular platform may in fact make it less efficient on other platforms and often makes it complex and hard to understand or maintain.
+An alternative is the use of tuned libraries of key numerical algorithms (e.g., BLAS~\cite{Dongarra:1990fk} and LAPACK~\cite{LAPACK} for dense linear algebra, SPARSPAK~\cite{sparspak} for sparse linear algebra, and BLITZ++~\cite{blitz} to support scientific computing in C++, or GPULib by Tech-X, an open source library for accelerating numerical computations on Graphics Processing Units~\cite{gpulib}, to name just a few).
+Such libraries are difficult to develop, and they generally cannot adapt to execution context: as we note in Section~\ref{sec:composed}, a series of calls to library routines can incur high memory costs.
 
-ATLAS produces dense linear algebra kernels that have portable high efficiency across a variety of platforms.
-Like PHiPAC, ATLAS employs a test and search strategy.
-The automatic output of ATLAS is highly efficient implementations of the BLAS and of a limited set of LAPACK routines.
-Even in this narrow incarnation, ATLAS produces valuable results, since these subprograms are the essential computational building blocks of many scientific computing codes and of many general-purpose libraries such as LAPACK or PETSc~\cite{BGC+98}.
-Moreover, ATLAS-based BLAS are immediately and directly reusable in software designed and written for distributed-memory concurrent computers, since optimized kernels are also heavily used locally by each processor in many scientific computing applications.
+Specialized code generators circumvent the high costs of manual code generation .
+They include tools for basic dense linear algebra operations (ATLAS~\cite{WN147}, PhiPAC~\cite{bilmes97optimizing,phipacwww}), sparse linear algebra (OSKI~\cite{vuduc05}), FFTs~\cite{FFTW,Spiral}), stencil-based operations~\cite{kamil06}, and tensor computations (TCE~\cite{TCE}) among others.
+It is often impossible to predict the performance of code precisely on modern computer architectures.
+Thus, many of these specialized code generators exploit search strategies to identify the best (or nearly best) code for a particular choice of problem parameters and machine.
+Such autotuning tools are not general but rather each is valuable in one specific problem domain.
+While not all applications can be written in terms of available libraries, the performance achieved with autotuning tools does convincingly demonstrate the advantage of source-to-source transformation over optimizing compilers.
 
-Another project that lies outside of the domain of linear algebra, is FFTW (Fastest FFT in the West)~\cite{frigo98, frigo99} for tuning discrete fast Fourier transforms.
-FFTW uses the Cooley-Tukey algorithm and relies on explicit recursion in order to improve locality.
-It also follows the test-and-search procedure of the linear algebra systems.
-FFTW creates sequences of instructions for the evaluation of the FFT called {\em plans}.
-Using a dynamic programming algorithm, it then efficiently searches through the plans to identify an optimal one.
+The best performance today is obtained through approaches like that taken by PERI~\cite{PERI}, which involves a labor intensive rewrite of portion of the code by a group of performance tools experts in collaboration with application scientists.
+Maintaining performance of the code as the application and hardware evolves means that the tuning process may need to be repeated.
+We propose to investigate ways to capture this performance tuning expertise in a way that can be integrated into the application.
+We will thus enable performance portability and ensure that the improvements to the application are long-lasting.
+
+% TODO: Paragraph about source-to-source transformation tools, including PLuTo, Orio
 
@@
 
-%These words swiped from the recent DOE proposal.
+Codes based on matrix algebra are generally constructed as a sequence of calls to the Basic Linear Algebra Subprograms (BLAS)~\cite{Dongarra:1988uq,Lawson:1979kx,Dongarra:1990fk} and similar sparse matrix libraries~\cite{Saad:fr,George:1981uq}.
+Writing programs in this way promotes readability and maintainability but can be costly in terms of memory efficiency.
+Specifically, the retrieval of a large order matrix at each routine call can have a profound effect on performance even when highly tuned implementations of the BLAS (e.g ~\cite{Bilmes:1997ye,Whaley:1998fk,IntelMath:oq,ESSL:kl,Goto:2006fk}) are used.
 
-To begin, we will focus on the matrix algebra that is a foundational part of many scientific applications.
-Codes based on matrix algebra are typically constructed as a sequence of calls to the Basic Linear Algebra Subprograms (BLAS)~\cite{Dongarra:1988uq,Lawson:1979kx,Dongarra:1990fk} and similar sparse matrix libraries~\cite{Saad:fr,George:1981uq}.
-Writing programs in this way promotes readability and maintainability but can be costly in terms of memory efficiency.
-While highly tuned implementations of the BLAS are available~\cite{Bilmes:1997ye,Whaley:1998fk,IntelMath:oq,ESSL:kl,Goto:2006fk}, the retrieval of a large order matrix at each routine call can have a profound effect on performance.
-Indeed, the PIs and other researchers have observed that a sequence of BLAS calls can be substantially less efficient than a call to a single specialized routine that performs the same operations (i.e, a {\it composed} BLAS routine)~\cite{Ashby:uq,Blackford:2002vn,Stanley:CSD-98-992,baker03blgmres,baker03lgmres,Dennis:2005tg,Howell:2008,gropp01,Vuduc:2003kl}.
-There are several ways to construct composed BLAS.
-As an example, consider the pair of matrix-vector products
-$q  = Ap$, $s  = A^Tr$, where $A$ is a matrix and $p$, $q$, $r$, and $s$ are vectors, that comprise the computational
-bottleneck of the BiConjugate Gradient Method~\cite{Barrett:1994kx,Saad:2003fk}.
-These two operations can be implemented as a pair of calls to the BLAS routine {\tt GEMV} or they can be rewritten as a single doubly nested loop encompassing both matrix-vector multiplies.
-In the former case, the matrix $A$ is accessed twice while, in the latter, is it accessed only a single time.
+The PIs and other researchers have observed that it can be much more efficient to call a single specialized routine that performs multiple operations than to make successive calls to separate BLAS ~\cite{Ashby:uq,Blackford:2002vn,Stanley:CSD-98-992,baker03blgmres,baker03lgmres,Dennis:2005tg,Howell:2008,gropp01,Vuduc:2003kl}.
+Single routines that carry out more than one linear algebra operation are known as {\t composed} BLAS.
+As an example, consider the pair of matrix-vector products $q  = Ap$, $s  = A^Tr$, where $A$ is a matrix and $p$, $q$, $r$, and $s$ are vectors, that comprise the computational
+bottlenecks of the BiConjugate Gradient Method~\cite{Barrett:1994kx,Saad:2003fk}  and of the GEMVER kernel examined in Section~\ref{sec:matlab}.
+These two operations can be implemented as a pair of calls to the BLAS routine {\tt GEMV} or they can be rewritten as a composed BLAS consisting of the doubly nested loop encompassing both matrix-vector multiplies.
+In the former case, the matrix $A$ is accessed twice while, in the latter, is it accessed only once.
 Our preliminary results indicate that this loop fusion leads to a routine that delivers 90\% more megaflops than does a pair of calls to the highly optimized Goto BLAS~\cite{Goto:2006fk} {\tt GEMV} routines for large matrix orders.
@@
 
-To generate highly optimized code from a MATLAB prototype, we will follow a three-step approach.    To begin, we will use a compiler that converts a MATLAB script into simple C code.
+To generate highly optimized code from a MATLAB prototype, we will follow a three-step approach, illustrated in Figure~\ref{fig:process}.
+To begin, we have developed a compiler that converts a MATLAB script into simple C code.
+We then use the source-to-source automatic parallelization tool  PLuTo\cite{Pluto} to optimize for coarse-grained parallelism and locality simultaneously.
+The result is an automatically generated OpenMP parallel code for multicores.
+Finally, we employ the extensible annotation system Orio to generate many tuned versions of the same operation using different optimization parameters and perform an empirical search for selecting the best among the multiple optimized code variants.
 
-Matlab compiler -> Pluto ->  Orio
+Probably need some words about WHY this 3 pronged approach is a good idea.
 
-\subsection{A Matlab Compiler}
+\begin{figure}[htbp]
+\centering
+\includegraphics[width=.7\textwidth]{figures/process.png}
+
+\caption{Overview of the code generation and tuning process.}
+\label{fig:process}
+\end{figure}
+
+\subsection{A {MATLAB} Compiler}
 \label{sec:matlab}
@@
 \centering
+\includegraphics[width=.7\textwidth]{figures/compile.png}
 
-  \caption{Overview of the Matlab compiler.}
+\caption{Overview of the MATLAB compiler.}
 \label{fig:compiler}
@@
 An example dataflow graph for the GEMVER kernel, defined below, is given in figure~\ref{fig:gemver-dataflow}.
-\begin{eqnarray*}
+\begin{align*}
   A &\gets A + u_1 v_1^T + u_2 v_2^T \\[-0.5ex]
@@
   w &\gets \alpha A x
-\end{eqnarray*}
+\end{align*}
 Each node represents a parameter of the kernel or an operation.
@@
 \centering
+\includegraphics[width=.7\textwidth]{figures/gemver-dataflow.png}
 
@@
 
-\begin{figure*}[btp]
+\begin{figure}[htbp]
 \centering
-
+\includegraphics[width=.9\textwidth]{figures/refine-mv-dot.png}
 \caption{The refinement of a matrix-vector product to a set of inner products and the refinement of inner products to scalar multiplications and additions.}
 \label{fig:refine-mv-dot}
-\end{figure*}
+\end{figure}
 
@@
 However, as mentioned above we are sometimes overly aggressive in applying optimizations.
-To solve this problem we plan to use a combination of analytic methods and empirical
-methods~\cite{Chen:2007vn}.
+To solve this problem we plan to use a combination of analytic methods and empirical methods~\cite{Chen:2007vn}.
 %Yotov:2005kx
@@
 \label{sec:plutonorio}
-% http://www.cse.ohio-state.edu/~bondhugu/pluto/
+% http://www.cse.ohio-state.edu/~bondhugu//
 
-PLuTo~\cite{uday07tr70,uday08cc} is a source-to-source transformation tool for optimizing sequences of (possibly nested) loops.
+PLuTo~\cite{Pluto,uday08cc} is a source-to-source transformation tool for optimizing sequences of (possibly nested) loops.
 
@@
 
-% TODO (Boyana): redo this figure
-
-\subsection{Putting it All Together}
-\label{sec:combining}
+\begin{figure}[htbp]
+\centering
+\includegraphics[width=.7\textwidth]{figures/orio.png}
+\caption{Overview of the Orio empirical tuning process.}
+\label{fig:orio}
+\end{figure}
 
@@
 
-\section{Multicore?}
-\label{sec:multicore}
+\subsection{VADD}
 
-\section{Conclusions}
+\subsection{WAXPBY $v = Ax + By$}
+perhaps add another term
 
+\subsection{GEMVER}
+
+%for (i=0; i<=A_nrows-1; i++)
+%  for (j=0; j<=A_ncols-1; j++)
+%    B[i][j] = A[i][j] + u1[i]*v1[j] + u2[i]*v2[j];
+%
+%for (i=0; i<=A_ncols-1; i++)
+%  for (j=0; j<=A_nrows-1; j++)
+%    x[i] = x[i] + b * B[j][i]*y[j];
+%
+%
+%for (i=0; i<=A_ncols-1; i++)
+%  x[i] = x[i] + z[i];
+%
+%for (i=0; i<=A_nrows-1; i++)
+%  for (j=0; j<=A_ncols-1; j++)
+%    w[i] = w[i] + a * B[i][j]*x[j];
+
+Experiments on Intel Xeon and BGP.
+
+% Placeholder for the new results, the loops tuned for the figure above were almost identical to the ones generated by the MATLAB compiler
+
+\subsection{GESUMMV $y = a * (A * x) + b * (B * x)$}
+
+\subsection{Bicgkernel}
+
+\section{Conclusions and Future Work}
+
+fully automate the process to get rid of the manual steps along the way
+
+extend beyond composed BLAS
+\singlespacing
+\section*{References}
+\bibliographystyle{siam}
 \bibliography{paper}
-\bibliographystyle{siam}
-
 \end{document}

### 2008-08-19 src/module/loop/submodule/composite/transformator.py

@@ -99,35 +99,37 @@ class Transformator:
         '''To apply loop unroll/jamming'''
         
         if isinstance(stmt, module.loop.ast.ExpStmt):
-            return (stmt, False)
+            return (stmt, [])
 
         elif isinstance(stmt, module.loop.ast.CompStmt):
             tstmts = []
-            body_unrolled = False
+            unrolled_loop_infos = []
             for s in stmt.stmts:
-                t,b = self.__unrollJam(s, tinfos)
+                t,l = self.__unrollJam(s, tinfos)
                 tstmts.append(t)
-                body_unrolled = body_unrolled or b
+                unrolled_loop_infos.extend(l)
             stmt.stmts = tstmts
-            return (stmt, body_unrolled)
+            return (stmt, unrolled_loop_infos)
             
         elif isinstance(stmt, module.loop.ast.IfStmt):
-            t,body_unrolled = self.__unrollJam(stmt.true_stmt, tinfos)
+            unrolled_loop_infos = []
+            t,l = self.__unrollJam(stmt.true_stmt, tinfos)
             stmt.true_stmt = t
+            unrolled_loop_infos.extend(l)
             if stmt.false_stmt:
-                t,b = self.__unrollJam(stmt.false_stmt, tinfos)
+                t,l = self.__unrollJam(stmt.false_stmt, tinfos)
                 stmt.false_stmt = t
-                body_unrolled = body_unrolled or b
-            return (stmt, body_unrolled)
-                
+                unrolled_loop_infos.extend(l)
+            return (stmt, unrolled_loop_infos)
+
         elif isinstance(stmt, module.loop.ast.ForStmt):
+            t,l = self.__unrollJam(stmt.stmt, tinfos)
+            stmt.stmt = t
+            unrolled_loop_infos = l[:]
 
-            # recursively transform the loop body
-            stmt.stmt, body_unrolled = self.__unrollJam(stmt.stmt, tinfos)
-
-            # apply unroll/jamming
             for_loop_info = self.flib.extractForLoopInfo(stmt)
             index_id, lbound_exp, ubound_exp, stride_exp, loop_body = for_loop_info
+            
             ufactor = -1
             for lid, uf in tinfos:
                 if lid == index_id.name:
@@ -135,13 +137,17 @@ class Transformator:
                     break
 
             if ufactor > 1:
-                do_jamming = not body_unrolled
+                do_jamming = True
+                for lbexp, ubexp, stexp in unrolled_loop_infos:
+                    do_jamming = (do_jamming and 
[…]

### 2008-08-19 testsuite/fdtd/cookie/fdtd-2d.src1.c

@@ -11,25 +11,25 @@
 #   [4,8,16,32,64,128];
 #   [1,4,8,16];
 
-    param T1_1[] = [16];
+    param T1_1[] = [32];
     param T1_2[] = [32];
     param T1_3[] = [32];
-    param T2_1[] = [1];
-    param T2_2[] = [1];
-    param T2_3[] = [1];
+    param T2_1[] = [8];
+    param T2_2[] = [8];
+    param T2_3[] = [8];
 
 #    constraint c1 = (T1_1*T2_1<=1024 and T1_2*T2_2<=1024 and T1_3*T2_3<=1024);
 #    constraint c2 = (T1_1==T1_3 and T2_1==T2_3);
     
-    param U1[] = [8];
-    param U2[] = [2];
-    param U3[] = [4];
+    param U1[] = [1];
+    param U2[] = [1];
+    param U3[] = [1];
 
     constraint c3 = (U1*U2*U3<=256);
 
     param PERM[] = [
-#      [0,1,2],
-      [0,2,1],
+      [0,1,2],
+#      [0,2,1],
 #      [1,0,2],
 #      [1,2,0],
 #      [2,0,1],
@@ -37,9 +37,9 @@
 
     ];
 
-    param PAR[] = [True];
+    param PAR[] = [False];
     param SCREP[] = [False];
-    param IVEC[] = [True];
+    param IVEC[] = [False];
     param RECTILE[] = [False];
   }
   

### 2008-08-19 testsuite/fdtd/cookie/exp/fdtd-2d.pluto.seq.l2tile.c
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@302 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@ -1,0 +1,590 @@
+
+#include <stdio.h>
+#include <stdlib.h>
+#include <sys/time.h>
+#include <math.h>
+
+#define tmax T
+#define nx N
+#define ny N
+double ex[nx][ny +1];
+double ey[nx +1][ny];
+double hz[nx][ny];
+
+void init_arrays()
+{
+    int i, j;
+    for (i=0; i<nx+1; i++)  {
+        for (j=0; j<ny; j++)  {
+            ey[i][j] = 0;
+        }
+    }
+    for (i=0; i<nx; i++)  {
+        for (j=0; j<ny+1; j++)  {
+            ex[i][j] = 0;
+        }
+    }
+    for (j=0; j<ny; j++)  {
+        ey[0][j] = ((double)j)/ny;
+    }
+    for (i=0; i<nx; i++)    {
+        for (j=0; j<ny; j++)  {
+            hz[i][j] = 0;
+        }
+    }
+}
+
+double rtclock()
+{
+  struct timezone tzp;
+  struct timeval tp;
+  int stat;
+  gettimeofday (&tp, &tzp);
+  return (tp.tv_sec + tp.tv_usec*1.0e-6);
+}
+
+int main()
+{
+  init_arrays();
+
+  double annot_t_start=0, annot_t_end=0, annot_t_total=0;
+  int annot_i;
+
+  for (annot_i=0; annot_i<REPS; annot_i++)
+  {
+    annot_t_start = rtclock();
+    
+#include <math.h>
+#include <assert.h>
+#define ceild(n,d)  ceil(((double)(n))/((double)(d)))
+#define floord(n,d) floor(((double)(n))/((double)(d)))
+#define max(x,y)    ((x) > (y)? (x) : (y))
+#define min(x,y)    ((x) < (y)? (x) : (y))
+
+
+
+
+
+ int c1, c2, c3, c4, c5, c6, c7, c8, c9;
+ register int lbv, ubv;
+
+for (c1=0;c1<=floord(tmax-1,256);c1++) {
[…]

### 2008-08-19 testsuite/fdtd/cookie/exp/runscript.py
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@303 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@ -99,11 +99,13 @@ def checkCorrectness(optflag = '-O0'):
     f.close()
     
     fnames = [
-        #'fdtd-2d.pluto.seq.c', 
-        #'fdtd-2d.pluto.par.c', 
-        #'fdtd-2d.orio.seq.small.c',
-        #'fdtd-2d.orio.seq.large.c',
-        #'fdtd-2d.orio.par.c',
+        'fdtd-2d.pluto.seq.l1tile.c', 
+        #'fdtd-2d.pluto.seq.l2tile.c', 
+        'fdtd-2d.pluto.par.l1tile.c', 
+        #'fdtd-2d.pluto.par.l2tile.c', 
+        'fdtd-2d.orio.seq.small.c',
+        'fdtd-2d.orio.seq.large.c',
+        'fdtd-2d.orio.par.c',
         ]
     for fname in fnames:
         compile_cmd = (('gcc %s -fopenmp -DREPS=1 -DT=%s -DN=%s -DTEST -o opt_test %s -lm') % 
@@ -149,23 +151,29 @@ if 1:
     flags = '-DREPS=%s -DT=%s -DN=%s' % (reps, T, N)
 
     mflopss_base = []
-    mflopss_pluto = []
+    mflopss_pluto_l1 = []
+    mflopss_pluto_l2 = []
     mflopss_orio = []
 
-    #rtimes_base = runExp([1,2,3,4,5,6,7,8], 'icc %s -parallel' % OPTFLAG, 
-    #                     'fdtd-2d.base.c', 'base_par', flags, '-lm')
-    #mflopss_base = countFlops(T,N,rtimes_base)
+    rtimes_base = runExp([1,2,3,4,5,6,7,8], 'icc %s -parallel' % OPTFLAG, 
+                         'fdtd-2d.base.c', 'base_par', flags, '-lm')
+    mflopss_base = countFlops(T,N,rtimes_base)
     
-    rtimes_pluto = runExp([1,2,3,4,5,6,7,8], 'icc %s -openmp' % OPTFLAG, 
-                          'fdtd-2d.pluto.par.c', 'pluto_par', flags, '-lm')
-    mflopss_pluto = countFlops(T,N,rtimes_pluto)
+    rtimes_pluto_l1 = runExp([1,2,3,4,5,6,7,8], 'icc %s -openmp' % OPTFLAG, 
[…]

### 2008-08-19 testsuite/seidel/cookie/exp/runscript.py

@@ -99,8 +99,10 @@ def checkCorrectness(optflag = '-O0'):
     f.close()
     
     fnames = [
-        'seidel.pluto.seq.c', 
-        'seidel.pluto.par.c', 
+        'seidel.pluto.seq.l1tile.c', 
+        'seidel.pluto.seq.l2tile.c', 
+        'seidel.pluto.par.l1tile.c', 
+        'seidel.pluto.par.l2tile.c', 
         'seidel.orio.seq.c',
         'seidel.orio.par.c',
         ]
@@ -148,23 +150,29 @@ if 1:
     flags = '-DREPS=%s -DT=%s -DN=%s' % (reps, T, N)
 
     mflopss_base = []
-    mflopss_pluto = []
+    mflopss_pluto_l1 = []
+    mflopss_pluto_l2 = []
     mflopss_orio = []
 
-    rtimes_base = runExp([1,2,3,4,5,6,7,8], 'icc %s -parallel' % OPTFLAG, 
-                         'seidel.base.c', 'base_par', flags, '-lm')
-    mflopss_base = countFlops(T,N,rtimes_base)
+    #rtimes_base = runExp([1,2,3,4,5,6,7,8], 'icc %s -parallel' % OPTFLAG, 
+    #                     'seidel.base.c', 'base_par', flags, '-lm')
+    #mflopss_base = countFlops(T,N,rtimes_base)
     
-    rtimes_pluto = runExp([1,2,3,4,5,6,7,8], 'icc %s -openmp' % OPTFLAG, 
-                          'seidel.pluto.par.c', 'pluto_par', flags, '-lm')
-    mflopss_pluto = countFlops(T,N,rtimes_pluto)
+    #rtimes_pluto_l1 = runExp([1,2,3,4,5,6,7,8], 'icc %s -openmp' % OPTFLAG, 
+    #                         'seidel.pluto.par.l1tile.c', 'pluto_par', flags, '-lm')
+    #mflopss_pluto_l1 = countFlops(T,N,rtimes_pluto_l1)
     
-    rtimes_orio = runExp([1,2,3,4,5,6,7,8], 'icc %s -openmp' % OPTFLAG, 
-                         'seidel.orio.par.c', 'orio_par', flags, '-lm')
-    mflopss_orio = countFlops(T,N,rtimes_orio)
+    rtimes_pluto_l2 = runExp([1,2,3,4,5,6,7,8], 'icc %s -openmp' % OPTFLAG, 
[…]

### 2008-08-19 testsuite/seidel/cookie/exp/runscript.py

@@ -107,7 +107,7 @@ def checkCorrectness(optflag = '-O0'):
         'seidel.orio.par.c',
         ]
     for fname in fnames:
-        compile_cmd = (('gcc %s -fopenmp -DREPS=1 -DT=%s -DN=%s -DTEST -o opt_test %s -lm') % 
+        compile_cmd = (('icc %s -openmp -DREPS=1 -DT=%s -DN=%s -DTEST -o opt_test %s -lm') % 
                        (optflag, T, N, fname))
         run_cmd = 'export OMP_NUM_THREADS=1; ./opt_test'
         print '***********************'

### 2008-08-19 testsuite/fdtd/cookie/exp/runscript.py
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@306 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@ -82,7 +82,7 @@ def myDiff(fname1, fname2):
     return (total_diffs==0, diffs, total_diffs, total_nums)
 
 def checkCorrectness(optflag = '-O0'):
-    T=100
+    T=10
     N=500
     compile_cmd = 'gcc -O0 -DREPS=1 -DT=%s -DN=%s -DTEST -o base_test fdtd-2d.base.c -lm' % (T,N)
     run_cmd = 'export OMP_NUM_THREADS=1; ./base_test'
@@ -108,7 +108,7 @@ def checkCorrectness(optflag = '-O0'):
         'fdtd-2d.orio.par.c',
         ]
     for fname in fnames:
-        compile_cmd = (('gcc %s -fopenmp -DREPS=1 -DT=%s -DN=%s -DTEST -o opt_test %s -lm') % 
+        compile_cmd = (('icc %s -openmp -DREPS=1 -DT=%s -DN=%s -DTEST -o opt_test %s -lm') % 
                        (optflag, T, N, fname))
         run_cmd = 'export OMP_NUM_THREADS=1; ./opt_test'
         print '***********************'

### 2008-08-19 doc/papers/cgo08/f77/bicgkernel.f
f77 cods
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@307 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@ -1,0 +1,31 @@
+      subroutine bicgkernel(lda, n, A, p, r, s, q)
+      integer lda, n
+      double precision zero, one
+      double precision p(*), r(*)
+      double precision A(lda, *)
+c
+c   BICG
+c   in
+c     A : column matrix, p : vector, r : vector
+c   out
+c     q : vector, s : vector
+c   {
+c     q = A * p
+c     s = A' * r
+c   }
+c
+      incx = 1
+      incy = 1
+      one = 1.0d0
+      zero = 0.0d0
+c
+c    Put A*p in q 
+c
+      call dgemv( 'n', n, n, one, a, lda, p, incx, zero, q, incy )	
+c
+c    Put A'*r in s 
+c
+      call dgemv( 't', n, n, one, a, lda, r, incx, zero, s, incy )	
+c
+      return
+      end

### 2008-08-19 doc/papers/cgo08/f77/gesummv.f
got rid of prints
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@308 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@ -24,13 +24,11 @@ c   Put beta*B*x into y.
 c
       zero = 0.0d0
       call dgemv( 'n', n, n, beta, b, lda, x, incx, zero, y, incy )
-      print*,'first y',(y(j),j=1,n)
 c
 c   Put y = beta*B*x + alpha*A*x into y.
 c
       one = 1.0d0
       call dgemv( 'n', n, n, alpha, a, lda, x, incx, one, y, incy )
-      print*,'second y',(y(j),j=1,n)
 c
       return
       end

### 2008-08-19 doc/papers/cgo08/f77/vadd.f
fix dims
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@309 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@ -1,7 +1,7 @@
-      subroutine vadd(n, x, w, y, z) 
+      subroutine vadd(n, x, w, y, z, yy) 
       integer n, incx, incy
       double precision one
-      double precision yy(n) 
+      double precision yy(*) 
       double precision w(*), x(*), y(*), z(*)
 c
 c   vadd

### 2008-08-19 doc/papers/cgo08/f77/waxpby.f

@@ -1,7 +1,7 @@
-      subroutine waxpby(n, w, alpha, x, beta, y)
+      subroutine waxpby(n, w, alpha, x, beta, y, yy)
       integer n, incx, incy
       double precision alpha
-      double precision yy(n)
+      double precision yy(*)
       double precision w(*), x(*), y(*)
 c
 c   WAXPBY

### 2008-08-19 doc/papers/cgo08/f77/testprogram.f
makefile and test
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@311 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@ -1,0 +1,82 @@
+      program testprogram
+      integer n
+      parameter (lda = 5, ldb = 5, n = 5)
+c
+      integer j
+      double precision alpha, beta
+      double precision w(n), x(n), y(n), z(n), xorig(n), q(n), s(n)
+      double precision u1(n), u2(n), v1(n), v2(n), worig(n), yorig(n)
+      double precision A(lda, n), B(lda, n)
+      data alpha/22/, beta/7/
+      data worig/1, 2, 3, 4, 5/, xorig/5, 10, 15, 20, 25/
+      data z/2, 4, 6, 8, 10/, yorig/0, 3, 5, 7, 9/
+      data u1/1, 2, 3, 4, 5/, u2/5, 10, 15, 20, 25/, v1/0, 3, 5, 7, 9/
+      data v2/2, 4, 6, 8, 10/ 
+      data A/1,2,3,4,5,2,3,4,5,6,0,1,0,1,1,1,1,1,1,1,2,1,2,1,2/
+      data B/10,2,3,4,5,2,3,4,5,6,0,1,0,1,1,1,1,1,1,1,2,1,2,1,20/
+c      
+c    x = w + y + z
+c      
+      do j = 1,n
+        x(j) = xorig(j)
+        y(j) = yorig(j)
+        w(j) = worig(j)
+      enddo
+      call vadd(n, x, w, y, z)
+      print*,'vadd result x',(x(j),j=1,n)
+      print*,'vadd result should be (3, 9, 14, 19, 24)' 
+      print*,' '
+      do j = 1,n
+        x(j) = xorig(j)
+      enddo
+c
+c     w = alpha * x + beta * y
+c
+      do j = 1,n
+        x(j) = xorig(j)
+        w(j) = worig(j)
+        y(j) = yorig(j)
+      enddo
+      call waxpby(n, w, alpha, x, beta, y)
+      print*,'waxpby result w',(w(j),j=1,n)
+      print*,'vadd result should be (110   241   365   489   613)'
+      print*,' '
+      do j = 1,n
+        w(j) = worig(j)
+      enddo
+c
+c     q = A * x
+c     s = A' * y
+c
+      call bicgkernel(lda, n, A, x, y, s, q)
+      print*,'bicgkernel result q',(q(j),j=1,n)
+      print*,'bicgkernel result s',(s(j),j=1,n)
+      print*,'bicgkernel result should be (95 100 125 130 170)'
[…]

### 2008-08-19 doc/papers/cgo08/f77/gemver.f
add
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@312 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@ -5,6 +5,8 @@
       double precision z(n)
       double precision A(lda,*), B(ldb,*)
 c
+c  NOTE COLUMN ORIENTED MATRIX COPY BELOW.
+c
 c   GEMVER
 c   in
 c     A : column matrix, u1 : vector, u2 : vector, v1 : vector, v2 : vector,

### 2008-08-20 testsuite/lu/cookie/exp/lu.orio.seq.small.c
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@313 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@ -1,0 +1,317 @@
+#include <stdio.h>
+#include <stdlib.h>
+#include <sys/time.h>
+#include <math.h>
+
+#define ceild(n,d)  ceil(((double)(n))/((double)(d)))
+#define floord(n,d) floor(((double)(n))/((double)(d)))
+#define max(x,y)    ((x) > (y)? (x) : (y))
+#define min(x,y)    ((x) < (y)? (x) : (y))
+
+double L[N][N];
+double U[N][N];
+double A[N][N+13];
+
+void init_arrays()
+{
+  int i, j, k;
+
+  /* have to initialize this matrix properly to prevent                                              
+   * division by zero                                                                                 
+   */
+  for (i=0; i<N; i++) {
+    for (j=0; j<N; j++) {
+      L[i][j] = 0.0;
+      U[i][j] = 0.0;
+    }
+  }
+
+  for (i=0; i<N; i++) {
+    for (j=0; j<=i; j++) {
+      L[i][j] = i+j+1;
+      U[j][i] = i+j+1;
+    }
+  }
+
+  for (i=0; i<N; i++) {
+    for (j=0; j<N; j++) {
+      for (k=0; k<N; k++) {
+	A[i][j] += L[i][k]*U[k][j];
+      }
+    }
+  }
+}
+
+double rtclock()
+{
+  struct timezone tzp;
+  struct timeval tp;
+  int stat;
+  gettimeofday (&tp, &tzp);
+  return (tp.tv_sec + tp.tv_usec*1.0e-6);
+}
+
+int main()
+{
+  init_arrays();
+
+  double annot_t_start=0, annot_t_end=0, annot_t_total=0;
+  int annot_i;
+
+  for (annot_i=0; annot_i<REPS; annot_i++)
+    {
+      annot_t_start = rtclock();
+
+
+
+/*@ begin PerfTuning (         
+  def build 
+  { 
[…]

### 2008-08-20 testsuite/lu/cookie/exp/template.begin.c

@@ -1,0 +1,66 @@
+#include <stdio.h>
+#include <stdlib.h>
+#include <sys/time.h>
+#include <math.h>
+
+#define ceild(n,d)  ceil(((double)(n))/((double)(d)))
+#define floord(n,d) floor(((double)(n))/((double)(d)))
+#define max(x,y)    ((x) > (y)? (x) : (y))
+#define min(x,y)    ((x) < (y)? (x) : (y))
+
+double L[N][N];
+double U[N][N];
+double A[N][N+13];
+
+void init_arrays()
+{
+  int i, j, k;
+
+  /* have to initialize this matrix properly to prevent                                              
+   * division by zero                                                                                 
+   */
+  for (i=0; i<N; i++) {
+    for (j=0; j<N; j++) {
+      L[i][j] = 0.0;
+      U[i][j] = 0.0;
+    }
+  }
+
+  for (i=0; i<N; i++) {
+    for (j=0; j<=i; j++) {
+      L[i][j] = i+j+1;
+      U[j][i] = i+j+1;
+    }
+  }
+
+  for (i=0; i<N; i++) {
+    for (j=0; j<N; j++) {
+      for (k=0; k<N; k++) {
+	A[i][j] += L[i][k]*U[k][j];
+      }
+    }
+  }
+}
+
+double rtclock()
+{
+  struct timezone tzp;
+  struct timeval tp;
+  int stat;
+  gettimeofday (&tp, &tzp);
+  return (tp.tv_sec + tp.tv_usec*1.0e-6);
+}
+
+int main()
+{
+  init_arrays();
+
+  double annot_t_start=0, annot_t_end=0, annot_t_total=0;
+  int annot_i;
+
+  for (annot_i=0; annot_i<REPS; annot_i++)
+    {
+      annot_t_start = rtclock();
+
+
+

### 2008-08-20 testsuite/lu/cookie/lu.src2.c

@@ -1,0 +1,74 @@
+/*@ begin PerfTuning (         
+  def build 
+  { 
+    arg build_command = 'icc -O3 -openmp -I/usr/local/icc/include -lm'; 
+  } 
+    
+  def performance_counter          
+  { 
+    arg repetitions = 50; 
+  }
+
+  def performance_params 
+  {
+    param PERM[] = [
+     ['i','j'],
+#     ['j','i'],
+    ];
+
+    param U1[] = [1];
+    param U2[] = [1];
+    param U3[] = [1];
+    param U4[] = [24];
+
+    param IVEC1[] = [False];
+    param IVEC2[] = [False];
+    param SCREP[] = [False];
+  }
+
+  def search 
+  { 
+    arg algorithm = 'Exhaustive'; 
+#    arg algorithm = 'Simplex'; 
+#    arg total_runs = 1;
+  } 
+   
+  def input_params 
+  {
+    param N[] = [500];
+  }
+
+  def input_vars
+  {
+    arg decl_file = 'decl_code.h';
+    arg init_file = 'init_code.c';
+  }
+) @*/ 
+
+register int i,j,k,it,jt,kt;
+
+/*@ begin Loop (
+transform Unroll(ufactor=U4)
+for (k=0; k<=N-1; k++) {
+
+  transform Composite (
+  scalarreplace = (SCREP, 'double'),
+  regtile = (['j'],[U1]),
+  vector = (IVEC1, ['ivdep','vector always']))
+    for (j=k+1; j<=N-1; j++)
+      A[k][j] = A[k][j]/A[k][k];
+
+  transform Composite (
+  permut = [PERM],
+  scalarreplace = (SCREP, 'double'),
+  regtile = (['i','j'],[U2,U3]),
+  vector = (IVEC2, ['ivdep','vector always']))
+    for(i=k+1; i<=N-1; i++)
+      for (j=k+1; j<=N-1; j++)
+        A[i][j] = A[i][j] - A[i][k]*A[k][j];
+  }
+) @*/
+/*@ end @*/
+
+/*@ end @*/
+

### 2008-08-20 testsuite/lu/cookie/lu_profiling.c

@@ -1,0 +1,46 @@
+#include <stdio.h>
+#include <stdlib.h>
+
+#define N 600
+double L[N][N];
+double U[N][N];
+double A[N][N+13];
+
+void init_array()
+{
+  int i, j, k;
+
+  /* have to initialize this matrix properly to prevent                                              
+   * division by zero                                                                                 
+   */
+  for (i=0; i<N; i++) {
+    for (j=0; j<N; j++) {
+      L[i][j] = 0.0;
+      U[i][j] = 0.0;
+    }
+  }
+
+  for (i=0; i<N; i++) {
+    for (j=0; j<=i; j++) {
+      L[i][j] = i+j+1;
+      U[j][i] = i+j+1;
+    }
+  }
+
+  for (i=0; i<N; i++) {
+    for (j=0; j<N; j++) {
+      for (k=0; k<N; k++) {
+	A[i][j] += L[i][k]*U[k][j];
+      }
+    }
+  }
+}
+
+int main()
+{
+  init_array();
+
+  /*@ profiled code @*/
+
+  return A[0][0];   //needed to avoid dead code elimination
+}

### 2008-08-20 testsuite/lu/cookie/decl_code.h
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@313 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@ -1,0 +1,3 @@
+double L[N][N];
+double U[N][N];
+double A[N][N+13];

### 2008-08-20 testsuite/lu/cookie/lu.src1.c

@@ -1,0 +1,99 @@
+/*@ begin PerfTuning (         
+  def build 
+  { 
+    arg build_command = 'icc -O3 -openmp -lm'; 
+  } 
+    
+  def performance_counter          
+  { 
+    arg repetitions = 1; 
+  }
+
+  def performance_params 
+  {
+    param T1_1[] = [16];
+    param T1_2[] = [512];
+    param T1_3[] = [16];
+    param T2_1[] = [16];
+    param T2_2[] = [1];
+    param T2_3[] = [16];
+
+    #constraint c1 = (T1_1*T2_1<=1024 and T1_1*T2_1<=1024 and T1_1*T2_1<=1024);
+    constraint c2 = ((T1_1 == T1_3) and (T2_1 == T2_3));
+
+    param U1[] = [7];
+    param U2[] = [1];
+    param U3[] = [5];
+
+    constraint c3 = (U1*U2*U3<=256);
+
+    param PERM[] = [
+#      [0,1,2],
+#      [0,2,1],
+#      [1,0,2],
+#      [1,2,0],
+      [2,0,1],
+#      [2,1,0],
+      ];
+
+    param PAR[] = [False];
+    param SCREP[] = [True];
+    param IVEC[] = [True];
+    param RECTILE[] = [False];
+  }
+
+  def search 
+  { 
+    arg algorithm = 'Exhaustive'; 
+#    arg algorithm = 'Simplex'; 
+#    arg total_runs = 1;
+  } 
+   
+  def input_params 
+  {
+    param N[] = [2000];
+  }
+
+  def input_vars
+  {
+    arg decl_file = 'decl_code.h';
+    arg init_file = 'init_code.c';
+  }
+) @*/ 
+
+register int i,j,k;
+register int c1t, c2t, c3t, c4t, c5t, c6t, c7t, c8t, c9t, c10t, c11t, c12t;
+register int newlb_c1, newlb_c2, newlb_c3, newlb_c4, newlb_c5, newlb_c6,
+  newlb_c7, newlb_c8, newlb_c9, newlb_c10, newlb_c11, newlb_c12;
+register int newub_c1, newub_c2, newub_c3, newub_c4, newub_c5, newub_c6,
+  newub_c7, newub_c8, newub_c9, newub_c10, newub_c11, newub_c12;
+
+
+/*@ begin PolySyn(    
[…]

### 2008-08-20 testsuite/gemver/cookie/exp/gemver.orio.par.c
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@314 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@ -1,0 +1,136 @@
+
+#include <stdio.h>
+#include <stdlib.h>
+#include <sys/time.h>
+#include <math.h>
+
+#define ceild(n,d)  ceil(((double)(n))/((double)(d)))
+#define floord(n,d) floor(((double)(n))/((double)(d)))
+#define max(x,y)    ((x) > (y)? (x) : (y))
+#define min(x,y)    ((x) < (y)? (x) : (y))
+
+double a;
+double b;
+int nx;
+int ny;
+double A[Nx][Ny];
+double B[Nx][Ny];
+double u1[Nx];
+double u2[Nx];
+double v1[Ny];
+double v2[Ny];
+double y[Nx];
+double z[Ny];
+double w[Nx];
+double x[Ny];
+
+void init_arrays()
+{
+  int i, j;
+  a = 1.5;
+  b = 2.5;
+  nx = Nx;
+  ny = Ny;
+  for (i=0; i<=Nx-1; i++) {
+    u1[i]=(i+1)/Nx/2.0;
+    u2[i]=(i+1)/Nx/4.0;
+    y[i]=(i+1)/Nx/6.0;
+    w[i]=(i+1)/Nx/8.0;
+    for (j=0; j<=Ny-1; j++) {
+      A[i][j]=(i*j)/Ny;
+      B[i][j]=0;
+    }
+  }
+  for (j=0; j<=Ny-1; j++) {
+    v1[j]=(j+1)/Ny/2.0;
+    v2[j]=(j+1)/Ny/4.0;
+    z[j]=(j+1)/Ny/6.0;
+    x[j]=(j+1)/Ny/8.0;
+  }
+}
+
+double rtclock()
+{
+  struct timezone tzp;
+  struct timeval tp;
+  int stat;
+  gettimeofday (&tp, &tzp);
+  return (tp.tv_sec + tp.tv_usec*1.0e-6);
+}
+
+int main()
+{
+  init_arrays();
+
+  double annot_t_start=0, annot_t_end=0, annot_t_total=0;
+  int annot_i;
+
+  for (annot_i=0; annot_i<REPS; annot_i++)
+  {
+    annot_t_start = rtclock();
+
+
+
+
+int i,j,ii,jj,it,jt;
+
+for (i=0; i<=min(nx-1,ny-1); i=i+1) {
+  x[i]=0;
+  w[i]=0;
+}
+for (i=nx; i<=ny-1; i=i+1) 
+  x[i]=0;
[…]

### 2008-08-20 testsuite/gemver/cookie/exp/runscript.py

@@ -1,0 +1,203 @@
+#!/usr/bin/env python
+
+import os, sys, re
+
+def printFloats(ls):
+    s=''
+    s+='['
+    for i,f in enumerate(ls):
+        if i: s+=', '
+        s+='%2.3f' % f
+    s+=']'
+    print s
+
+def runExp(nthreadss, cc, src_file, out_file, flags, libs):
+    compile_cmd = '%s %s -o %s %s %s' % (cc, flags, out_file, src_file, libs)
+    print '*************************************'
+    print compile_cmd
+    print '*************************************'
+    os.system(compile_cmd)
+
+    rtimes = []
+    for nt in nthreadss:
+        run_cmd = 'export OMP_NUM_THREADS=%s; ./%s' % (nt, out_file)
+        print '*************************************'
+        print run_cmd
+        print '*************************************'
+        f = os.popen(run_cmd)
+        output = f.read()
+        f.close()
+        rtime = eval(output)
+        rtimes.append(rtime)
+    return rtimes
+
+def countFlops(N, rtimes):
+    ops = (10*N*N+N)
+    mflopss = []
+    for rtime in rtimes:
+        mflops = 1.0*ops/(rtime*1000000)
+        mflopss.append(mflops)
+    return mflopss
+
+def myDiff(fname1, fname2):
+    f1 = open(fname1)
+    c1 = f1.read()
+    f1.close()
+    f2 = open(fname2)
+    c2 = f2.read()
+    f2.close()
+    cls1 = c1.split('\n')
+    cls2 = c2.split('\n')
+    maxlength=100
+    diffs = []
+    total_diffs = 0
+    total_nums = 0
+    for l1,l2 in zip(cls1, cls2):
+        l1 = l1.split()
+        l2 = l2.split()
+        for n1,n2 in zip(l1,l2):
+            total_nums+=1
+            try:
+                n1=eval(n1)
+                n2=eval(n2)
+            except:
+                print 'error: cannot evaluate'
+                print 'n1=%s' % n1
+                print 'n2=%s' % n2
+                sys.exit(1)
+            d=n1-n2
+            if not (-1<d<1):
+                total_diffs+=1
+                if len(diffs) == 0:
+                    diffs.append(d)
+                    continue
+                maxdiff=diffs[-1]
+                if len(diffs) < maxlength:
+                    diffs.append(d)
[…]

### 2008-08-20 testsuite/gemver/cookie/gemver.src.c

@@ -1,0 +1,114 @@
+/*@ begin PerfTuning (  
+  def build
+  { 
+  arg build_command = 'icc -O3 -openmp -lm'; 
+  } 
+  
+  def performance_counter  
+  { 
+  arg repetitions = 5;
+  }
+  
+  def performance_params 
+  {
+  param U1[]  = [1];
+  param U2[]  = [1];
+  param U3[]  = [1];
+  param U4i[] = [1];
+  param U4j[] = [1];
+  param U5[]  = [1];
+  param U6i[] = [1];
+  param U6j[] = [1];
+  param U7[]  = [1];
+
+  param PAR1[] = [False];
+  param PAR2[] = [False];
+  param PAR3[] = [False];
+  param PAR4[] = [False];
+  param PAR5[] = [False];
+  param PAR6[] = [True];
+  param PAR7[] = [False];
+
+  param SCR4[] = [False];
+
+  param VEC4[] = [True];
+  param VEC6[] = [False];
+
+  }
+			 
+  def search 
+  { 
+    arg algorithm = 'Exhaustive'; 
+#    arg algorithm = 'Simplex'; 
+#    arg total_runs = 1;
+  } 
+  
+  def input_params 
+  {
+  let N = [10000];
+  param Nx[] = N;
+  param Ny[] = N;
+  }
+  
+  def input_vars
+  {
+  arg decl_file = 'decl_code.h';
+  arg init_file = 'init_code.c';
+  }
+) @*/
+
+#define max(x,y)    ((x) > (y)? (x) : (y))
+#define min(x,y)    ((x) < (y)? (x) : (y))
+
+int i,j,ii,jj,it,jt;
+
+/*@ begin Loop(
+
+transform UnrollJam(ufactor=U1, parallelize=PAR1)
+for (i=0;i<=min(nx-1,ny-1);i++) {
+  x[i]=0; 
+  w[i]=0;
+}
+
+transform UnrollJam(ufactor=U2, parallelize=PAR2)
+for (i=nx;i<=ny-1;i++)
+  x[i]=0;
+
+transform UnrollJam(ufactor=U3, parallelize=PAR3)
+for (i=ny;i<=nx-1;i++)
+  w[i]=0;
+
+transform Composite(
+ scalarreplace = (SCR4, 'double'),
[…]

### 2008-08-20 testsuite/gemver/cookie/init_code.c

@@ -1,0 +1,26 @@
+
+void init_input_vars() {
+
+  int i, j;
+  a = 1.5;
+  b = 2.5;
+  nx = Nx;
+  ny = Ny;
+  for (i=0; i<=Nx-1; i++) {
+    u1[i]=(i+1)/Nx/2.0;
+    u2[i]=(i+1)/Nx/4.0;
+    y[i]=(i+1)/Nx/6.0;
+    w[i]=(i+1)/Nx/8.0;
+    for (j=0; j<=Ny-1; j++) {
+      A[i][j]=(i*j)/Ny;
+      B[i][j]=0;
+    }
+  }
+  for (j=0; j<=Ny-1; j++) {
+    v1[j]=(j+1)/Ny/2.0;
+    v2[j]=(j+1)/Ny/4.0;
+    z[j]=(j+1)/Ny/6.0;
+    x[j]=(j+1)/Ny/8.0;
+  }
+
+}

### 2008-08-20 testsuite/gemver/cookie/decl_code.h

@@ -1,0 +1,18 @@
+
+
+double a;
+double b;
+int nx;
+int ny;
+double A[Nx][Ny];
+double B[Nx][Ny];
+double u1[Nx];
+double u2[Nx];
+double v1[Ny];
+double v2[Ny];
+double y[Nx];
+double z[Ny];
+double w[Nx];
+double x[Ny];
+
+

### 2008-08-20 testsuite/gemver/cookie/exp/gemver.base.c

@@ -1,0 +1,123 @@
+
+#include <stdio.h>
+#include <stdlib.h>
+#include <sys/time.h>
+#include <math.h>
+
+#define ceild(n,d)  ceil(((double)(n))/((double)(d)))
+#define floord(n,d) floor(((double)(n))/((double)(d)))
+#define max(x,y)    ((x) > (y)? (x) : (y))
+#define min(x,y)    ((x) < (y)? (x) : (y))
+
+double a;
+double b;
+int nx;
+int ny;
+double A[Nx][Ny];
+double B[Nx][Ny];
+double u1[Nx];
+double u2[Nx];
+double v1[Ny];
+double v2[Ny];
+double y[Nx];
+double z[Ny];
+double w[Nx];
+double x[Ny];
+
+void init_arrays()
+{
+  int i, j;
+  a = 1.5;
+  b = 2.5;
+  nx = Nx;
+  ny = Ny;
+  for (i=0; i<=Nx-1; i++) {
+    u1[i]=(i+1)/Nx/2.0;
+    u2[i]=(i+1)/Nx/4.0;
+    y[i]=(i+1)/Nx/6.0;
+    w[i]=(i+1)/Nx/8.0;
+    for (j=0; j<=Ny-1; j++) {
+      A[i][j]=(i*j)/Ny;
+      B[i][j]=0;
+    }
+  }
+  for (j=0; j<=Ny-1; j++) {
+    v1[j]=(j+1)/Ny/2.0;
+    v2[j]=(j+1)/Ny/4.0;
+    z[j]=(j+1)/Ny/6.0;
+    x[j]=(j+1)/Ny/8.0;
+  }
+}
+
+double rtclock()
+{
+  struct timezone tzp;
+  struct timeval tp;
+  int stat;
+  gettimeofday (&tp, &tzp);
+  return (tp.tv_sec + tp.tv_usec*1.0e-6);
+}
+
+int main()
+{
+  init_arrays();
+
+  double annot_t_start=0, annot_t_end=0, annot_t_total=0;
+  int annot_i;
+
+  for (annot_i=0; annot_i<REPS; annot_i++)
+  {
+    annot_t_start = rtclock();
+
+
+    int i,j;
+
+
+for (i=0; i<=nx-1; i++) 
+  for (j=0; j<=ny-1; j++) 
+    B[i][j] = A[i][j] + u1[i]*v1[j] + u2[i]*v2[j];
+
+for (j=0; j<=ny-1; j++) 
+  x[j] = 0;
+
[…]

### 2008-08-20 testsuite/gemver/cookie/exp/gemver.orio.seq.c

@@ -1,0 +1,128 @@
+
+#include <stdio.h>
+#include <stdlib.h>
+#include <sys/time.h>
+#include <math.h>
+
+#define ceild(n,d)  ceil(((double)(n))/((double)(d)))
+#define floord(n,d) floor(((double)(n))/((double)(d)))
+#define max(x,y)    ((x) > (y)? (x) : (y))
+#define min(x,y)    ((x) < (y)? (x) : (y))
+
+double a;
+double b;
+int nx;
+int ny;
+double A[Nx][Ny];
+double B[Nx][Ny];
+double u1[Nx];
+double u2[Nx];
+double v1[Ny];
+double v2[Ny];
+double y[Nx];
+double z[Ny];
+double w[Nx];
+double x[Ny];
+
+void init_arrays()
+{
+  int i, j;
+  a = 1.5;
+  b = 2.5;
+  nx = Nx;
+  ny = Ny;
+  for (i=0; i<=Nx-1; i++) {
+    u1[i]=(i+1)/Nx/2.0;
+    u2[i]=(i+1)/Nx/4.0;
+    y[i]=(i+1)/Nx/6.0;
+    w[i]=(i+1)/Nx/8.0;
+    for (j=0; j<=Ny-1; j++) {
+      A[i][j]=(i*j)/Ny;
+      B[i][j]=0;
+    }
+  }
+  for (j=0; j<=Ny-1; j++) {
+    v1[j]=(j+1)/Ny/2.0;
+    v2[j]=(j+1)/Ny/4.0;
+    z[j]=(j+1)/Ny/6.0;
+    x[j]=(j+1)/Ny/8.0;
+  }
+}
+
+double rtclock()
+{
+  struct timezone tzp;
+  struct timeval tp;
+  int stat;
+  gettimeofday (&tp, &tzp);
+  return (tp.tv_sec + tp.tv_usec*1.0e-6);
+}
+
+int main()
+{
+  init_arrays();
+
+  double annot_t_start=0, annot_t_end=0, annot_t_total=0;
+  int annot_i;
+
+  for (annot_i=0; annot_i<REPS; annot_i++)
+  {
+    annot_t_start = rtclock();
+
+
+
+
+int i,j,ii,jj,it,jt;
+
+for (i=0; i<=min(nx-1,ny-1); i=i+1) {
+  x[i]=0;
+  w[i]=0;
+}
+for (i=nx; i<=ny-1; i=i+1) 
[…]

### 2008-08-20 testsuite/gemver/cookie/gemver.c

@@ -1,0 +1,25 @@
+
+/* pluto start (nx,ny,a,b) */
+
+for (i=0; i<=nx-1; i++) 
+  for (j=0; j<=ny-1; j++) 
+    B[i][j] = A[i][j] + u1[i]*v1[j] + u2[i]*v2[j];
+
+for (j=0; j<=ny-1; j++) 
+  x[j] = 0;
+
+for (i=0; i<=nx-1; i++) 
+  for (j=0; j<=ny-1; j++) 
+    x[j] = x[j] + y[i]*B[i][j];
+
+for (j=0; j<=ny-1; j++) 
+  x[j] = b*x[j] + z[j];
+    
+for (i=0; i<=nx-1; i++) {
+  w[i] = 0;
+  for (j=0; j<=ny-1; j++) 
+    w[i] = w[i] + B[i][j]*x[j];
+  w[i] = a*w[i];
+ }
+
+/* pluto end */

### 2008-08-20 testsuite/gemver/cookie/exp/gemver.pluto.c
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@314 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@ -1,0 +1,122 @@
+
+#include <stdio.h>
+#include <stdlib.h>
+#include <sys/time.h>
+#include <math.h>
+
+#define ceild(n,d)  ceil(((double)(n))/((double)(d)))
+#define floord(n,d) floor(((double)(n))/((double)(d)))
+#define max(x,y)    ((x) > (y)? (x) : (y))
+#define min(x,y)    ((x) < (y)? (x) : (y))
+
+double a;
+double b;
+int nx;
+int ny;
+double A[Nx][Ny];
+double B[Nx][Ny];
+double u1[Nx];
+double u2[Nx];
+double v1[Ny];
+double v2[Ny];
+double y[Nx];
+double z[Ny];
+double w[Nx];
+double x[Ny];
+
+void init_arrays()
+{
+  int i, j;
+  a = 1.5;
+  b = 2.5;
+  nx = Nx;
+  ny = Ny;
+  for (i=0; i<=Nx-1; i++) {
+    u1[i]=(i+1)/Nx/2.0;
+    u2[i]=(i+1)/Nx/4.0;
+    y[i]=(i+1)/Nx/6.0;
+    w[i]=(i+1)/Nx/8.0;
+    for (j=0; j<=Ny-1; j++) {
+      A[i][j]=(i*j)/Ny;
+      B[i][j]=0;
+    }
+  }
+  for (j=0; j<=Ny-1; j++) {
+    v1[j]=(j+1)/Ny/2.0;
+    v2[j]=(j+1)/Ny/4.0;
+    z[j]=(j+1)/Ny/6.0;
+    x[j]=(j+1)/Ny/8.0;
+  }
+}
+
+double rtclock()
+{
+  struct timezone tzp;
+  struct timeval tp;
+  int stat;
+  gettimeofday (&tp, &tzp);
+  return (tp.tv_sec + tp.tv_usec*1.0e-6);
+}
+
+int main()
+{
+  init_arrays();
+
+  double annot_t_start=0, annot_t_end=0, annot_t_total=0;
+  int annot_i;
+
+  for (annot_i=0; annot_i<REPS; annot_i++)
+  {
+    annot_t_start = rtclock();
+
+
+int i,j,ii,jj,it,jt;
+
+for (i=0; i<=min(nx-1,ny-1); i=i+1) {
+  x[i]=0;
+  w[i]=0;
+}
+for (i=nx; i<=ny-1; i++ ) 
+  x[i]=0;
[…]

### 2008-08-20 testsuite/gemver/cookie/exp/template.begin.c

@@ -1,0 +1,72 @@
+
+#include <stdio.h>
+#include <stdlib.h>
+#include <sys/time.h>
+#include <math.h>
+
+#define ceild(n,d)  ceil(((double)(n))/((double)(d)))
+#define floord(n,d) floor(((double)(n))/((double)(d)))
+#define max(x,y)    ((x) > (y)? (x) : (y))
+#define min(x,y)    ((x) < (y)? (x) : (y))
+
+double a;
+double b;
+int nx;
+int ny;
+double A[Nx][Ny];
+double B[Nx][Ny];
+double u1[Nx];
+double u2[Nx];
+double v1[Ny];
+double v2[Ny];
+double y[Nx];
+double z[Ny];
+double w[Nx];
+double x[Ny];
+
+void init_arrays()
+{
+  int i, j;
+  a = 1.5;
+  b = 2.5;
+  nx = Nx;
+  ny = Ny;
+  for (i=0; i<=Nx-1; i++) {
+    u1[i]=(i+1)/Nx/2.0;
+    u2[i]=(i+1)/Nx/4.0;
+    y[i]=(i+1)/Nx/6.0;
+    w[i]=(i+1)/Nx/8.0;
+    for (j=0; j<=Ny-1; j++) {
+      A[i][j]=(i*j)/Ny;
+      B[i][j]=0;
+    }
+  }
+  for (j=0; j<=Ny-1; j++) {
+    v1[j]=(j+1)/Ny/2.0;
+    v2[j]=(j+1)/Ny/4.0;
+    z[j]=(j+1)/Ny/6.0;
+    x[j]=(j+1)/Ny/8.0;
+  }
+}
+
+double rtclock()
+{
+  struct timezone tzp;
+  struct timeval tp;
+  int stat;
+  gettimeofday (&tp, &tzp);
+  return (tp.tv_sec + tp.tv_usec*1.0e-6);
+}
+
+int main()
+{
+  init_arrays();
+
+  double annot_t_start=0, annot_t_end=0, annot_t_total=0;
+  int annot_i;
+
+  for (annot_i=0; annot_i<REPS; annot_i++)
+  {
+    annot_t_start = rtclock();
+
+

### 2008-08-20 testsuite/gemver/cookie/exp/template.end.c

@@ -1,0 +1,28 @@
+
+
+
+
+    annot_t_end = rtclock();
+    annot_t_total += annot_t_end - annot_t_start;
+  }
+  
+  annot_t_total = annot_t_total / REPS;
+
+#ifndef TEST
+  printf("%f\n", annot_t_total);
+#else
+  {
+    int i, j;
+    for (i=0; i<nx; i++) {
+      if (i%100==0)
+	printf("\n");
+      printf("%f ",w[i]);
+    }
+    printf("\n");
+  }
+#endif
+
+  return ((int) w[0]); 
+}
+                                
+

### 2008-09-11 doc/papers/cgo08/f77/bicgkernel.f
code fix
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@342 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@ -1,7 +1,8 @@
       subroutine bicgkernel(lda, n, A, p, r, s, q)
-      integer lda, n
+      integer lda, n, incx, incy
       double precision zero, one
       double precision p(*), r(*)
+      double precision s(*), q(*)
       double precision A(lda, *)
 c
 c   BICG

### 2008-09-11 doc/papers/cgo08/paper.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@300 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb
current gdoc version
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@347 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@
 % Full paper: Sep. 19, 2008
-
 \documentclass[11pt]{article}
@@
 \usepackage{html}
+\usepackage{setspace}
 \usepackage{graphicx}
-\usepackage{subfigure}
+\usepackage{cite}
 \usepackage{amsmath}
-\usepackage{cite}
-\usepackage{times}
-\usepackage{fullpage}
-\usepackage{setspace}
 \usepackage[margin=1in]{geometry}
@@
 \vspace{-.05in}
-{\small $^1$University of Colorado at Boulder Dept. of Computer Science, Boulder, CO~80302}\\
+{\small $^1$University of Colorado at Boulder Dept. of Computer Science, Boulder, CO~80309}\\
 \vspace{-.05in}
@@
 \vspace{-.05in}
-{\small $^3$Ohio State University Dept. of Computer Science and Engineering, Columbus, OH 43210}\\
+{\small $^3$Ohio State University Dept. of Computer Science and Engineering, Columbus, OH~43210}\\
 \vspace{-.05in}
@@
 \vspace{-.05in}
-{\small $^4$University of Colorado at Boulder Dept. of Electrical and Computer Engineering, Boulder, CO~80302}\\
+{\small $^4$University of Colorado at Boulder Dept. of Electrical and Computer Engineering, Boulder, CO~80309}\\
 \vspace{-.05in}
@@
 \abstract{TODO}
+
 \doublespacing
@@
 
-\subsection{A {MATLAB} Compiler}
+\subsection{A MATLAB Compiler}
 \label{sec:matlab}
@@
 
-\begin{figure}[htbp]
+\begin{figure}[hbtp]
   \centering
-\includegraphics[width=.9\textwidth]{figures/refine-mv-dot.png}
+  \includegraphics[width=\textwidth]{figures/refine-mv-dot.png}
+
   \caption{The refinement of a matrix-vector product to a set of inner products and the refinement of inner products to scalar multiplications and additions.}
@@
 
+In this section we present performance results for a number of composed BLAS operations. We performed experiments on an dual quad-core 2/88 Xeon with 2.88 GHz, 16GB DDR2 FBDIMM
+
 \subsection{VADD}
 
-\subsection{WAXPBY $v = Ax + By$}
-perhaps add another term
+$x = w + y + z$
+
+\subsection{ATAX}
+
+$y = A' * (A * x)$
+
+\begin{figure}[htp]
+\centering
+\includegraphics[width=.8\textwidth]{figures/atax.png}
+\caption{Wall-clock time (in seconds) for the ATAX operation using different code versions.}
+\label{fig:atax}
+\end{figure}
 
@@
 
-%for (i=0; i<=A_nrows-1; i++)
-%  for (j=0; j<=A_ncols-1; j++)
-%    B[i][j] = A[i][j] + u1[i]*v1[j] + u2[i]*v2[j];
+\begin{figure}[htb]
+\centering
+\scriptsize
+\begin{verbatim}
+for (i=0; i<=A_nrows-1; i++)
+  for (j=0; j<=A_ncols-1; j++)
+   B[i][j] = A[i][j] + u1[i]*v1[j] + u2[i]*v2[j];
+
+for (i=0; i<=A_ncols-1; i++)
+  for (j=0; j<=A_nrows-1; j++)
+    x[i] = x[i] + b * B[j][i]*y[j];
+
+for (i=0; i<=A_ncols-1; i++)
+  x[i] = x[i] + z[i];
+
+for (i=0; i<=A_nrows-1; i++)
+  for (j=0; j<=A_ncols-1; j++)
+    w[i] = w[i] + a * B[i][j]*x[j];
+\end{verbatim}
+\caption{GEMVER C code.}
+\label{fig:gemver}
+\end{figure}
+
+%--- Sequential: seconds (dynamic arrays) ---
+%cmatlab= [0.038, 0.157, 0.352, 0.626, 0.976, 1.406, 1.909, 2.494, 3.143, 3.886]
+%orio= [0.029, 0.118, 0.263, 0.480, 0.759, 1.075, 1.447, 1.899, 2.338, 2.978]
+%blas= [0.064, 0.265, 0.598, 1.091, 1.711, 2.456, 3.340, 4.345, 5.480, 6.840]
+%mkl=  [0.073, 0.147, 0.331, 0.591, 0.923, 1.326, 1.803, 2.350, 2.975, 3.676]
+%atlas=[0.039, 0.163, 0.401, 0.796, 1.225, 1.916, 2.582, 3.381, 4.210, 4.756]
 %
-%for (i=0; i<=A_ncols-1; i++)
-%  for (j=0; j<=A_nrows-1; j++)
-%    x[i] = x[i] + b * B[j][i]*y[j];
-%
-%
-%for (i=0; i<=A_ncols-1; i++)
-%  x[i] = x[i] + z[i];
-%
-%for (i=0; i<=A_nrows-1; i++)
-%  for (j=0; j<=A_ncols-1; j++)
-%    w[i] = w[i] + a * B[i][j]*x[j];
+%--- Parallel: seconds (dynamic arrays) --- N=10,000, cores=1-8
+%cmatlab= [0.976, 0.976, 0.976, 0.976, 0.976, 0.976, 0.975, 0.976]
+%orio= [0.754, 0.738, 0.666, 0.649, 0.664, 0.664, 0.647, 0.668]
 
@@
 
-% Placeholder for the new results, the loops tuned for the figure above were almost identical to the ones generated by the MATLAB compiler
+\begin{figure}[htp]
+\centering
+\includegraphics[width=.8\textwidth]{figures/gemver.png}
+\caption{Wall-clock time (in seconds) for the GEMVER operation using different code versions.}
+\label{fig:gemver}
+\end{figure}
 
-\subsection{GESUMMV $y = a * (A * x) + b * (B * x)$}
+\subsection{GESUMMV}
+$y = a * (A * x) + b * (B * x)$
 
-\subsection{Bicgkernel}
+\begin{figure}[htp]
+\centering
+\includegraphics[width=.8\textwidth]{figures/gesummv.png}
+
+\caption{Wall-clock time (in seconds) for the GESUMMV operation using different code versions.}
+\label{fig:gesummv}
+\end{figure}
+\subsection{BiCG Kernel}
+
+$q = A * p$, $s = A' * r$
+
+\begin{figure}[htp]
+\centering
+
+\includegraphics[width=.8\textwidth]{figures/bicgkernel.png}
+\caption{Wall-clock time (in seconds) for the BiCG Kernel operation using different code versions.}
+\label{fig:bicgkernel}
+\end{figure}
+
+%NOTES about the experiments
+
+%Results already obtained:
+% -gemver (orio/testsuite/gemver/cookie directory)
+% -gesummv (orio/testsuite/gesummv/cookie directory)
+% -bicgkernel (orio/testsuite/bicgkernel/cookie directory)
+% -atax (orio/testsuite/atax/cookie directory)
+
+%* All results were obtained on the cookie machine (eight-core Intel Xeon)
+
+%* Compiler used: icc 10.1
+%* Optimization flags used: -O3 (-parallel for parallel matlab code) (-openmp for parallel orio code)
+
+%* We compare two types of codes:
+%  1. matlab
+%  2. orio
+%1. 'matlab' code is the original code produced by the matlab c compiler.
+%2. 'orio' is obtained first from the Pluto-generated code (use loop fusion, NO loop tiling,
+%   NO unrolling, NO parallelization, NO vectorization), then was decorated with
+%   performance annotations before being tuned by Orio.
+
+%* Each code has two versions: one with static arrays, and another with dynamic arrays.
+
+%* Conversion from/to static arrays to/from dynamic arrays was done manually.
+
+%* I run Pluto first to maximally fuse all loops (using polycc <file>
+%--maxfuse option) and see the resulting fused loop(s). And then tune the
+%fused loop using Orio. Sometimes, the fused loop gets very complicated
+%because of Pluto's various transformations (e.g. skewing). In that case, I
+%don't use Pluto's loop fusion. For example, gemver and gesummv results use
+%Pluto's loop fusion, but bicgkernel doesn't use Pluto's loop fusion at
+%all.
+
+%* All experiment files are in <kernel-name>/cookie directory
+
+%* All generated codes (the base Matlab codes, and the Orio-tuned codes),
+%results, script for running the experiments are in
+%<kernel-name>/cookie/exp
+
+%* I tuned 4 different versions for each kernel:
+% 1. sequential (static arrays)
+% 2. sequential (dynamic arrays)
+% 3. parallel (static arrays)
+% 4. parallel (dynamic arrays)
+
+%The annotated source code for 1 and 3 is in <kernel-name>.src1.c
+%The annotated source code for 3 and 4 is in <kernel-name>.src2.c
+
+%The final code for 1 and 2 is in <kernel-name>.orio.seq.c
+%The final code for 3 and 4 is in <kernel-name>.orio.par.c
+
+%* <kernel-name>.matlab.c contains the original Matlab code (for both
+%static and dynamics arrays)
+
+%* The best performance parameters for each version can be seen in
+%results.txt file.
 
@@
 extend beyond composed BLAS
+
+Ultimately, the tuning process should result in more than one version of the computation since different versions are optimal for different problem sizes.
+
 \singlespacing
-\section*{References}
 \bibliographystyle{siam}

### 2008-09-12 doc/papers/cgo08/paper.tex
version with abstract
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@348 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@
 \usepackage{graphicx}
+\usepackage{subfigure}
 \usepackage{cite}
@@
 
-\abstract{TODO}
+\abstract{
+The growing demand for higher levels of detail and accuracy in results means that the size and complexity of scientific computations is increasing at least as fast as the improvements in processor technology.
+Programming scientific applications is hard, and optimizing them for high performance is even harder.
+The development of optimized codes requires extensive knowledge, not only of the costs of floating-point arithmetic but also of  memory access issues and compiler optimizations.
+Experiments show that the complexity of this hardware-software system means that performance is difficult to predict fully.
+Therefore, computational scientists are often forced to choose between investing too much time in tuning code or  accepting performance that is significantly lower than the best achievable performance on a given architecture.
+
+In this paper, we describe the first steps toward a fully automated system for the optimization of the matrix algebra kernels that are a foundational part of many scientific applications.
+To generate highly optimized code from a high-level MATLAB prototype, we will follow a three-step approach.
+To begin, we have developed a compiler that converts a MATLAB script into simple C code.
+We then use the source transformation tool PLuTo to optimize that code for coarse-grained parallelism and locality simultaneously.
+The result is an automatically generated code containing loop transformations and OpenMP annotations for exploiting multicore parallelism.
+Finally, we employ the empirical performance tuning system Orio to generate many tuned versions of the same operation using different optimization techniques, such as loop unrolling and memory alignment.
+Orio performs an automated empirical search for selecting the best among the multiple optimized code variants.
+
+We discuss performance results showing that the code generated by using our system significantly outperforms not only the original simple C code but also code based on source BLAS, ATLAS-optimized BLAS, and Intel MKL routines.
+}
 
@@
 
+This motivates further work on automating more parts of the process, so that these tools can be used by application developers to automatically produce custom linear algebra libraries tuned for specific architectures.
+
 \singlespacing

### 2008-09-19 doc/papers/cgo08/paper.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@349 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@
 % Max 22 pages, double-spaced, 11pt or larger font
-% Abstract: Sep. 12, 2008
+% Abstract 300 words: Sep. 12, 2008
 % Full paper: Sep. 19, 2008
+
 \documentclass[11pt]{article}
@@
 \abstract{
+
 The growing demand for higher levels of detail and accuracy in results means that the size and complexity of scientific computations is increasing at least as fast as the improvements in processor technology.
@@
 
-%Words here in Intro and Back are lifted from a 2004 proposal by Liz, Bill, and Jack, written almost completely by Liz.   Perhaps dated but
-% at least a jumping off point.
-
-The growing demand for higher levels of detail and accuracy in results means that the size and complexity of scientific computations is increasing at least as fast as the improvements in processor technology.
-Programming scientific applications is hard, and optimizing them for high performance is even harder.
-The development of optimized codes requires extensive knowledge, not only of the costs of floating-point arithmetic but also of  memory access issues and of the compiler being used.
-Hardware platforms are continually evolving.
-In addition, experiments show that the complexity of this  hardware-software system means that performance is difficult to predict fully.
-Therefore, computational scientists are often forced to choose between investing too much time in tuning code or  accepting performance that is lower (and sometimes significantly lower) than peak.
+The development of high-performance numerical codes is challenging because performance is determined by complex interactions between the algorithm, data structure, programming language, compiler, and computer architecture.
+Scientists seeking high performance are thus required to master advanced concepts in computer science and carry out intricate programming tasks in addition to managing the scientific content of their work.
+They must either invest substantial time in tuning their software or accept low performance.
+In either case, the productivity of the scientist degrades.
 
@@
 
-In this paper, we describe the first steps toward a fully automated system for the optimization of the matrix algebra kernels that are a foundational part of many scientific applications.
-[Ugly] Beginning with a MATLAB prototype, we first generate C or Fortran then run it through Pluto~\cite{} to optimize the sequence of loops and then Pluto~\cite{} to generate a collection of organizations of code and to identify the best choice.
-Comment on manual parts of this.
+We are working on a combination of the two approaches designed to overcome some of their shortcomings.
+This paper described our three-pronged approach toward the development of an automatically tuned library for matrix algebra computations.
+In Section~\ref{sec:matlab}, we describe the compiler we have developed that converts a MATLAB script into simple C code.
+In Section~\ref{sec:orio}, we demonstrate how to employ the empirical performance tuning system Orio in conjunction with the PLuTo tool to generate many tuned versions of the same operation using different optimization techniques, such as loop unrolling and memory alignment.
+The result is an automatically generated code containing loop transformations and OpenMP annotations for exploiting multicore parallelism.
+In Section~\ref{sec:experiments}, we provide performance results showing that the code generated by using our system significantly outperforms not only the original simple C code but also code based on source BLAS, ATLAS-optimized BLAS, and Intel MKL routines.
+%Text in intro repeats abstract some, but that might not be a big deal for a conf paper.
 
@@
 
-The best performance today is obtained through approaches like that taken by PERI~\cite{PERI}, which involves a labor intensive rewrite of portion of the code by a group of performance tools experts in collaboration with application scientists.
+The best performance today is obtained through approaches like that taken by PERI~\cite{PERI} which involves a labor intensive rewrite of portion of the code by a group of performance tools experts in collaboration with application scientists.
 Maintaining performance of the code as the application and hardware evolves means that the tuning process may need to be repeated.
@@
 
-% TODO: Paragraph about source-to-source transformation tools, including PLuTo, Orio
+A number of source or binary transformation tools for performance-improving optimizations exist.
+LoopTool~\cite{LoopTool}, developed at Rice University, supports annotation-based loop fusion, unroll-and-jam, skewing and tiling.
+A new tool, POET~\cite{POET} also supports a number of loop transformations.
+POET offers a complex template-based syntax for defining transformations in a language-independent manner.
+Other recent research efforts whose goal, at least in part, is to enable optimizations of source code to be augmented with performance-related information include the X language~\cite{XLanguage} (a macro C-like language for annotating C code), the Broadway~\cite{broadway} compiler,  telescoping languages~\cite{telescopingurl,teleoverview,Ken99}, and various meta-programming techniques~\cite{veldhuizen95,weise93,kiczales91,chiba95}.
+PLuTo~\cite{Pluto,uday08cc} is a source-to-source transformation tool for optimizing sequences of nested loops.
+PLuTo employs a polyhedral model of nested loops, where the dynamic instance (iteration) of each statement is viewed as an integer point in a well-defined space, called the statement's polyhedron.
+Combined with a characterization of data dependences, this representation allows the construction of mathematically correct complex loop transformations.
+The transformations target both improved cache locality and parallelism.
 
@@
 
-The PIs and other researchers have observed that it can be much more efficient to call a single specialized routine that performs multiple operations than to make successive calls to separate BLAS ~\cite{Ashby:uq,Blackford:2002vn,Stanley:CSD-98-992,baker03blgmres,baker03lgmres,Dennis:2005tg,Howell:2008,gropp01,Vuduc:2003kl}.
+The authors and other researchers have observed that it can be much more efficient to call a single specialized routine that performs multiple operations than to make successive calls to separate BLAS ~\cite{Ashby:uq,Blackford:2002vn,Stanley:CSD-98-992,baker03blgmres,baker03lgmres,Dennis:2005tg,Howell:2008,gropp01,Vuduc:2003kl}.
 Single routines that carry out more than one linear algebra operation are known as {\t composed} BLAS.
@@
 We then use the source-to-source automatic parallelization tool  PLuTo\cite{Pluto} to optimize for coarse-grained parallelism and locality simultaneously.
-The result is an automatically generated OpenMP parallel code for multicores.
+The result is an automatically generated code containing transformed loops and OpenMP directives for compiler parallelization.
 Finally, we employ the extensible annotation system Orio to generate many tuned versions of the same operation using different optimization parameters and perform an empirical search for selecting the best among the multiple optimized code variants.
+%Does this repeat abstract/intro too closely?
 
-Probably need some words about WHY this 3 pronged approach is a good idea.
+%Probably need some words about WHY this 3 pronged approach is a good idea.
 
@@
 
+We have developed a compiler that translates MATLAB scripts into C code.
 Figure~\ref{fig:compiler} gives an overview of the compilation process.
@@
 
-\subsubsection{Search Space Exploration and Evaluation}
-\label{sec:search}
-
-This will need updating depending on how we decide to address this topic. -Jeremy 8/6/08 10:19 AM 
-The set of all possible combinations of implementation and optimization choices is called the search space~\cite{Kisuki:2000uq,Triantafyllis:2003uq, Cooper:2005kx}.
-Although the dataflow graphs are quite small in our setting (usually less than 100 nodes), it is still important to prune the search space to avoid an exponential compilation time.
-Our prototype already performs some pruning: the optimization rewrite rules only apply under certain circumstances.
-However, as mentioned above we are sometimes overly aggressive in applying optimizations.
-To solve this problem we plan to use a combination of analytic methods and empirical methods~\cite{Chen:2007vn}.
-%Yotov:2005kx
-We plan to use a fast analytic cost function to prune out regions of the search space that will not produce competitive implementations.
-The cost function is based on a model of the computer architecture, but does not necessarily account for all the details.
-Once the search space is narrowed, we will use empirical testing with representative data sets to determine the exact performance of the alternatives and choose the best one.
-
 \subsubsection{Translate the Dataflow Graph to C}
@@
 The generator outputs a C loop for each subgraph.
-The order in which the loops, and scalar operations within the loops, appear in the output is determined by performing a topological sort.
-
-\subsection{PLuTo}
-\label{sec:plutonorio}
-% http://www.cse.ohio-state.edu/~bondhugu//
-
-PLuTo~\cite{Pluto,uday08cc} is a source-to-source transformation tool for optimizing sequences of (possibly nested) loops.
-
+The order in which the loops, and scalar operations within the loops, appear in the output is determined by performing a topological sort
 \subsection{Orio}
@@
 Orio does not parse the application code; rather, it relies on annotations expressed as structured comments, which contain a simplified expression of the computation, as well as directives on what transformations to apply.
+
+% Explain how Pluto is used
+
+% Forward ref. to search approaches
 
@@
 
+\subsection{Search Space Exploration and Evaluation}
+\label{sec:search}
+
+This will need updating depending on how we decide to address this topic. -Jeremy 8/6/08 10:19 AM
+% BN: moved to its own section since multiple tools deal with the search space problem.
+The set of all possible combinations of implementation and optimization choices is called the search space~\cite{Kisuki:2000uq,Triantafyllis:2003uq, Cooper:2005kx}.
+Although the dataflow graphs are quite small in our setting (usually less than 100 nodes), it is still important to prune the search space to avoid an exponential compilation time.
+Our prototype Matlab compiler already performs some pruning: the optimization rewrite rules only apply under certain circumstances.
+%
+% BN: commented out below since it's more future work
+%However, as mentioned above we are sometimes overly aggressive in applying optimizations.
+%To solve this problem we plan to use a combination of analytic methods and empirical methods~\cite{Chen:2007vn}.
+%
+Orio also offers a number of different general strategies for narrowing the search space over loop-level optimizations, including random, simplex, and simulated annealing searches.
+
 \section{Experimental Results}
@@
 
-In this section we present performance results for a number of composed BLAS operations. We performed experiments on an dual quad-core 2/88 Xeon with 2.88 GHz, 16GB DDR2 FBDIMM
+In this section we present performance results for a number of composed BLAS operations.
+We performed experiments on an Intel Xeon 2.80 GHz Intel(R) E5462 dual quad-core workstation with 16GB DDR2 FBDIMM RAM.
+% Include peak numbers?
 
@@
 
+BN: working on this
+
 \subsection{ATAX}
 
-$y = A' * (A * x)$
+The extended Matlab code for the ATAX operation, $y = A' * (A * x)$,
+and the corresponding performance results are shown in Figure~\ref{fig:gesummv}.
 
@@
 \centering
-\includegraphics[width=.8\textwidth]{figures/atax.png}
-\caption{Wall-clock time (in seconds) for the ATAX operation using different code versions.}
+\begin{tabular}{cc}
+\begin{minipage}[b]{.36\textwidth}
+\footnotesize
+\begin{verbatim}
+ATAX
+in
+  A : row matrix, x : vector
+out
+  y : vector
+{
+  y = A' * (A * x)
+}
+
+\end{verbatim}
+\end{minipage}
+&
+\begin{minipage}[b]{.6\textwidth}
+\includegraphics[width=\textwidth]{figures/atax.png}
+\end{minipage}\\
+\end{tabular}
+\caption{Annotated MATLAB code (left) and wall-clock time in seconds (right) for the ATAX operation using different code versions.}
 \label{fig:atax}
@@
 
-\begin{figure}[htb]
+The extended Matlab code for the GEMVER operation and the corresponding performance results are shown in Figure~\ref{fig:gemver}.
+
+\begin{figure}[htp]
 \centering
-\scriptsize
+\begin{tabular}{cc}
+\begin{minipage}[b]{.36\textwidth}
+\footnotesize
 \begin{verbatim}
-for (i=0; i<=A_nrows-1; i++)
-  for (j=0; j<=A_ncols-1; j++)
-   B[i][j] = A[i][j] + u1[i]*v1[j] + u2[i]*v2[j];
-
-for (i=0; i<=A_ncols-1; i++)
-  for (j=0; j<=A_nrows-1; j++)
-    x[i] = x[i] + b * B[j][i]*y[j];
-
-for (i=0; i<=A_ncols-1; i++)
-  x[i] = x[i] + z[i];
-
-for (i=0; i<=A_nrows-1; i++)
-  for (j=0; j<=A_ncols-1; j++)
-    w[i] = w[i] + a * B[i][j]*x[j];
+GEMVER
+in
+  A : column matrix,
+  u1 : vector, u2 : vector,
+  v1 : vector, v2 : vector,
+  a : scalar, b : scalar,
+  y : vector, z : vector
+out
+  B : column matrix, 
+  x : vector, w : vector
+{
+  B = A + u1 * v1' + u2 * v2'
+  x = b * (B' * y) + z
+  w = a * (B * x)
+}
 \end{verbatim}
-\caption{GEMVER C code.}
-\label{fig:gemver}
+\end{minipage}
+&
+\begin{minipage}[b]{.6\textwidth}
+\includegraphics[width=\textwidth]{figures/gemver.png}
+\end{minipage}\\
+\end{tabular}
+\caption{Annotated MATLAB code (left) and wall-clock time in seconds (right) for the GEMVER operation using different code versions.}
+\label{fig:gesummv}
 \end{figure}
 
+% Data (Dual quad-core Intel Xeon E5462, 16GB RAM)
 %--- Sequential: seconds (dynamic arrays) ---
@@
 
-Experiments on Intel Xeon and BGP.
+\subsection{GESUMMV}
+
+The extended Matlab code for the GESUMMV operation, $y = a * (A * x) + b * (B * x)$,
+and the corresponding performance results are shown in Figure~\ref{fig:gesummv}.
 
@@
 \centering
-\includegraphics[width=.8\textwidth]{figures/gemver.png}
-\caption{Wall-clock time (in seconds) for the GEMVER operation using different code versions.}
-\label{fig:gemver}
+\begin{tabular}{cc}
+\begin{minipage}[b]{.36\textwidth}
+\footnotesize
+\begin{verbatim}
+GESUMMV
+in
+  A : row matrix,
+  B : row matrix,
+  x : vector,
+  a : scalar,
+  b : scalar
+out
+  y : vector
+{
+  y = a * (A * x) + b * (B * x)
+}
+
+\end{verbatim}
+\end{minipage}
+&
+\begin{minipage}[b]{.6\textwidth}
+\includegraphics[width=\textwidth]{figures/gesummv.png}
+\end{minipage}\\
+\end{tabular}
+\caption{Annotated MATLAB code (left) and wall-clock time in seconds (right) for the GESUMMV operation using different code versions.}
+\label{fig:gesummv}
 \end{figure}
 
-\subsection{GESUMMV}
-$y = a * (A * x) + b * (B * x)$
+% BN: ATLAS's dgemv is annihilating everything else, even though it's called twice... wow.
+% I did check the result and it was correct
+% The Time and MFLOPS data for the above are (with icc):
+% (Dual quad-core Intel Xeon E5462, 16GB RAM)
+% ops = (4*N*N+3*N)
+% --- Parallel: seconds (dynamic arrays) ---
+%matlabc= [0.019, 0.075, 0.170, 0.301, 0.471, 0.678, 0.923, 1.205, 1.529, 1.887]
+%blas= [0.00006, 0.00018, 0.00039, 0.00060, 0.00090, 0.00126, 0.00168, 0.00216, 0.00270, 0.00330]
+%orio= [0.013, 0.035, 0.076, 0.133, 0.210, 0.300, 0.410, 0.533, 0.676, 0.831]
+%--- Parallel: Mflops/sec (dynamic arrays) ---
+%matlabc= [846.565, 848.425, 849.028, 849.503, 849.690, 849.968, 849.605, 849.946, 847.893, 848.023]
+%blas= [271288.136, 347891.304, 367392.857, 421785.832, 480228.091, 497440.415, 508787.800, 516153.226, 524930.741, 530172.300]
+%orio= [1231.989, 1846.163, 1885.595, 1918.746, 1907.976, 1920.818, 1910.867, 1920.502, 1915.970, 1924.623]
+%
+% and gcc behaves similarly, so it's not icc being smart and detecting 2 calls to dgemv
+% time: blas= [0.000, 0.000, 0.000, 0.001, 0.001, 0.001, 0.002, 0.002, 0.003, 0.003]
+% mflops: blas= [254063.492, 346010.811, 393491.803, 421785.832, 439593.407, 453214.792, 463108.092, 470394.120, 462381.020, 469088.244]
 
-\begin{figure}[htp]
-\centering
-\includegraphics[width=.8\textwidth]{figures/gesummv.png}
-
-\caption{Wall-clock time (in seconds) for the GESUMMV operation using different code versions.}
-\label{fig:gesummv}
-\end{figure}
 \subsection{BiCG Kernel}
@@
 
+% BN: moved from search space section above
+%Yotov:2005kx
+We plan to use a fast analytic cost function to prune out regions of the search space that will not produce competitive implementations.
+The cost function is based on a model of the computer architecture, but does not necessarily account for all the details.
+Once the search space is narrowed, we will use empirical testing with representative data sets to determine the exact performance of the alternatives and choose the best one.
+
 \singlespacing

### 2008-09-19 doc/papers/cgo08/paper.tex
done for tonight
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@350 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@
 
+\setlength{\textwidth}{6.45in}
+\setlength{\oddsidemargin}{0.05in}
+\setlength{\evensidemargin}{0.05in}
+
+\setlength{\topmargin}{-0.5in}
+\setlength{\textheight}{9in}
+
 \usepackage{url}
@@
 \usepackage{amsmath}
-\usepackage[margin=1in]{geometry}
+%\usepackage[margin=1in]{geometry}
 
@@
 
-$x = w + y + z$
+We optimized the performance for the VADD operation, $x = w + y + z$, on the Blue Gene/P at Argonne National Laboratory using the IBM XLC compiler, version 9.0.
+We measured the performance for two scenarios: using one core per processor and using all cores.
+In the single-core case, the codes were compiled with the following options: \texttt{-O3 -qstrict -qarch=450d -qtune=450 -qhot -qsmp=noauto}.
+For the multicore case, the only difference in the compilation was that for the non-Orio versions, the \texttt{-qsmp=auto} option was used.
 
-BN: working on this
+\begin{figure}[htp]
+\centering
+\begin{tabular}{cc}
+\begin{minipage}[b]{.25\textwidth}
+\footnotesize
+\begin{verbatim}
+VADD
+in
+  w : vector,
+  y : vector,
+  z : vector
+out
+  x : vector
+{
+  x = w + y + z
+}
+
+\end{verbatim}
+\end{minipage}
+&
+\begin{minipage}[b]{.6\textwidth}
+\includegraphics[width=\textwidth]{figures/vadd_bgp.png}
+\end{minipage}\\
+\end{tabular}
+\caption{Annotated MATLAB code (left) and MFLOP/s (right) for the VADD operation using different code versions.}
+\label{fig:atax}
+\end{figure}
+
+The "Base" label designates the simple loop C implementation generated by the Matlab compiler, which was simply compiled with the options mentioned earlier and executed.
+We tested the performance of the IBM ESSL library and Goto BLAS for the two calls to \texttt{daxpy} used in the BLAS version of the code.
+Finally, we annotated the simple C loop version with memory alignment and loop unrolling directives, and then used Orio to select the best tuned version for the single-core and multicore scenarios.
+We observe that even for a very simple operation, such as the vector addition above, the compiler alone is unable to obtain the same level of performance as the empirically tuned versions.
+Furthermore, as expected, the BLAS performance was worse than that of the single loop implementation, however, selecting the correct compiler options was crucial for achieving good performance of the base version.
 
@@
 \begin{tabular}{cc}
-\begin{minipage}[b]{.36\textwidth}
+\begin{minipage}[b]{.3\textwidth}
 \footnotesize
@@
 \centering
+\begin{tabular}{cc}
+\begin{minipage}[b]{.3\textwidth}
+\footnotesize
+\begin{verbatim}
+BICG
+in
+  A : column matrix,
+  p : vector, r : vector
+out
+  q : vector, s : vector
+{
+  q = A * p
+  s = A' * r
+}
 
-\includegraphics[width=.8\textwidth]{figures/bicgkernel.png}
-\caption{Wall-clock time (in seconds) for the BiCG Kernel operation using different code versions.}
-\label{fig:bicgkernel}
+\end{verbatim}
+\end{minipage}
+&
+\begin{minipage}[b]{.6\textwidth}
+\includegraphics[width=\textwidth]{figures/bicgkernel.png}
+\end{minipage}\\
+\end{tabular}
+\caption{Annotated MATLAB code (left) and wall-clock time in seconds (right) for the BiCG kernel operation using different code versions.}
+\label{fig:gesummv}
 \end{figure}

### 2008-09-19 doc/papers/cgo08/paper.tex
done for tonight
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@351 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb
current version
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@353 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@
 \maketitle
+\doublespacing
 
@@
 
-\doublespacing
 \section{Introduction}
@@
 
-%Modified from recent DOE proposal
+Existing optimizing MATLAB compilers, such as the MaJIC MATLAB compiler, include limited local optimizations for matrix expressions, but do not perform optimizations such as loop fusion across multiple operations as we do with the tools described in this paper.
+The telescoping languages project~\cite{telescopingurl,teleoverview,Ken99} uses techniques such as strength reduction, vectorization, and procedure specialization to optimize MATLAB scripts, but do not generate reusable optimized linear algebra routines as described in this paper.
 
@@
 POET offers a complex template-based syntax for defining transformations in a language-independent manner.
-Other recent research efforts whose goal, at least in part, is to enable optimizations of source code to be augmented with performance-related information include the X language~\cite{XLanguage} (a macro C-like language for annotating C code), the Broadway~\cite{broadway} compiler,  telescoping languages~\cite{telescopingurl,teleoverview,Ken99}, and various meta-programming techniques~\cite{veldhuizen95,weise93,kiczales91,chiba95}.
+Other recent research efforts whose goal, at least in part, is to enable optimizations of source code to be augmented with performance-related information include the X language~\cite{XLanguage} (a macro C-like language for annotating C code), the Broadway~\cite{broadway} compiler, and various meta-programming techniques~\cite{veldhuizen95,weise93,kiczales91,chiba95}.
 PLuTo~\cite{Pluto,uday08cc} is a source-to-source transformation tool for optimizing sequences of nested loops.
@@
 We then use the source-to-source automatic parallelization tool  PLuTo\cite{Pluto} to optimize for coarse-grained parallelism and locality simultaneously.
-The result is an automatically generated code containing transformed loops and OpenMP directives for compiler parallelization.
-Finally, we employ the extensible annotation system Orio to generate many tuned versions of the same operation using different optimization parameters and perform an empirical search for selecting the best among the multiple optimized code variants.
-%Does this repeat abstract/intro too closely?
+Using the results of the PLuTo analysis, we insert annotations into the C code, which are then processed by the extensible annotation system Orio to generate many tuned versions of the same operation using different optimization parameters.
+Orio then performs an empirical search to select the best among the multiple optimized code variants.
 
 %Probably need some words about WHY this 3 pronged approach is a good idea.
+MATLAB optimizing compilers traditionally have not been able to achieve
 
@@
 
+% Explain how Orio is used
+
 % Forward ref. to search approaches
@@
 \caption{Annotated MATLAB code (left) and MFLOP/s (right) for the VADD operation using different code versions.}
-\label{fig:atax}
+\label{fig:vadd}
 \end{figure}
@@
 The extended Matlab code for the ATAX operation, $y = A' * (A * x)$,
-and the corresponding performance results are shown in Figure~\ref{fig:gesummv}.
+and the corresponding performance results are shown in Figure~\ref{fig:atax}.
 
@@
 \caption{Annotated MATLAB code (left) and wall-clock time in seconds (right) for the GEMVER operation using different code versions.}
-\label{fig:gesummv}
+\label{fig:gemver}
 \end{figure}
@@
 \caption{Annotated MATLAB code (left) and wall-clock time in seconds (right) for the BiCG kernel operation using different code versions.}
-\label{fig:gesummv}
+\label{fig:bicgkernel}
 \end{figure}
@@
 
-fully automate the process to get rid of the manual steps along the way
-
-extend beyond composed BLAS
+The positive initial results from our three-pronged approach to generating tuned linear algebra routines motivate several future lines of investigation.
+At present there are a number of manual steps involved in obtaining a compiled tuned linear algebra library from the high-level extended MATLAB descriptions.
+We will next delve into each manual step and create the software that automates it and in effect serves as the ``glue'' between the different tools.
+We will also investigate extending the supported functionality beyond composed BLAS operations, thus benefiting a wider spectrum of applications.
 

### 2008-09-19 doc/papers/cgo08/paper.tex
added missing bib entries
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@354 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@
 The generator outputs a C loop for each subgraph.
-The order in which the loops, and scalar operations within the loops, appear in the output is determined by performing a topological sort
+A topological sort is performed to determine the order in which the loops and scalar operations within the loops appear in the output.
 \subsection{Orio}
@@
 
-% Explain how Orio is used
-
-% Forward ref. to search approaches
+Figure~\ref{fig:orio} illustrates the tuning process using Orio.
+In the simplest case, Orio can be used to improve code performance by performing source-to-source transformations, such as loop unrolling and memory alignment optimizations.
+First, the input to Orio is C code, which contains syntactically structured comments that contain both the computation to be tuned, as well as various performance-tuning directives.
+Orio scans the annotated input code and extracts all annotated regions.
+Each annotated region is then passed to the code transformation modules for potential optimizations.
+The code generator then produces the final code incorporating the newly generated optimized C code corresponding to the annotation.
+The original C code is not parsed (only the annotation comments are processed) and is preserved in the new version.
+The user has the option of selecting either the original or the tuned code by using a compile-time macro.
 
@@
 
+Orio can also be used as an automatic performance tuning tool.
+The system uses its code transformation modules and code generator to generate an optimized code version for each distinct combination of performance parameters.
+Then the optimized code versions are executed and their performance measured, which is subsequently compared to the performances of other previously tested code variants.
+After iteratively evaluating all code variants under consideration, the best-performing code is selected as the final output of Orio.
+The tuning specifications, written by users in the form of annotations (either embedded in the source code or in separate files), are parsed and used by Orio to guide the search and tuning process.
+These specifications include important information such as the used compilers, the search strategy, the program transformation parameters, the input data sizes, etc.
+
 \subsection{Search Space Exploration and Evaluation}
@@
 
-This will need updating depending on how we decide to address this topic. -Jeremy 8/6/08 10:19 AM
 % BN: moved to its own section since multiple tools deal with the search space problem.
@@
 %
-Orio also offers a number of different general strategies for narrowing the search space over loop-level optimizations, including random, simplex, and simulated annealing searches.
+Orio provides a number of different search heuristics to narrow the search space for near-optimal performance, including random, simplex, and simulated annealing searches.
+The search algorithm is one of the user-defined values in the Orio tuning specification.
 
@@
 In this section we present performance results for a number of composed BLAS operations.
-We performed experiments on an Intel Xeon 2.80 GHz Intel(R) E5462 dual quad-core workstation with 16GB DDR2 FBDIMM RAM.
+The experiment described in Section~\ref{sec:vadd} were performed on Blue Gene/P, while the remaining experiments were all performed on an Intel Xeon 2.80 GHz Intel(R) E5462 dual quad-core workstation with a 1600MHz front-side bus and 16GB DDR2 FBDIMM RAM.
+
 % Include peak numbers?
@@
 \subsection{VADD}
+\label{sec:vadd}
 
-We optimized the performance for the VADD operation, $x = w + y + z$, on the Blue Gene/P at Argonne National Laboratory using the IBM XLC compiler, version 9.0.
+We optimized the performance for the VADD operation, $x = w + y + z$, on the Blue Gene/P at Argonne National Laboratory using the IBM XLC compiler version 9.0.
 We measured the performance for two scenarios: using one core per processor and using all cores.
@@
 
-The extended Matlab code for the ATAX operation, $y = A' * (A * x)$,
-and the corresponding performance results are shown in Figure~\ref{fig:atax}.
+The extended Matlab code for the ATAX operation, $y = A' * (A * x)$, and the corresponding performance results are shown in Figure~\ref{fig:atax}.
+The Orio-tuned version that incorporates OpenMP directives achieves the best performance for most problem sizes, while the
 
@@
 % BN: ATLAS's dgemv is annihilating everything else, even though it's called twice... wow.
-% I did check the result and it was correct
+% I did check the result and it was correct, i'm still somewhat baffled
 % The Time and MFLOPS data for the above are (with icc):
@@
 
-%NOTES about the experiments
-
-%Results already obtained:
-% -gemver (orio/testsuite/gemver/cookie directory)
-% -gesummv (orio/testsuite/gesummv/cookie directory)
-% -bicgkernel (orio/testsuite/bicgkernel/cookie directory)
-% -atax (orio/testsuite/atax/cookie directory)
-
-%* All results were obtained on the cookie machine (eight-core Intel Xeon)
-
-%* Compiler used: icc 10.1
-%* Optimization flags used: -O3 (-parallel for parallel matlab code) (-openmp for parallel orio code)
-
-%* We compare two types of codes:
-%  1. matlab
-%  2. orio
-%1. 'matlab' code is the original code produced by the matlab c compiler.
-%2. 'orio' is obtained first from the Pluto-generated code (use loop fusion, NO loop tiling,
-%   NO unrolling, NO parallelization, NO vectorization), then was decorated with
-%   performance annotations before being tuned by Orio.
-
-%* Each code has two versions: one with static arrays, and another with dynamic arrays.
-
-%* Conversion from/to static arrays to/from dynamic arrays was done manually.
-
-%* I run Pluto first to maximally fuse all loops (using polycc <file>
-%--maxfuse option) and see the resulting fused loop(s). And then tune the
-%fused loop using Orio. Sometimes, the fused loop gets very complicated
-%because of Pluto's various transformations (e.g. skewing). In that case, I
-%don't use Pluto's loop fusion. For example, gemver and gesummv results use
-%Pluto's loop fusion, but bicgkernel doesn't use Pluto's loop fusion at
-%all.
-
-%* All experiment files are in <kernel-name>/cookie directory
-
-%* All generated codes (the base Matlab codes, and the Orio-tuned codes),
-%results, script for running the experiments are in
-%<kernel-name>/cookie/exp
-
-%* I tuned 4 different versions for each kernel:
-% 1. sequential (static arrays)
-% 2. sequential (dynamic arrays)
-% 3. parallel (static arrays)
-% 4. parallel (dynamic arrays)
-
-%The annotated source code for 1 and 3 is in <kernel-name>.src1.c
-%The annotated source code for 3 and 4 is in <kernel-name>.src2.c
-
-%The final code for 1 and 2 is in <kernel-name>.orio.seq.c
-%The final code for 3 and 4 is in <kernel-name>.orio.par.c
-
-%* <kernel-name>.matlab.c contains the original Matlab code (for both
-%static and dynamics arrays)
-
-%* The best performance parameters for each version can be seen in
-%results.txt file.
-
 \section{Conclusions and Future Work}

### 2008-09-20 doc/papers/cgo08/paper.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@355 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@
 To begin, we have developed a compiler that converts a MATLAB script into simple C code.
-We then use the source-to-source automatic parallelization tool  PLuTo\cite{Pluto} to optimize for coarse-grained parallelism and locality simultaneously.
-Using the results of the PLuTo analysis, we insert annotations into the C code, which are then processed by the extensible annotation system Orio to generate many tuned versions of the same operation using different optimization parameters.
+We then (optionally) use the source-to-source automatic parallelization tool PLuTo\cite{Pluto} developed at Ohio State University to optimize for coarse-grained parallelism and locality simultaneously.
+Using the results of the PLuTo analysis, we insert annotations into the C code, which are then processed by our extensible annotation system Orio to generate many tuned versions of the same operation using different optimization parameters.
 Orio then performs an empirical search to select the best among the multiple optimized code variants.
@@
 %Probably need some words about WHY this 3 pronged approach is a good idea.
-MATLAB optimizing compilers traditionally have not been able to achieve
 
@@
 The analysis algorithm makes implementation choices using the most-constrained first strategy (also called minimum remaining values)
-from the artificial intelligence literature~\cite{Russell:2003mz}.This may need to be updated depending on how we decide to talk about the empirical search. -Jeremy 8/6/08 10:15 AM 
+from the artificial intelligence literature~\cite{Russell:2003mz}.
+%This may need to be updated depending on how we decide to talk about the empirical search. -Jeremy 8/6/08 10:15 AM 
 The compiler chooses the node with the fewest matching implementations (in the linear algebra database) and assigns a matching implementation to the node.
@@
 Figure~\ref{fig:refine-mv-dot} shows two steps of refinement for a matrix-vector multiplication.
-The first step expands the matrix-vector multiplication according to the mv-dot algorithm from figure~\ref{fig:algos}.
+The first step expands the matrix-vector multiplication according to the mv-dot algorithm: $(Ax)(i) = A(i,:)x$.
 The refinement step replaces the $\times$ node with a subgraph that computes the inner product of a row of the matrix---the node labeled $(i,:)$---with the vector, storing the result in the $i$th element of the result vector---the node labeled $(i)$.
@@
 %
-The second pass of refinement introduces another subgraph to implement the inner products (the dot algorithm from figure~\ref{fig:algos}).
+The second pass of refinement introduces another subgraph to implement the inner products.
 The new subgraph iterates through each dense row, as indicated by the $j=1:n$ annotation.
@@
 
+\begin{figure}[thp]
+\centering
+\begin{tabular}{cc}
+\begin{minipage}[b]{.45\textwidth}
+\scriptsize
+\begin{verbatim}
+void vadd(int n, double *y, double *x1,
+               double *x2, double *x3)
+{
+    int i;
+
+/*@ begin Align (x1[],x2[],x3[],y[]) @*/
+/*@ begin Loop (
+  transform Unroll(ufactor=20, parallelize=True)
+    for (i=0; i < n; i++)
+      y[i] = x1[i] + x2[i] + x3[i];
+      ) @*/
+
+    for (i=0; i < n ; i++)
+        y[i] = x1[i] + x2[i] + x3[i];
+
+/*@ end @*/
+/*@ end @*/
+
+}
+
+\end{verbatim}
+\end{minipage}
+&
+\begin{minipage}[b]{.45\textwidth}
+\scriptsize
+\begin{verbatim}
+spec align_unroll {
+ def build  
+ { 
+   arg build_command = 'mpixlc -O3 -qstrict -lm'; 
+ } 
+ def performance_counter  
+ { 
+   arg method = 'basic timer'; 
+   arg repetitions = 10000;
+ } 
+ def performance_params 
+ { 
+   param UF[] = range(1,20);
+ } 
+ def input_params 
+ { 
+   param SIZE = @THESIZE@;
+   decl int n = SIZE;
+   decl double x1[n] = random;   
+   decl double x2[n] = random;   
+   decl double x3[n] = random;   
+   decl double y[n] = 0;
+ } 
+ def search
+ {
+   arg algorithm = 'Exhaustive';
+   arg time_limit = 20;
+   arg run_command = 'cqsub -n 64 -t 10 -q short ';
+   arg num_processes = 64;
+ }
+}
+\end{verbatim}
+\end{minipage}\\
+\end{tabular}
+\caption{Orio example: annotated C source code (left) and a tuning specification for the Blue Gene/P (right).}
+\label{fig:orio-example}
+\end{figure}
+
+We use the VADD operation, $x = w + y + z$, as a simple example of the annotations and tuning specification used by Orio.
+Figure~\ref{fig:orio-example} shows the C code with annotation comments indicating that Orio should perform memory alignment and loop transformations.
+In addition to simple source transformation annotations in the example above, Orio supports a number of other transformations, including loop unroll and jam, composite optimizations, and architecture-specific directives (e.g., for generation of calls to SIMD intrinsics on the Blue Gene/P).
+On the right-hand side of Figure~\ref{fig:orio-example}, a separate tuning specification contains the information required for creating complete executable tests and running them, including variable types, initializations (e.g., random), the search heuristic (in this case, exhaustive search is indicated, other options are random, simplex, and simulated annealing), and execution environment details.
+When parallel resources are available, Orio executes multiple variants in the same parallel job (the granularity is indicated by the \texttt{num\_processes} field in the search portion of the tuning specification.
+At present the tuning specification must be created manually.
+When Orio is used in conjunction with compiler tools, such as the MATLAB compiler described in this paper, it should be possible to generate most of this tuning specification automatically.
+
 \subsection{Search Space Exploration and Evaluation}
@@
 Although the dataflow graphs are quite small in our setting (usually less than 100 nodes), it is still important to prune the search space to avoid an exponential compilation time.
-Our prototype Matlab compiler already performs some pruning: the optimization rewrite rules only apply under certain circumstances.
+Our prototype MATLAB compiler already performs some pruning: the optimization rewrite rules only apply under certain circumstances.
 %
@@
 In this section we present performance results for a number of composed BLAS operations.
-The experiment described in Section~\ref{sec:vadd} were performed on Blue Gene/P, while the remaining experiments were all performed on an Intel Xeon 2.80 GHz Intel(R) E5462 dual quad-core workstation with a 1600MHz front-side bus and 16GB DDR2 FBDIMM RAM.
+The experiment described in Section~\ref{sec:vadd} were performed on Blue Gene/P, while the remaining experiments described in this section were all performed on an Intel Xeon 2.80 GHz Intel(R) E5462 dual quad-core workstation with a 1600Mhz front-side bus and 16GB DDR2 FBDIMM RAM, running Ubuntu 8.04.
 
 % Include peak numbers?
+
+For the Xeon experiments, we used the Intel icc compiler (version 10.1) with the \texttt{-O3 -parallel} optimization flags.
+We generated and tested the following different implementations of each operation:
+\begin{itemize}
+\item C from MATLAB: C code generated by the MATLAB compiler described in Section~\ref{sec:matlab}.
+\item BLAS: A version implemented using multiple BLAS calls and linked with the default BLAS library available on Ubutnu 8.04.
+\item Intel MKL: The same code as the BLAS version, but linked to the Intel MKL libraries.
+\item ATLAS: The same code as the BLAS version, but linked to the ATLAS libraries tuned for this machine.
+\item Orio (Seq.): The C code generated by the MATLAB compiler annotated with sequential-only optimization directives.
+\item Orio (Par.) : The C code generated by the MATLAB compiler annotated with both sequential and parallel optimization directives. This version was compiled with the \texttt{-O3 -openmp} options.
+\end{itemize}
 
@@
 
-We optimized the performance for the VADD operation, $x = w + y + z$, on the Blue Gene/P at Argonne National Laboratory using the IBM XLC compiler version 9.0.
+The VADD operation is simply the addition of three arrays, storing the result in a fourth: $x = w + y + z$.
+We optimized the performance for the VADD operation on the Blue Gene/P at Argonne National Laboratory using the IBM XLC compiler version 9.0.
 We measured the performance for two scenarios: using one core per processor and using all cores.
@@
 The extended Matlab code for the ATAX operation, $y = A' * (A * x)$, and the corresponding performance results are shown in Figure~\ref{fig:atax}.
-The Orio-tuned version that incorporates OpenMP directives achieves the best performance for most problem sizes, while the
+The Orio-tuned version that incorporates PLuTo-generated loop fusion optimizations and Orio unrolling and parallelization directives achieves the best performance for most problem sizes, outperforming the Intel MKL version by a factor of 2 to 5.7 and the compiler-optimized C version by a factor of 4 to 7.
 
@@
 
-The extended Matlab code for the GEMVER operation and the corresponding performance results are shown in Figure~\ref{fig:gemver}.
+The extended MATLAB code for the GEMVER operation and the corresponding performance results are shown in Figure~\ref{fig:gemver}.
+As in the ATAX example, the parallel Orio version achieves the best performance, although in this case the sequential Orio version performance almost as well.
 
@@
 
-$q = A * p$, $s = A' * r$
+% No pluto
+The extended Matlab code for the BiCG Kernel operation,$q = A * p$, $s = A' * r$,
+and the corresponding performance results are shown in Figure~\ref{fig:gesummv}.
+For this operation, the PLuTo analysis did not result in performance improvement, thus we are showing the results obtained only through Orio transformations, which included vectorization, scalar replacement, and unroll and jam.
+Again the best performance was exhibited by the parallel Orio version, while all the BLAS versions performed worse than the compiler-optimized C loop version.
 
@@
 We will next delve into each manual step and create the software that automates it and in effect serves as the ``glue'' between the different tools.
-We will also investigate extending the supported functionality beyond composed BLAS operations, thus benefiting a wider spectrum of applications.
+This would enable application developers to easily produce custom linear algebra libraries tuned for specific architectures.
+We will also investigate extending the supported functionality of our MATLAB compiler beyond composed BLAS operations, thus benefiting a wider spectrum of applications.
 
 Ultimately, the tuning process should result in more than one version of the computation since different versions are optimal for different problem sizes.
-
-This motivates further work on automating more parts of the process, so that these tools can be used by application developers to automatically produce custom linear algebra libraries tuned for specific architectures.
 
@@
 \end{document}
+
+% Current bibtex warnings:
+This is BibTeX, Version 0.99c (Web2C 7.5.4)
+The top-level auxiliary file: paper.aux
+The style file: siam.bst
+Database file #1: paper.bib
+Warning--can't use both volume and number fields in TCE
+(There was 1 warning)
+
+% Put bibtex here only if you do not have access to the subversion repository, otherwise insert them into paper.bib

### 2008-09-20 doc/papers/cgo08/paper.tex
submitted version
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@358 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@
-% CGO 09: http://www.cgo.org/
+%, CGO 09: http://www.cgo.org/
 % svn co https://svn.mcs.anl.gov/repos/performance/orio/doc/papers/cgo08
@@
 
-\setlength{\textwidth}{6.45in}
-\setlength{\oddsidemargin}{0.05in}
-\setlength{\evensidemargin}{0.05in}
-
-\setlength{\topmargin}{-0.5in}
-\setlength{\textheight}{9in}
-
 \usepackage{url}
@@
 \usepackage{cite}
+\usepackage{times}
 \usepackage{amsmath}
-%\usepackage[margin=1in]{geometry}
+\usepackage[margin=1in]{geometry}
+
+\newenvironment{itemizer}{\begin{itemize}\setlength{\parsep}{0cm}\setlength{\itemsep}{-.3em}}{\end{itemize}}
 
@@
 \maketitle
-\doublespacing
-
-\abstract{
-
+%
+\onehalfspacing
+\begin{abstract}
 The growing demand for higher levels of detail and accuracy in results means that the size and complexity of scientific computations is increasing at least as fast as the improvements in processor technology.
@@
 Therefore, computational scientists are often forced to choose between investing too much time in tuning code or  accepting performance that is significantly lower than the best achievable performance on a given architecture.
-
 In this paper, we describe the first steps toward a fully automated system for the optimization of the matrix algebra kernels that are a foundational part of many scientific applications.
-To generate highly optimized code from a high-level MATLAB prototype, we will follow a three-step approach.
+To generate highly optimized code from a high-level MATLAB prototype, we define a three-step approach.
 To begin, we have developed a compiler that converts a MATLAB script into simple C code.
-We then use the source transformation tool PLuTo to optimize that code for coarse-grained parallelism and locality simultaneously.
-The result is an automatically generated code containing loop transformations and OpenMP annotations for exploiting multicore parallelism.
-Finally, we employ the empirical performance tuning system Orio to generate many tuned versions of the same operation using different optimization techniques, such as loop unrolling and memory alignment.
-Orio performs an automated empirical search for selecting the best among the multiple optimized code variants.
-
+We then use the polyhedral optimization system PLuTo to optimize that code for coarse-grained parallelism and locality simultaneously.
+Finally, we annotate the resulting code with performance tuning directives and use the empirical performance tuning system Orio to generate many tuned versions of the same operation using different optimization techniques, such as loop unrolling and memory alignment.
+Orio performs an automated empirical search to select the best among the multiple optimized code variants.
 We discuss performance results showing that the code generated by using our system significantly outperforms not only the original simple C code but also code based on source BLAS, ATLAS-optimized BLAS, and Intel MKL routines.
-}
-
+\end{abstract}
+%
+\doublespacing
+%
 \section{Introduction}
@@
 Despite the immense effort that has been poured into that approach, its success has been limited both by practical constraints (users will not tolerate compile times that extend into several days) and by the amount of detailed information the compiler can obtain about the software to be compiled and the hardware on which it is to execute~\cite{Ken99,Wol89,Wol96}.
-The second path has been to identify kernel routines that constitute the dominant costs of a wide variety of applications.
+The second path is to identify kernel routines that constitute the dominant costs of a wide variety of applications.
 When such kernels can be identified and agreed upon by the members of the community, programmers with the required level of technical knowledge can concentrate on producing these optimized kernel libraries for architectures of interest.
@@
 We are working on a combination of the two approaches designed to overcome some of their shortcomings.
-This paper described our three-pronged approach toward the development of an automatically tuned library for matrix algebra computations.
-In Section~\ref{sec:matlab}, we describe the compiler we have developed that converts a MATLAB script into simple C code.
-In Section~\ref{sec:orio}, we demonstrate how to employ the empirical performance tuning system Orio in conjunction with the PLuTo tool to generate many tuned versions of the same operation using different optimization techniques, such as loop unrolling and memory alignment.
-The result is an automatically generated code containing loop transformations and OpenMP annotations for exploiting multicore parallelism.
+This paper describes our multipronged approach toward the development of software infrastructure for \emph{generating} automatically tuned libraries for matrix algebra computations.
+We briefly overview relevant prior and current research efforts in Section~\ref{sec:background}.
+In Section~\ref{sec:approach} we describe our MATLAB to C compiler and the empirical performance tuning system Orio and its use in conjunction with the PLuTo tool suite to generate and empirically evaluate many tuned versions of the C code generated by the MATLAB compiler.
 In Section~\ref{sec:experiments}, we provide performance results showing that the code generated by using our system significantly outperforms not only the original simple C code but also code based on source BLAS, ATLAS-optimized BLAS, and Intel MKL routines.
@@
 An alternative is the use of tuned libraries of key numerical algorithms (e.g., BLAS~\cite{Dongarra:1990fk} and LAPACK~\cite{LAPACK} for dense linear algebra, SPARSPAK~\cite{sparspak} for sparse linear algebra, and BLITZ++~\cite{blitz} to support scientific computing in C++, or GPULib by Tech-X, an open source library for accelerating numerical computations on Graphics Processing Units~\cite{gpulib}, to name just a few).
-Such libraries are difficult to develop, and they generally cannot adapt to execution context: as we note in Section~\ref{sec:composed}, a series of calls to library routines can incur high memory costs.
+Such libraries are difficult to develop, and they generally cannot adapt to execution context: as we note in Section~\ref{sec:approach}, a series of calls to library routines can incur high memory costs.
 
@@
 Maintaining performance of the code as the application and hardware evolves means that the tuning process may need to be repeated.
-We propose to investigate ways to capture this performance tuning expertise in a way that can be integrated into the application.
+Our efforts aim to capture this performance tuning expertise in a way that can be integrated into the application.
 We will thus enable performance portability and ensure that the improvements to the application are long-lasting.
@@
 LoopTool~\cite{LoopTool}, developed at Rice University, supports annotation-based loop fusion, unroll-and-jam, skewing and tiling.
-A new tool, POET~\cite{POET} also supports a number of loop transformations.
+A relatively new tool, POET~\cite{POET} also supports a number of loop transformations.
 POET offers a complex template-based syntax for defining transformations in a language-independent manner.
@@
 
-\section{Composed Operations}
-\label{sec:composed}
+\section{Our Approach to Optimizing Composed BLAS Operations}
+\label{sec:approach}
+
+%\section{Optimizing Composed Operations}
+%\label{sec:composed}
 
@@
 Writing programs in this way promotes readability and maintainability but can be costly in terms of memory efficiency.
-Specifically, the retrieval of a large order matrix at each routine call can have a profound effect on performance even when highly tuned implementations of the BLAS (e.g ~\cite{Bilmes:1997ye,Whaley:1998fk,IntelMath:oq,ESSL:kl,Goto:2006fk}) are used.
+Specifically, the retrieval of a large order matrix at each routine call can have a profound effect on performance even when highly tuned implementations of the BLAS (e.g.~\cite{Bilmes:1997ye,Whaley:1998fk,IntelMath:oq,ESSL:kl,Goto:2006fk}) are used.
 
-The authors and other researchers have observed that it can be much more efficient to call a single specialized routine that performs multiple operations than to make successive calls to separate BLAS ~\cite{Ashby:uq,Blackford:2002vn,Stanley:CSD-98-992,baker03blgmres,baker03lgmres,Dennis:2005tg,Howell:2008,gropp01,Vuduc:2003kl}.
-Single routines that carry out more than one linear algebra operation are known as {\t composed} BLAS.
+The authors and other researchers have observed that it can be much more efficient to call a single specialized routine that performs multiple operations than to make successive calls to separate BLAS routines ~\cite{Ashby:uq,Blackford:2002vn,baker03blgmres,baker03lgmres,Dennis:2005tg,Howell:2008,gropp01,Vuduc:2003kl}.
+Single routines that carry out more than one linear algebra operation are known as \emph{composed} BLAS.
 As an example, consider the pair of matrix-vector products $q  = Ap$, $s  = A^Tr$, where $A$ is a matrix and $p$, $q$, $r$, and $s$ are vectors, that comprise the computational
@@
 These two operations can be implemented as a pair of calls to the BLAS routine {\tt GEMV} or they can be rewritten as a composed BLAS consisting of the doubly nested loop encompassing both matrix-vector multiplies.
-In the former case, the matrix $A$ is accessed twice while, in the latter, is it accessed only once.
-Our preliminary results indicate that this loop fusion leads to a routine that delivers 90\% more megaflops than does a pair of calls to the highly optimized Goto BLAS~\cite{Goto:2006fk} {\tt GEMV} routines for large matrix orders.
-
+In the former case, the matrix $A$ is accessed twice, while in the latter it is accessed only once.
+Our preliminary results in prior research and in this paper indicate that this loop fusion leads to a routine that delivers significantly better performance than does a pair of calls to the best optimized BLAS {\tt GEMV} routines for large matrix orders.
 Composed routines are the focus of the work presented here, but much of what we discuss generalizes to a much broader array of computations.
-Extending our approach will be the subject of future work.
 
@@
 
-\section{Our Approach}
-\label{sec:approach}
-
-To generate highly optimized code from a MATLAB prototype, we will follow a three-step approach, illustrated in Figure~\ref{fig:process}.
-To begin, we have developed a compiler that converts a MATLAB script into simple C code.
-We then (optionally) use the source-to-source automatic parallelization tool PLuTo\cite{Pluto} developed at Ohio State University to optimize for coarse-grained parallelism and locality simultaneously.
-Using the results of the PLuTo analysis, we insert annotations into the C code, which are then processed by our extensible annotation system Orio to generate many tuned versions of the same operation using different optimization parameters.
-Orio then performs an empirical search to select the best among the multiple optimized code variants.
-
-%Probably need some words about WHY this 3 pronged approach is a good idea.
-
 \begin{figure}[htbp]
@@
 
-\caption{Overview of the code generation and tuning process.}
+\caption{Overview of the code generation and tuning process, optional steps are indicated with dashed arrows.}
 \label{fig:process}
 \end{figure}
+To generate highly optimized code from a MATLAB prototype of the composed BLAS operation, we follow a three-step approach, illustrated in Figure~\ref{fig:process}.
+To begin, we have developed a compiler that converts a MATLAB script into simple C code~\cite{Siek}.
+After generating the C code from the high-level MATLAB prototype, we (optionally) use the source-to-source automatic parallelization tool PLuTo\cite{Pluto} developed at Ohio State University to optimize for coarse-grained parallelism and locality simultaneously.
+Using the results of the PLuTo analysis, we insert annotations into the C code, which are then processed by our extensible annotation system Orio to generate many tuned versions of the same operation using different optimization parameters.
+Orio then performs an empirical search to select the best among the multiple optimized code variants.
+
+%Probably need some words about WHY this multipronged approach is a good idea.
+
+In the remainder of this section we describe each of the tools developed by the authors of this paper, namely the MATLAB to C compiler and the Orio empirical tuning tool.
 
@@
 
-Orio~\cite{Norris:2007} is an empirical tuning tool that takes as input annotated C code and generates multiple tuned versions of the annotated code.
+Orio~\cite{Norris:2007} is an empirical tuning tool that takes as input annotated C code and generates multiple tuned versions.
 Orio does not parse the application code; rather, it relies on annotations expressed as structured comments, which contain a simplified expression of the computation, as well as directives on what transformations to apply.
-
-% Explain how Pluto is used
 
@@
 
-\begin{figure}[thp]
+We use the VADD operation, $x = w + y + z$, as a simple example of the annotations and tuning specification used by Orio.
+Figure~\ref{fig:orio-example} shows the C code with annotation comments indicating that Orio should perform memory alignment and loop transformations.
+In addition to simple source transformation annotations in the example above, Orio supports a number of other transformations, including loop unroll and jam, composite optimizations, and architecture-specific directives (e.g., for generation of calls to SIMD intrinsics on the Blue Gene/P).
+
+\begin{figure}[htp]
 \centering
@@
 \end{tabular}
-\caption{Orio example: annotated C source code (left) and a tuning specification for the Blue Gene/P (right).}
+\caption{Orio example: Annotated C source code (left) and a sample tuning specification for the Blue Gene/P (right).}
 \label{fig:orio-example}
@@
 
-We use the VADD operation, $x = w + y + z$, as a simple example of the annotations and tuning specification used by Orio.
-Figure~\ref{fig:orio-example} shows the C code with annotation comments indicating that Orio should perform memory alignment and loop transformations.
-In addition to simple source transformation annotations in the example above, Orio supports a number of other transformations, including loop unroll and jam, composite optimizations, and architecture-specific directives (e.g., for generation of calls to SIMD intrinsics on the Blue Gene/P).
-On the right-hand side of Figure~\ref{fig:orio-example}, a separate tuning specification contains the information required for creating complete executable tests and running them, including variable types, initializations (e.g., random), the search heuristic (in this case, exhaustive search is indicated, other options are random, simplex, and simulated annealing), and execution environment details.
+On the right-hand side of Figure~\ref{fig:orio-example}, we show a separate tuning specification example, which contains the information required for creating complete executable tests and running them, including variable types, initializations (e.g., random), the search heuristic (in this case, exhaustive search is indicated, other options are random, simplex, and simulated annealing), and execution environment details.
 When parallel resources are available, Orio executes multiple variants in the same parallel job (the granularity is indicated by the \texttt{num\_processes} field in the search portion of the tuning specification.
 At present the tuning specification must be created manually.
-When Orio is used in conjunction with compiler tools, such as the MATLAB compiler described in this paper, it should be possible to generate most of this tuning specification automatically.
+When Orio is used in conjunction with compiler tools, such as the MATLAB compiler described in this paper, it should eventually be possible to generate most of this tuning specification automatically.
 
@@
 We generated and tested the following different implementations of each operation:
-\begin{itemize}
+\begin{itemizer}
 \item C from MATLAB: C code generated by the MATLAB compiler described in Section~\ref{sec:matlab}.
@@
 \item Orio (Seq.): The C code generated by the MATLAB compiler annotated with sequential-only optimization directives.
-\item Orio (Par.) : The C code generated by the MATLAB compiler annotated with both sequential and parallel optimization directives. This version was compiled with the \texttt{-O3 -openmp} options.
-\end{itemize}
+\item Orio (Par.) : The C code generated by the MATLAB compiler annotated with both sequential and parallel optimization directives (note that the code is the same as for Orio (Seq.), the only difference is the value of a parallelization parameter in the separate tuning specification).
+This version was compiled with the \texttt{-O3 -openmp} options.
+\end{itemizer}
 
@@
 
-The VADD operation is simply the addition of three arrays, storing the result in a fourth: $x = w + y + z$.
+The VADD operation computes the sum of three arrays, storing the result in a fourth: $x = w + y + z$.
 We optimized the performance for the VADD operation on the Blue Gene/P at Argonne National Laboratory using the IBM XLC compiler version 9.0.
-We measured the performance for two scenarios: using one core per processor and using all cores.
+We measured the performance for two scenarios: using one core per processor and using all 4 cores on each processor.
 In the single-core case, the codes were compiled with the following options: \texttt{-O3 -qstrict -qarch=450d -qtune=450 -qhot -qsmp=noauto}.
 For the multicore case, the only difference in the compilation was that for the non-Orio versions, the \texttt{-qsmp=auto} option was used.
+The parallel Orio version differs from the sequential in that it contains parallelization directives, which result in the placement of OpenMP pragmas in the generated code.
 
@@
 The extended Matlab code for the ATAX operation, $y = A' * (A * x)$, and the corresponding performance results are shown in Figure~\ref{fig:atax}.
-The Orio-tuned version that incorporates PLuTo-generated loop fusion optimizations and Orio unrolling and parallelization directives achieves the best performance for most problem sizes, outperforming the Intel MKL version by a factor of 2 to 5.7 and the compiler-optimized C version by a factor of 4 to 7.
+The Orio-tuned version that incorporates PLuTo-generated loop fusion optimizations and Orio parallelization directives achieves the best performance for most problem sizes, outperforming the Intel MKL version by a factor of 2 to 5.7 and the compiler-optimized C version by a factor of 4 to 7.
+The optimizations performed by both Orio versions include scalar replacement, vectorization, and unroll and jam.
 
@@
 The extended MATLAB code for the GEMVER operation and the corresponding performance results are shown in Figure~\ref{fig:gemver}.
-As in the ATAX example, the parallel Orio version achieves the best performance, although in this case the sequential Orio version performance almost as well.
-
+Here we used the same PLuTo and Orio optimizations as for the ATAX example and similarly the parallel Orio version achieves the best performance, although in this case the sequential Orio version performs almost as well suggesting that the compiler was not able to parallelize the code very effectively.
+For this operation, there are substantial performance differences between different BLAS versions, with the Intel MKL version achieving performance close to that of the simple compiler code.
 \begin{figure}[htp]
@@
 and the corresponding performance results are shown in Figure~\ref{fig:gesummv}.
+Unlike all the other operations we tested, the best performing version is the ATLAS one, which substantially outperforms all the alternatives.
+The next best performance is exhibited by the parallel Orio version.
+This result motivates the development of transformations that enable the generation of code with calls to existing optimizing libraries in addition to the low-level loop transformations targeted by the current tools (PLuTo and Orio).
 
@@
 
-The positive initial results from our three-pronged approach to generating tuned linear algebra routines motivate several future lines of investigation.
+We have described an approach to generating tuned linear algebra libraries from high-level annotated MATLAB code that involves a suite of tools to (1) translate the MATLAB code to C, (2) analyze the resulting loops and identify locatility and parallelism-enhancing optimizations using PLuTo, and (3) annotate the resulting C code with syntactic performance directives and use Orio to generate multiple optimized versions and empirically select the one with the best performance.
+Preliminary results from experiments with several composed BLAS operations show that in almost all cases, the optimized code generated by this suite of tools significantly outperforms the versions using tuned BLAS and aggressive compiler optimizations.
+
+The positive initial results from our approach to generating tuned linear algebra routines motivate several future lines of investigation.
 At present there are a number of manual steps involved in obtaining a compiled tuned linear algebra library from the high-level extended MATLAB descriptions.
-We will next delve into each manual step and create the software that automates it and in effect serves as the ``glue'' between the different tools.
+We will work on interfacing our separately developed tools more closely and automating their interactions.
 This would enable application developers to easily produce custom linear algebra libraries tuned for specific architectures.
@@
 
-Ultimately, the tuning process should result in more than one version of the computation since different versions are optimal for different problem sizes.
+Orio is growing in a number of directions.
+While it automates much of the tuning process, a number of steps still require nontrivial manual input.
+Orio's modular design allows new transformations to be added easily, which presents many opportunities for expanding the existing set of optimizations with new ones implemented from scratch or through interfaces to other tools, as we have successfully done with PLuTo.
+We are also working on generating code that contains special cases for different input parameter values since in many cases different optimizations are optimal for different input sizes.
+We will also continue developing the parallel empirical search infrastructure; it is clear that more feedback between the MATLAB compiler and Orio will result in automation of the performance directive annotation process and a much more effective exploration of the search space.
+We plan to use a fast analytic cost function in the MATLAB compiler implementation to prune out regions of the search space that will not produce competitive implementations.
+The cost function is based on a model of the computer architecture, but does not necessarily account for all the details.
+Once the search space is narrowed, we will use the Orio empirical testing results with representative data sets to determine the exact performance of the alternatives and choose the best one.
 
-% BN: moved from search space section above
-%Yotov:2005kx
-We plan to use a fast analytic cost function to prune out regions of the search space that will not produce competitive implementations.
-The cost function is based on a model of the computer architecture, but does not necessarily account for all the details.
-Once the search space is narrowed, we will use empirical testing with representative data sets to determine the exact performance of the alternatives and choose the best one.
+Finally, while Orio currently generates various loop transformations, it does not generate calls to existing library implementations.
+For this paper, we implemented the BLAS versions of all operations manually.
+Our past success with identifying Blue Gene/L intrinsics from expression graphs motivate further research on identifying BLAS and other established library operations from the expression graphs of the MATLAB code.
 
 \singlespacing
-\bibliographystyle{siam}
+%% or stuff it in a footnote
+\section*{Acknowledgments}
+This work was supported by the Mathematical, Information, and
+Computational Sciences Division subprogram of the Office of Advanced
+Scientific Computing Research, Office of Science, U.S. Dept. of Energy,
+under Contract DE-AC02-06CH11357. 
+%TODO: need Colorado and Ohio acknowledgments
+
+\bibliographystyle{acm}
 \bibliography{paper}
+
+\vfill
+\begin{flushright}
+\scriptsize
+\framebox{\parbox{2.4in}{The submitted manuscript has been created by
+UChicago Argonne, LLC, Operator of Argonne National Laboratory
+("Argonne").  Argonne, a U.S. Department of Energy Office
+of Science laboratory, is operated under Contract No.
+DE-AC02-06CH11357.  The U.S. Government retains for itself, and
+others acting on its behalf, a paid-up, nonexclusive, irrevocable
+worldwide license in said article to reproduce, prepare derivative works,
+distribute copies to the public, and perform publicly and display
+publicly, by or on behalf of the Government.}}
+\normalsize
+\end{flushright}
+
 \end{document}

### 2008-10-13 doc/papers/cgo09/paper.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@361 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

--- 2008-09-20 doc/papers/cgo08/paper.tex
@@
-%, CGO 09: http://www.cgo.org/
+% CGO 09: http://www.cgo.org/
 % svn co https://svn.mcs.anl.gov/repos/performance/orio/doc/papers/cgo08
@@
 
-\author{
-Elizabeth Jessup,$^1$ Boyana Norris,$^2$ Albert Hartono,$^3$ and Jeremy Siek$^4$\\
-\\
-\vspace{-.05in}
-{\small $^1$University of Colorado at Boulder Dept. of Computer Science, Boulder, CO~80309}\\
-\vspace{-.05in}
-{\small \url{jessup@cs.colorado.edu}}\\
-\vspace{-.05in}
-{\small $^2$Argonne National Laboratory Mathematics and Computer Science Division, Argonne, IL~60439}\\
-\vspace{-.05in}
-{\small \url{norris@mcs.anl.gov}}\\
-\vspace{-.05in}
-{\small $^3$Ohio State University Dept. of Computer Science and Engineering, Columbus, OH~43210}\\
-\vspace{-.05in}
-{\small \url{hartonoa@cse.ohio-state.edu}}\\
-\vspace{-.05in}
-{\small $^4$University of Colorado at Boulder Dept. of Electrical and Computer Engineering, Boulder, CO~80309}\\
-\vspace{-.05in}
-{\small \url{jeremy.siek@colorado.edu}}
-}
+%\author{
+%Elizabeth Jessup,$^1$ Boyana Norris,$^2$ Albert Hartono,$^3$ and Jeremy Siek$^4$\\
+%\\
+%\vspace{-.05in}
+%{\small $^1$University of Colorado at Boulder Dept. of Computer Science, Boulder, CO~80309}\\
+%\vspace{-.05in}
+%{\small \url{jessup@cs.colorado.edu}}\\
+%\vspace{-.05in}
+%{\small $^2$Argonne National Laboratory Mathematics and Computer Science Division, Argonne, IL~60439}\\
+%\vspace{-.05in}
+%{\small \url{norris@mcs.anl.gov}}\\
+%\vspace{-.05in}
+%{\small $^3$Ohio State University Dept. of Computer Science and Engineering, Columbus, OH~43210}\\
+%\vspace{-.05in}
+%{\small \url{hartonoa@cse.ohio-state.edu}}\\
+%\vspace{-.05in}
+%{\small $^4$University of Colorado at Boulder Dept. of Electrical and Computer Engineering, Boulder, CO~80309}\\
+%\vspace{-.05in}
+%{\small \url{jeremy.siek@colorado.edu}}
+%}
 
@@
 To generate highly optimized code from a MATLAB prototype of the composed BLAS operation, we follow a three-step approach, illustrated in Figure~\ref{fig:process}.
-To begin, we have developed a compiler that converts a MATLAB script into simple C code~\cite{Siek}.
-After generating the C code from the high-level MATLAB prototype, we (optionally) use the source-to-source automatic parallelization tool PLuTo\cite{Pluto} developed at Ohio State University to optimize for coarse-grained parallelism and locality simultaneously.
+To begin, we have developed a compiler that converts a MATLAB script into simple C code.
+%~\cite{Siek}. 
+After generating the C code from the high-level MATLAB prototype, we (optionally) use the source-to-source automatic parallelization tool PLuTo\cite{Pluto} %developed at Ohio State University
+to optimize for coarse-grained parallelism and locality simultaneously.
 Using the results of the PLuTo analysis, we insert annotations into the C code, which are then processed by our extensible annotation system Orio to generate many tuned versions of the same operation using different optimization parameters.
@@
 
-We have developed a compiler that translates MATLAB scripts into C code.
+In this section we describe the design and operation of the MATLAB to C compiler we have developed.
 Figure~\ref{fig:compiler} gives an overview of the compilation process.
@@
 
-Orio~\cite{Norris:2007} is an empirical tuning tool that takes as input annotated C code and generates multiple tuned versions.
+%~\cite{Norris:2007}
+Orio is an empirical tuning tool that takes as input annotated C code and generates multiple tuned versions.
 Orio does not parse the application code; rather, it relies on annotations expressed as structured comments, which contain a simplified expression of the computation, as well as directives on what transformations to apply.
@@
 % BN: moved to its own section since multiple tools deal with the search space problem.
-The set of all possible combinations of implementation and optimization choices is called the search space~\cite{Kisuki:2000uq,Triantafyllis:2003uq, Cooper:2005kx}.
+The set of all possible combinations of implementation and optimization choices is called the search space~\cite{Kisuki:2000uq,Triantafyllis:2003uq,Cooper:2005kx}.
 Although the dataflow graphs are quite small in our setting (usually less than 100 nodes), it is still important to prune the search space to avoid an exponential compilation time.
@@
 The VADD operation computes the sum of three arrays, storing the result in a fourth: $x = w + y + z$.
-We optimized the performance for the VADD operation on the Blue Gene/P at Argonne National Laboratory using the IBM XLC compiler version 9.0.
+We optimized the performance for the VADD operation on a Blue Gene/P
+%at Argonne National Laboratory
+using the IBM XLC compiler version 9.0.
 We measured the performance for two scenarios: using one core per processor and using all 4 cores on each processor.
@@
 in
-  A : row matrix, x : vector
+  A : row matrix,
+  x : vector
 out
@@
 {
-  B = A + u1 * v1' + u2 * v2'
+  B = A + u1 * v1'
+           + u2 * v2'
   x = b * (B' * y) + z
@@
 {
-  y = a * (A * x) + b * (B * x)
+  y = a * (A * x)
+       + b * (B * x)
 }
@@
 %% or stuff it in a footnote
-\section*{Acknowledgments}
-This work was supported by the Mathematical, Information, and
-Computational Sciences Division subprogram of the Office of Advanced
-Scientific Computing Research, Office of Science, U.S. Dept. of Energy,
-under Contract DE-AC02-06CH11357. 
+%\section*{Acknowledgments}
+%This work was supported by the Mathematical, Information, and
+%Computational Sciences Division subprogram of the Office of Advanced
+%Scientific Computing Research, Office of Science, U.S. Dept. of Energy,
+%under Contract DE-AC02-06CH11357. 
 %TODO: need Colorado and Ohio acknowledgments
@@
 
-\vfill
-\begin{flushright}
-\scriptsize
-\framebox{\parbox{2.4in}{The submitted manuscript has been created by
-UChicago Argonne, LLC, Operator of Argonne National Laboratory
-("Argonne").  Argonne, a U.S. Department of Energy Office
-of Science laboratory, is operated under Contract No.
-DE-AC02-06CH11357.  The U.S. Government retains for itself, and
-others acting on its behalf, a paid-up, nonexclusive, irrevocable
-worldwide license in said article to reproduce, prepare derivative works,
-distribute copies to the public, and perform publicly and display
-publicly, by or on behalf of the Government.}}
-\normalsize
-\end{flushright}
+%\vfill
+%\begin{flushright}
+%\scriptsize
+%\framebox{\parbox{2.4in}{The submitted manuscript has been created by
+%UChicago Argonne, LLC, Operator of Argonne National Laboratory
+%("Argonne").  Argonne, a U.S. Department of Energy Office
+%of Science laboratory, is operated under Contract No.
+%DE-AC02-06CH11357.  The U.S. Government retains for itself, and
+%others acting on its behalf, a paid-up, nonexclusive, irrevocable
+%worldwide license in said article to reproduce, prepare derivative works,
+%distribute copies to the public, and perform publicly and display
+%publicly, by or on behalf of the Government.}}
+%\normalsize
+%\end{flushright}
 

### 2008-10-13 doc/papers/ipdps09/abstract.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@363 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

In many scientific applications, significant time is spent in tuning codes for a particular high-performance architecture. Tuning approaches range from the relatively nonintrusive (e.g., by using compiler options) to extensive code modifications that attempt to exploit specific architecture features. Intrusive techniques often result in code changes that are not easily reversible, which can negatively impact readability, maintainability, and performance on different architectures. We introduce an extensible annotation-based empirical tuning system called Orio, which is aimed at improving both performance and productivity by enabling software developers to insert annotations in the form of structured comments into their source code that trigger a number of low-level performance optimizations on a specified code fragment.  To maximize the performance tuning opportunities, the annotation processing infrastructure is designed to support both architecture-independent and architecture-specific code optimizations. Given the annotated code as input, Orio generates many tuned versions of the same operation and empirically evaluates the alternatives to select the best performing version for production use. We have also enabled the use of the PLuTo automatic parallelization tool in conjunction with Orio to generate efficient OpenMP-based parallel code. We describe our experimental results involving a number of computational kernels, including dense array and sparse matrix operations.

### 2008-10-13 doc/papers/ipdps09/conclusions.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@363 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

\section{Conclusions and Future Work}
\label{sec:conclusions}

%Future directions:
%- Support for Fortran
%- Supports for various architecture-specific optimizations (e.g., SSE)
%- Automatic tuning for various problem sizes
%- Automatic parallelization for multi-core machines
%- Tuning MPI code (communication minimization)
%- Tuning for GPGPU

%- Build a decision tree that automatically directs to the best tuned code for a given problem size

We have described the design and implementation of Orio, an extensible Python
software system for defining annotation-based performance-improving
transformations. Our experiments with a number of different types of
computations on two different architectures show that Orio can deliver
performance improvements when used alone or in conjunction with other source
transformation tools.

Orio is a new tool under active development; future work includes (but is not
limited to) providing support for annotating and generating Fortran code,
defining new annotation languages and corresponding transformation modules,
e.g., using matrix notation for linear algebra operations, and integration
with other source transformation tools through new optimization modules.

\vspace{-.1in} 
\paragraph{Acknowledgments.} We would like to thank
Uday Bondhugula of Ohio State University for many productive discussions and
his valuable help with PLuTo.

\vspace{-.1in}

### 2008-10-13 doc/papers/ipdps09/integration.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@363 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

\section{PLuTo-Orio Integration}
\label{sec:integration}

A number of source-to-source transformation tools for performance
optimization exist. Using these tools to achieve (near) optimal performance
on different architectures, however, is still a nontrivial task that requires
significant architectural and compiler expertise. By combining code
transformation tools with an empirical tuning system, such as Orio, we can
reduce the amount of manual effort and automatically determine the
transformations that result in the best performance. We show in this section
how Orio has been extended with an external transformation program, called
PLuTo~\cite{uday08pldi}. This integration also demonstrates the easy
extensibility of Orio and the ability to leverage other source transformation
approaches.

PLuTo is a source-to-source automatic transformation tool aimed at optimizing
a sequence of nested loops for data locality and coarse-grained parallelism
simultaneously. PLuTo employs a polyhedral model of arbitrary loop nests,
where the dynamic instance (iteration) of each statement is viewed as an
integer point in a well-defined space called the statement's polyhedron. This
statement representation and a precise characterization of data dependences
enable PLuTo to construct mathematically correct complex loop
transformations. PLuTo's polyhedral-based transformations result in improved
cache locality and loops parallelized for multicore architecture.

Figure~\ref{fig:pluto-orio} outlines the overall structure of the PLuTo-Orio
integrated system, which is implemented as a new optimization module in
Orio.
%% BN: the detailed description below is good, but commented out for now for space.
%Initially, the transformation process takes two kinds of input:
%the loop code to be optimized and the optimization parameters code. The loop
%code is embedded in the annotation body block, while the performance
%parameters code is written in the module body block. A simple parser
%implemented inside the new module parses the performance parameters code to
%extract their values. 
In addition to the loop code to be optimized, the module parses performance
parameters, which include tile sizes, unroll factors, loop order
(permutation), as well as several boolean values for triggering OpenMP
parallelization, scalar replacement, and auto-vectorization. Using these
parameter values, PLuTo then performs polyhedral transformations (e.g.,
two-level tiling and OpenMP parallelization) on the input loop code. The
resulting generated code is subsequently passed to the widely-available
profiling tool gprof~\cite{gprof} for hotspot detection. 
%The statistical
%information produced by gprof provides accurate locations (line numbers) of
%the loop nests where the PLuTo-generated code spends most of its execution
%time. 
The identified hotspot loop nests are then decorated with Orio's performance
annotations for complementary syntactic transformations (e.g., loop
permutation, loop unroll/jamming, scalar replacement, and explicit
auto-vectorization). Finally, Orio optimizes the annotated hotspots as 
described in Section~\ref{sec:implm}.

\begin{figure}[tb]
\begin{center}
\includegraphics[width=0.6\textwidth]{figures/pluto-orio.eps}    
\end{center}   
\vspace{-.2in}
\caption{Integration of PLuTo into Orio's code transformation module.}
\label{fig:pluto-orio}
\end{figure}

PLuTo also performs syntactic loop unroll/jamming as a post-processing
pass. The target loops, however, are limited only to innermost loops with a
maximum depth of two (i.e., 1-D unrolling and 2-D unroll/jamming). So we
choose to use the loop unroll/jamming transformation already available in
Orio, which is more flexible because it can be applied to loop nests of
depths larger than two.

At present Orio does not employ any data dependency analysis when performing
syntactic transformations. Therefore, there is no guarantee that the code
generated after syntactic transformations is correct. Because of this, the
tuning process using the integrated PLuTo-Orio system is currently
semi-automatic---user involvement is required to decide whether it is safe
to apply a syntactic transformation on an identified hotspot.
%The user only needs
%to check the transformation legalities once, since the hotspots
%consistently have similar basic loop structure for different
%PLuTo-generated code versions. Once all illegal syntactic
%transformations are known by the user, they must be switched off so
%that the tuning process can proceed fully automatic with no more human
%interventions.

### 2008-10-13 doc/papers/ipdps09/paper.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@363 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

\documentclass[12pt]{article}
%\usepackage{latex8}
%\usepackage[]{rotating} 
\usepackage{ifpdf}
\usepackage{url} 
\usepackage{comment} 
\usepackage{cite} 
\usepackage{graphicx} 
\usepackage{subfigure}  
\usepackage{wrapfig}
\usepackage{times}
\usepackage{amsmath,amsthm}
\usepackage[letterpaper,margin=1in]{geometry}

\ifpdf
\DeclareGraphicsExtensions{.pdf,.png,.jpg}
\else
\DeclareGraphicsExtensions{.eps,.ps}
\fi

\newenvironment{itemizer}{\begin{itemize}\setlength{\parsep}{0cm}\setlength{\itemsep}{-.3em}}{\end{itemize}}

\long\def\comment#1{}
\date{}

\title{\vspace{-.6in}Annotation-Based Empirical Performance Tuning\\ Using Orio}

\author{
Albert Hartono$^1$, Boyana Norris$^2$, and P. Sadayappan$^1$ \\
{\small $^1$Ohio State University, Dept. of Computer Science and Engineering, Columbus, OH 43210}\\
{\small \texttt{\{hartonoa, saday\}@cse.ohio-state.edu}}\\
{\small $^2$Argonne National Laboratory, Mathematics and Computer Science Division, Argonne, IL 60439}\\
{\small \texttt{norris@mcs.anl.gov}}
}

\begin{document}

\maketitle

\begin{abstract}
\input{abstract}
\end{abstract}

\input{motivation}
\input{related}
\input{implementation}
\input{integration}
\input{results} 
\input{conclusions}

\small
\bibliographystyle{abbrv} 
\bibliography{paper}

\end{document}

### 2008-10-13 doc/papers/ipdps09/results.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@363 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

\section{Experimental Results} 
\label{sec:results}

In this section, we discuss the performance results of several experiments on
a multicore Intel Xeon workstation and a Blue Gene/P (both at Argonne). The
workstation has dual quad-core E5462 Xeon processors (8 cores total) running
at 2.8 GHz (1600 MHz FSB) with 32 KB L1 cache, 12 MB of L2 cache (6 MB shared
per core pair), and 2 GB of DDR2 FBDIMM RAM, running Linux kernel version
2.6.25 (x86-64). Each compute node of the Blue Gene/P is equipped with four
850MHz IBM PowerPC 450 processors with a dual floating-point unit and 2GB
total memory per node, private L1 (32 KB) and L2 (4 MB) caches, a shared L3
cache (8 MB), and running a proprietary lightweight operating system. We used
version 10.1 of the Intel and version 9.0 of the IBM XL C/C++ V9.0 compilers
on the Xeon and Blue Gene/P, respectively. Because of space considerations,
here we discuss a limited number of computational kernels; more performance
results for different computations are available at the Orio project
website~\cite{OrioURL}.

\subsection{Sequence of Linear Algebra Operations} 
\label{sec:axpy4-results}

In this experiment, we tuned the performance of the AXPY-4 operation (see
Figure~\ref{fig:orio-example}) on a single node of the Blue Gene/P machine
using the IBM xlc compiler.  We measured the performance for two scenarios:
using a single core per node and using all four cores. The results are shown
in Figures~\ref{fig:axpy4-bgp-seq} and~\ref{fig:axpy4-bgp-par},
respectively. The single-core scenario was compiled with the following options:
\texttt{-O3 -qstrict -qarch=450d -qtune=450 -qhot -qsmp=noauto};  the
multicore scenario differs in the use of \texttt{-qsmp=auto} for the non-Orio
versions. The parallel Orio version contains OpenMP parallelization
directives in the generated code, thus necessitating the use of the
\texttt{-qsmp=omp:noauto} compiler option. Included are performance numbers for four
code variants: a simple loop implementation without any library calls
(labeled ``Compiler-optimized''), two BLAS-based implementations that use
Goto BLAS~\cite{Goto:2006fk} and the ESSL~\cite{ESSL} libraries, and the
Orio-tuned version.

\begin{figure}[thb] 
\begin{center} 
  \subfigure[Sequential (single core)]{ 
  \includegraphics[width=.4\textwidth]{figures/axpy4_bgp/seq.eps}  
  \label{fig:axpy4-bgp-seq} 
  } 
  \subfigure[Parallel (four cores)]{ 
  \includegraphics[width=.4\textwidth]{figures/axpy4_bgp/par.eps}  
  \label{fig:axpy4-bgp-par} 
  } 
\end{center}
\vspace{-.2in} 
\caption{Performance of AXPY-4 operations on the Blue Gene/P.} 
\label{fig:axpy4-bgp-results} 
\end{figure}

The performance results shown in Figure~\ref{fig:axpy4-bgp-results} indicate
that the code tuned by Orio consistently outperforms the other three versions
for both the sequential and parallel cases. We observe that even for a simple
algebraic operation, such as the composed AXPY routines, the compiler alone
is unable to yield performance comparable to the empirically tuned
version. Moreover, implementations that rely on calls to multiple tuned
library routines (e.g., Goto BLAS and ESSL) suffer from loss of both spatial
and temporal localities, resulting in inferior memory performance.
%The sequential BLAS-based implementation perform better than the
%sequential single loop implementation. while for the parallel case, the 
%simple loop implementation is more efficient than the BLAS codes.

\subsection{Sparse Matrix Computations}

In this section we examine the effectiveness of Orio in optimizing key
computations in PETSc~\cite{petsc-user-ref}, a toolkit for the parallel
numerical solution of partial differential equations, by empirically tuning
one of its heavily used kernels, sparse matrix-vector multiplication
(SpMV). SpMV dominates the performance of various scientific applications;
yet, traditional implementations of sparse kernels exhibit relatively poor
performance because of the use of indirect addressing for accessing the
matrix data elements.
%In
%order to attain higher performance, SpMV requires selecting a compact data
%structure and code transformations that best exploit properties of both the
%sparse matrix and the underlying architecture. Hence, the need for
%optimization and runtime tuning is a big distinction from the dense
%matrix-vector multiplication.

The SpMV operation computes $\forall_{A_{i,j}} \neq 0:y_{i}
\leftarrow y_{i} + A_{i,j} \cdot x_{j}$, where $A$ is a sparse matrix,
and $x$, $y$ are dense vectors. Each element of $A$ is used precisely once
and element reuse is only possible for $x$ and $y$. Thus, to optimize SpMV,
one should use compact data structures for $A$ and try to maximize temporal
reuse of $x$ and $y$. One of the most commonly used data structures for
storing a sparse matrix is compressed sparse row (CSR) storage,
%~\cite{vuduc-thesis} 
illustrated in Figure~\ref{fig:spmv}(a). Elements in each row of $A$ are
packed together in a dense array, \texttt{Aval}, and a corresponding array of
integers \texttt{Aind} stores the column indices. The \texttt{Aptr} array
designates where each sparse row begins in \texttt{Aval} and
\texttt{Aind}. An implementation of SpMV using CSR storage is
shown in Figure~\ref{fig:spmv}(b).

\begin{figure}[tb]
\centering
\vspace{-.2in}
\begin{tabular}{cc}
\begin{minipage}[b]{.5\textwidth}
\centering
\includegraphics[width=.8\textwidth]{figures/spmv.eps}  
\end{minipage}
&
\begin{minipage}[b]{.5\textwidth}
\scriptsize
\begin{verbatim}
for (i=0; i<num_rows; i++)
  for (j=Aptr[i]; j<Aptr[i+1]; j++)
    y[i] += Aval[j]*x[Aind[j]];

\end{verbatim}
\end{minipage}\\
(a) & (b)\\
\end{tabular}
\vspace{-.2in}
\caption{(a) Compressed sparse row (CSR) format (b) Basic implementation of CSR-based SpMV.}
\label{fig:spmv}
%\vspace{.2in}
\end{figure}

PETSc's implementation of SpMV used in this experiment exploits matrix
structure by using \textit{inodes}, which represent rows with identical
nonzero structure.  PETSc detects inodes automatically in arbitrary sparse
matrices. For example, the sparse matrix shown in Figure~\ref{fig:spmv}(a)
contains three inodes. The first inode is of size two because it holds two
adjacent rows (i.e., the first and the second rows) with the same nonzero
structure. The second inode contains three consecutive rows with identical
nonzero structure, while the last inode only has a single row. Knowing the
properties of each inode at runtime enables maximum reuse of vector $x$ since
multiple elements of $x$ can be loaded and used exactly once for all rows in
the same inode structure. To accomplish this, one must use a
register-blocking transformation. To optimize the inode-based SpMV routine,
we incorporated a new transformation module inside Orio that implements
various optimization strategies including register blocking, SIMDization,
memory alignment optimization, loop-control optimization, accumulator
expansion, and thread-level parallelization (with OpenMP). We then used Orio
to automatically select the best version of the optimized inode SpMV. Only
the outer loop which iterates over inodes was parallelized by using OpenMP.

To evaluate the performance of the tuned SpMV routine, we conducted the
experiment using a 2-D driven cavity flow simulation
application~\cite{coff:kell:keye} (SNES ex27 in PETSc), on both the Xeon and
Blue Gene/P. The ``Compiler-optimized'' label is used as a base case that
represents the performance of the naive implementation of SpMV (shown in
Figure~\ref{fig:spmv}(b)) optimized by the native compiler. We also tested
the performance of the hand-tuned (by PETSc developers) SpMV code, which is
included in PETSc releases.

Figure~\ref{fig:ex27-cookie-results} shows the performance results on the
Intel Xeon. The problem size is the number of grid points in the $x$ and $y$
directions, not the size of the sparse input matrix. The matrix dimension is
based the grid size and the number of field components computed for each grid
point, for example, an $8 \times 8$ problem involves sparse matrices of
dimension 900 with 17,040 nonzero entries. In this experiment, we ran in
three different multicore modes: SMP (one MPI process, eight
threads/process), dual (four MPI processes, two threads/process), and virtual
node or VN (eight MPI processes, one thread/process). The code tuned by Orio
consistently outperforms the alternatives. Furthermore, the performance of
the hand-tuned code is almost equivalent to that of the icc-optimized simple
code. Performance differences between the Orio version and the hand-tuned
version become more significant as more threads per process are employed to
facilitate OpenMP parallelization. The SMP and VN results show that
optimizations using loop-level parallelism (through OpenMP) achieve much
better performance than using MPI coarse-grained parallelism.
%Moreover, for the multithread cases, tuning using Orio determined that
%the performance benefit of SIMDization is more significant than that of
%thread-level parallelization only when the problem sizes are small. The
%reason for this is due to the overhead introduced by OpenMP from coordinating
%its threads across processors.
%So, each optimized
%code for the many-thread cases comprises two distinct tuned codes to handle
\begin{figure} [ht]
\begin{center} 
  \subfigure[SMP: $p=1$, $t/p=8$]{
  \includegraphics[width=.3\textwidth]{figures/ex27_cookie/smp.eps}
  \label{fig:ex27-cookie-smp} } \subfigure[Dual: $p=4$, $t/p=2$]{
  \includegraphics[width=.3\textwidth]{figures/ex27_cookie/dual.eps}
  \label{fig:ex27-cookie-dual} } \subfigure[VN: $p=8$, $t/p=1$]{
  \includegraphics[width=.3\textwidth]{figures/ex27_cookie/vn.eps}
  \label{fig:ex27-cookie-vn} }
\end{center}
\vspace{-.2in} 
\caption{Performance of inode SpMV on eight-core Intel machine; $p$ is the number of processes, and $t/p$ is the threads per process.} 
\label{fig:ex27-cookie-results} 
\end{figure}

The Blue Gene/P results in Figure~\ref{fig:ex27-bgp-results1} again show
%Blue Gene/P was done for a varied number of nodes (i.e., 1, 8, and 32)
%and the same three different parallel modes as before (i.e., SMP,
%dual, and VN). 
that empirical optimization using Orio produces the best performance for
all cases. 
%outperforming both the hand-tuned and the xlc-optimized codes. 
On this architecture, the performance gap between the xlc-optimized and the
hand-tuned codes is now larger than in the Intel experiment. Similarly, by
exploiting thread-level parallelism, the Orio-optimized code performs better
than the hand-tuned version when the number of threads per process increases
and the number of nodes decreases. For the single-thread cases (VN mode),
thread-level parallelism is not exploited; nevertheless, the Orio version
still performs better than the other two alternatives.
%This validates the
%effectiveness of Orio in empirically tuning sequential codes.

\begin{figure} [htb]
\begin{center} 
  \subfigure[1 node, SMP mode]{ 
  \includegraphics[width=.3\textwidth]{figures/ex27_bgp/n1_smp.eps}  
  \label{fig:ex27-bgp-smp-n1} 
  } 
  \subfigure[1 node, Dual mode]{ 
  \includegraphics[width=.3\textwidth]{figures/ex27_bgp/n1_dual.eps}  
  \label{fig:ex27-bgp-dual-n1} 
  } 
  \subfigure[1 node, VN mode]{ 
  \includegraphics[width=.3\textwidth]{figures/ex27_bgp/n1_vn.eps}  
  \label{fig:ex27-bgp-vn-n1} 
  } 
  \subfigure[8 nodes, SMP mode]{ 
  \includegraphics[width=.3\textwidth]{figures/ex27_bgp/n8_smp.eps}  
  \label{fig:ex27-bgp-smp-n8} 
  } 
  \subfigure[8 nodes, Dual mode]{ 
  \includegraphics[width=.3\textwidth]{figures/ex27_bgp/n8_dual.eps}  
  \label{fig:ex27-bgp-dual-n8} 
  } 
  \subfigure[8 nodes, VN mode]{ 
  \includegraphics[width=.3\textwidth]{figures/ex27_bgp/n8_vn.eps}  
  \label{fig:ex27-bgp-vn-n8} 
  } 
%  \subfigure[32 nodes, SMP mode]{ 
%  \includegraphics[width=.3\textwidth]{figures/ex27_bgp/n32_smp.eps}  
%  \label{fig:ex27-bgp-smp-n32} 
%  } 
%  \subfigure[32 nodes, Dual mode]{ 
%  \includegraphics[width=.3\textwidth]{figures/ex27_bgp/n32_dual.eps}  
%  \label{fig:ex27-bgp-dual-n32} 
%  } 
%  \subfigure[32 nodes, VN mode]{ 
%  \includegraphics[width=.3\textwidth]{figures/ex27_bgp/n32_vn.eps}  
%  \label{fig:ex27-bgp-vn-n32} 
%  } 
\end{center}
\vspace{-.2in} 
\caption{Performance of inode SpMV on Blue Gene/P for 1, 8, and 32 nodes.} 
\label{fig:ex27-bgp-results1} 
\end{figure}

\comment{
\begin{figure} 
\begin{center} 
  \subfigure[32x32, SMP mode]{ 
  \includegraphics[width=.3\textwidth]{figures/ex27_bgp/s32_smp.eps}  
  \label{fig:ex27-bgp-smp-32x32} 
  } 
  \subfigure[32x32, Dual mode]{ 
  \includegraphics[width=.3\textwidth]{figures/ex27_bgp/s32_dual.eps}  
  \label{fig:ex27-bgp-dual-32x32} 
  } 
  \subfigure[32x32, VN mode]{ 
  \includegraphics[width=.3\textwidth]{figures/ex27_bgp/s32_vn.eps}  
  \label{fig:ex27-bgp-vn-32x32} 
  } 
  \subfigure[64x64, SMP mode]{ 
  \includegraphics[width=.3\textwidth]{figures/ex27_bgp/s64_smp.eps}  
  \label{fig:ex27-bgp-smp-64x64} 
  } 
  \subfigure[64x64, Dual mode]{ 
  \includegraphics[width=.3\textwidth]{figures/ex27_bgp/s64_dual.eps}  
  \label{fig:ex27-bgp-dual-64x64} 
  } 
  \subfigure[64x64, VN mode]{ 
  \includegraphics[width=.3\textwidth]{figures/ex27_bgp/s64_vn.eps}  
  \label{fig:ex27-bgp-vn-64x64} 
  } 
  \subfigure[128x128, SMP mode]{ 
  \includegraphics[width=.3\textwidth]{figures/ex27_bgp/s128_smp.eps}  
  \label{fig:ex27-bgp-smp-128x128} 
  } 
  \subfigure[128x128, Dual mode]{ 
  \includegraphics[width=.3\textwidth]{figures/ex27_bgp/s128_dual.eps}  
  \label{fig:ex27-bgp-dual-128x128} 
  } 
  \subfigure[128x128, VN mode]{ 
  \includegraphics[width=.3\textwidth]{figures/ex27_bgp/s128_vn.eps}  
  \label{fig:ex27-bgp-vn-128x128} 
  } 
\end{center}
\vspace{-.2in}
\caption{Performance of inode SpMV on Blue Gene/P for problem sizes 32x32, 64x64, and 128x128.} 
\label{fig:ex27-bgp-results2} 
\end{figure} 
}

\subsection{Evaluation of the PLuTo-Orio Integrated System}

This section discusses the performance evaluation of the integrated
PLuTo-Orio system (discussed in Section~\ref{sec:integration}) on the
multicore Intel Xeon by using a number of application kernels that are
nontrivial to optimize and parallelize. We compare the performance of the
code tuned by Orio with the base code and the PLuTo-generated codes. The
PLuTo code was obtained by running the base code with
PLuTo-0.3.0~\cite{pluto030} using \texttt{--tile} to employ loop tiling for
L1 cache, \texttt{--l2tile} to employ loop tiling for L2 cache, and
\texttt{--unroll} to employ loop unrolling, and for parallel code generation,
an additional \texttt{--parallel} option. We used PLuTo's default tile
sizes (L1: 32x32 or 32x32x32; L2: 256x256 or 256x256x256) and default
unroll factors (64 for 1-D unrolling and 8x8 for 2-D
unroll/jamming). All codes were compiled with the Intel C Compiler
using the \texttt{-O3} optimization flag to enable auto-vectorization
and other advanced optimizations, and
\texttt{-parallel} (or \texttt{-openmp} in the case of manual OpenMP parallelization) 
to enable code parallelization.

In the following sections, we refer to the icc-optimized base code as
``ICC,'' the PLuTo-generated code with L1 tiling and
unroll/jamming as ``PLuTo (L1 tiling),'' and the PLuTo-generated code with L1
and L2 tilings and unroll/jamming as ``PLuTo (L1+L2 tiling).'' The code
tuned by Orio is referred to as ``PLuTo+Orio.''

\subsubsection{2-D Finite-Difference Time-Domain Method for Computational Electromagnetics}  
We consider the two-dimensional Finite Difference Time Domain (FDTD) algorithm, a
popular method for solving the time-dependent Maxwell's equations in the
context of computational 
%
\begin{wrapfigure}{r}{3in}
\vspace{-0.2in}
\begin{center}
\begin{minipage}{3in} 
\scriptsize
\begin{verbatim} 
for(t=0; t<tmax; t++) { 
  for (j=0; j<ny; j++) ey[0][j] = t; 
  for (i=1; i<nx; i++) 
    for (j=0; j<ny; j++) 
      ey[i][j] -= 0.5*(hz[i][j] - hz[i-1][j]); 
  for (i=0; i<nx; i++) 
    for (j=1; j<ny; j++) 
      ex[i][j] -= 0.5*(hz[i][j] - hz[i][j-1]); 
  for (i=0; i<nx; i++) 
    for (j=0; j<ny; j++) 
      hz[i][j] -= 0.7*(ex[i][j+1] - ex[i][j]
                  + ey[i+1][j] - ey[i][j]); 
} 
\end{verbatim} 
\end{minipage} 
\end{center}
\vspace{-0.2in}
\caption{2-D FDTD code.} 
\label{fig:fdtd-2d-code} 
\vspace{-.1in}
\end{wrapfigure}
%
electrodynamic problems. As shown in Figure~\ref{fig:fdtd-2d-code}, the 2-D
FDTD method is implemented as an outer iteration over time containing four imperfectly
nested loops.  The arrays $ex$ and $ey$ denote the electric field components,
and the array $hz$ denotes the magnetic field.

The performance of the sequential 2-D FDTD code for $tmax=500$ and $nx=ny$ is
shown in Figure~\ref{fig:fdtd-2d-cookie-seq}. The base code optimized by icc
alone performs better than PLuTo for small problem sizes since all input
arrays fit in the L2 cache (insufficient computation to offset PLuTo's tiling
overhead). As the array sizes increase, the lack of data reuse impairs the
base code's performance, whereas the PLuTo performance remains about the
same.
%We used Orio to generate two code variants: one for small problem sizes
%and another for large problem sizes. 
When the input arrays are small, Orio discovers that applying
PLuTo's polyhedral transformations is not beneficial, and therefore,
it employs only its syntactic transformations on the original FDTD
code. For large problem sizes, Orio exploits some of the PLuTo's code
transformations and enhances it further with its syntactic
optimizations, resulting in performance consistently and significantly
higher than both the base and PLuTo codes (up to 86\% over PLuTo).

%By tuning the sequential code with Orio,
%the attainable performance enhancement over PLuTo is up to 86\%.

Figure~\ref{fig:fdtd-2d-cookie-par} shows the multicore performance obtained
for $tmax=500$ and $nx=ny=2000$. The results indicate that whereas icc is
unable to auto-parallelize the code, PLuTo detects the existence of pipelined
parallelism and then successfully parallelizes the code. Optimizing and
tuning by Orio additionally improves the PLuTo performance by up to
78\%. Moreover, we observe that because of memory contention between the two
quad-core Intel processors, the speedup of both the PLuTo and Orio codes
slightly decreases when the number of cores used is greater than four.

\begin{figure}[htb] 
\begin{center} 
  \subfigure[Sequential (T=500)]{ 
  \includegraphics[width=.4\textwidth]{figures/fdtd-2d-cookie/seq.eps} 
  \label{fig:fdtd-2d-cookie-seq} } \subfigure[Parallel (T=500, 
  N=2000)]{ 
  \includegraphics[width=.4\textwidth]{figures/fdtd-2d-cookie/par.eps} 
  \label{fig:fdtd-2d-cookie-par} } 
\end{center}
\vspace{-.2in} 
\caption{2-D FDTD performance on an eight-core Intel Xeon.} 
\label{fig:fdtd-2d-cookie-results} 
%\vspace{-.2in}
\end{figure}

%%BN: commented out since this needs more explanation (no room in this version)
%It is to be noted that collecting performance results of the PLuTo 
%code with L1 and L2 tilings was not possible because of the large code 
%size of the generated loop nest, which led to an impractically huge 
%compile time.

%% BN: comment for space
\comment{
\subsubsection{3-D Gauss-Seidel Successive Over-Relaxation Method}  
Figure~\ref{fig:seidel-code} shows the 3-D Gauss-Seidel computation, 
which is sometimes is referred to as
\emph{successive displacement method}, indicating the dependence of the 
iterations on the ordering. If the ordering is
%
%\begin{figure}[ht]
\begin{wrapfigure}{r}{3in}
\vspace{-0.3in}
\begin{center}
\begin{minipage}{3in} 
\scriptsize
\begin{verbatim} 
for (t=0; t<=T-1; t++) 
  for (i=1; i<=N-2; i++) 
    for (j=1; j<=N-2; j++) 
      A[i][j] = (A[i-1][j-1] + A[i-1][j] 
          + A[i-1][j+1] + A[i][j-1] + A[i][j] 
          + A[i][j+1] + A[i+1][j-1] + A[i+1][j] 
          + A[i+1][j+1]) / 9.0; 
\end{verbatim} 
\end{minipage} 
\end{center}
\vspace{-0.2in}
\caption{3-D Gauss-Seidel code} 
\label{fig:seidel-code} 
\vspace{-.1in}
\end{wrapfigure}
%
changed, the components of the new iterations will change as well.

Figure~\ref{fig:seidel-cookie-seq} contains the sequential performance
results for $T=500$, which show that applying PLuTo's polyhedral tiling on
the original code always delivers performance boosts, which range from 126\%
to 142\%. When using PLuTo, performing two-level tiling (for both L1 and L2
caches) yields 13\% lower performance than performing only L1 tiling. 
This is also reflected in the best sequential code found by Orio, where
one-level tiling (for L1 cache only) is always performed. The sequential
speedup obtained from tuning the PLuTo code with Orio is significant, ranging
from 51\% to 133\%.

\begin{figure} [htb]
\begin{center} 
  \subfigure[Sequential (T=500)]{ 
  \includegraphics[width=.4\textwidth]{figures/seidel-cookie/seq.eps} 
  \label{fig:seidel-cookie-seq} } \subfigure[Parallel (T=500, 
  N=4000)]{ 
  \includegraphics[width=.4\textwidth]{figures/seidel-cookie/par.eps} 
  \label{fig:seidel-cookie-par} } 
\end{center}
\vspace{-.2in} 
\caption{3-D Gauss-Seidel performance on eight-core Intel machine.} 
\label{fig:seidel-cookie-results} 
\vspace{-.1in}
\end{figure}

The parallel performance results for $T=500$ and $N=4000$ are shown in
Figure~\ref{fig:seidel-cookie-par}. Again we observe that icc failed to
parallelize the code, whereas PLuTo was able to parallelize it and Orio
improved the performance of the one-level tiled PLuTo code further by up to
128\% (and up to 548\% for the two-level tiled PLuTo code). We observed that
tiling over both L1 and L2 can result in worse performance than L1-only
tiling because the number of tiles can be smaller than the number of cores.
%Similar earlier trend also exists
%in these performance results: PLuTo parallelizes the code through its
%polyhedral transformations (i.e., skewing and tiling), while the icc
%compiler alone does not parallelize the loop. Empirical tuning via
%Orio creates up to 122\% speedups over the one-level tiled PLuTo code,
%and up to 548\% speedups over the two-level tiled PLuTo code. Another
%finding is that tiling for both L1 and L2 caches contributes to a poor
%scalability because the number of tiles can be less than the number of
%processors, resulting in underutilization of the available resources.
} % end comment

\subsubsection{LU Factorization}

LU factorization or decomposition is a numerical method for the solution of linear
systems 
%
\begin{wrapfigure}{r}{3in}
\begin{center}
\vspace{-.2in}
\begin{minipage}{2.8in} 
\scriptsize
\begin{verbatim} 
for (k=0; k<=N-1; k++) { 
  for (j=k+1; j<=N-1; j++) 
    A[k][j] /= A[k][k]; 
  for(i=k+1; i<=N-1; i++) 
    for (j=k+1; j<=N-1; j++) 
      A[i][j] -= A[i][k]*A[k][j]; 
} 
\end{verbatim} 
\end{minipage} 
\end{center}
\vspace{-0.2in}
\caption{LU Decomposition code.} 
\label{fig:lu-code} 
\end{wrapfigure}
%
of equations; a simple implementation is shown in Figure~\ref{fig:lu-code}.

The sequential performance results are shown in
Figure~\ref{fig:lu-cookie-seq}. Similarly to the 2-D FDTD results, the icc-optimized code is more
efficient than the PLuTo-generated codes when the input arrays are small and
fit in the L2 cache. Because of better data locality, however, the
performance of the PLuTo-tiled codes is better than that of the base code as
the input arrays get larger. 
%Again, there are two different codes generated
%by Orio: one for small problem sizes and one for large problem sizes. And
Consequently, Orio employs PLuTo's polyhedral transformations only for large
input sizes prior to applying its own syntactic transformations. 
%Unlike 3-D
%Gauss-Seidel, this time Orio chooses to apply both L1 and L2 tilings when
%generating the best tuned code. 
Compared to the two PLuTo codes, the
Orio-tuned code yields performance improvements ranging between 26\% and 277\%.

The parallel performance results shown in Figure~\ref{fig:lu-cookie-par} were
obtained for $N=4000$. We observe that icc is not able to parallelize the
code, whereas PLuTo achieves higher performance by exploiting multicore
parallelism. Furthermore, both PLuTo-generated codes have comparable
performance, with slightly better scalability exhibited by the single-level tiled
code. Orio further improved the performance of the PLuTo-generated versions
by a factor of 51\% to 120\%.

\begin{figure}[htb]
\begin{center} 
  \subfigure[Sequential]{ 
  \includegraphics[width=.4\textwidth]{figures/lu-cookie/seq.eps} 
  \label{fig:lu-cookie-seq} } \subfigure[Parallel (N=4000)]{ 
  \includegraphics[width=.4\textwidth]{figures/lu-cookie/par.eps} 
  \label{fig:lu-cookie-par} } 
\end{center} 
\vspace{-.2in}
\caption{LU Decomposition performance on eight-core Intel machine.} 
\label{fig:lu-cookie-results} 
\end{figure}

### 2008-10-13 doc/papers/ipdps09/implementation.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@363 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

\section{Orio Design and Implementation}
\label{sec:implm}

Orio~\cite{OrioURL} is an empirical performance tuning system that takes
annotated C source code as input, generates many optimized code variants of
the annotated code, and empirically evaluates the performance of the
generated codes, ultimately selecting the best performing version to use for
production runs.

The Orio annotation approach differs from existing compiler- and
annotation-based systems in the following significant ways.
%
First, by not committing to a single general-purpose language, we can define 
annotation grammars that \emph{restrict} the original syntax, enabling more
effective performance transformations (e.g., disallowing pointer arithmetic
in a C-like annotation language); furthermore, it enables the definition of
new high-level languages which retain domain-specific
information that is normally lost in low-level C or Fortran
implementations. This in turn expands the range of possible performance-improving 
transformations.
%
Second, Orio was conceived and designed with the following requirements in mind: 
portability (which precludes extensive dependencies on external packages),
extensibility (new functionality must require little or no change to the
existing Orio implementation and interfaces that enable integration with
other source transformation tools must be defined), and automation
(ultimately Orio should provide tools that manage
\emph{all} the steps of the performance tuning process, automating each step 
as much as possible).
%
Finally, Orio is usable in real scientific applications without
requiring reimplementation. This ensures that the significant investments in
the development of complex scientific codes is leveraged to the greatest
extent possible.

%\subsection{Overall Approach} 
%\label{sec:overall}
\begin{figure}[tb]
\vspace{-.2in}
\begin{center}  
\includegraphics[width=0.6\textwidth]{figures/orio.eps}   
\end{center}  
\vspace{-.2in}
\caption{Overview of Orio's code generation and empirical tuning process.}  
\label{fig:orio}  
\end{figure}

Figure~\ref{fig:orio} contains a high-level graphical depiction of the code
generation and tuning process implemented in Orio.  Orio can be used to
improve performance by source-to-source transformations such as loop
unrolling, loop tiling, and loop permutation. The input to Orio is C code
containing structured comments that include a simplified expression of the
computation, as well as various performance-tuning directives. Orio scans the
marked-up input code and extracts all annotation regions. Each annotation
region is then processed by transformation modules. The code generator then
produces the final C code with various optimizations that correspond to the
specified annotations. Unlike compiler approaches, we do not implement a
full-blown C compiler; rather, we use a precompiler that parses only the
language-independent annotations.

Orio can also be used as an automatic performance tuning tool.  The code
transformation modules and code generator produce an optimized code version
for each distinct combination of performance parameter values. Then each
optimized code version is executed and its performance evaluated.  After
iteratively evaluating all code variants, the best-performing code is picked
as the final output of Orio. Because the search space of all possible
optimized code versions can be huge, a brute force search strategy is not
always feasible. Hence, Orio provides various search heuristics for reducing
the size of the search space and thus the empirical testing time.

%% BN: The info in the paragraph below is repeated later in the discussion.
%The tuning specifications written by users in the form of annotations
%are parsed and used by Orio to guide the transformations and the
%empirical tuning process. These specifications include essential
%information such as the used base compilers, the search strategy, the
%program transformation parameters, the input data sizes, etc.

\subsection{Annotation Language Syntax} 
\label{sec:ann-lang}

Orio annotations are embedded into the source code as structured C comments
that always start with \texttt{/*@} and end with \texttt{@*/}. For example,
the annotation \texttt{/*@ end @*/} is used to indicate the end of an
annotated code region. A simple grammar illustrating the basic syntax of Orio
annotations is depicted in Figure~\ref{fig:ann-lang}. An annotation region
consists of three main parts: leader annotation, annotation body, and trailer
annotation. The annotation body can be either empty or contain C code that
possibly includes other nested annotation regions. A leader annotation
contains the name of the code transformation module used to transform and
generate the annotated application code. A high-level description of the
computation and several performance hints are coded in the module body inside
the leader annotation. A trailer annotation, which has a fixed form (i.e.,
\texttt{/*@ end @*/}), closes an annotation region.

\begin{figure}[thb]
\begin{center} 
\begin{tabular}{rrl} 
$<$annotation-region$>$ & ::= & $<$leader-annotation$>$ $<$annotation-body$>$ $<$trailer-annotation$>$\\ 
$<$leader-annotation$>$ & ::= & \texttt{/*@ begin} $<$module-name$>$ \texttt{(} $<$module-body$>$ \texttt{) @*/} \\
$<$trailer-annotation$>$ & ::= & \texttt{/*@ end @*/} \\
\end{tabular} 
\end{center}  
\vspace{-.1in}
\caption{Annotation language grammar excerpt.}  
\label{fig:ann-lang}
\end{figure}

\subsection{Orio Input Example}
\label{sec:example}

Figure~\ref{fig:orio-example} shows a concrete annotation example that
empirically optimizes a C function for the Blue Gene/P
architecture. This is an instance of an AXPY operation, i.e., one that computes
$y=y+a_1 x_1+\cdots+a_n x_n$, where $a_1,\ldots,a_n$ are scalars and
$y,x_1,\ldots,x_n$ are one-dimensional arrays. The specific AXPY operation
considered in this example corresponds to $n=4$. The first annotation contains the
\texttt{BGP\_Align}\footnote{Architecture-specific annotations are simply ignored when the
code is being tuned on an architecture that doesn't support them.} directive,
which instructs Orio to dynamically load its memory-alignment optimization
module and then generate preprocessor directives, such as pragmas and calls
to memory alignment intrinsics, including a check for data alignment. The
main purpose of these alignment optimizations is to enable the use of the
dual floating-point unit (Double Hummer) of the Blue Gene/P, which requires
16-byte alignment. As discussed later in Section~\ref{sec:axpy4-results},
even these simple alignment optimizations can lead to potentially significant
performance improvements. This example also shows the use of Orio's loop
transformation module (named \texttt{Loop}) to optimize the AXPY-4 loop by
unrolling and generating OpenMP parallelization directives for exploiting
multicore parallelism. In addition to the simple source transformations in
this example, Orio also supports other optimizations, such as register
tiling and scalar replacement.

Whereas the \texttt{BGP\_Align} and \texttt{Loop} annotations in this example
guide the source-to-source transformations, the purpose of the
\texttt{PerfTuning} annotation is primarily to control the empirical
performance tuning process. Details of the tuning specifications for
optimizing the AXPY-4 code on Blue Gene/P are shown in the right-hand side of
Figure~\ref{fig:orio-example}.  The tuning specification contains data
required for building, initializing, and running experiments, including input
variable information, the search algorithm, performance counting technique,
performance parameters values, and execution environment details.  The tuning
specifications can be either integrated in the source code or defined in a
separate file, as in this example.

\begin{figure}[thb] 
\centering 
\begin{tabular}{cc} 
\begin{minipage}{.5\textwidth}  
\scriptsize
\begin{verbatim}  
void axpy4(int N, double *y, 
 double a1, double *x1, double a2, double *x2, 
 double a3, double *x3, double a4, double *x4) {

 /*@ begin PerfTuning (
   import spec axpy4_tune_spec;
 ) @*/

 register int i;

 /*@ begin BGP_Align (x1[],x2[],x3[],x4[],y[]) @*/ 
 /*@ begin Loop ( 
  transform Unroll (ufactor=UF, parallelize=PAR)
  for (i=0; i<=N-1; i++) 
    y[i] += a1*x1[i]+a2*x2[i]+a3*x3[i]+a4*x4[i]; 
 ) @*/

 for (i=0; i<=N-1; i++) 
   y[i] += a1*x1[i]+a2*x2[i]+a3*x3[i]+a4*x4[i];

 /*@ end @*/ 
 /*@ end @*/
 /*@ end @*/
}

\end{verbatim}  
\end{minipage}
&
\begin{minipage}{.5\textwidth}  
\scriptsize
\begin{verbatim}  
spec axpy4_tune_spec {
 def build { 
  arg build_command = 
   'mpixlc_r -O3 -qstrict -qhot -qsmp=omp:noauto'; 
  arg batch_command = 
   'qsub -n 128 -t 20 --env "OMP_NUM_THREADS=4"'; 
  arg status_command = 'qstat';
  arg num_procs = 128; 
 } 
 def performance_counter { 
  arg method = 'basic timer';
  arg repetitions = 10000;
 } 
 def performance_params { 
  param UF[] = range(1,33);
  param PAR[] = [True,False]; 
 }
 def input_params { 
  param N[] = [10,100,1000,10**4,10**5,
               10**6,10**7]; 
 }
 def input_vars { 
  decl dynamic double y[N] = 0; 
  decl dynamic double x1[N] = random; 
  decl double a1 = random; 
  decl double a2 = random; 
  # ... ommitted ...
 } 
 def search { 
  arg algorithm = 'Exhaustive'; 
  arg time_limit = 20;
 }
}
\end{verbatim}  
\end{minipage}
\\
%(a) Annotated AXPY-4 source code
%& 
%(b) Tuning specification for the Blue Gene/P\\
\end{tabular}
\caption{Orio input example; annotated AXPY-4 source code (left) and tuning specification for the Blue Gene/P (right).}
\label{fig:orio-example}  
%\vspace{-.2in}
\end{figure}

The annotated AXPY-4 code (left side of Figure~\ref{fig:orio-example}) uses two
performance parameters whose values are defined in the tuning specification:
the unroll factor (\texttt{UF})
% which signifies how many times the loop body will be
%replicated in the generated code.
and the boolean variable \texttt{PAR}, which is used to activate or
deactivate OpenMP parallelization. 
%Such boolean parameter is needed by Orio
%due to the following basis: when both the number of loop iterations and the
%total amount of computation in the loop body are small, parallelizing a loop
%using OpenMP can give poor performance because of the dominating overhead
%incurred from thread coordination and task scheduling. Therefore, based on a
%given vector size, Orio can determine at runtime whether it is beneficial to
%parallelize the loop.
These parameters are used by Orio to determine at runtime whether it is
beneficial to parallelize the loop, i.e., whether there is enough work per
thread to offset the OpenMP overhead or not.

Achieving the best performance for different input problem sizes may require
different tuning approaches; thus, the entire tuning process is
repeated for each specified problem size. In the AXPY-4 example, the search
space includes seven different input problem sizes (variable~\texttt{N}). 
%Orio repeats the tuning process for each problem size, resulting
%in the generation of seven final optimized codes.

\subsection{Annotation Parsing and Code Generation}

The Orio system consists of several optimization modules, each implemented as
a Python module. As mentioned earlier in Section~\ref{sec:example}, given the
module name in the leader annotation, Orio dynamically loads the
corresponding code transformation module and uses it for both annotation
parsing and code generation. If the pertinent module cannot be found in the
transformation modules directory, an error message is emitted and the tuning
process is terminated. This name-based dynamic loading provides flexibility
and easy extensibility without requiring detailed knowledge or modification
of the existing Orio software. Therefore, varied approaches to code
transformations ranging from low-level loop optimizations for cache
performance to composed linear algebra operations and new specialized
algorithms can easily be integrated into Orio.

%Taking as input the information supplied in both the annotation body
%and the module body, 
After parsing the annotation, each module performs a distinct optimization
transformation prior to generating the optimized code.  The transformation
module can either reuse an existing annotation parser or define new language
syntax and a corresponding parser component.

%%BN: the details below are probably not essential for this paper
%In some cases, the annotation is seemingly redundant, containing a version of
%the computation that is very similar to the original code. 
%We took this
%approach for two reasons.  First, basing the transformation modules on the
%actual application code demands a full-fledged C compiler
%infrastructure. Requiring this complex compiler infrastructure to avoid a
%relatively small manual effort in creating annotations runs counter to our
%goal of creating the annotation system portable and easy to install and
%use. Second, requiring in effect a rewrite of the code to be optimized
%encourages simplification and enables the code optimization effort to start
%with a cleaner, rather than an already hand-tuned, version of the
%code. Moreover, we are investigating higher-level, domain-specific syntax for
%annotations, thus capturing semantics without imposing tuning constraints
%resulting from the use of the general-purpose C language. 
In some cases, the annotation is seemingly redundant, containing a version of
the computation that is very similar to the original code.  As mentioned
earlier, we took this approach so that the annotation language can be defined
in a way that enables more effective transformations (through restrictions or
high-level domain information).

Current optimizations supported by Orio span different types of code
transformations that are not provided by production compilers in some
cases. Available optimizations include simple loop unrolling, memory
alignment optimization, loop unroll/jamming, loop tiling, loop permutation,
scalar replacement, register tiling, loop-bound replacement, array copy
optimization, multicore parallelization (using OpenMP), and other
architecture-specific optimizations (e.g., generating calls to SIMD
intrinsics on Intel and Blue Gene/P architectures).

\subsection{Search Space Exploration and Evaluation}
\label{sec:search-space}

As briefly discussed earlier, our empirical tuning approach systematically
measures the performance costs of automatically-generated code variants in
order to find the most optimal available version. In the context of empirical
optimization, code variants are alternative, semantically equivalent,
implementations of the same computation. Each implementation variant is
associated with a collection of different optimization parameters that
correspond to source-to-source code transformations such as, for instance,
unroll factors, tile sizes, and loop permutation order. So, each coordinate
in the search space of empirical tuning problem represents a distinct
combination of performance parameter values. Both the dimension and the size
of the search space depend on the total number and the value ranges of used
performance parameters, respectively.

The conceptually straightforward approach to exploring the space of the
parameter values is to use an exhaustive search procedure that is guaranteed
to determine the optimal code version. Normally the size of the search space
is too large, making full coverage impractical.  Thus, in addition to
supporting exhaustive search, we have implemented several search heuristics.
The simplest search heuristic is a random search, which picks a random
coordinate in the search space at each step and then measures its
performance; random search is not guaranteed to return close-to-optimal results.
We have also developed two other search heuristics to effectively
narrow the search space for close-to-optimal performance, including a
heuristic based on the \textit{Nelder-Mead simplex}
method~\cite{Lagarias98simplex,Lewis00directsearch}, 
a popular non-derivative direct search method for optimization, and \textit{simulated
annealing}\cite{Kirkpatrick83optimizationby}.
%with the additional inclusion of
%a gradually-reduced ``temperature'' used to control the randomness of
%selecting the next move in the search space. 
%By default Orio uses exhaustive
%search, but users can set the search parameters in the tuning specification
%to use and configure one of the available heuristics.  
Similarly to the implementation of Orio's optimization modules, each search
technique is implemented as an independent Python module in Orio and is dynamically
loaded given only the algorithm's name as one of the fields in the tuning
specification.

%%BN: commenting for space
%The key idea of a basic simplex method
%is to construct a nondegenerate simplex (i.e., a convex hull of $n+1$ points,
%where $n$ is the dimension of the search space), and then obtain its
%associated function values to steer the search to the next move by replacing
%the worst vertex with a point reflected through the centroid of the remaining
%$n$ points. The simplex search is improved in the Nelder-Mead method by
%adding more moves, making the search more robust and faster to converge. \

%The exhaustive approach is chosen as the default space exploration
%method of Orio; however, users can switch to the other algorithms by
%manually indicating his preferred search strategy in the tuning
%specifications. Users can also specify terminating criteria of the
%chosen search strategy by providing numerical values to the search
%time limit and the maximum total number of search trials. Each time
%the search exceeds the specified time limit or the predetermined
%number of full search runs, we suspend the search and return the best
%implementation variant up to that point of time. Also, users can
%optionally adjust the search by altering the values of some
%algorithm-specific parameters in the tuning specifications. As an
%example, users can change four types of coefficients that the
%Nelder-Mead simplex method has: reflection, expansion, contraction,
%and shrinkage, in order to control the search convergence speed and
%the quality of the search outcome.

To improve the quality of the search result further, each search heuristic is
enhanced by applying a local search after the global search completes. The
local search compares the best performance with neighboring coordinates. If a
better coordinate is discovered, the local search continues recursively until
no further performance improvement is possible or a user-specified
termination criterion is met.

Further pruning of the search space is enabled through user-specified
constraints in the tuning specifications. We use loop the
unroll/jamming transformation to make the discussion more concrete. Loop
unroll/jamming~\cite{sarkar-unrolljam} (coupled with scalar replacement) is
mainly intended to increase data reuse at the register level. So, when
unroll/jamming is applied to multiple loops, the values of unroll factors
must be such that register locality is maximized while satisfying the
register capacity constraints to avoid unnecessary register
spills. Figure~\ref{fig:par-constraint} shows an example of analytically
enforcing register capacity constraints on matrix-matrix multiplication code
that is optimized with loop unroll/jamming, assuming 32
registers. \texttt{Ui}, \texttt{Uj}, \texttt{Uk} are unroll factors for loops
\texttt{i}, \texttt{j}, \texttt{k}, respectively. With the specified
parameter constraints, the search space of performance parameter values is
drastically trimmed down from 32,768 points (i.e., $32*32*32$) to only 65
points. This example also demonstrates that a set of constraints can also be
imposed on input parameters to decide what input problem sizes to consider.

\begin{figure}[htp] 
\centering 
\begin{tabular}{cc} 
\begin{minipage}{.4\textwidth}  
\scriptsize
\begin{verbatim}  
/*@ begin Loop (
transform UnrollJam(ufactor=Ui)
for (i=0; i<=M-1; i++)
  transform UnrollJam(ufactor=Uj)
  for (j=0; j<=N-1; j++)
    transform UnrollJam(ufactor=Uk)   
    for (k=0; k<=O-1; k++)
      A[i][j] += B[i][k]*C[k][j];
) @*/
for (i=0; i<=M-1; i++)
 for (j=0; j<=N-1; j++)
  for (k=0; k<=O-1; k++)
    A[i][j] += B[i][k]*C[k][j];
/*@ end @*/
\end{verbatim}  
\end{minipage}
&
\begin{minipage}{.45\textwidth}  
\scriptsize
\begin{verbatim}  
def performance_params {
  param Ui[] = range(1,33);
  param Uj[] = range(1,33);
  param Uk[] = range(1,33);
  constraint reg_capacity = Ui*Uj+Ui*Uk+Uk*Uj<=32;
}

def input_params {
  param M[] = [10,50,100,500,1000];
  param N[] = [10,50,100,500,1000];
  param O[] = [10,50,100,500,1000];
  constraint square_matrices = (M==N) and (N==O);
}

\end{verbatim}  
\end{minipage}
\\
%(a) Annotated code of matrix-matrix multiplication
%& 
%(b) Orio's parameter specifications\\
\end{tabular}
\caption{Example of specifying parameter constraints in Orio; annotated code for matrix-matrix multiplication (left) and constraint specification (right).}
\label{fig:par-constraint}  
\end{figure}

Orio also supports parallel search when parallel resources are available,
e.g., when tuning on the Blue Gene/P. The parallel Orio driver concurrently
executes multiple independent code variants in the same parallel job. After
Orio submits a parallel job, each node in the target machine executes a
distinct generated code variant to collect the code performance. The
resulting performance results are collected and stored in a temporary file,
which is later processed by Orio to determine the best performing variant.
Figure~\ref{fig:orio-example} has an example of using parallel search by
specifying the number of nodes to use (per job) with the \texttt{num\_procs}
variable. The remaining fields used by the parallel search are the
\texttt{batch\_command} and the \texttt{status\_command} fields, which specify how Orio
should submit a parallel job and query its status, respectively.

### 2008-10-13 doc/papers/ipdps09/motivation.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@363 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

\section{Motivation}
\label{sec:motivation}

%The use of empirical performance tuning and annotations approaches for
%the following purposes:
%- improve productivity of scientific computation applications
%- improve high performance by (semi-)automating the tuning process
%- improve code readability
%- alleviate performance portability issues

The size and complexity of scientific computations are increasing at least as
fast as the improvements in processor technology.  Programming such
scientific applications is hard, and optimizing them for high performance is
even harder.  This results in a potentially large gap between the achieved
performance of applications and the peak available performance, with
many applications achieving 10\% or less of the peak (e.g.,~
%\cite{Oliker,
\cite{conf/vecpar/CarterOS06}).  
A greater concern is the inability of existing languages, compilers, and
systems to deliver the available performance for the application through
fully automated code optimizations.

Delivering performance without degrading productivity is crucial for the
success of scientific computing. Scientific code developers generally attempt
to improve performance by applying one or more of the following three
approaches: manually optimizing code fragments, using tuned libraries for key
numerical algorithms, and, less frequently, using compiler-based source
transformation tools for loop-level optimizations. Manual tuning is
time-consuming and impedes readability and performance portability. Tuned
libraries often deliver great performance without requiring significant
programming effort, but can only provide limited
functionality. General-purpose source transformation tools for performance
optimizations are few and have not yet gained popularity among computational
scientists, at least in part because of poor portability and steep learning
curves.

### 2008-10-13 doc/papers/ipdps09/related.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@363 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

\section{Related Work}
\label{sec:related}

%Optimization abstraction:
%- POET (Qing Yi)
%- Apan Qasem's thesis work (Ken Kennedy's student)
%- ...
%
%Automated performance tuning:
%- ATLAS
%- SPIRAL
%- FFTW
%-...

Ideally, a developer should only have to specify a few simple command-line
options and then rely on the compiler to optimize the performance of an
application on any architecture. Compilers alone, however, cannot fully
satisfy the performance needs of scientific applications.  First,
compilers must operate in a black-box fashion and at a very low level,
limiting both the type and number of optimizations that can be done.  Second,
static analysis of general-purpose languages, such as C, C++, and Fortran, is
necessarily conservative, thereby precluding many possible optimizations.
%Third, in the process of transforming a mathematical model into a computer
%program, much potentially useful (for optimization purposes) information is
%lost since it cannot be represented by the programming language.  
Finally, extensive manual tuning of a code may prevent certain compiler optimizations
and result in worse performance on new architectures, resulting in 
loss of performance portability.

%As briefly discussed in Section~\ref{sec:motivation}, performance tuning is
%generally approached in three ways: by performing manual optimizations of key
%portions of the code; by using compiler-based source transformation tools for
%loop optimizations; and by using tuned libraries for key numerical
%algorithms.
%% Libraries
An alternative to manual or automated tuning of application codes is the use
of tuned libraries. The two basic approaches to supplying high-performance
libraries include providing a library of hand-coded options (e.g.,
\cite{BLAS,ESSL,Goto:2006fk}) and generating optimized code automatically
for the given problem and machine parameters. 
ATLAS~\cite{atlas_sc98}
%,WN147} 
for BLAS and LAPACK
%~\cite{BLAS} and LAPACK~\cite{laug}
routines, OSKI~\cite{OSKI} for sparse linear algebra,
PHiPAC~\cite{bilmes97optimizing} for matrix-matrix products, and
domain-specific libraries such as FFTW~\cite{frigo98} and
SPIRAL~\cite{SPIRAL} are all examples of the latter approach. Most 
automatic tuning approaches perform empirical
parameter searches on the target platform.  
%FFTW uses a combination of static
%models and empirical techniques to optimize FFTs. SPIRAL generates optimized
%digital signal processing libraries by an extensive empirical search over
%implementation variants.  GotoBLAS~\cite{Goto:2006fk,Goto:fk}, on the other
%hand, achieves great performance results on several architectures by using
%hand-tuned data structures and kernel operations.  
These auto- or hand-tuned
approaches can deliver performance that can be five times as fast as that
produced by many optimizing compilers \cite{WN147}.  The library approach,
however, is limited by the fact that optimizations are highly problem- and
machine-dependent. Furthermore, at this time, the functionality of the
currently available automated tuning systems is quite limited.

%% Other annotation-based source transformation approaches
General-purpose tools for optimizing loop performance are also available.
LoopTool~\cite{LoopTool} supports annotation-based loop fusion,
unroll/jamming, skewing and tiling.  The Matrix Template Library
\cite{Siek:1998ys} uses template metaprograms to tile at both the register
and cache levels.  A new tool, POET~\cite{POET} also supports a number of
loop transformations. 
%POET offers a complex template-based syntax for
%defining transformations in a language-independent manner. 
Other research efforts whose goal, at least in part, is to enable
optimizations of source code to be augmented with performance-related
information include the X language~\cite{XLanguage} (a macro C-like language
for annotating C code), the Broadway~\cite{broadway} compiler, and
telescoping languages~\cite{telescopingurl}.
%,teleoverview,Ken99}
%, and various
%meta-programming techniques~\cite{veldhuizen95,weise93,kiczales91,chiba95}.

\comment{
Emerging annotation-based tools are normally designed by compiler researchers
and thus the interfaces are not necessarily based on concepts accessible to
computational scientists. The complexity of existing annotation languages and lack
of common syntaxes for transformations (e.g., loop unrolling) result
in steep learning curves and the inability to take advantage of more than one
approach at a time. Furthermore, at present, there is no good way for
\emph{users} to learn about the tools available and compare their
capabilities and performance.
}

### 2008-10-13 doc/papers/ipdps09/results-long.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@363 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

\section{Experimental Results} 
\label{sec:results}

In this section, we discuss the performance of several codes tuned with
Orio. We ran experiments on a multicore Intel Xeon machine and a Blue Gene/P
(both at Argonne National Laboratory). The Intel machine has dual quad-core
E5462 Xeon processors (eight cores in total) clocked at 2.8 GHz (1600 MHz
FSB) with 32 KB L1 cache, 12 MB of L2 cache (6 MB shared per core pair), and
2 GB of DDR2 FBDIMM RAM, running Linux kernel version 2.6.25 (x86-64). Each
compute node of the Blue Gene/P is equipped with four cores, each one a
850MHz IBM PowerPC 450 processors with a dual floating-point unit. While L1
cache (32 KB) and L2 cache (4 MB) are private for each processor, L3 cache (8
MB) is shared on each Blue Gene/P node. The total memory per node is 2GB. The
operating system on Blue Gene/P is based on Linux 2.6.16 (ppc64). ICC 10.1
and IBM XL C/C++ V9.0 were the primary compilers used on the Intel machine
and the Blue Gene/P, respectively.

\subsection{Sequence of Linear Algebra Operations} 
\label{sec:axpy4-results}

In this experiment, we tuned the performance of the AXPY-4 operation
(seen previously in Figure~\ref{fig:orio-example}) on a single node of
the Blue Gene/P machine using the IBM XLC compiler.  The performance
measurements were done for two cases: using a single core per node and using
all four cores, where the results can be found in
Figure~\ref{fig:axpy4-bgp-seq} and Figure~\ref{fig:axpy4-bgp-par},
respectively. The single-core scenario used the following compiler
options: \texttt{-O3 -qstrict -qarch=450d -qtune=450 -qhot
-qsmp=noauto}, whereas the multicore scenario differs in the use of
\texttt{-qsmp=auto} for the non-Orio versions. The parallel Orio
version contains OpenMP parallelization directives in generated code,
thus demanding the use of compiler option
\texttt{-qsmp=omp:noauto}. Included are performance numbers for four
code variants: a simple loop implementation with no any library calls
(labeled ``Compiler-optimized''), two BLAS-based implementations that
utilize the Goto BLAS~\cite{Goto:2006fk,Goto:fk} and the
ESSL~\cite{ESSL}, and the Orio-tuned version.

\begin{figure}[thb] 
\begin{center} 
  \subfigure[Sequential (single core)]{ 
  \includegraphics[width=.4\textwidth]{figures/axpy4_bgp/seq.eps}  
  \label{fig:axpy4-bgp-seq} 
  } 
  \subfigure[Parallel (four cores)]{ 
  \includegraphics[width=.4\textwidth]{figures/axpy4_bgp/par.eps}  
  \label{fig:axpy4-bgp-par} 
  } 
\end{center}
\vspace{-.2in} 
\caption{Performance of AXPY-4 operations on Blue Gene/P} 
\label{fig:axpy4-bgp-results} 
\end{figure}

The performance results shown in Figure~\ref{fig:axpy4-bgp-results}
indicate that the code version tuned by Orio consistently outperforms
all the other three versions for both the sequential and parallel
cases. It can be observed that even for a simple algebraic operation,
such as the composed AXPY routines, the compiler alone is unable to
yield performance comparable to the empirically tuned
version. Moreover, implementations that rely on calls to multiple
tuned library routines (e.g., Goto BLAS and ESSL) suffer from loss
both spatial and temporal localities, resulting in inferior memory
performance. The sequential performance results of the BLAS-based
implementations are higher than that of the single loop
implementation; while for the parallel case with high degree of
parallelism, the simple loop implementation is more efficient than the
BLAS codes.

\subsection{Sparse Matrix Computations}

We examined the effectiveness of Orio in optimizing
PETSc~\cite{petsc-efficient}, a toolkit for the parallel numerical
solution of partial differential equations, by empirically tuning one
of its heavily used kernels: sparse matrix-vector multiplication
(SpMV). SpMV dominates the performance of various scientific
applications; yet, traditional implementations of sparse kernels are
relatively poor in performance. In order to attain higher performance,
SpMV requires selecting a compact data structure and code
transformations that best exploit properties of both the sparse matrix
and the underlying architecture. Hence, the need for optimization and
runtime tuning is a big distinction from the dense matrix-vector
multiplication.

SpMV operation takes in the form of $\forall_{A_{i,j}} \neq 0:y_{i}
\leftarrow y_{i} + A_{i,j} \cdot x_{j}$, where $A$ is a sparse matrix,
and $x$, $y$ are dense vectors. Each element of $A$ is used precisely
once and element reuse is only possible for $x$ and $y$. So,
optimizing SpMV should seek data structures for $A$ that are small and
maximize temporal reuse of $x$ and $y$. The most widespread data
structure used for storing a sparse matrix is compressed sparse row
(CSR)~\cite{vuduc-thesis} storage, as shown in
Figure~\ref{fig:spmv}(a). Elements in each row of $A$ are packed
together in a dense array, \texttt{Aval}, and a corresponding array of
integers \texttt{Aind} stores the column indices. The \texttt{Aptr}
array designates where each sparse row begins in \texttt{Aval} and
\texttt{Aind}. A direct implementation of SpMV using CSR storage is
presented in Figure~\ref{fig:spmv}(b).

\begin{figure}[htp]
\centering
\vspace{.2in}
\begin{tabular}{cc}
\begin{minipage}[b]{.5\textwidth}
\centering
\includegraphics[width=.8\textwidth]{figures/spmv.eps}  
\end{minipage}
&
\begin{minipage}[b]{.5\textwidth}
\scriptsize
\begin{verbatim}
for (i=0; i<num_rows; i++)
  for (j=Aptr[i]; j<Aptr[i+1]; j++)
    y[i] += Aval[j]*x[Aind[j]];

\end{verbatim}
\end{minipage}\\
(a) & (b)\\
\end{tabular}
\vspace{-.1in}
\caption{(a) Compressed sparse row (CSR) format (b) Basic implementation of CSR-based SpMV}
\label{fig:spmv}
\end{figure}

The PETSc's SpMV used in this experiment exploits \textit{inode}
format, which takes advantage of rows with identical nonzero
structure. For example, the sparse matrix portrayed in
Figure~\ref{fig:spmv}(a) consists of three inodes. The first inode has
the size of two as it holds two adjacent rows (i.e., the first and the
second rows) with the same nonzero structure. The second inode
contains three consecutive rows with identical nonzero structure,
while the last inode only has a single row. Knowing the properties of
each inode at runtime enables maximum reuse of vector $x$ since
multiple elements of $x$ can be loaded and used exactly once for all
rows in the same inode structure. To do this requires the usage of
register-blocking transformation. In order to optimize inode-based
SpMV routine, we incorporated a new transformation module inside Orio
that implements various optimization strategies including register
blocking, SIMDization, memory alignment optimization, loop-control
optimization, accumulator expansion, and thread-level parallelization
(with OpenMP). We then used Orio to automatically select the best
version of the optimized inode SpMV. It is noted that the loop to be
parallelized with OpenMP is the outer most loop that iterates over
each inode structure.

\begin{figure} [thb]
\begin{center} 
  \subfigure[1 node, SMP mode]{ 
  \includegraphics[width=.3\textwidth]{figures/ex27_bgp/n1_smp.eps}  
  \label{fig:ex27-bgp-smp-n1} 
  } 
  \subfigure[1 node, Dual mode]{ 
  \includegraphics[width=.3\textwidth]{figures/ex27_bgp/n1_dual.eps}  
  \label{fig:ex27-bgp-dual-n1} 
  } 
  \subfigure[1 node, VN mode]{ 
  \includegraphics[width=.3\textwidth]{figures/ex27_bgp/n1_vn.eps}  
  \label{fig:ex27-bgp-vn-n1} 
  } 
  \subfigure[8 nodes, SMP mode]{ 
  \includegraphics[width=.3\textwidth]{figures/ex27_bgp/n8_smp.eps}  
  \label{fig:ex27-bgp-smp-n8} 
  } 
  \subfigure[8 nodes, Dual mode]{ 
  \includegraphics[width=.3\textwidth]{figures/ex27_bgp/n8_dual.eps}  
  \label{fig:ex27-bgp-dual-n8} 
  } 
  \subfigure[8 nodes, VN mode]{ 
  \includegraphics[width=.3\textwidth]{figures/ex27_bgp/n8_vn.eps}  
  \label{fig:ex27-bgp-vn-n8} 
  } 
  \subfigure[32 nodes, SMP mode]{ 
  \includegraphics[width=.3\textwidth]{figures/ex27_bgp/n32_smp.eps}  
  \label{fig:ex27-bgp-smp-n32} 
  } 
  \subfigure[32 nodes, Dual mode]{ 
  \includegraphics[width=.3\textwidth]{figures/ex27_bgp/n32_dual.eps}  
  \label{fig:ex27-bgp-dual-n32} 
  } 
  \subfigure[32 nodes, VN mode]{ 
  \includegraphics[width=.3\textwidth]{figures/ex27_bgp/n32_vn.eps}  
  \label{fig:ex27-bgp-vn-n32} 
  } 
\end{center}
\vspace{-.2in} 
\caption{Performance of inode SpMV on Blue Gene/P for 1, 8, and 32 nodes.} 
\label{fig:ex27-bgp-results1} 
\end{figure}

\comment{
\begin{figure} 
\begin{center} 
  \subfigure[32x32, SMP mode]{ 
  \includegraphics[width=.3\textwidth]{figures/ex27_bgp/s32_smp.eps}  
  \label{fig:ex27-bgp-smp-32x32} 
  } 
  \subfigure[32x32, Dual mode]{ 
  \includegraphics[width=.3\textwidth]{figures/ex27_bgp/s32_dual.eps}  
  \label{fig:ex27-bgp-dual-32x32} 
  } 
  \subfigure[32x32, VN mode]{ 
  \includegraphics[width=.3\textwidth]{figures/ex27_bgp/s32_vn.eps}  
  \label{fig:ex27-bgp-vn-32x32} 
  } 
  \subfigure[64x64, SMP mode]{ 
  \includegraphics[width=.3\textwidth]{figures/ex27_bgp/s64_smp.eps}  
  \label{fig:ex27-bgp-smp-64x64} 
  } 
  \subfigure[64x64, Dual mode]{ 
  \includegraphics[width=.4\textwidth]{figures/ex27_bgp/s64_dual.eps}  
  \label{fig:ex27-bgp-dual-64x64} 
  } 
  \subfigure[64x64, VN mode]{ 
  \includegraphics[width=.3\textwidth]{figures/ex27_bgp/s64_vn.eps}  
  \label{fig:ex27-bgp-vn-64x64} 
  } 
  \subfigure[128x128, SMP mode]{ 
  \includegraphics[width=.3\textwidth]{figures/ex27_bgp/s128_smp.eps}  
  \label{fig:ex27-bgp-smp-128x128} 
  } 
  \subfigure[128x128, Dual mode]{ 
  \includegraphics[width=.3\textwidth]{figures/ex27_bgp/s128_dual.eps}  
  \label{fig:ex27-bgp-dual-128x128} 
  } 
  \subfigure[128x128, VN mode]{ 
  \includegraphics[width=.3\textwidth]{figures/ex27_bgp/s128_vn.eps}  
  \label{fig:ex27-bgp-vn-128x128} 
  } 
\end{center}
\vspace{-.2in}
\caption{Performance of inode SpMV on Blue Gene/P for problem sizes 32x32, 64x64, and 128x128.} 
\label{fig:ex27-bgp-results2} 
\end{figure} 
}

To evaluate the performance of the tuned SpMV routine, we conduct the
experiment using one of actual application simulations of PETSc,
called \textit{Nonlinear 2-D Driven Cavity} problem, on both the
multicore Intel machine and Blue Gene/P. The ``Compiler-optimized''
label is used as a base comparison that represents the performance of
the naive implementation of SpMV (shown in Figure~\ref{fig:spmv}(b))
optimized by the native compiler. We also tested the performance of
the hand-tuned SpMV code that is already integrated in the PETSc
releases.

Figure~\ref{fig:ex27-cookie-results} exhibits the performance speedups
accomplished by Orio on the Intel Xeon machine. To be noted that the
input problem size signifies the number of grid points in $x$ and $y$
directions, not the size of the sparse input matrix. In this
experiment, we utilized all available cores by running PETSc in three
different modes: SMP (one MPI process, eight threads/process), dual
(four MPI processes, two threads/process), and virtual node or VN
(eight MPI processes, one thread/process). The results demonstrate
that the code tuned by Orio constantly gives the best performance
costs for all cases. And the performances of the hand-tuned code are
almost equivalent to those of the ICC-optimized code for most
cases. We observe that performance differences between the Orio
version and the hand-tuned version become larger as more threads per
process are employed into the system to facilitate OpenMP
parallelization. By comparing the SMP and VN results, it is
interesting to see that optimization using loop-level parallelism (by
OpenMP) offers much better performance than using coarse-grained
parallelism (by MPI). Moreover, for the multithread cases, tuning
using Orio determined that performance benefit from SIMDization is
more significant than thread-level parallelization only when the
problem sizes are small. The reason for this is due to the overhead
introduced by OpenMP from coordinating its threads across
processors. So, each optimized code for the many-thread cases
comprises two distinct tuned codes to handle both small and large
problem sizes.

As we can see in Figure~\ref{fig:ex27-bgp-results1}, our experiment on
Blue Gene/P was done for a varied number of nodes (i.e., 1, 8, and 32)
and the same three different parallel modes as before (i.e., SMP,
dual, and VN). Again, empirical optimization using Orio gives the best
performance costs for all cases, outperforming both the hand-tuned and
the ICC-optimized codes. We can also see from the results that the
performance gap between the ICC-optimized and the hand-tuned codes is
now larger than that in the earlier Intel experiment. Similarly, due
to thread-level parallelism, the Orio-optimized code has larger
performance differences than the hand-tuned code when the number of
threads per process increases and also when the number of nodes
decreases. For the single-thread cases (i.e., VN mode), thread-level
parallelism is unexploited and yet, the Orio version still performs
better than the other two alternatives. This validates the
effectiveness of Orio in empirically tuning sequential codes.

\begin{figure} [thb]
\begin{center} 
  \subfigure[SMP: $p=1$, $t/p=8$]{
  \includegraphics[width=.3\textwidth]{figures/ex27_cookie/smp.eps}
  \label{fig:ex27-cookie-smp} } \subfigure[Dual: $p=4$, $t/p=2$]{
  \includegraphics[width=.3\textwidth]{figures/ex27_cookie/dual.eps}
  \label{fig:ex27-cookie-dual} } \subfigure[VN: $p=8$, $t/p=1$]{
  \includegraphics[width=.3\textwidth]{figures/ex27_cookie/vn.eps}
  \label{fig:ex27-cookie-vn} }
\end{center}
\vspace{-.2in} 
\caption{Performance of Inode SpMV on eight-core Intel machine; $p$ is the number of processes, $t/p$ is the threads per process.} 
\label{fig:ex27-cookie-results} 
\end{figure}

\subsection{Irregular Computation Kernels}

This section discusses the performance evaluations of the integrated
PLuTo-Orio system (explained earlier in Section~\ref{sec:integration})
on the multicore Intel machine by using a number of application
kernels that are non-trivial to optimize and parallelize. We compare
the performance of the code tuned by Orio with the base code and the
PLuTo-generated codes. The PLuTo code was obtained via running the
base code with PLuTo-0.3.0~\cite{pluto030} using \texttt{--tile} to
employ loop tiling for L1 cache,
\texttt{--l2tile} to employ loop tiling for L2 cache, and
\texttt{--unroll} to employ loop unrolling. Generating a parallel
version of PLuTo code needs an additional \texttt{--parallel}
option. Note that we used PLuTo's default tile sizes and unroll
factors to produce all PLuTo codes. By default, PLuTo always sets the
L1 tile sizes to 32x32 (or 32x32x32) and L2 tile sizes to 256x256 (or
256x256x256) depending on the loop nest depth. The default unroll
factors are always 64 for 1-D unrolling and 8x8 for 2-D
unroll/jamming. Furthermore, all codes were compiled with Intel C
Compiler using \texttt{-O3} optimization flag to ensure
auto-vectorization and other ICC optimizations, and \texttt{-parallel}
(or \texttt{-openmp}, when manual parallelization using OpenMP was
used) to enable code parallelization.

In the following sections, we refer to the ICC-optimized base code as
'Compiler-optimized', the PLuTo-generated code with L1 tiling and
unroll/jamming as 'PLuTo (L1 tiling)', and the PLuTo-generated code
with L1 and L2 tilings and unroll/jamming as 'PLuTo (L1+L2
tiling)'. Also, the code tuned by Orio is referred to as 'PLuTo+Orio'.

\subsubsection{2-D Finite-Difference Time-Domain Method for Computational Electromagnetics}  
We consider the two-dimensional Finite Difference Time Domain (FDTD),
a popular method for solving computational electrodynamic problems. As
illustrated in Figure~\ref{fig:fdtd-2d-code}, the code of 
%
\begin{wrapfigure}{r}{3in}
\vspace{-0.2in}
\begin{center}
\begin{minipage}{3in} 
\scriptsize
\begin{verbatim} 
for(t=0; t<tmax; t++) { 
  for (j=0; j<ny; j++) 
    ey[0][j] = t; 
  for (i=1; i<nx; i++) 
    for (j=0; j<ny; j++) 
      ey[i][j] -= 0.5*(hz[i][j] - hz[i-1][j]); 
  for (i=0; i<nx; i++) 
    for (j=1; j<ny; j++) 
      ex[i][j] -= 0.5*(hz[i][j] - hz[i][j-1]); 
  for (i=0; i<nx; i++) 
    for (j=0; j<ny; j++) 
      hz[i][j] -= 0.7*(ex[i][j+1] - ex[i][j]
                  + ey[i+1][j] - ey[i][j]); 
} 
\end{verbatim} 
\end{minipage} 
\end{center}
\vspace{-0.2in}
\caption{2-D FDTD code} 
\label{fig:fdtd-2d-code} 
\end{wrapfigure}
%
2-D FDTD consists of four statements that are imperfectly nested. The arrays
$ex$ and $ey$ denote the electric fields, while the array $hz$ denotes
the magnetic field.

The performance of sequential 2-D FDTD codes for $tmax=500$ and
$nx=ny$ is presented in Figure~\ref{fig:fdtd-2d-cookie-seq}. The base
code optimized by ICC performs better than PLuTo for small problem
sizes since all input arrays fit in the L2 cache, thus tiling
performed by PLuTo will only introduce more overhead. As the array
sizes become larger, the lack of data reuse impairs the base code's
performance, whereas the PLuTo performance remains steady. Code tuning
using Orio generates two code variants: one for small problem sizes
and another for large problem sizes. When the input array sizes are
small, Orio discovers that applying PLuTo's polyhedral transformations
is not beneficial, and therefore, it employs only its syntactic
transformations on the original FDTD code. As to tuning for large
problem size, Orio exploits some of the PLuTo's code transformations
and enhances it further with its syntactic optimizations, resulting in
performance constantly higher than both the base and PLuTo codes. By
tuning the sequential code with Orio, the attainable performance
enhancement over PLuTo is up to 86\%.

Figure~\ref{fig:fdtd-2d-cookie-par} displays the multicore performance
obtained for $tmax=500$ and $nx=ny=2000$. The results indicate that
while ICC is unable to auto-parallelize the code, PLuTo detects the
existence of pipelined parallelism and then parallelizes the
code. Optimizing and tuning by Orio additionally improves the PLuTo
performance by up to 78\%. Moreover, it can be seen that due to memory
contention between the two quad-core Intel processors, the performance
speedup of both the PLuTo and Orio codes slightly decreases when the
number of used cores is greater than four.

\begin{figure}[htb] 
\begin{center} 
  \subfigure[Sequential (T=500)]{ 
  \includegraphics[width=.4\textwidth]{figures/fdtd-2d-cookie/seq.eps} 
  \label{fig:fdtd-2d-cookie-seq} } \subfigure[Parallel (T=500, 
  N=2000)]{ 
  \includegraphics[width=.4\textwidth]{figures/fdtd-2d-cookie/par.eps} 
  \label{fig:fdtd-2d-cookie-par} } 
\end{center}
\vspace{-.2in} 
\caption{2-D FDTD performance on eight-core Intel machine} 
\label{fig:fdtd-2d-cookie-results} 
\vspace{-.2in}
\end{figure}

%%BN: commented out since this needs more explanation (no room in this version)
%It is to be noted that collecting performance results of the PLuTo 
%code with L1 and L2 tilings was not possible because of the large code 
%size of the generated loop nest, which led to an impractically huge 
%compile time.

\subsubsection{3-D Gauss-Seidel Successive Over-Relaxation Method}  
Figure~\ref{fig:seidel-code} shows the 3-D Gauss-Seidel computation, 
which sometimes is referred to as
%
\begin{wrapfigure}{r}{3in}
\vspace{-0.2in}
\begin{center}
\begin{minipage}{3in} 
\scriptsize
\begin{verbatim} 
for (t=0; t<=T-1; t++) 
  for (i=1; i<=N-2; i++) 
    for (j=1; j<=N-2; j++) 
      A[i][j] = (A[i-1][j-1] + A[i-1][j] 
                + A[i-1][j+1] + A[i][j-1] 
                +A[i][j] + A[i][j+1] 
                + A[i+1][j-1] + A[i+1][j] 
                + A[i+1][j+1]) / 9.0; 
\end{verbatim} 
\end{minipage} 
\end{center}
\vspace{-0.1in}
\caption{3-D Gauss-Seidel code} 
\label{fig:seidel-code} 
\end{wrapfigure}
%
\emph{successive displacement method} to indicate the dependence of the iterations on the 
ordering. If the ordering is changed, the components of the new 
iterations will change as well.

The sequential performance results shown in
Figure~\ref{fig:seidel-cookie-seq} are when $T=500$. As the figure
shows, applying PLuTo's polyhedral tiling on the original code always
delivers performance boosts, which range from 126\% to
142\%. Performing two-level tiling (for both L1 and L2 caches) by
PLuTo yields 13\% lower performance than only using L1 tiling; and
this is also reflected in the best sequential code found by Orio,
where one-level tiling (for L1 cache only) is always performed. The
sequential speedup obtained from tuning the PLuTo code with Orio is
significant, ranging from 51\% to 133\%.

\begin{figure} [htb]
\begin{center} 
  \subfigure[Sequential (T=500)]{ 
  \includegraphics[width=.4\textwidth]{figures/seidel-cookie/seq.eps} 
  \label{fig:seidel-cookie-seq} } \subfigure[Parallel (T=500, 
  N=4000)]{ 
  \includegraphics[width=.4\textwidth]{figures/seidel-cookie/par.eps} 
  \label{fig:seidel-cookie-par} } 
\end{center}
\vspace{-.2in} 
\caption{3-D Gauss-Seidel performance on eight-core Intel machine} 
\label{fig:seidel-cookie-results} 
\end{figure}

Parallel performance for $T=500$ and $N=4000$ can be viewed in
Figure~\ref{fig:seidel-cookie-par}. Similar earlier trend also exists
in these performance results: PLuTo parallelizes the code through its
polyhedral transformations (i.e., skewing and tiling), while the ICC
compiler alone does not parallelize the loop. Empirical tuning via
Orio creates up to 122\% speedups over the one-level tiled PLuTo code,
and up to 548\% speedups over the two-level tiled PLuTo code. Another
finding is that tiling for both L1 and L2 caches contributes to a poor
scalability because the number of tiles can be less than the number of
processors, resulting in underutilization of the available resources.

\subsubsection{LU Decomposition}

LU Decomposition is a popular numerical method to solve linear
equation systems, and the 
%
\begin{wrapfigure}{r}{3in}
\begin{center}
\vspace{-.25in}
\begin{minipage}{2.8in} 
\scriptsize
\begin{verbatim} 
for (k=0; k<=N-1; k++) { 
  for (j=k+1; j<=N-1; j++) 
    A[k][j] /= A[k][k]; 
  for(i=k+1; i<=N-1; i++) 
    for (j=k+1; j<=N-1; j++) 
      A[i][j] -= A[i][k]*A[k][j]; 
} 
\end{verbatim} 
\end{minipage} 
\end{center}
\vspace{-0.2in}
\caption{LU Decomposition code} 
\label{fig:lu-code} 
\end{wrapfigure}
%
corresponding code is available in Figure~\ref{fig:lu-code}.

The sequential performance results of LU Decomposition are presented
in Figure~\ref{fig:lu-cookie-seq}. Like 2-D FDTD, the ICC-optimized
code is more efficient than the PLuTo-generated codes when the input
arrays are small and fit in the L2 cache. However, due to better data
locality, the performance of the PLuTo-tiled codes exceeds the base
code as the input arrays get larger. Again, there are two different
codes generated by Orio: one for small problem sizes and one for large
problem sizes. And only the code tuned for large input arrays employs
PLuTo's polyhedral transformations, prior to applying Orio's syntactic
transformations. Unlike 3-D Gauss-Seidel, this time Orio chooses to
apply both L1 and L2 tilings when generating the best tuned
code. Compared to the two PLuTo codes, the Orio-tuned code yields
performance speedups in the range of 26\% to 277\%.

Parallel performance results shown in Figure~\ref{fig:lu-cookie-par}
were obtained for $N=4000$. From the results, we observe that ICC is
not able to parallelize the code, whereas PLuTo achieves higher
performance by exploitation of multicore parallelism. Furthermore,
both PLuTo-generated codes have comparable performance numbers, with a
slightly better scalability on the one-level tiled code. By the use of
Orio tuning, the performances of the PLuTo base codes were improved by
a factor of 51\% to 120\%.

\begin{figure}[htb]
\begin{center} 
  \subfigure[Sequential]{ 
  \includegraphics[width=.4\textwidth]{figures/lu-cookie/seq.eps} 
  \label{fig:lu-cookie-seq} } \subfigure[Parallel (N=4000)]{ 
  \includegraphics[width=.4\textwidth]{figures/lu-cookie/par.eps} 
  \label{fig:lu-cookie-par} } 
\end{center} 
\vspace{-.2in}
\caption{LU Decomposition performance on eight-core Intel machine} 
\label{fig:lu-cookie-results} 
\end{figure}

### 2008-10-14 src/module/loop/loop.py

@@ -33,7 +33,7 @@ class Loop(module.module.Module):
         indent = ' ' * self.indent_size
         extra_indent = '  '
         cgen = codegen.CodeGen()
-        transformed_code = ''
+        transformed_code = '\n'
         for s in transformed_stmts:
             transformed_code += cgen.generate(s, indent, extra_indent)
 

### 2008-11-19 doc/papers/iccs09/f77/waxpby.f
initial version copied from cgo directory, but using the new style
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@367 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@ -1,0 +1,41 @@
+      subroutine waxpby(n, w, alpha, x, beta, y)
+      integer n, incx, incy
+      double precision alpha
+      double precision yy(n)
+      double precision w(*), x(*), y(*)
+c
+c   WAXPBY
+c   in
+c     alpha : scalar,
+c     x : vector,
+c     beta : scalar,
+c     y : vector
+c   out
+c     w : vector
+c   {
+c     w = alpha * x + beta * y
+c   }
+c
+      incx = 1
+      incy = 1
+c
+c  Copy y into yy so that input is not overwritten.
+c
+      call dcopy(n, y, incx, yy, incy)
+c
+c  Put beta*yy into yy
+c
+      call dscal(n, beta, yy, incx )
+c
+c  Put x + yy into yy
+c
+      call daxpy(n, alpha, x, incx, yy, incy)
+c
+c  Copy yy into w.
+c
+      call dcopy(n, yy, incx, w, incy)
+c
+c
+      return
+      end
+

### 2008-11-19 doc/papers/iccs09/paper.tex
initial version copied from cgo directory, but using the new style
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@367 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

--- 2008-10-13 doc/papers/cgo09/paper.tex
@@
 
-\documentclass[11pt]{article}
+\documentclass[runningheads]{llncs}
 
+\usepackage{amssymb}
+\setcounter{tocdepth}{3}
+\usepackage{graphicx}
 \usepackage{url}
@@
 \usepackage{cite}
-\usepackage{times}
 \usepackage{amsmath}
-\usepackage[margin=1in]{geometry}
+
+\urldef{\mailsa}\path|{alfred.hofmann,ursula.barth,ingrid.beyer,natalie.brecht,|
+\urldef{\mailsb}\path|christine.guenther,frank.holzwarth,piamaria.karbach,|
+\urldef{\mailsc}\path|anna.kramer,erika.siebert-cole,lncs}@springer.com|
+\newcommand{\keywords}[1]{\par\addvspace\baselineskip
+\noindent\keywordname\enspace\ignorespaces#1}
 
@@
 
+\begin{document}
+
+\mainmatter  % start of an individual contribution
+
+% first the title is needed
 \title{Generating Empirically Optimized Numerical Software from MATLAB Prototypes}
 
-%\author{
-%Elizabeth Jessup,$^1$ Boyana Norris,$^2$ Albert Hartono,$^3$ and Jeremy Siek$^4$\\
-%\\
-%\vspace{-.05in}
-%{\small $^1$University of Colorado at Boulder Dept. of Computer Science, Boulder, CO~80309}\\
-%\vspace{-.05in}
-%{\small \url{jessup@cs.colorado.edu}}\\
-%\vspace{-.05in}
-%{\small $^2$Argonne National Laboratory Mathematics and Computer Science Division, Argonne, IL~60439}\\
-%\vspace{-.05in}
-%{\small \url{norris@mcs.anl.gov}}\\
-%\vspace{-.05in}
-%{\small $^3$Ohio State University Dept. of Computer Science and Engineering, Columbus, OH~43210}\\
-%\vspace{-.05in}
-%{\small \url{hartonoa@cse.ohio-state.edu}}\\
-%\vspace{-.05in}
-%{\small $^4$University of Colorado at Boulder Dept. of Electrical and Computer Engineering, Boulder, CO~80309}\\
-%\vspace{-.05in}
-%{\small \url{jeremy.siek@colorado.edu}}
-%}
+        % a short form should be given in case it is too long for the running head
+        \titlerunning{Empiriclaly Optimized Numerical Software from MATLAB}
 
-\begin{document}
+        % the name(s) of the author(s) follow(s) next
+        %
+        % NB: Chinese authors should write their first names(s) in front of
+        % their surnames. This ensures that the names appear correctly in
+        % the running heads and the author index.
+        %
+\author{Elizabeth Jessup\inst{1} 
+  \and Boyana Norris\inst{2}
+  \and Albert Hartono\inst{3}
+  \and Jeremy Siek\inst{1}}
 
+\authorrunning{Jessup, Norris, Hartono, Siek}
+
+\institute{University of Colorado at Boulder \\
+  \email{\{elisabeth.jessup,jeremy.siek\}@colorado.edu}
+  \and Argonne National Laboratory \\
+  \email{\{norris\}@mcs.anl.gov}
+  \and Ohio State University \\
+  \email{\{hartonoa\}@cse.ohio-state.edu}}
+
+%\toctitle{Lecture Notes in Computer Science}
+\tocauthor{Elizabeth Jessup, Boyana Norris, Albert Hartono, and Jeremy Siek}
 \maketitle
-%
-\onehalfspacing
+
 \begin{abstract}
+% between 75 and 150 words
 The growing demand for higher levels of detail and accuracy in results means that the size and complexity of scientific computations is increasing at least as fast as the improvements in processor technology.
@@
 We discuss performance results showing that the code generated by using our system significantly outperforms not only the original simple C code but also code based on source BLAS, ATLAS-optimized BLAS, and Intel MKL routines.
+\keywords{Code generation, empirical performance tuning}
 \end{abstract}
-%
-\doublespacing
-%
+
 \section{Introduction}
@@
 
-\singlespacing
+%\singlespacing
 %% or stuff it in a footnote
@@
 
-\bibliographystyle{acm}
+\bibliographystyle{splncs}
 \bibliography{paper}

### 2008-11-19 doc/papers/iccs09/C/gesummv.c

@@ -1,0 +1,72 @@
+/* gesummv.f -- translated by f2c (version 20050501).
+   You must link the resulting object file with libf2c:
+	on Microsoft Windows system, link with libf2c.lib;
+	on Linux or Unix systems, link with .../path/to/libf2c.a -lm
+	or, if you install libf2c.a in a standard place, with -lf2c -lm
+	-- in that order, at the end of the command line, as in
+		cc *.o -lf2c -lm
+	Source for libf2c is in /netlib/f2c/libf2c.zip, e.g.,
+
+		http://www.netlib.org/f2c/libf2c.zip
+*/
+
+#include "f2c.h"
+
+/* Subroutine */ int gesummv_(doublereal *alpha, doublereal *beta, integer *n,
+	 integer *lda, doublereal *a, integer *ldb, doublereal *b, doublereal 
+	*x, doublereal *y)
+{
+    /* System generated locals */
+    integer a_dim1, a_offset, b_dim1, b_offset;
+
+    /* Local variables */
+    static doublereal one;
+    static integer incx, incy;
+    static doublereal zero;
+    extern /* Subroutine */ int dgemv_(char *, integer *, integer *, 
+	    doublereal *, doublereal *, integer *, doublereal *, integer *, 
+	    doublereal *, doublereal *, integer *, ftnlen);
+
+
+/*   GESUMMV */
+/*   in */
+/*     A : column matrix, */
+/*     B : column matrix, */
+/*     x : vector, */
+/*     a : scalar, */
+/*     b : scalar */
+/*   out */
+/*     y : vector */
+/*   { */
+/*     y = alpha * (A * x) + beta * (B * x) */
+/*   } */
+
+    /* Parameter adjustments */
+    a_dim1 = *lda;
+    a_offset = 1 + a_dim1;
+    a -= a_offset;
+    b_dim1 = *ldb;
+    b_offset = 1 + b_dim1;
+    b -= b_offset;
+    --x;
+    --y;
+
+    /* Function Body */
+    incx = 1;
+    incy = 1;
+
+/*   Put beta*B*x into y. */
+
+    zero = 0.;
+    dgemv_("n", n, n, beta, &b[b_offset], lda, &x[1], &incx, &zero, &y[1], &
+	    incy, (ftnlen)1);
+
+/*   Put y = beta*B*x + alpha*A*x into y. */
+
+    one = 1.;
[…]

### 2008-11-19 doc/papers/iccs09/f77/gemver.f

@@ -1,0 +1,54 @@
+      subroutine gemver(alpha,beta,lda,n,A,ldb,B,u1,v1,u2,v2,w,x,y,z)
+      integer incx, incy, i, j, n
+      double precision alpha, beta, zero, one
+      double precision w(*), x(*), y(*), u1(*), u2(*), v1(*), v2(*)
+      double precision z(n)
+      double precision A(lda,*), B(ldb,*)
+c
+c   GEMVER
+c   in
+c     A : column matrix, u1 : vector, u2 : vector, v1 : vector, v2 : vector,
+c     alpha : scalar, beta : scalar,
+c     y : vector, z : vector
+c   out
+c     B : column matrix, x : vector, w : vector
+c
+      incx = 1
+      incy = 1
+      one = 1.0d0
+      zero = 0.0d0
+c
+c     B = A + u1 * v1' + u2 * v2'
+c     x = beta * (B' * y) + z
+c     w = alpha * (B * x)
+c
+c
+c  Copy A so that it doesn't get overwritten.
+c
+      do j = 1,n
+         do  i = 1,n
+            B(i,j) = A(i,j)
+         enddo
+c      do j = 1,n
+c         call dcopy(n, A(1,j), incx, B(1,j), incx)
+      enddo
+c
+c  Rank 1 updates
+c
+      call dger(n, n, one, u1, incx, v1, incx, B, ldb )	
+      call dger(n, n, one, u2, incx, v2, incx, B, ldb )	
+c
+c  Copy z so that it doesn't get overwritten.
+c
+      call dcopy(n, z, incx, x, incy)
+c
+c     x = beta * (B' * y) + z  (z is x here)
+c
+      call dgemv( 't', n, n, beta, B, lda, y, incx, one, x, incx )
+c
+c     w = alpha * (B * x)
+c
+      call dgemv( 'n', n, n, alpha, B, lda, x, incx, zero, w, incx )
+c
+      return
+      end

### 2008-11-19 doc/papers/iccs09/f77/vadd.f

@@ -1,0 +1,38 @@
+      subroutine vadd(n, x, w, y, z) 
+      integer n, incx, incy
+      double precision one
+      double precision yy(n) 
+      double precision w(*), x(*), y(*), z(*)
+c
+c   vadd
+c   in
+c   w : vector, y : vector, z : vector
+c   out
+c   x : vector
+c   {
+c    x = w + y + z
+c   }
+c
+      incx = 1
+      incy = 1
+      one = 1.0d0
+c
+c  Copy y into yy so that input is not overwritten.
+c
+      call dcopy(n, y, incx, yy, incy)
+c
+c  y = w + y
+c
+      call daxpy(n, one, w, incx, yy, incy)
+c
+c  y = z + y
+c
+      call daxpy(n, one, z, incx, yy, incy)
+c
+c  Copy result into x.
+c
+      call dcopy(n, yy, incx, x, incy)
+c
+      return
+      end
+

### 2008-12-09 src/module/ortil/transformator.py

@@ -56,7 +56,7 @@ class SimpleLoops:
                 is_done = True
                 break
 
-        # create a new loop and add it into the sequence of loops 
+        # fusing is not possible, thus create a new loop and add it into the sequence of loops 
         if not is_done:
             inames_seq.reverse()
             l = stmt
@@ -67,7 +67,7 @@ class SimpleLoops:
     #----------------------------------------------
 
     def convertToASTs(self, loop_info_table, node = None):
-        '''To generate a sequence of ASTs that correspond to this sequence of loops'''
+        '''To generate a sequence of ASTs that correspond to this sequence of "simple" loops'''
 
         if node == None:
             node = self.loops
@@ -100,23 +100,31 @@ class Transformator:
         num_level, tiling_table = tiling_info
         tiled_loop_inames = tiling_table.keys()
         tile_size_table = tiling_table
-
+        
         self.perf_params = perf_params
         self.num_level = num_level
         self.tiled_loop_inames = tiled_loop_inames
         self.tile_size_table = tile_size_table
         self.ast_util = ast_util.ASTUtil()
+
+        # used for variable name generation
         self.counter = 1
+
+        # to turn on/off the static determination of loop bound values for scanning the full tiles
+        self.affine_lbound_exps = False
 
     #----------------------------------------------
 
     def __getIterName(self, iter_name, level):
+        '''Generate a new variable name for inter-tile loop iterator'''
         return iter_name + ('t%s' % level)
 
     def __getTileSizeName(self, iter_name, level):
+        '''Generate a new variable name for tile size variable'''
         return ('T%s' % level) + iter_name
 
     def __getLoopBoundNames(self):
+        '''Generate new variable names for both lower and upper loop bounds'''
         lb_name = 'lb%s' % self.counter
         ub_name = 'ub%s' % self.counter
         self.counter += 1
@@ -187,10 +195,133 @@ class Transformator:
     
     #----------------------------------------------
 
-    def __getLoopBoundScanningStmts(self, stmts, outer_loop_inames, loop_info_table):
+    def __findMinMaxVal(self, min_or_max, exp, var_names, val_table, up_sign = 1):
         '''
-        Generate an explicit loop-bound scanning code used at runtime to determine the latest start
-        and the earliest end of scanning full tiles.
+        To statically find the actual min/max value of the given expression, based on the given 
[…]

### 2008-12-09 testsuite/ortil.c

@@ -1,5 +1,17 @@
 
 //--------------------------------------------------
+// DGEMM
+//--------------------------------------------------
+/*@ begin OrTil ((i,j,k):2) @*/
+
+  for (i=0;i<=M-1;i++) 
+    for (j=0;j<=N-1;j++) 
+      for (k=0;k<=O-1;k++) 
+	X[i][j]=X[i][j]+A[i][k]*B[k][j];
+
+/*@ end @*/
+
+//--------------------------------------------------
 // 1-d Jacobi
 //--------------------------------------------------
 /*@ begin OrTil ((t,i):3) @*/

### 2008-12-11 doc/papers/iccs09/paper.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@382 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@383 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@
 
+% formatting cheats
+\renewcommand{\baselinestretch}{0.92}  
+\renewcommand\floatpagefraction{.9} 
+\renewcommand\topfraction{.9} 
+\renewcommand\bottomfraction{.9} 
+\renewcommand\textfraction{.1} 
+\setcounter{totalnumber}{50} 
+\setcounter{topnumber}{50} 
+\setcounter{bottomnumber}{50} 
+\setlength{\textwidth}{12.5cm} 
+\setlength{\textheight}{19.5cm}
+
 \begin{document}
@@
         % a short form should be given in case it is too long for the running head
-        \titlerunning{Empiriclaly Optimized Numerical Software from MATLAB}
+        \titlerunning{Empirically Optimized Numerical Software from MATLAB}
 
@@
 \centering
-\includegraphics[width=.7\textwidth]{figures/process.png}
-
+\includegraphics[width=0.7\textwidth]{figures/process.eps}
 \caption{Overview of the code generation and tuning process, optional steps are indicated with dashed arrows.}
@@
 \centering
-\includegraphics[width=.7\textwidth]{figures/compile.png}
+\includegraphics[width=.7\textwidth]{figures/compile.eps}
 
@@
 \centering
-\includegraphics[width=.7\textwidth]{figures/gemver-dataflow.png}
+\includegraphics[width=.7\textwidth]{figures/gemver-dataflow.eps}
 
@@
   \centering
-  \includegraphics[width=\textwidth]{figures/refine-mv-dot.png}
+  \includegraphics[width=\textwidth]{figures/refine-mv-dot.eps}
 
@@
 %~\cite{Norris:2007}
-Orio is an empirical tuning tool that takes as input annotated C code and generates multiple tuned versions.
-Orio does not parse the application code; rather, it relies on annotations expressed as structured comments, which contain a simplified expression of the computation, as well as directives on what transformations to apply.
+Orio is an empirical tuning tool that takes annotated C code as input,
+generates multiple transformed versions of the annotated code, and
+empirically evaluates the performance of the generated codes,
+ultimately choosing the best-performing version to use in production
+runs.
 
-Figure~\ref{fig:orio} illustrates the tuning process using Orio.
-In the simplest case, Orio can be used to improve code performance by performing source-to-source transformations, such as loop unrolling and memory alignment optimizations.
-First, the input to Orio is C code, which contains syntactically structured comments that contain both the computation to be tuned, as well as various performance-tuning directives.
-Orio scans the annotated input code and extracts all annotated regions.
-Each annotated region is then passed to the code transformation modules for potential optimizations.
-The code generator then produces the final code incorporating the newly generated optimized C code corresponding to the annotation.
-The original C code is not parsed (only the annotation comments are processed) and is preserved in the new version.
-The user has the option of selecting either the original or the tuned code by using a compile-time macro.
+Figure~\ref{fig:orio} illustrates the tuning process implemented in
+Orio. The input to Orio is C code containing semantic comments that
+describe both the computation, as well as various performance-tuning
+directives. Orio first extracts all annotation code regions by parsing
+the marked-up input code. Each annotated region is then passed to code
+transformator and code generator for potential optimizations. Then the
+transformed C code with various incorporated optimizations
+corresponding to the specified annotations is produced. The
+source-to-source transformation system of Orio generates an optimized
+code version for each distinct combination of performance parameter
+values. Each generated code variant is then executed and its
+performance cost evaluated. After iteratively testing all code
+variants, the best-performing code is selected as the final output of
+Orio. Because the search space of all possible optimized code variants
+can be exponentially large, the search engine implements a number of
+search heuristics (i.e., random, simplex, and simulated annealing) to
+effectively narrow the search for near-optimal performance. The tuning
+specifications, written by users in the form of annotations, contains
+information required for guiding the transformation and tuning
+process.
 
-\begin{figure}[htbp]
-\centering
-\includegraphics[width=.7\textwidth]{figures/orio.png}
+\begin{figure}[tb]
+\vspace{-.1in}
+\begin{center}
+\includegraphics[width=.65\textwidth]{figures/orio.eps}
+\end{center}
+\vspace{-.25in}
 \caption{Overview of the Orio empirical tuning process.}
@@
 
-Orio can also be used as an automatic performance tuning tool.
-The system uses its code transformation modules and code generator to generate an optimized code version for each distinct combination of performance parameters.
-Then the optimized code versions are executed and their performance measured, which is subsequently compared to the performances of other previously tested code variants.
-After iteratively evaluating all code variants under consideration, the best-performing code is selected as the final output of Orio.
-The tuning specifications, written by users in the form of annotations (either embedded in the source code or in separate files), are parsed and used by Orio to guide the search and tuning process.
-These specifications include important information such as the used compilers, the search strategy, the program transformation parameters, the input data sizes, etc.
-
-We use the VADD operation, $x = w + y + z$, as a simple example of the annotations and tuning specification used by Orio.
-Figure~\ref{fig:orio-example} shows the C code with annotation comments indicating that Orio should perform memory alignment and loop transformations.
-In addition to simple source transformation annotations in the example above, Orio supports a number of other transformations, including loop unroll and jam, composite optimizations, and architecture-specific directives (e.g., for generation of calls to SIMD intrinsics on the Blue Gene/P).
-
-\begin{figure}[htp]
+\begin{figure}[htb]
 \centering
@@
 void vadd(int n, double *y, double *x1,
-               double *x2, double *x3)
-{
-    int i;
+          double *x2, double *x3) {
+/*@ begin PerfTuning(
+  import spec vadd_tune_spec;
+) @*/
 
-/*@ begin Align (x1[],x2[],x3[],y[]) @*/
-/*@ begin Loop (
-  transform Unroll(ufactor=20, parallelize=True)
-    for (i=0; i < n; i++)
+ register int i;
+
+/*@ begin BGP_Align(y[],x1[],
+                    x2[],x3[]) @*/
+/*@ begin Loop(
+ transform Unroll(ufactor=UF, 
+                  parallelize=PAR)
+ for (i=0; i<=n-1; i++)
    y[i] = x1[i] + x2[i] + x3[i];
@@
 
-    for (i=0; i < n ; i++)
+ for (i=0; i<=n-1; i++)
    y[i] = x1[i] + x2[i] + x3[i];
@@
 /*@ end @*/
-
+/*@ end @*/
 }
@@
 \begin{verbatim}
-spec align_unroll {
- def build  
- { 
+spec vadd_tune_spec {
+ def build { 
   arg build_command = 'mpixlc -O3 -qstrict -lm';
+  arg batch_command = 'qsub -n 64 -t 10';
+  arg status_command = 'qstat';
+  arg num_procs = 64;
  }
- def performance_counter  
- { 
-   arg method = 'basic timer'; 
-   arg repetitions = 10000;
+ def performance_params { 
+  param UF[] = range(1,32);
+  param PAR[] = [True, False];
  } 
- def performance_params 
- { 
-   param UF[] = range(1,20);
+ def input_params { 
+  param N = [10,100,1000,10**4,10**5,10**6,10**7];
  } 
- def input_params 
- { 
-   param SIZE = @THESIZE@;
-   decl int n = SIZE;
-   decl double x1[n] = random;   
-   decl double x2[n] = random;   
-   decl double x3[n] = random;   
-   decl double y[n] = 0;
+ def input_vars {
+  decl int n = N;
+  decl double y[N] = 0;
+  decl double x1[N] = random;   
+  decl double x2[N] = random;   
+  decl double x3[N] = random;   
  }
- def search
- {
+ def search {
   arg algorithm = 'Exhaustive';
-   arg time_limit = 20;
-   arg run_command = 'cqsub -n 64 -t 10 -q short ';
-   arg num_processes = 64;
  }
@@
 \end{tabular}
-\caption{Orio example: Annotated C source code (left) and a sample tuning specification for the Blue Gene/P (right).}
+\vspace{-.1in}
+\caption{Orio example: Annotated C source code (left) and tuning specifications for Blue Gene/P (right).}
 \label{fig:orio-example}
+\vspace{-.2in}
 \end{figure}
 
-On the right-hand side of Figure~\ref{fig:orio-example}, we show a separate tuning specification example, which contains the information required for creating complete executable tests and running them, including variable types, initializations (e.g., random), the search heuristic (in this case, exhaustive search is indicated, other options are random, simplex, and simulated annealing), and execution environment details.
-When parallel resources are available, Orio executes multiple variants in the same parallel job (the granularity is indicated by the \texttt{num\_processes} field in the search portion of the tuning specification.
-At present the tuning specification must be created manually.
-When Orio is used in conjunction with compiler tools, such as the MATLAB compiler described in this paper, it should eventually be possible to generate most of this tuning specification automatically.
+Figure~\ref{fig:orio-example} shows an annotation example used by Orio
+to empirically optimize VADD operation on Blue Gene/P. The annotations
+contain performance hints that instruct Orio to perform memory
+alignment optimization, loop unrolling, and multicore parallelization
+(using OpenMP). In addition to these simple optimizations, Orio
+supports other transformations such as loop blocking, loop
+permutation, scalar replacement, array copy optimization, and other
+architecture-dependent optimizations (e.g., SIMDization for Intel and
+Blue Gene/P architectures). The right-hand side of
+Figure~\ref{fig:orio-example} shows separate tuning specifications
+used for building and running executable tests, including performance
+parameter values, execution environment details, input variable
+information, and the search algorithm. Orio also supports parallel
+search when parallel resources are available. In this example, the
+parallel Orio driver simultaneously executes 64 code variants in the
+same parallel job. The commands used by Orio to submit a parallel job
+and to query its status are also specified in the tuning
+specifications. At present, users must create the tuning
+specifications manually. When Orio is used in conjunction with
+compiler tools, such as the MATLAB compiler described in this paper,
+it should eventually be possible to automatically generate the tuning
+specifications.
 
-\subsection{Search Space Exploration and Evaluation}
-\label{sec:search}
-
-% BN: moved to its own section since multiple tools deal with the search space problem.
-The set of all possible combinations of implementation and optimization choices is called the search space~\cite{Kisuki:2000uq,Triantafyllis:2003uq,Cooper:2005kx}.
-Although the dataflow graphs are quite small in our setting (usually less than 100 nodes), it is still important to prune the search space to avoid an exponential compilation time.
-Our prototype MATLAB compiler already performs some pruning: the optimization rewrite rules only apply under certain circumstances.
-%
-% BN: commented out below since it's more future work
-%However, as mentioned above we are sometimes overly aggressive in applying optimizations.
-%To solve this problem we plan to use a combination of analytic methods and empirical methods~\cite{Chen:2007vn}.
-%
-Orio provides a number of different search heuristics to narrow the search space for near-optimal performance, including random, simplex, and simulated annealing searches.
-The search algorithm is one of the user-defined values in the Orio tuning specification.
+\vspace{-0.1in}
 
@@
 
-In this section we present performance results for a number of composed BLAS operations.
-The experiment described in Section~\ref{sec:vadd} were performed on Blue Gene/P, while the remaining experiments described in this section were all performed on an Intel Xeon 2.80 GHz Intel(R) E5462 dual quad-core workstation with a 1600Mhz front-side bus and 16GB DDR2 FBDIMM RAM, running Ubuntu 8.04.
+We evaluated our approach by running experiments on an Intel Xeon
+workstation and the Blue Gene/P at Argonne. The Intel machine has dual
+quad-core E5462 Xeon processors (8 cores total) running at 2.8 GHz
+(1600 MHz FSB) with 2 GB RAM, running Ubuntu 8.04. Intel C compiler
+(v10.1) was used with \texttt{-O3} option (and
+\texttt{-parallel}/\texttt{-openmp} for automatic/manual
+parallelization). Each node of the Blue Gene/P has four 850 MHz
+PowerPC 450 processors with a dual floating-point unit and 2 GB total
+memory per node, running a proprietary operating system. On Blue
+Gene/P, we used IBM XLC compiler (v9.0), with \texttt{-O3 -qstrict
+-qarch=450d -qtune=450 -qhot} options (and
+\texttt{-qsmp=auto}/\texttt{-qsmp=noauto} for automatic/manual
+parallelization).
 
-% Include peak numbers?
+Table~\ref{tbl:blas-ops} lists the composed BLAS operations used in
+our experiments, along with their input and output variables. Vectors
+are typeset in lowercase with an overhead arrow. Scalars and matrices
+are represented as lowercase and uppercase letters, respectively. A
+regular uppercase denotes a row matrix, whereas a bold uppercase
+symbolizes a column matrix. The extended MATLAB code that corresponds
+to each operation can be seen in the last column of
+Table~\ref{tbl:blas-ops}.
 
-For the Xeon experiments, we used the Intel icc compiler (version 10.1) with the \texttt{-O3 -parallel} optimization flags.
-We generated and tested the following different implementations of each operation:
-\begin{itemizer}
-\item C from MATLAB: C code generated by the MATLAB compiler described in Section~\ref{sec:matlab}.
-\item BLAS: A version implemented using multiple BLAS calls and linked with the default BLAS library available on Ubutnu 8.04.
-\item Intel MKL: The same code as the BLAS version, but linked to the Intel MKL libraries.
-\item ATLAS: The same code as the BLAS version, but linked to the ATLAS libraries tuned for this machine.
-\item Orio (Seq.): The C code generated by the MATLAB compiler annotated with sequential-only optimization directives.
-\item Orio (Par.) : The C code generated by the MATLAB compiler annotated with both sequential and parallel optimization directives (note that the code is the same as for Orio (Seq.), the only difference is the value of a parallelization parameter in the separate tuning specification).
-This version was compiled with the \texttt{-O3 -openmp} options.
-\end{itemizer}
-
-\subsection{VADD}
-\label{sec:vadd}
-
-The VADD operation computes the sum of three arrays, storing the result in a fourth: $x = w + y + z$.
-We optimized the performance for the VADD operation on a Blue Gene/P
-%at Argonne National Laboratory
-using the IBM XLC compiler version 9.0.
-We measured the performance for two scenarios: using one core per processor and using all 4 cores on each processor.
-In the single-core case, the codes were compiled with the following options: \texttt{-O3 -qstrict -qarch=450d -qtune=450 -qhot -qsmp=noauto}.
-For the multicore case, the only difference in the compilation was that for the non-Orio versions, the \texttt{-qsmp=auto} option was used.
-The parallel Orio version differs from the sequential in that it contains parallelization directives, which result in the placement of OpenMP pragmas in the generated code.
+\begin{table}[htb] 
+\vspace{-.1in} 
+\caption{Composed BLAS operations used in our experiments}
+\label{tbl:blas-ops}
+\small
+\centering 
+\vspace{-.1in} 
+\begin{tabular}{|c|c|c|c|} 
+\hline 
+Name & Input & Output & C code from MATLAB compiler \\
+\hline 
+\hline 
+VADD & $\overrightarrow{w}$,$\overrightarrow{y}$,$\overrightarrow{z}$ & $\overrightarrow{x}$ &
+$\overrightarrow{x} = \overrightarrow{w} + \overrightarrow{y} + \overrightarrow{z}$ \\ 
+\hline 
+ATAX & $A$,$\overrightarrow{x}$ & $\overrightarrow{y}$ & 
+$\overrightarrow{y} = A' * (A * \overrightarrow{x})$ \\ 
+\hline 
+BiCG Kernel & ${\bf A}$,$\overrightarrow{p}$,$\overrightarrow{r}$ & $\overrightarrow{q}$,$\overrightarrow{s}$ & $\overrightarrow{q} = {\bf A} * \overrightarrow{p}$ \\
+& & & $\overrightarrow{s} = {\bf A}' * \overrightarrow{r}$\\
+\hline 
+ & ${\bf A}$,$a$,$b$, & ${\bf B}$, & ${\bf B} = {\bf A} + \overrightarrow{u_1} * \overrightarrow{v_1}' + \overrightarrow{u_2} * \overrightarrow{v_2}'$\\
+GEMVER & $\overrightarrow{u_1}$,$\overrightarrow{u_2}$,$\overrightarrow{v_1}$,$\overrightarrow{v_2}$, & $\overrightarrow{x}$,$\overrightarrow{w}$ & $\overrightarrow{x} = b * ({\bf B}' * \overrightarrow{y}) + \overrightarrow{z}$\\
+& $\overrightarrow{y}$,$\overrightarrow{z}$ & & $\overrightarrow{w} = a * ({\bf B} * \overrightarrow{x})$\\
+\hline 
+\end{tabular} 
+\vspace{-.2in} 
+\end{table}
 
 \begin{figure}[htp]
+{\small
 \centering
 \begin{tabular}{cc}
-\begin{minipage}[b]{.25\textwidth}
-\footnotesize
-\begin{verbatim}
-VADD
-in
-  w : vector,
-  y : vector,
-  z : vector
-out
-  x : vector
-{
-  x = w + y + z
-}
 
-\end{verbatim}
+\begin{minipage}[b]{.515\textwidth}
+\includegraphics[width=\textwidth]{figures/vadd_bgp.eps}
 \end{minipage}
 &
-\begin{minipage}[b]{.6\textwidth}
-\includegraphics[width=\textwidth]{figures/vadd_bgp.png}
+\begin{minipage}[b]{.515\textwidth}
+\includegraphics[width=\textwidth]{figures/atax.eps}
 \end{minipage}\\
+
+(a) VADD & (b) ATAX \\
+
+\begin{minipage}[b]{.515\textwidth}
+\includegraphics[width=\textwidth]{figures/gemver.eps}
+\end{minipage}
+&
+\begin{minipage}[b]{.515\textwidth}
+\includegraphics[width=\textwidth]{figures/bicgkernel.eps}
+\end{minipage}\\
+
+(c) GEMVER & (d) BiCG Kernel \\
+
 \end{tabular}
-\caption{Annotated MATLAB code (left) and MFLOP/s (right) for the VADD operation using different code versions.}
-\label{fig:vadd}
+}
+\caption{Performance numbers for different implementations of various composed of BLAS operations}
+\label{fig:results}
 \end{figure}
 
-The "Base" label designates the simple loop C implementation generated by the Matlab compiler, which was simply compiled with the options mentioned earlier and executed.
-We tested the performance of the IBM ESSL library and Goto BLAS for the two calls to \texttt{daxpy} used in the BLAS version of the code.
-Finally, we annotated the simple C loop version with memory alignment and loop unrolling directives, and then used Orio to select the best tuned version for the single-core and multicore scenarios.
-We observe that even for a very simple operation, such as the vector addition above, the compiler alone is unable to obtain the same level of performance as the empirically tuned versions.
-Furthermore, as expected, the BLAS performance was worse than that of the single loop implementation, however, selecting the correct compiler options was crucial for achieving good performance of the base version.
+The performance results of tuning the VADD operation are given in
+Figure~\ref{fig:results}(a). The ``Base'' label designates the simple
+loop C implementation generated by the MATLAB compiler, which was
+simply compiled with the options mentioned earlier and executed. We
+tested the performance of the IBM ESSL library and Goto BLAS for the
+two calls to \texttt{daxpy} used in the BLAS version of the code.
+Finally, we tuned the simple C loop version using Orio, with the
+performance annotations previously shown in
+Figure~\ref{fig:orio-example}. In this experiment, we measured the
+performance for both the sequential and parallel scenarios. We observe
+that even for a very simple operation, such as the vector addition
+above, the compiler alone is unable to obtain the same level of
+performance as the empirically tuned versions. Furthermore, as
+expected, the BLAS performance was worse than that of the single loop
+implementation, however, selecting the correct compiler options was
+crucial for achieving good performance of the base version.
 
-\subsection{ATAX}
+The experiments of the remaining operations were performed on the
+multicore Intel Xeon. Included in these experiments are performance
+numbers for six code variants: the C code generated by the MATLAB
+compiler (''C from MATLAB''), three BLAS-based implementations that
+use Intel MKL, ATLAS, and the default BLAS library on Ubuntu 8.04, and
+the sequential and parallel code variants tuned by Orio (``Orio
+(Seq.)'' and ``Orio (Par.)'', respectively).
 
-The extended Matlab code for the ATAX operation, $y = A' * (A * x)$, and the corresponding performance results are shown in Figure~\ref{fig:atax}.
-The Orio-tuned version that incorporates PLuTo-generated loop fusion optimizations and Orio parallelization directives achieves the best performance for most problem sizes, outperforming the Intel MKL version by a factor of 2 to 5.7 and the compiler-optimized C version by a factor of 4 to 7.
-The optimizations performed by both Orio versions include scalar replacement, vectorization, and unroll and jam.
+The performance results of ATAX are shown in
+Figure~\ref{fig:results}(b). The Orio-tuned version that incorporates
+PLuTo-generated loop fusion optimizations and Orio parallelization
+directives achieves the best performance for most problem sizes,
+outperforming the Intel MKL version by a factor of 2 to 5.7 and the
+compiler-optimized C version by a factor of 4 to 7. The optimizations
+performed by both Orio versions include scalar replacement,
+vectorization, and loop unroll/jam.
 
-\begin{figure}[htp]
-\centering
-\begin{tabular}{cc}
-\begin{minipage}[b]{.3\textwidth}
-\footnotesize
-\begin{verbatim}
-ATAX
-in
-  A : row matrix,
-  x : vector
-out
-  y : vector
-{
-  y = A' * (A * x)
-}
+Figure~\ref{fig:results}(c) exhibits the performance results of the
+GEMVER operation. Here we used the same PLuTo and Orio optimizations
+as for the ATAX example and, similarly, the parallel Orio version
+achieves the best performance, although in this case the sequential
+Orio version performs almost the same, suggesting that the compiler was
+not able to parallelize the code very effectively. For this operation,
+there are substantial performance differences between different BLAS
+versions, with the Intel MKL version achieving performance close to
+that of the simple compiler code.
 
-\end{verbatim}
-\end{minipage}
-&
-\begin{minipage}[b]{.6\textwidth}
-\includegraphics[width=\textwidth]{figures/atax.png}
-\end{minipage}\\
-\end{tabular}
-\caption{Annotated MATLAB code (left) and wall-clock time in seconds (right) for the ATAX operation using different code versions.}
-\label{fig:atax}
-\end{figure}
-
-\subsection{GEMVER}
-
-The extended MATLAB code for the GEMVER operation and the corresponding performance results are shown in Figure~\ref{fig:gemver}.
-Here we used the same PLuTo and Orio optimizations as for the ATAX example and similarly the parallel Orio version achieves the best performance, although in this case the sequential Orio version performs almost as well suggesting that the compiler was not able to parallelize the code very effectively.
-For this operation, there are substantial performance differences between different BLAS versions, with the Intel MKL version achieving performance close to that of the simple compiler code.
-\begin{figure}[htp]
-\centering
-\begin{tabular}{cc}
-\begin{minipage}[b]{.3\textwidth}
-\footnotesize
-\begin{verbatim}
-GEMVER
-in
-  A : column matrix,
-  u1 : vector, u2 : vector,
-  v1 : vector, v2 : vector,
-  a : scalar, b : scalar,
-  y : vector, z : vector
-out
-  B : column matrix,
-  x : vector, w : vector
-{
-  B = A + u1 * v1'
-           + u2 * v2'
-  x = b * (B' * y) + z
-  w = a * (B * x)
-}
-\end{verbatim}
-\end{minipage}
-&
-\begin{minipage}[b]{.6\textwidth}
-\includegraphics[width=\textwidth]{figures/gemver.png}
-\end{minipage}\\
-\end{tabular}
-\caption{Annotated MATLAB code (left) and wall-clock time in seconds (right) for the GEMVER operation using different code versions.}
-\label{fig:gemver}
-\end{figure}
-
-% Data (Dual quad-core Intel Xeon E5462, 16GB RAM)
-%--- Sequential: seconds (dynamic arrays) ---
-%cmatlab= [0.038, 0.157, 0.352, 0.626, 0.976, 1.406, 1.909, 2.494, 3.143, 3.886]
-%orio= [0.029, 0.118, 0.263, 0.480, 0.759, 1.075, 1.447, 1.899, 2.338, 2.978]
-%blas= [0.064, 0.265, 0.598, 1.091, 1.711, 2.456, 3.340, 4.345, 5.480, 6.840]
-%mkl=  [0.073, 0.147, 0.331, 0.591, 0.923, 1.326, 1.803, 2.350, 2.975, 3.676]
-%atlas=[0.039, 0.163, 0.401, 0.796, 1.225, 1.916, 2.582, 3.381, 4.210, 4.756]
-%
-%--- Parallel: seconds (dynamic arrays) --- N=10,000, cores=1-8
-%cmatlab= [0.976, 0.976, 0.976, 0.976, 0.976, 0.976, 0.975, 0.976]
-%orio= [0.754, 0.738, 0.666, 0.649, 0.664, 0.664, 0.647, 0.668]
-
-\subsection{GESUMMV}
-
-The extended Matlab code for the GESUMMV operation, $y = a * (A * x) + b * (B * x)$,
-and the corresponding performance results are shown in Figure~\ref{fig:gesummv}.
-Unlike all the other operations we tested, the best performing version is the ATLAS one, which substantially outperforms all the alternatives.
-The next best performance is exhibited by the parallel Orio version.
-This result motivates the development of transformations that enable the generation of code with calls to existing optimizing libraries in addition to the low-level loop transformations targeted by the current tools (PLuTo and Orio).
-
-\begin{figure}[htp]
-\centering
-\begin{tabular}{cc}
-\begin{minipage}[b]{.3\textwidth}
-\footnotesize
-\begin{verbatim}
-GESUMMV
-in
-  A : row matrix,
-  B : row matrix,
-  x : vector,
-  a : scalar,
-  b : scalar
-out
-  y : vector
-{
-  y = a * (A * x)
-       + b * (B * x)
-}
-
-\end{verbatim}
-\end{minipage}
-&
-\begin{minipage}[b]{.6\textwidth}
-\includegraphics[width=\textwidth]{figures/gesummv.png}
-\end{minipage}\\
-\end{tabular}
-\caption{Annotated MATLAB code (left) and wall-clock time in seconds (right) for the GESUMMV operation using different code versions.}
-\label{fig:gesummv}
-\end{figure}
-
-% BN: ATLAS's dgemv is annihilating everything else, even though it's called twice... wow.
-% I did check the result and it was correct, i'm still somewhat baffled
-% The Time and MFLOPS data for the above are (with icc):
-% (Dual quad-core Intel Xeon E5462, 16GB RAM)
-% ops = (4*N*N+3*N)
-% --- Parallel: seconds (dynamic arrays) ---
-%matlabc= [0.019, 0.075, 0.170, 0.301, 0.471, 0.678, 0.923, 1.205, 1.529, 1.887]
-%blas= [0.00006, 0.00018, 0.00039, 0.00060, 0.00090, 0.00126, 0.00168, 0.00216, 0.00270, 0.00330]
-%orio= [0.013, 0.035, 0.076, 0.133, 0.210, 0.300, 0.410, 0.533, 0.676, 0.831]
-%--- Parallel: Mflops/sec (dynamic arrays) ---
-%matlabc= [846.565, 848.425, 849.028, 849.503, 849.690, 849.968, 849.605, 849.946, 847.893, 848.023]
-%blas= [271288.136, 347891.304, 367392.857, 421785.832, 480228.091, 497440.415, 508787.800, 516153.226, 524930.741, 530172.300]
-%orio= [1231.989, 1846.163, 1885.595, 1918.746, 1907.976, 1920.818, 1910.867, 1920.502, 1915.970, 1924.623]
-%
-% and gcc behaves similarly, so it's not icc being smart and detecting 2 calls to dgemv
-% time: blas= [0.000, 0.000, 0.000, 0.001, 0.001, 0.001, 0.002, 0.002, 0.003, 0.003]
-% mflops: blas= [254063.492, 346010.811, 393491.803, 421785.832, 439593.407, 453214.792, 463108.092, 470394.120, 462381.020, 469088.244]
-
-\subsection{BiCG Kernel}
-
-% No pluto
-The extended Matlab code for the BiCG Kernel operation,$q = A * p$, $s = A' * r$,
-and the corresponding performance results are shown in Figure~\ref{fig:gesummv}.
-For this operation, the PLuTo analysis did not result in performance improvement, thus we are showing the results obtained only through Orio transformations, which included vectorization, scalar replacement, and unroll and jam.
-Again the best performance was exhibited by the parallel Orio version, while all the BLAS versions performed worse than the compiler-optimized C loop version.
-
-\begin{figure}[htp]
-\centering
-\begin{tabular}{cc}
-\begin{minipage}[b]{.3\textwidth}
-\footnotesize
-\begin{verbatim}
-BICG
-in
-  A : column matrix,
-  p : vector, r : vector
-out
-  q : vector, s : vector
-{
-  q = A * p
-  s = A' * r
-}
-
-\end{verbatim}
-\end{minipage}
-&
-\begin{minipage}[b]{.6\textwidth}
-\includegraphics[width=\textwidth]{figures/bicgkernel.png}
-\end{minipage}\\
-\end{tabular}
-\caption{Annotated MATLAB code (left) and wall-clock time in seconds (right) for the BiCG kernel operation using different code versions.}
-\label{fig:bicgkernel}
-\end{figure}
+The experimental results of BiCG Kernel operation are provided in
+Figure~\ref{fig:results}(d). For this operation, the PLuTo analysis
+did not result in performance improvement, thus we are showing the
+results obtained only through Orio transformations, which included
+vectorization, scalar replacement, and loop unroll/jam. Again the best
+performance was demonstrated by the parallel Orio version, while all
+the BLAS versions performed worse than the compiler-optimized C loop
+version.
 

### 2008-12-11 doc/papers/iccs09/paper.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@384 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@
 \centering
-\includegraphics[width=0.7\textwidth]{figures/process.eps}
+\includegraphics[width=0.7\textwidth]{figures/process.png}
 \caption{Overview of the code generation and tuning process, optional steps are indicated with dashed arrows.}
@@
 \centering
-\includegraphics[width=.7\textwidth]{figures/compile.eps}
+\includegraphics[width=.7\textwidth]{figures/compile.png}
 
@@
 \centering
-\includegraphics[width=.7\textwidth]{figures/gemver-dataflow.eps}
+\includegraphics[width=.7\textwidth]{figures/gemver-dataflow.png}
 
@@
   \centering
-  \includegraphics[width=\textwidth]{figures/refine-mv-dot.eps}
+  \includegraphics[width=\textwidth]{figures/refine-mv-dot.png}
 
@@
 \begin{center}
-\includegraphics[width=.65\textwidth]{figures/orio.eps}
+\includegraphics[width=.65\textwidth]{figures/orio.png}
 \end{center}
@@
 \texttt{-parallel}/\texttt{-openmp} for automatic/manual
-parallelization). Each node of the Blue Gene/P has four 850 MHz
-PowerPC 450 processors with a dual floating-point unit and 2 GB total
-memory per node, running a proprietary operating system. On Blue
-Gene/P, we used IBM XLC compiler (v9.0), with \texttt{-O3 -qstrict
--qarch=450d -qtune=450 -qhot} options (and
+parallelization, respectively). Each node of the Blue Gene/P has four
+850 MHz PowerPC 450 processors with a dual floating-point unit and 2
+GB total memory per node, running a proprietary operating system. On
+Blue Gene/P, we used IBM XLC compiler (v9.0), with \texttt{-O3
+-qstrict -qarch=450d -qtune=450 -qhot} options (and
 \texttt{-qsmp=auto}/\texttt{-qsmp=noauto} for automatic/manual
-parallelization).
+parallelization, respectively).
 
@@
 \hline 
-BiCG Kernel & ${\bf A}$,$\overrightarrow{p}$,$\overrightarrow{r}$ & $\overrightarrow{q}$,$\overrightarrow{s}$ & $\overrightarrow{q} = {\bf A} * \overrightarrow{p}$ \\
-& & & $\overrightarrow{s} = {\bf A}' * \overrightarrow{r}$\\
-\hline 
  & ${\bf A}$,$a$,$b$, & ${\bf B}$, & ${\bf B} = {\bf A} + \overrightarrow{u_1} * \overrightarrow{v_1}' + \overrightarrow{u_2} * \overrightarrow{v_2}'$\\
@@
 & $\overrightarrow{y}$,$\overrightarrow{z}$ & & $\overrightarrow{w} = a * ({\bf B} * \overrightarrow{x})$\\
+\hline 
+BiCG Kernel & ${\bf A}$,$\overrightarrow{p}$,$\overrightarrow{r}$ & $\overrightarrow{q}$,$\overrightarrow{s}$ & $\overrightarrow{q} = {\bf A} * \overrightarrow{p}$ \\
+& & & $\overrightarrow{s} = {\bf A}' * \overrightarrow{r}$\\
 \hline 
@@
 \begin{minipage}[b]{.515\textwidth}
-\includegraphics[width=\textwidth]{figures/vadd_bgp.eps}
+\includegraphics[width=\textwidth]{figures/vadd_bgp.png}
 \end{minipage}
@@
 \begin{minipage}[b]{.515\textwidth}
-\includegraphics[width=\textwidth]{figures/atax.eps}
+\includegraphics[width=\textwidth]{figures/atax.png}
 \end{minipage}\\
@@
 \begin{minipage}[b]{.515\textwidth}
-\includegraphics[width=\textwidth]{figures/gemver.eps}
+\includegraphics[width=\textwidth]{figures/gemver.png}
 \end{minipage}
@@
 \begin{minipage}[b]{.515\textwidth}
-\includegraphics[width=\textwidth]{figures/bicgkernel.eps}
+\includegraphics[width=\textwidth]{figures/bicgkernel.png}
 \end{minipage}\\

### 2008-12-12 doc/papers/iccs09/abstract.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@385 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

--- 2008-12-11 doc/papers/iccs09/paper.tex
@@
-% CGO 09: http://www.cgo.org/
-% svn co https://svn.mcs.anl.gov/repos/performance/orio/doc/papers/cgo08
-% Max 22 pages, double-spaced, 11pt or larger font
-% Abstract 300 words: Sep. 12, 2008
-% Full paper: Sep. 19, 2008
-
-\documentclass[runningheads]{llncs}
-
-\usepackage{amssymb}
-\setcounter{tocdepth}{3}
-\usepackage{graphicx}
-\usepackage{url}
-\usepackage{html}
-\usepackage{setspace}
-\usepackage{graphicx}
-\usepackage{subfigure}
-\usepackage{cite}
-\usepackage{amsmath}
-
-\urldef{\mailsa}\path|{alfred.hofmann,ursula.barth,ingrid.beyer,natalie.brecht,|
-\urldef{\mailsb}\path|christine.guenther,frank.holzwarth,piamaria.karbach,|
-\urldef{\mailsc}\path|anna.kramer,erika.siebert-cole,lncs}@springer.com|
-\newcommand{\keywords}[1]{\par\addvspace\baselineskip
-\noindent\keywordname\enspace\ignorespaces#1}
-
-\newenvironment{itemizer}{\begin{itemize}\setlength{\parsep}{0cm}\setlength{\itemsep}{-.3em}}{\end{itemize}}
-
-% formatting cheats
-\renewcommand{\baselinestretch}{0.92}  
-\renewcommand\floatpagefraction{.9} 
-\renewcommand\topfraction{.9} 
-\renewcommand\bottomfraction{.9} 
-\renewcommand\textfraction{.1} 
-\setcounter{totalnumber}{50} 
-\setcounter{topnumber}{50} 
-\setcounter{bottomnumber}{50} 
-\setlength{\textwidth}{12.5cm} 
-\setlength{\textheight}{19.5cm}
-
-\begin{document}
-
-\mainmatter  % start of an individual contribution
-
-% first the title is needed
-\title{Generating Empirically Optimized Numerical Software from MATLAB Prototypes}
-
-        % a short form should be given in case it is too long for the running head
-        \titlerunning{Empirically Optimized Numerical Software from MATLAB}
-
-        % the name(s) of the author(s) follow(s) next
-        %
-        % NB: Chinese authors should write their first names(s) in front of
-        % their surnames. This ensures that the names appear correctly in
-        % the running heads and the author index.
-        %
-\author{Elizabeth Jessup\inst{1} 
-  \and Boyana Norris\inst{2}
-  \and Albert Hartono\inst{3}
-  \and Jeremy Siek\inst{1}}
-
-\authorrunning{Jessup, Norris, Hartono, Siek}
-
-\institute{University of Colorado at Boulder \\
-  \email{\{elisabeth.jessup,jeremy.siek\}@colorado.edu}
-  \and Argonne National Laboratory \\
-  \email{\{norris\}@mcs.anl.gov}
-  \and Ohio State University \\
-  \email{\{hartonoa\}@cse.ohio-state.edu}}
-
-%\toctitle{Lecture Notes in Computer Science}
-\tocauthor{Elizabeth Jessup, Boyana Norris, Albert Hartono, and Jeremy Siek}
-\maketitle
-
 \begin{abstract}
 % between 75 and 150 words
-The growing demand for higher levels of detail and accuracy in results means that the size and complexity of scientific computations is increasing at least as fast as the improvements in processor technology.
-Programming scientific applications is hard, and optimizing them for high performance is even harder.
-The development of optimized codes requires extensive knowledge, not only of the costs of floating-point arithmetic but also of  memory access issues and compiler optimizations.
-Experiments show that the complexity of this hardware-software system means that performance is difficult to predict fully.
-Therefore, computational scientists are often forced to choose between investing too much time in tuning code or  accepting performance that is significantly lower than the best achievable performance on a given architecture.
-In this paper, we describe the first steps toward a fully automated system for the optimization of the matrix algebra kernels that are a foundational part of many scientific applications.
-To generate highly optimized code from a high-level MATLAB prototype, we define a three-step approach.
-To begin, we have developed a compiler that converts a MATLAB script into simple C code.
-We then use the polyhedral optimization system PLuTo to optimize that code for coarse-grained parallelism and locality simultaneously.
-Finally, we annotate the resulting code with performance tuning directives and use the empirical performance tuning system Orio to generate many tuned versions of the same operation using different optimization techniques, such as loop unrolling and memory alignment.
-Orio performs an automated empirical search to select the best among the multiple optimized code variants.
-We discuss performance results showing that the code generated by using our system significantly outperforms not only the original simple C code but also code based on source BLAS, ATLAS-optimized BLAS, and Intel MKL routines.
+%The growing demand for higher levels of detail and accuracy in results means
+%that the size and complexity of scientific computations is increasing at
+%least as fast as the improvements in processor technology. 
+%Programming
+%scientific applications is hard, and optimizing them for high performance is
+%even harder.  
+The development of optimized codes is time-consuming and requires extensive
+architecture, compiler, and language expertise; therefore, computational
+scientists are often forced to choose between investing too much time in
+tuning code or accepting lower performance.
+%
+In this paper, we describe the first steps toward a fully automated system
+for the optimization of the matrix algebra kernels that are a foundational
+part of many scientific applications.  To generate highly optimized code from
+a high-level MATLAB prototype, we define a three-step approach.  To begin, we
+have developed a compiler that converts a MATLAB script into simple C code.
+We then use the polyhedral optimization system PLuTo to optimize that code
+for coarse-grained parallelism and locality simultaneously. Finally, we
+annotate the resulting code with performance tuning directives and use the
+empirical performance tuning system Orio to generate many tuned versions of
+the same operation using different optimization techniques, such as loop
+unrolling and memory alignment. Orio performs an automated empirical search
+to select the best among the multiple optimized code variants. We discuss
+performance results on two architectures.
+%showing that the code generated by using our system
+%significantly outperforms not only the original simple C code but also code
+%based on source BLAS, ATLAS-optimized BLAS, and Intel MKL routines.
 \keywords{Code generation, empirical performance tuning}
 \end{abstract}
-
-\section{Introduction}
-\label{sec:intro}
-
-The development of high-performance numerical codes is challenging because performance is determined by complex interactions between the algorithm, data structure, programming language, compiler, and computer architecture.
-Scientists seeking high performance are thus required to master advanced concepts in computer science and carry out intricate programming tasks in addition to managing the scientific content of their work.
-They must either invest substantial time in tuning their software or accept low performance.
-In either case, the productivity of the scientist degrades.
-
-Historically, the research community has pursued two separate paths toward the goal of making software run at near-peak levels.
-The first of these builds on research into compilers and their associated technologies.
-The holy grail of compilation research is to take an arbitrary code as input and produce optimal code as output for a given language and hardware platform.
-Despite the immense effort that has been poured into that approach, its success has been limited both by practical constraints (users will not tolerate compile times that extend into several days) and by the amount of detailed information the compiler can obtain about the software to be compiled and the hardware on which it is to execute~\cite{Ken99,Wol89,Wol96}.
-The second path is to identify kernel routines that constitute the dominant costs of a wide variety of applications.
-When such kernels can be identified and agreed upon by the members of the community, programmers with the required level of technical knowledge can concentrate on producing these optimized kernel libraries for architectures of interest.
-A prime example is the high-performance BLAS libraries produced by a combination of hardware vendors, independent software vendors, and researchers.
-Developers who write their codes calling these routines can achieve high performance across all supported architectures.
-The efforts in high-performance library development aptly demonstrate, however, that the research into compiler technologies has not been fully successful.
-If compilers could take the relatively simple code even for basic tasks such as a dense matrix-matrix product and produce optimal code, this second approach would be unnecessary.
-Like compiler optimization, the kernel library-oriented approach has significant limitations.
-Support for a variety of architectures is the most serious problem because the optimizations necessary to achieve near-peak performance are by nature not portable.
-Lack of portability is particularly problematic because processor designs are constantly changing,  and constant updates are nearly impossible to maintain by hand.
-It is a sad fact of the computing industry that by the time highly-optimized code is available for a given architecture, that architecture is generally well on its way to obsolescence.
-
-We are working on a combination of the two approaches designed to overcome some of their shortcomings.
-This paper describes our multipronged approach toward the development of software infrastructure for \emph{generating} automatically tuned libraries for matrix algebra computations.
-We briefly overview relevant prior and current research efforts in Section~\ref{sec:background}.
-In Section~\ref{sec:approach} we describe our MATLAB to C compiler and the empirical performance tuning system Orio and its use in conjunction with the PLuTo tool suite to generate and empirically evaluate many tuned versions of the C code generated by the MATLAB compiler.
-In Section~\ref{sec:experiments}, we provide performance results showing that the code generated by using our system significantly outperforms not only the original simple C code but also code based on source BLAS, ATLAS-optimized BLAS, and Intel MKL routines.
-%Text in intro repeats abstract some, but that might not be a big deal for a conf paper.
-
-\section{Background}
-\label{sec:background}
-
-Existing optimizing MATLAB compilers, such as the MaJIC MATLAB compiler, include limited local optimizations for matrix expressions, but do not perform optimizations such as loop fusion across multiple operations as we do with the tools described in this paper.
-The telescoping languages project~\cite{telescopingurl,teleoverview,Ken99} uses techniques such as strength reduction, vectorization, and procedure specialization to optimize MATLAB scripts, but do not generate reusable optimized linear algebra routines as described in this paper.
-
-The most common approach  to tuning numerical codes is for an expert to transform the source manually, unrolling loops, blocking for multiple levels of cache, and inserting prefetch instructions.
-The pitfalls of this approach are well-understood~\cite{Goedecker01}.
-It requires a significant amount of time and effort.
-Optimizing code for one particular platform may in fact make it less efficient on other platforms and often makes it complex and hard to understand or maintain.
-An alternative is the use of tuned libraries of key numerical algorithms (e.g., BLAS~\cite{Dongarra:1990fk} and LAPACK~\cite{LAPACK} for dense linear algebra, SPARSPAK~\cite{sparspak} for sparse linear algebra, and BLITZ++~\cite{blitz} to support scientific computing in C++, or GPULib by Tech-X, an open source library for accelerating numerical computations on Graphics Processing Units~\cite{gpulib}, to name just a few).
-Such libraries are difficult to develop, and they generally cannot adapt to execution context: as we note in Section~\ref{sec:approach}, a series of calls to library routines can incur high memory costs.
-
-Specialized code generators circumvent the high costs of manual code generation .
-They include tools for basic dense linear algebra operations (ATLAS~\cite{WN147}, PhiPAC~\cite{bilmes97optimizing,phipacwww}), sparse linear algebra (OSKI~\cite{vuduc05}), FFTs~\cite{FFTW,Spiral}), stencil-based operations~\cite{kamil06}, and tensor computations (TCE~\cite{TCE}) among others.
-It is often impossible to predict the performance of code precisely on modern computer architectures.
-Thus, many of these specialized code generators exploit search strategies to identify the best (or nearly best) code for a particular choice of problem parameters and machine.
-Such autotuning tools are not general but rather each is valuable in one specific problem domain.
-While not all applications can be written in terms of available libraries, the performance achieved with autotuning tools does convincingly demonstrate the advantage of source-to-source transformation over optimizing compilers.
-
-The best performance today is obtained through approaches like that taken by PERI~\cite{PERI} which involves a labor intensive rewrite of portion of the code by a group of performance tools experts in collaboration with application scientists.
-Maintaining performance of the code as the application and hardware evolves means that the tuning process may need to be repeated.
-Our efforts aim to capture this performance tuning expertise in a way that can be integrated into the application.
-We will thus enable performance portability and ensure that the improvements to the application are long-lasting.
-
-A number of source or binary transformation tools for performance-improving optimizations exist.
-LoopTool~\cite{LoopTool}, developed at Rice University, supports annotation-based loop fusion, unroll-and-jam, skewing and tiling.
-A relatively new tool, POET~\cite{POET} also supports a number of loop transformations.
-POET offers a complex template-based syntax for defining transformations in a language-independent manner.
-Other recent research efforts whose goal, at least in part, is to enable optimizations of source code to be augmented with performance-related information include the X language~\cite{XLanguage} (a macro C-like language for annotating C code), the Broadway~\cite{broadway} compiler, and various meta-programming techniques~\cite{veldhuizen95,weise93,kiczales91,chiba95}.
-PLuTo~\cite{Pluto,uday08cc} is a source-to-source transformation tool for optimizing sequences of nested loops.
-PLuTo employs a polyhedral model of nested loops, where the dynamic instance (iteration) of each statement is viewed as an integer point in a well-defined space, called the statement's polyhedron.
-Combined with a characterization of data dependences, this representation allows the construction of mathematically correct complex loop transformations.
-The transformations target both improved cache locality and parallelism.
-
-\section{Our Approach to Optimizing Composed BLAS Operations}
-\label{sec:approach}
-
-%\section{Optimizing Composed Operations}
-%\label{sec:composed}
-
-Codes based on matrix algebra are generally constructed as a sequence of calls to the Basic Linear Algebra Subprograms (BLAS)~\cite{Dongarra:1988uq,Lawson:1979kx,Dongarra:1990fk} and similar sparse matrix libraries~\cite{Saad:fr,George:1981uq}.
-Writing programs in this way promotes readability and maintainability but can be costly in terms of memory efficiency.
-Specifically, the retrieval of a large order matrix at each routine call can have a profound effect on performance even when highly tuned implementations of the BLAS (e.g.~\cite{Bilmes:1997ye,Whaley:1998fk,IntelMath:oq,ESSL:kl,Goto:2006fk}) are used.
-
-The authors and other researchers have observed that it can be much more efficient to call a single specialized routine that performs multiple operations than to make successive calls to separate BLAS routines ~\cite{Ashby:uq,Blackford:2002vn,baker03blgmres,baker03lgmres,Dennis:2005tg,Howell:2008,gropp01,Vuduc:2003kl}.
-Single routines that carry out more than one linear algebra operation are known as \emph{composed} BLAS.
-As an example, consider the pair of matrix-vector products $q  = Ap$, $s  = A^Tr$, where $A$ is a matrix and $p$, $q$, $r$, and $s$ are vectors, that comprise the computational
-bottlenecks of the BiConjugate Gradient Method~\cite{Barrett:1994kx,Saad:2003fk}  and of the GEMVER kernel examined in Section~\ref{sec:matlab}.
-These two operations can be implemented as a pair of calls to the BLAS routine {\tt GEMV} or they can be rewritten as a composed BLAS consisting of the doubly nested loop encompassing both matrix-vector multiplies.
-In the former case, the matrix $A$ is accessed twice, while in the latter it is accessed only once.
-Our preliminary results in prior research and in this paper indicate that this loop fusion leads to a routine that delivers significantly better performance than does a pair of calls to the best optimized BLAS {\tt GEMV} routines for large matrix orders.
-Composed routines are the focus of the work presented here, but much of what we discuss generalizes to a much broader array of computations.
-
-% While large speedups are probable, composition can  also force memory accesses to be consecutive, and it can
-
-%lead to the need to bring more data into cache. These bad
-
-% effects can be remedied by combining composition with other tuning techniques like cache blocking and software pipelining.
-
-\begin{figure}[htbp]
-\centering
-\includegraphics[width=0.7\textwidth]{figures/process.eps}
-\caption{Overview of the code generation and tuning process, optional steps are indicated with dashed arrows.}
-\label{fig:process}
-\end{figure}
-To generate highly optimized code from a MATLAB prototype of the composed BLAS operation, we follow a three-step approach, illustrated in Figure~\ref{fig:process}.
-To begin, we have developed a compiler that converts a MATLAB script into simple C code.
-%~\cite{Siek}. 
-After generating the C code from the high-level MATLAB prototype, we (optionally) use the source-to-source automatic parallelization tool PLuTo\cite{Pluto} %developed at Ohio State University
-to optimize for coarse-grained parallelism and locality simultaneously.
-Using the results of the PLuTo analysis, we insert annotations into the C code, which are then processed by our extensible annotation system Orio to generate many tuned versions of the same operation using different optimization parameters.
-Orio then performs an empirical search to select the best among the multiple optimized code variants.
-
-%Probably need some words about WHY this multipronged approach is a good idea.
-
-In the remainder of this section we describe each of the tools developed by the authors of this paper, namely the MATLAB to C compiler and the Orio empirical tuning tool.
-
-\subsection{A MATLAB Compiler}
-\label{sec:matlab}
-
-In this section we describe the design and operation of the MATLAB to C compiler we have developed.
-Figure~\ref{fig:compiler} gives an overview of the compilation process.
-The MATLAB kernel specification is parsed into a high-level intermediate representation in the form of a dataflow graph.
-This dataflow graph is then iteratively processed until all of the implementation choices have been made.
-The compilation process consists of three phases: analysis, lowering, and optimization, which are together iterated until all of the implementation decisions have been made.
-The graph is then translated into C code.
-
-\begin{figure}[htbp]
-\centering
-\includegraphics[width=.7\textwidth]{figures/compile.eps}
-
-\caption{Overview of the MATLAB compiler.}
-\label{fig:compiler}
-\end{figure}
-
-\subsubsection{The Dataflow Graph}
-
-An example dataflow graph for the GEMVER kernel, defined below, is given in figure~\ref{fig:gemver-dataflow}.
-\begin{align*}
-  A &\gets A + u_1 v_1^T + u_2 v_2^T \\[-0.5ex]
-  x &\gets \beta A^T y + z \\[-0.5ex]
-  w &\gets \alpha A x
-\end{align*}
-Each node represents a parameter of the kernel or an operation.
-The arrows indicate the flow of data.
-At first the graph specifies what operations are to be performed but does not contain any implementation details.
-The symbol $\times$ in the depicted graph stands variously for outer product, matrix-vector multiplication, and scalar-vector multiplication, and it does not yet specify, for example, whether the outer products compute a row or a column at a time of the result matrix.
-
-\begin{figure}[htbp]
-\centering
-\includegraphics[width=.7\textwidth]{figures/gemver-dataflow.eps}
-
-\caption{Dataflow graph for the GEMVER kernel.}
-\label{fig:gemver-dataflow}
-\end{figure}
-
-\subsubsection{Analyze the Dataflow Graph}
-
-At the beginning of the analysis phase, each input and output node of the dataflow graph is labeled with its kind (scalar, vector, matrix, or unknown) and storage format (dense row-major, dense column-major, or unknown) according to the annotation given by the user.
-During analysis, the kind and storage format are computed for the intermediate nodes and algorithms are assigned to operation nodes.
-The choice of algorithm and the determination of kinds of storage formats are interrelated so they must be computed together.
-For example, the $\times$ symbol in the graph in Figure~\ref{fig:gemver-dataflow} having the inputs $u_1$ and $v_1^T$ can be implemented by iterating over rows first, or over columns first.
-The choice depends on how the result is used downstream in the dataflow graph.
-In this case, the result is added to the outer product of $u_2$ and $v_2^T$, so we still could choose either option as long as we make the same choice for both outer products.
-Going one more step downstream, there is an addition with $A$, which was annotated in Figure~\ref{fig:gemver} to be column major.
-At this point it is clear that the outer-products should be computed in column-major order.
-
-We do not hard code this knowledge of basic linear algebra in the compiler algorithm itself, but instead use a data-driven approach and store information regarding how to implement basic linear algebra in a separate database, which we refer to as the \emph{linear algebra database}.
-This separation allows us to add new matrix formats, operations, and basic linear algebra algorithms without changes to the
-compiler algorithm.
-
-The analysis algorithm makes implementation choices using the most-constrained first strategy (also called minimum remaining values)
-from the artificial intelligence literature~\cite{Russell:2003mz}.
-%This may need to be updated depending on how we decide to talk about the empirical search. -Jeremy 8/6/08 10:15 AM 
-The compiler chooses the node with the fewest matching implementations (in the linear algebra database) and assigns a matching implementation to the node.
-If there is more than one match, the prototype compiler picks the first.
-Once we add backtracking we will push each alternative onto the stack to be explored later.
-Once the choice is made, the kind and storage format labels for the operation node itself and for the input nodes are updated with the labels specified in the linear algebra database for the chosen algorithm.
-The algorithm then goes on to find the next node with the fewest matching implementations and repeats.
-In this phase, only the name of the algorithm is assigned to the operation node.
-The details of the implementation are not specified until the refinement phase.
-
-\subsection{Refine the Dataflow Graph}
-\label{sec:refine}
-
-The refinement phase resolves the implementation for each operation node in the graph into a subgraph defining the details of the chosen algorithm.
-Each subgraph is an abstract representation of the loop that implements the given operation.
-In each case, the subgraph has an iteration strategy that specifies how to traverse the elements of the matrix or vector.
-
-Figure~\ref{fig:refine-mv-dot} shows two steps of refinement for a matrix-vector multiplication.
-The first step expands the matrix-vector multiplication according to the mv-dot algorithm: $(Ax)(i) = A(i,:)x$.
-The refinement step replaces the $\times$ node with a subgraph that computes the inner product of a row of the matrix---the node labeled $(i,:)$---with the vector, storing the result in the $i$th element of the result vector---the node labeled $(i)$.
-The subgraph is labeled with $i=1\!:\!m$ indicating that the iteration strategy is to traverse the rows of the dense matrix.
-%
-The second pass of refinement introduces another subgraph to implement the inner products.
-The new subgraph iterates through each dense row, as indicated by the $j=1:n$ annotation.
-The sign $\Sigma$ indicates that the sum of all of the iterations is computed.
-
-\begin{figure}[hbtp]
-  \centering
-  \includegraphics[width=\textwidth]{figures/refine-mv-dot.eps}
-
-  \caption{The refinement of a matrix-vector product to a set of inner products and the refinement of inner products to scalar multiplications and additions.}
-  \label{fig:refine-mv-dot}
-\end{figure}
-
-\subsubsection{Optimize the Dataflow Graph}
-
-In the optimization step, the dataflow graph is optimized by applying conditional graph rewrite rules.
-One example rewrite rule is \emph{Merge Operand Sharing Subgraphs}: if two subgraphs share a common operand but do not depend on one another, and if the iteration strategies of the two subgraphs are compatible, then merge the subgraphs into a single subgraph.
-This rule is responsible for fusing the loops of the two matrix-vector products in the GEMVER kernel.
-Another example rule is \emph{Merge Dependent Subgraphs}: if one subgraph depends on another subgraph, and the iteration strategies of the two subgraphs are compatible, then merge them into a single subgraph.
-This rule, for example, is responsible for fusing the loops of the two outer products in the GEMVER kernel.
-
-The choice of which optimizations to apply to various parts of the graph depends on the characteristics of the kernel, the algorithm, the matrix order and kind, and architectural features.
-The following is an instructive example: $y \gets A^T A (x + z + w)$.
-Assume that the compiler has already merged the two vector additions into one subgraph, call it $S_1$, and the two matrix products into another subgraph, call it $S_2$.
-The question then is whether it is profitable to apply the Merge Dependent Subgraphs rule one more time to merge $S_1$ and $S_2$.
-If the three vectors plus a row of $A$ do not fit in cache, then the merger will not be profitable because it will cause each row of $A$ to fall out of cache before the row gets used a second time.
-On the other hand, if all three vectors plus a row of $A$ do it in cache, then $S_1$ and $S_2$ should be merged.
-In general, we need a way to search through the many different combinations of optimization choices and choose a combination that minimizes memory traffic.
-
-\subsubsection{Translate the Dataflow Graph to C}
-
-A graph that cannot be further refined has been reduced to a collection of subgraphs representing loops over scalar operations.
-The generator outputs a C loop for each subgraph.
-A topological sort is performed to determine the order in which the loops and scalar operations within the loops appear in the output.
-\subsection{Orio}
-\label{sec:orio}
-
-%~\cite{Norris:2007}
-Orio is an empirical tuning tool that takes annotated C code as input,
-generates multiple transformed versions of the annotated code, and
-empirically evaluates the performance of the generated codes,
-ultimately choosing the best-performing version to use in production
-runs.
-
-Figure~\ref{fig:orio} illustrates the tuning process implemented in
-Orio. The input to Orio is C code containing semantic comments that
-describe both the computation, as well as various performance-tuning
-directives. Orio first extracts all annotation code regions by parsing
-the marked-up input code. Each annotated region is then passed to code
-transformator and code generator for potential optimizations. Then the
-transformed C code with various incorporated optimizations
-corresponding to the specified annotations is produced. The
-source-to-source transformation system of Orio generates an optimized
-code version for each distinct combination of performance parameter
-values. Each generated code variant is then executed and its
-performance cost evaluated. After iteratively testing all code
-variants, the best-performing code is selected as the final output of
-Orio. Because the search space of all possible optimized code variants
-can be exponentially large, the search engine implements a number of
-search heuristics (i.e., random, simplex, and simulated annealing) to
-effectively narrow the search for near-optimal performance. The tuning
-specifications, written by users in the form of annotations, contains
-information required for guiding the transformation and tuning
-process.
-
-\begin{figure}[tb]
-\vspace{-.1in}
-\begin{center}
-\includegraphics[width=.65\textwidth]{figures/orio.eps}
-\end{center}
-\vspace{-.25in}
-\caption{Overview of the Orio empirical tuning process.}
-\label{fig:orio}
-\end{figure}
-
-\begin{figure}[htb]
-\centering
-\begin{tabular}{cc}
-\begin{minipage}[b]{.45\textwidth}
-\scriptsize
-\begin{verbatim}
-void vadd(int n, double *y, double *x1,
-          double *x2, double *x3) {
-/*@ begin PerfTuning(
-  import spec vadd_tune_spec;
-) @*/
-
- register int i;
-
-/*@ begin BGP_Align(y[],x1[],
-                    x2[],x3[]) @*/
-/*@ begin Loop(
- transform Unroll(ufactor=UF, 
-                  parallelize=PAR)
- for (i=0; i<=n-1; i++)
-   y[i] = x1[i] + x2[i] + x3[i];
-) @*/
-
- for (i=0; i<=n-1; i++)
-   y[i] = x1[i] + x2[i] + x3[i];
-
-/*@ end @*/
-/*@ end @*/
-/*@ end @*/
-}
-
-\end{verbatim}
-\end{minipage}
-&
-\begin{minipage}[b]{.45\textwidth}
-\scriptsize
-\begin{verbatim}
-spec vadd_tune_spec {
- def build { 
-  arg build_command = 'mpixlc -O3 -qstrict -lm';
-  arg batch_command = 'qsub -n 64 -t 10';
-  arg status_command = 'qstat';
-  arg num_procs = 64;
- }
- def performance_params { 
-  param UF[] = range(1,32);
-  param PAR[] = [True, False];
- } 
- def input_params { 
-  param N = [10,100,1000,10**4,10**5,10**6,10**7];
- } 
- def input_vars {
-  decl int n = N;
-  decl double y[N] = 0;
-  decl double x1[N] = random;   
-  decl double x2[N] = random;   
-  decl double x3[N] = random;   
- }
- def search {
-  arg algorithm = 'Exhaustive';
- }
-}
-\end{verbatim}
-\end{minipage}\\
-\end{tabular}
-\vspace{-.1in}
-\caption{Orio example: Annotated C source code (left) and tuning specifications for Blue Gene/P (right).}
-\label{fig:orio-example}
-\vspace{-.2in}
-\end{figure}
-
-Figure~\ref{fig:orio-example} shows an annotation example used by Orio
-to empirically optimize VADD operation on Blue Gene/P. The annotations
-contain performance hints that instruct Orio to perform memory
-alignment optimization, loop unrolling, and multicore parallelization
-(using OpenMP). In addition to these simple optimizations, Orio
-supports other transformations such as loop blocking, loop
-permutation, scalar replacement, array copy optimization, and other
-architecture-dependent optimizations (e.g., SIMDization for Intel and
-Blue Gene/P architectures). The right-hand side of
-Figure~\ref{fig:orio-example} shows separate tuning specifications
-used for building and running executable tests, including performance
-parameter values, execution environment details, input variable
-information, and the search algorithm. Orio also supports parallel
-search when parallel resources are available. In this example, the
-parallel Orio driver simultaneously executes 64 code variants in the
-same parallel job. The commands used by Orio to submit a parallel job
-and to query its status are also specified in the tuning
-specifications. At present, users must create the tuning
-specifications manually. When Orio is used in conjunction with
-compiler tools, such as the MATLAB compiler described in this paper,
-it should eventually be possible to automatically generate the tuning
-specifications.
-
-\vspace{-0.1in}
-
-\section{Experimental Results}
-\label{sec:experiments}
-
-We evaluated our approach by running experiments on an Intel Xeon
-workstation and the Blue Gene/P at Argonne. The Intel machine has dual
-quad-core E5462 Xeon processors (8 cores total) running at 2.8 GHz
-(1600 MHz FSB) with 2 GB RAM, running Ubuntu 8.04. Intel C compiler
-(v10.1) was used with \texttt{-O3} option (and
-\texttt{-parallel}/\texttt{-openmp} for automatic/manual
-parallelization). Each node of the Blue Gene/P has four 850 MHz
-PowerPC 450 processors with a dual floating-point unit and 2 GB total
-memory per node, running a proprietary operating system. On Blue
-Gene/P, we used IBM XLC compiler (v9.0), with \texttt{-O3 -qstrict
--qarch=450d -qtune=450 -qhot} options (and
-\texttt{-qsmp=auto}/\texttt{-qsmp=noauto} for automatic/manual
-parallelization).
-
-Table~\ref{tbl:blas-ops} lists the composed BLAS operations used in
-our experiments, along with their input and output variables. Vectors
-are typeset in lowercase with an overhead arrow. Scalars and matrices
-are represented as lowercase and uppercase letters, respectively. A
-regular uppercase denotes a row matrix, whereas a bold uppercase
-symbolizes a column matrix. The extended MATLAB code that corresponds
-to each operation can be seen in the last column of
-Table~\ref{tbl:blas-ops}.
-
-\begin{table}[htb] 
-\vspace{-.1in} 
-\caption{Composed BLAS operations used in our experiments}
-\label{tbl:blas-ops}
-\small
-\centering 
-\vspace{-.1in} 
-\begin{tabular}{|c|c|c|c|} 
-\hline 
-Name & Input & Output & C code from MATLAB compiler \\
-\hline 
-\hline 
-VADD & $\overrightarrow{w}$,$\overrightarrow{y}$,$\overrightarrow{z}$ & $\overrightarrow{x}$ &
-$\overrightarrow{x} = \overrightarrow{w} + \overrightarrow{y} + \overrightarrow{z}$ \\ 
-\hline 
-ATAX & $A$,$\overrightarrow{x}$ & $\overrightarrow{y}$ & 
-$\overrightarrow{y} = A' * (A * \overrightarrow{x})$ \\ 
-\hline 
-BiCG Kernel & ${\bf A}$,$\overrightarrow{p}$,$\overrightarrow{r}$ & $\overrightarrow{q}$,$\overrightarrow{s}$ & $\overrightarrow{q} = {\bf A} * \overrightarrow{p}$ \\
-& & & $\overrightarrow{s} = {\bf A}' * \overrightarrow{r}$\\
-\hline 
- & ${\bf A}$,$a$,$b$, & ${\bf B}$, & ${\bf B} = {\bf A} + \overrightarrow{u_1} * \overrightarrow{v_1}' + \overrightarrow{u_2} * \overrightarrow{v_2}'$\\
-GEMVER & $\overrightarrow{u_1}$,$\overrightarrow{u_2}$,$\overrightarrow{v_1}$,$\overrightarrow{v_2}$, & $\overrightarrow{x}$,$\overrightarrow{w}$ & $\overrightarrow{x} = b * ({\bf B}' * \overrightarrow{y}) + \overrightarrow{z}$\\
-& $\overrightarrow{y}$,$\overrightarrow{z}$ & & $\overrightarrow{w} = a * ({\bf B} * \overrightarrow{x})$\\
-\hline 
-\end{tabular} 
-\vspace{-.2in} 
-\end{table}
-
-\begin{figure}[htp]
-{\small
-\centering
-\begin{tabular}{cc}
-
-\begin{minipage}[b]{.515\textwidth}
-\includegraphics[width=\textwidth]{figures/vadd_bgp.eps}
-\end{minipage}
-&
-\begin{minipage}[b]{.515\textwidth}
-\includegraphics[width=\textwidth]{figures/atax.eps}
-\end{minipage}\\
-
-(a) VADD & (b) ATAX \\
-
-\begin{minipage}[b]{.515\textwidth}
-\includegraphics[width=\textwidth]{figures/gemver.eps}
-\end{minipage}
-&
-\begin{minipage}[b]{.515\textwidth}
-\includegraphics[width=\textwidth]{figures/bicgkernel.eps}
-\end{minipage}\\
-
-(c) GEMVER & (d) BiCG Kernel \\
-
-\end{tabular}
-}
-\caption{Performance numbers for different implementations of various composed of BLAS operations}
-\label{fig:results}
-\end{figure}
-
-The performance results of tuning the VADD operation are given in
-Figure~\ref{fig:results}(a). The ``Base'' label designates the simple
-loop C implementation generated by the MATLAB compiler, which was
-simply compiled with the options mentioned earlier and executed. We
-tested the performance of the IBM ESSL library and Goto BLAS for the
-two calls to \texttt{daxpy} used in the BLAS version of the code.
-Finally, we tuned the simple C loop version using Orio, with the
-performance annotations previously shown in
-Figure~\ref{fig:orio-example}. In this experiment, we measured the
-performance for both the sequential and parallel scenarios. We observe
-that even for a very simple operation, such as the vector addition
-above, the compiler alone is unable to obtain the same level of
-performance as the empirically tuned versions. Furthermore, as
-expected, the BLAS performance was worse than that of the single loop
-implementation, however, selecting the correct compiler options was
-crucial for achieving good performance of the base version.
-
-The experiments of the remaining operations were performed on the
-multicore Intel Xeon. Included in these experiments are performance
-numbers for six code variants: the C code generated by the MATLAB
-compiler (''C from MATLAB''), three BLAS-based implementations that
-use Intel MKL, ATLAS, and the default BLAS library on Ubuntu 8.04, and
-the sequential and parallel code variants tuned by Orio (``Orio
-(Seq.)'' and ``Orio (Par.)'', respectively).
-
-The performance results of ATAX are shown in
-Figure~\ref{fig:results}(b). The Orio-tuned version that incorporates
-PLuTo-generated loop fusion optimizations and Orio parallelization
-directives achieves the best performance for most problem sizes,
-outperforming the Intel MKL version by a factor of 2 to 5.7 and the
-compiler-optimized C version by a factor of 4 to 7. The optimizations
-performed by both Orio versions include scalar replacement,
-vectorization, and loop unroll/jam.
-
-Figure~\ref{fig:results}(c) exhibits the performance results of the
-GEMVER operation. Here we used the same PLuTo and Orio optimizations
-as for the ATAX example and, similarly, the parallel Orio version
-achieves the best performance, although in this case the sequential
-Orio version performs almost the same, suggesting that the compiler was
-not able to parallelize the code very effectively. For this operation,
-there are substantial performance differences between different BLAS
-versions, with the Intel MKL version achieving performance close to
-that of the simple compiler code.
-
-The experimental results of BiCG Kernel operation are provided in
-Figure~\ref{fig:results}(d). For this operation, the PLuTo analysis
-did not result in performance improvement, thus we are showing the
-results obtained only through Orio transformations, which included
-vectorization, scalar replacement, and loop unroll/jam. Again the best
-performance was demonstrated by the parallel Orio version, while all
-the BLAS versions performed worse than the compiler-optimized C loop
-version.
-
-\section{Conclusions and Future Work}
-
-We have described an approach to generating tuned linear algebra libraries from high-level annotated MATLAB code that involves a suite of tools to (1) translate the MATLAB code to C, (2) analyze the resulting loops and identify locatility and parallelism-enhancing optimizations using PLuTo, and (3) annotate the resulting C code with syntactic performance directives and use Orio to generate multiple optimized versions and empirically select the one with the best performance.
-Preliminary results from experiments with several composed BLAS operations show that in almost all cases, the optimized code generated by this suite of tools significantly outperforms the versions using tuned BLAS and aggressive compiler optimizations.
-
-The positive initial results from our approach to generating tuned linear algebra routines motivate several future lines of investigation.
-At present there are a number of manual steps involved in obtaining a compiled tuned linear algebra library from the high-level extended MATLAB descriptions.
-We will work on interfacing our separately developed tools more closely and automating their interactions.
-This would enable application developers to easily produce custom linear algebra libraries tuned for specific architectures.
-We will also investigate extending the supported functionality of our MATLAB compiler beyond composed BLAS operations, thus benefiting a wider spectrum of applications.
-
-Orio is growing in a number of directions.
-While it automates much of the tuning process, a number of steps still require nontrivial manual input.
-Orio's modular design allows new transformations to be added easily, which presents many opportunities for expanding the existing set of optimizations with new ones implemented from scratch or through interfaces to other tools, as we have successfully done with PLuTo.
-We are also working on generating code that contains special cases for different input parameter values since in many cases different optimizations are optimal for different input sizes.
-We will also continue developing the parallel empirical search infrastructure; it is clear that more feedback between the MATLAB compiler and Orio will result in automation of the performance directive annotation process and a much more effective exploration of the search space.
-We plan to use a fast analytic cost function in the MATLAB compiler implementation to prune out regions of the search space that will not produce competitive implementations.
-The cost function is based on a model of the computer architecture, but does not necessarily account for all the details.
-Once the search space is narrowed, we will use the Orio empirical testing results with representative data sets to determine the exact performance of the alternatives and choose the best one.
-
-Finally, while Orio currently generates various loop transformations, it does not generate calls to existing library implementations.
-For this paper, we implemented the BLAS versions of all operations manually.
-Our past success with identifying Blue Gene/L intrinsics from expression graphs motivate further research on identifying BLAS and other established library operations from the expression graphs of the MATLAB code.
-
-%\singlespacing
-%% or stuff it in a footnote
-%\section*{Acknowledgments}
-%This work was supported by the Mathematical, Information, and
-%Computational Sciences Division subprogram of the Office of Advanced
-%Scientific Computing Research, Office of Science, U.S. Dept. of Energy,
-%under Contract DE-AC02-06CH11357. 
-%TODO: need Colorado and Ohio acknowledgments
-
-\bibliographystyle{splncs}
-\bibliography{paper}
-
-%\vfill
-%\begin{flushright}
-%\scriptsize
-%\framebox{\parbox{2.4in}{The submitted manuscript has been created by
-%UChicago Argonne, LLC, Operator of Argonne National Laboratory
-%("Argonne").  Argonne, a U.S. Department of Energy Office
-%of Science laboratory, is operated under Contract No.
-%DE-AC02-06CH11357.  The U.S. Government retains for itself, and
-%others acting on its behalf, a paid-up, nonexclusive, irrevocable
-%worldwide license in said article to reproduce, prepare derivative works,
-%distribute copies to the public, and perform publicly and display
-%publicly, by or on behalf of the Government.}}
-%\normalsize
-%\end{flushright}
-
-\end{document}
-
-% Current bibtex warnings:
-This is BibTeX, Version 0.99c (Web2C 7.5.4)
-The top-level auxiliary file: paper.aux
-The style file: siam.bst
-Database file #1: paper.bib
-Warning--can't use both volume and number fields in TCE
-(There was 1 warning)
-
-% Put bibtex here only if you do not have access to the subversion repository, otherwise insert them into paper.bib

### 2008-12-12 doc/papers/iccs09/paper.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@385 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@
 \usepackage{html}
-\usepackage{setspace}
-\usepackage{graphicx}
 \usepackage{subfigure}
 \usepackage{cite}
+\usepackage{wrapfig}
 \usepackage{amsmath}
@@
 % first the title is needed
-\title{Generating Empirically Optimized Numerical Software from MATLAB Prototypes}
+\title{Generating Empirically Optimized Composed Matrix Kernles from MATLAB Prototypes}
 
@@
 
-        % the name(s) of the author(s) follow(s) next
-        %
-        % NB: Chinese authors should write their first names(s) in front of
-        % their surnames. This ensures that the names appear correctly in
-        % the running heads and the author index.
-        %
 \author{Elizabeth Jessup\inst{1} 
@@
 
-%\toctitle{Lecture Notes in Computer Science}
 \tocauthor{Elizabeth Jessup, Boyana Norris, Albert Hartono, and Jeremy Siek}
@@
 
-\begin{abstract}
-% between 75 and 150 words
-The growing demand for higher levels of detail and accuracy in results means that the size and complexity of scientific computations is increasing at least as fast as the improvements in processor technology.
-Programming scientific applications is hard, and optimizing them for high performance is even harder.
-The development of optimized codes requires extensive knowledge, not only of the costs of floating-point arithmetic but also of  memory access issues and compiler optimizations.
-Experiments show that the complexity of this hardware-software system means that performance is difficult to predict fully.
-Therefore, computational scientists are often forced to choose between investing too much time in tuning code or  accepting performance that is significantly lower than the best achievable performance on a given architecture.
-In this paper, we describe the first steps toward a fully automated system for the optimization of the matrix algebra kernels that are a foundational part of many scientific applications.
-To generate highly optimized code from a high-level MATLAB prototype, we define a three-step approach.
-To begin, we have developed a compiler that converts a MATLAB script into simple C code.
-We then use the polyhedral optimization system PLuTo to optimize that code for coarse-grained parallelism and locality simultaneously.
-Finally, we annotate the resulting code with performance tuning directives and use the empirical performance tuning system Orio to generate many tuned versions of the same operation using different optimization techniques, such as loop unrolling and memory alignment.
-Orio performs an automated empirical search to select the best among the multiple optimized code variants.
-We discuss performance results showing that the code generated by using our system significantly outperforms not only the original simple C code but also code based on source BLAS, ATLAS-optimized BLAS, and Intel MKL routines.
-\keywords{Code generation, empirical performance tuning}
-\end{abstract}
+\input{abstract}
 
@@
 
-The development of high-performance numerical codes is challenging because performance is determined by complex interactions between the algorithm, data structure, programming language, compiler, and computer architecture.
-Scientists seeking high performance are thus required to master advanced concepts in computer science and carry out intricate programming tasks in addition to managing the scientific content of their work.
-They must either invest substantial time in tuning their software or accept low performance.
-In either case, the productivity of the scientist degrades.
+The development of high-performance numerical codes is challenging because
+performance is determined by complex interactions between the algorithm, data
+structure, programming language, compiler, and computer
+architecture. Scientists seeking high performance are thus required to master
+advanced concepts in computer science and carry out intricate programming
+tasks in addition to managing the scientific content of their work.  They
+must either invest substantial time in tuning their software or accept low
+performance. In either case, the productivity of the scientist degrades.
 
-Historically, the research community has pursued two separate paths toward the goal of making software run at near-peak levels.
-The first of these builds on research into compilers and their associated technologies.
-The holy grail of compilation research is to take an arbitrary code as input and produce optimal code as output for a given language and hardware platform.
-Despite the immense effort that has been poured into that approach, its success has been limited both by practical constraints (users will not tolerate compile times that extend into several days) and by the amount of detailed information the compiler can obtain about the software to be compiled and the hardware on which it is to execute~\cite{Ken99,Wol89,Wol96}.
-The second path is to identify kernel routines that constitute the dominant costs of a wide variety of applications.
-When such kernels can be identified and agreed upon by the members of the community, programmers with the required level of technical knowledge can concentrate on producing these optimized kernel libraries for architectures of interest.
-A prime example is the high-performance BLAS libraries produced by a combination of hardware vendors, independent software vendors, and researchers.
-Developers who write their codes calling these routines can achieve high performance across all supported architectures.
-The efforts in high-performance library development aptly demonstrate, however, that the research into compiler technologies has not been fully successful.
-If compilers could take the relatively simple code even for basic tasks such as a dense matrix-matrix product and produce optimal code, this second approach would be unnecessary.
-Like compiler optimization, the kernel library-oriented approach has significant limitations.
-Support for a variety of architectures is the most serious problem because the optimizations necessary to achieve near-peak performance are by nature not portable.
-Lack of portability is particularly problematic because processor designs are constantly changing,  and constant updates are nearly impossible to maintain by hand.
-It is a sad fact of the computing industry that by the time highly-optimized code is available for a given architecture, that architecture is generally well on its way to obsolescence.
+Historically, the research community has pursued two separate paths toward
+the goal of making software run at near-peak levels.  The first path builds
+on research into compilers and their associated technologies.  One of the
+main goals of compilation research is to take an arbitrary code as input and
+produce optimal code as output for a given language and hardware platform.
+The success of this approach has been limited by a number of factors: (i)
+optimal mappings between the computation graph and the hardware are expensive
+(often NP-complete) to compute; (ii) potentially useful information that
+could aid optimization cannot be represented in general-purpose languages
+such as C and Fortran; and (iii) user control of compiler optimizations is
+limited and varies from compiler to compiler, and (iv) apart from differences
+in execution time, it is difficult for users to evaluate the effectiveness of
+different compiler optimizations.
 
-We are working on a combination of the two approaches designed to overcome some of their shortcomings.
-This paper describes our multipronged approach toward the development of software infrastructure for \emph{generating} automatically tuned libraries for matrix algebra computations.
-We briefly overview relevant prior and current research efforts in Section~\ref{sec:background}.
-In Section~\ref{sec:approach} we describe our MATLAB to C compiler and the empirical performance tuning system Orio and its use in conjunction with the PLuTo tool suite to generate and empirically evaluate many tuned versions of the C code generated by the MATLAB compiler.
-In Section~\ref{sec:experiments}, we provide performance results showing that the code generated by using our system significantly outperforms not only the original simple C code but also code based on source BLAS, ATLAS-optimized BLAS, and Intel MKL routines.
+The second path to performance optimization is to identify kernel
+routines that dominate the execution time of a wide variety of
+applications.  
+%When such kernels can be identified and agreed upon by the
+%members of the community, programmers with the required level of technical
+%knowledge can concentrate on producing these optimized kernel libraries for
+%architectures of interest. 
+An example is the high-performance BLAS libraries produced by a combination
+of hardware vendors, independent software vendors, and researchers.
+Developers who write their codes calling these routines can achieve high
+performance across all supported architectures. The efforts in
+high-performance library development aptly demonstrate, however, that the
+research into compiler technologies has not been fully successful.  If
+compilers could take the relatively simple code such as a dense matrix-matrix
+product and produce optimal code, this second approach would be unnecessary.
+Like compiler optimization, the kernel library-oriented approach has
+significant limitations.  Support for a variety of architectures is the most
+serious problem because the optimizations necessary to achieve near-peak
+performance are by nature not portable, requiring manual update or
+reimplementation for each new architecture.
+
+This paper describes a combination of the two approaches designed to overcome
+some of their shortcomings. We take a multipronged approach
+toward the development of software infrastructure for \emph{generating}
+automatically tuned libraries for matrix algebra computations.  We briefly
+overview relevant prior and current research efforts in
+Section~\ref{sec:background}. In Section~\ref{sec:approach} we describe our
+MATLAB~\cite{matlab:webpage} to C compiler and the empirical performance tuning system Orio and its
+use in conjunction with the PLuTo tool suite to generate and empirically
+evaluate many tuned versions of the C code generated by the MATLAB compiler.
+In Section~\ref{sec:experiments}, we provide performance results showing that
+the code generated by using our system significantly outperforms not only the
+original simple C code but also code based on source BLAS, ATLAS-optimized
+BLAS, and Intel MKL routines.
 %Text in intro repeats abstract some, but that might not be a big deal for a conf paper.
@@
 
-Existing optimizing MATLAB compilers, such as the MaJIC MATLAB compiler, include limited local optimizations for matrix expressions, but do not perform optimizations such as loop fusion across multiple operations as we do with the tools described in this paper.
-The telescoping languages project~\cite{telescopingurl,teleoverview,Ken99} uses techniques such as strength reduction, vectorization, and procedure specialization to optimize MATLAB scripts, but do not generate reusable optimized linear algebra routines as described in this paper.
+Existing optimizing MATLAB compilers, such as the MaJIC MATLAB compiler~\cite{MaJIC},
+include limited local optimizations for matrix expressions, but do not
+perform optimizations such as loop fusion across multiple operations as we do
+with the tools described in this paper. The telescoping languages
+project~\cite{telescopingurl,teleoverview,Ken99} uses techniques such as
+strength reduction, vectorization, and procedure specialization to optimize
+MATLAB scripts, but do not generate reusable optimized linear algebra
+routines as described in this paper.
 
-The most common approach  to tuning numerical codes is for an expert to transform the source manually, unrolling loops, blocking for multiple levels of cache, and inserting prefetch instructions.
-The pitfalls of this approach are well-understood~\cite{Goedecker01}.
-It requires a significant amount of time and effort.
-Optimizing code for one particular platform may in fact make it less efficient on other platforms and often makes it complex and hard to understand or maintain.
-An alternative is the use of tuned libraries of key numerical algorithms (e.g., BLAS~\cite{Dongarra:1990fk} and LAPACK~\cite{LAPACK} for dense linear algebra, SPARSPAK~\cite{sparspak} for sparse linear algebra, and BLITZ++~\cite{blitz} to support scientific computing in C++, or GPULib by Tech-X, an open source library for accelerating numerical computations on Graphics Processing Units~\cite{gpulib}, to name just a few).
-Such libraries are difficult to develop, and they generally cannot adapt to execution context: as we note in Section~\ref{sec:approach}, a series of calls to library routines can incur high memory costs.
+The most common approach to tuning numerical codes is for an expert to
+transform the source manually, unrolling loops, blocking for multiple levels
+of cache, and inserting prefetch instructions.  The pitfalls of this approach
+are well-understood~\cite{Goedecker01}.  It requires a significant amount of
+time and effort. Optimizing code for one particular platform may in fact make
+it less efficient on other platforms and often makes it complex and hard to
+understand or maintain.  An alternative is the use of tuned libraries of key
+numerical algorithms (e.g., BLAS~\cite{Dongarra:1990fk} and
+LAPACK~\cite{LAPACK} for dense linear algebra, SPARSPAK~\cite{sparspak} for
+sparse linear algebra, and BLITZ++~\cite{blitz} to support scientific
+computing in C++, or GPULib by Tech-X, an open source library for
+accelerating numerical computations on Graphics Processing
+Units~\cite{gpulib}, to name just a few). Such libraries are difficult to
+develop, and they generally cannot adapt to execution context: as we note in
+Section~\ref{sec:approach}, a series of calls to library routines can incur
+high memory costs.
 
-Specialized code generators circumvent the high costs of manual code generation .
-They include tools for basic dense linear algebra operations (ATLAS~\cite{WN147}, PhiPAC~\cite{bilmes97optimizing,phipacwww}), sparse linear algebra (OSKI~\cite{vuduc05}), FFTs~\cite{FFTW,Spiral}), stencil-based operations~\cite{kamil06}, and tensor computations (TCE~\cite{TCE}) among others.
-It is often impossible to predict the performance of code precisely on modern computer architectures.
-Thus, many of these specialized code generators exploit search strategies to identify the best (or nearly best) code for a particular choice of problem parameters and machine.
-Such autotuning tools are not general but rather each is valuable in one specific problem domain.
-While not all applications can be written in terms of available libraries, the performance achieved with autotuning tools does convincingly demonstrate the advantage of source-to-source transformation over optimizing compilers.
+Specialized code generators circumvent the high costs of manual code
+generation . They include tools for basic dense linear algebra operations
+(ATLAS~\cite{WN147}, PhiPAC~\cite{bilmes97optimizing,phipacwww}), sparse
+linear algebra (OSKI~\cite{vuduc05}), FFTs~\cite{FFTW,Spiral}), stencil-based
+operations~\cite{kamil06}, and tensor computations (TCE~\cite{TCE}) among
+others.  While these libraries target a specific kernel, our approach aims 
+at enabling the definition of arbitrary kernels involving dense matrix linear algebra.
+%It is often impossible to predict the performance of code precisely
+%on modern computer architectures. Thus, many of these specialized code
+%generators exploit search strategies to identify the best (or nearly best)
+%code for a particular choice of problem parameters and machine. Such
+%autotuning tools are not general but rather each is valuable in one specific
+%problem domain. 
+%While not all applications can be written in terms of available libraries,
+%the performance achieved with autotuning tools does convincingly demonstrate
+%the benefits of automated source code transformations.
 
-The best performance today is obtained through approaches like that taken by PERI~\cite{PERI} which involves a labor intensive rewrite of portion of the code by a group of performance tools experts in collaboration with application scientists.
-Maintaining performance of the code as the application and hardware evolves means that the tuning process may need to be repeated.
-Our efforts aim to capture this performance tuning expertise in a way that can be integrated into the application.
-We will thus enable performance portability and ensure that the improvements to the application are long-lasting.
+%The best performance today is obtained through approaches like that taken by
+%PERI~\cite{PERI} which involves a labor intensive rewrite of portion of the
+%code by a group of performance tools experts in collaboration with
+%application scientists. Maintaining performance of the code as the
+%application and hardware evolves means that the tuning process may need to be
+%repeated. Our efforts aim to capture this performance tuning expertise in a
+%way that can be integrated into the application. We will thus enable
+%performance portability and ensure that the improvements to the application
+%are long-lasting.
 
-A number of source or binary transformation tools for performance-improving optimizations exist.
-LoopTool~\cite{LoopTool}, developed at Rice University, supports annotation-based loop fusion, unroll-and-jam, skewing and tiling.
-A relatively new tool, POET~\cite{POET} also supports a number of loop transformations.
-POET offers a complex template-based syntax for defining transformations in a language-independent manner.
-Other recent research efforts whose goal, at least in part, is to enable optimizations of source code to be augmented with performance-related information include the X language~\cite{XLanguage} (a macro C-like language for annotating C code), the Broadway~\cite{broadway} compiler, and various meta-programming techniques~\cite{veldhuizen95,weise93,kiczales91,chiba95}.
-PLuTo~\cite{Pluto,uday08cc} is a source-to-source transformation tool for optimizing sequences of nested loops.
-PLuTo employs a polyhedral model of nested loops, where the dynamic instance (iteration) of each statement is viewed as an integer point in a well-defined space, called the statement's polyhedron.
-Combined with a characterization of data dependences, this representation allows the construction of mathematically correct complex loop transformations.
-The transformations target both improved cache locality and parallelism.
+A number of source or binary transformation tools for general
+performance-improving optimizations exist. LoopTool~\cite{LoopTool},
+developed at Rice University, supports annotation-based loop fusion,
+unroll-and-jam, skewing and tiling.  A relatively new tool, POET~\cite{POET}
+also supports a number of loop transformations. POET offers a complex
+template-based syntax for defining transformations in a language-independent
+manner. Other recent research efforts whose goal, at least in part, is to
+enable optimizations of source code to be augmented with performance-related
+information include the X language~\cite{XLanguage} (a macro C-like language
+for annotating C code), the Broadway~\cite{broadway} compiler, and various
+meta-programming
+techniques~\cite{veldhuizen95,weise93,kiczales91,chiba95}. PLuTo~\cite{Pluto,uday08cc}
+is a source-to-source transformation tool for optimizing sequences of nested
+loops. PLuTo employs a polyhedral model of nested loops, where the dynamic
+instance (iteration) of each statement is viewed as an integer point in a
+well-defined space, called the statement's polyhedron. Combined with a
+characterization of data dependences, this representation allows the
+construction of mathematically correct complex loop transformations. The
+transformations target both improved cache locality and parallelism.
 
-\section{Our Approach to Optimizing Composed BLAS Operations}
+\section{Optimizing Composed BLAS Operations}
 \label{sec:approach}
 
-%\section{Optimizing Composed Operations}
-%\label{sec:composed}
+Codes based on matrix algebra are generally constructed as a sequence of
+calls to the Basic Linear Algebra Subprograms
+(BLAS)~\cite{Dongarra:1988uq,Lawson:1979kx,Dongarra:1990fk} and similar
+sparse matrix libraries~\cite{Saad:fr,George:1981uq}. Writing programs in
+this way promotes readability and maintainability but can be costly in terms
+of memory efficiency. Specifically, the retrieval of a large order matrix at
+each routine call can have a profound effect on performance even when highly
+tuned implementations of the BLAS
+(e.g.~\cite{Bilmes:1997ye,Whaley:1998fk,IntelMath:oq,ESSL:kl,Goto:2006fk})
+are used.
 
-Codes based on matrix algebra are generally constructed as a sequence of calls to the Basic Linear Algebra Subprograms (BLAS)~\cite{Dongarra:1988uq,Lawson:1979kx,Dongarra:1990fk} and similar sparse matrix libraries~\cite{Saad:fr,George:1981uq}.
-Writing programs in this way promotes readability and maintainability but can be costly in terms of memory efficiency.
-Specifically, the retrieval of a large order matrix at each routine call can have a profound effect on performance even when highly tuned implementations of the BLAS (e.g.~\cite{Bilmes:1997ye,Whaley:1998fk,IntelMath:oq,ESSL:kl,Goto:2006fk}) are used.
+The authors and other researchers have observed that it can be much more
+efficient to call a single specialized routine that performs multiple
+operations than to make successive calls to separate BLAS routines (e.g., see
+~\cite{Blackford:2002vn,baker03lgmres,,gropp01,Vuduc:2003kl}.
+%~\cite{Ashby:uq,Blackford:2002vn,baker03blgmres,baker03lgmres,Dennis:2005tg,Howell:2008,gropp01,Vuduc:2003kl}.
+Single routines that carry out more than one linear algebra operation are
+known as \emph{composed} BLAS. As an example, consider the pair of
+matrix-vector products $q = Ap$, $s = A^Tr$, where $A$ is a matrix and $p$,
+$q$, $r$, and $s$ are vectors, that comprise the computational bottlenecks of
+the BiConjugate Gradient Method~\cite{Barrett:1994kx,Saad:2003fk} and of the
+GEMVER kernel examined in Section~\ref{sec:matlab}.  These two operations can
+be implemented as a pair of calls to the BLAS routine {\tt GEMV} or they can
+be rewritten as a composed BLAS consisting of the doubly nested loop
+encompassing both matrix-vector products. In the former case, the matrix
+$A$ is accessed twice, while in the latter it is accessed only once. Our
+preliminary results in prior research and in this paper indicate that 
+loop fusion leads to a routine that delivers significantly better performance
+than does a pair of calls to the best optimized BLAS {\tt GEMV} routines for
+large matrix orders.  Composed routines are the focus of the work presented
+here, but much of what we discuss generalizes to a much broader array of
+computations.
 
-The authors and other researchers have observed that it can be much more efficient to call a single specialized routine that performs multiple operations than to make successive calls to separate BLAS routines ~\cite{Ashby:uq,Blackford:2002vn,baker03blgmres,baker03lgmres,Dennis:2005tg,Howell:2008,gropp01,Vuduc:2003kl}.
-Single routines that carry out more than one linear algebra operation are known as \emph{composed} BLAS.
-As an example, consider the pair of matrix-vector products $q  = Ap$, $s  = A^Tr$, where $A$ is a matrix and $p$, $q$, $r$, and $s$ are vectors, that comprise the computational
-bottlenecks of the BiConjugate Gradient Method~\cite{Barrett:1994kx,Saad:2003fk}  and of the GEMVER kernel examined in Section~\ref{sec:matlab}.
-These two operations can be implemented as a pair of calls to the BLAS routine {\tt GEMV} or they can be rewritten as a composed BLAS consisting of the doubly nested loop encompassing both matrix-vector multiplies.
-In the former case, the matrix $A$ is accessed twice, while in the latter it is accessed only once.
-Our preliminary results in prior research and in this paper indicate that this loop fusion leads to a routine that delivers significantly better performance than does a pair of calls to the best optimized BLAS {\tt GEMV} routines for large matrix orders.
-Composed routines are the focus of the work presented here, but much of what we discuss generalizes to a much broader array of computations.
-
-% While large speedups are probable, composition can  also force memory accesses to be consecutive, and it can
-
-%lead to the need to bring more data into cache. These bad
-
-% effects can be remedied by combining composition with other tuning techniques like cache blocking and software pipelining.
-
-\begin{figure}[htbp]
+To generate highly optimized code from a MATLAB prototype of the composed BLAS operation, we follow a three-step approach, illustrated in Figure~\ref{fig:process}.  To begin,
+%
+\begin{wrapfigure}{r}{3in}
+\vspace{-.2in}
 \centering
-\includegraphics[width=0.7\textwidth]{figures/process.png}
+\includegraphics[width=3in]{figures/process.png}
 \caption{Overview of the code generation and tuning process, optional steps are indicated with dashed arrows.}
 \label{fig:process}
-\end{figure}
-To generate highly optimized code from a MATLAB prototype of the composed BLAS operation, we follow a three-step approach, illustrated in Figure~\ref{fig:process}.
-To begin, we have developed a compiler that converts a MATLAB script into simple C code.
+\vspace{-.1in}
+\end{wrapfigure}
+%
 %~\cite{Siek}.  
+we have developed a compiler that converts a MATLAB script into simple C code.
 After generating the C code from the high-level MATLAB prototype, we (optionally) use the source-to-source automatic parallelization tool PLuTo\cite{Pluto} %developed at Ohio State University
@@
 
+\vspace{-.2in}
 \subsection{A MATLAB Compiler}
 \label{sec:matlab}
-
-In this section we describe the design and operation of the MATLAB to C compiler we have developed.
-Figure~\ref{fig:compiler} gives an overview of the compilation process.
+%\input{matlab}
+In this section we describe the design and operation of the MATLAB to C compiler we have developed~\cite{Siek}.    Figure~\ref{fig:compiler} gives an overview of the compilation process. 
+%
+\begin{wrapfigure}{r}{2in}
+\vspace{-.2in}
+\centering
+\includegraphics[width=2in]{figures/compile.png}
+\caption{The MATLAB compiler.}
+\label{fig:compiler}
+\vspace{-.3in}
+\end{wrapfigure}
 The MATLAB kernel specification is parsed into a high-level intermediate representation in the form of a dataflow graph.
@@
 
-\begin{figure}[htbp]
-\centering
-\includegraphics[width=.7\textwidth]{figures/compile.png}
-
-\caption{Overview of the MATLAB compiler.}
-\label{fig:compiler}
-\end{figure}
-
-\subsubsection{The Dataflow Graph}
+\subsubsection{The Dataflow Graph.}
 
 An example dataflow graph for the GEMVER kernel, defined below, is given in figure~\ref{fig:gemver-dataflow}.
+\begin{figure}[ht]
+\begin{minipage}{.3\textwidth}
 \begin{align*}
@@
 \end{align*} 
-Each node represents a parameter of the kernel or an operation.
-The arrows indicate the flow of data.
+\end{minipage}
+\begin{minipage}{.6\textwidth}
+\includegraphics[width=3in]{figures/gemver-dataflow.png}
+\end{minipage}
+\caption{Dataflow graph (right) for the GEMVER kernel (left).}
+\label{fig:gemver-dataflow}
+\end{figure}
+
+Each node represents a parameter of the kernel or an operation.  The arrows 
+indicate the flow of data.
 At first the graph specifies what operations are to be performed but does not contain any implementation details.
@@
 
-\begin{figure}[htbp]
-\centering
-\includegraphics[width=.7\textwidth]{figures/gemver-dataflow.png}
-
-\caption{Dataflow graph for the GEMVER kernel.}
-\label{fig:gemver-dataflow}
-\end{figure}
-
-\subsubsection{Analyze the Dataflow Graph}
+\subsubsection{Analyze the Dataflow Graph.}
 
@@
 
-\subsection{Refine the Dataflow Graph}
+\subsubsection{Refine the Dataflow Graph.}
 \label{sec:refine}
@@
 
-\subsubsection{Optimize the Dataflow Graph}
+\subsubsection{Optimize the Dataflow Graph.}
 
@@
 
-\subsubsection{Translate the Dataflow Graph to C}
+\subsubsection{Translate the Dataflow Graph to C.}
 
@@
 %~\cite{Norris:2007}
-Orio is an empirical tuning tool that takes annotated C code as input,
+Orio\cite{Norris:2007,ipdps09} is an empirical tuning tool that takes annotated C code as input,
 generates multiple transformed versions of the annotated code, and
@@
 Our past success with identifying Blue Gene/L intrinsics from expression graphs motivate further research on identifying BLAS and other established library operations from the expression graphs of the MATLAB code.
-
+\cite{Siek}
 %\singlespacing

### 2008-12-12 doc/papers/iccs09/paper.tex
fits in 10 pages, more editing needed, though
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@386 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@
-% CGO 09: http://www.cgo.org/
-% svn co https://svn.mcs.anl.gov/repos/performance/orio/doc/papers/cgo08
-% Max 22 pages, double-spaced, 11pt or larger font
-% Abstract 300 words: Sep. 12, 2008
-% Full paper: Sep. 19, 2008
+% ICCS'09 http://www.iccs-meeting.org/iccs2009/
+% svn co https://svn.mcs.anl.gov/repos/performance/orio/doc/papers/iccs09
+% Important dates:
+% Full papers submission (extended)       December 20, 2008
+% Notification of acceptance of papers    February 2, 2009
+% Camera ready papers                        February 15, 2009
+% Max 10 pages, llncs style
 
@@
 with the tools described in this paper. The telescoping languages
-project~\cite{telescopingurl,teleoverview,Ken99} uses techniques such as
+%~\cite{telescopingurl,teleoverview,Ken99} 
+project~\cite{telescopingurl} uses techniques such as
 strength reduction, vectorization, and procedure specialization to optimize
@@
 understand or maintain.  An alternative is the use of tuned libraries of key
-numerical algorithms (e.g., BLAS~\cite{Dongarra:1990fk} and
-LAPACK~\cite{LAPACK} for dense linear algebra, SPARSPAK~\cite{sparspak} for
-sparse linear algebra, and BLITZ++~\cite{blitz} to support scientific
-computing in C++, or GPULib by Tech-X, an open source library for
-accelerating numerical computations on Graphics Processing
-Units~\cite{gpulib}, to name just a few). Such libraries are difficult to
-develop, and they generally cannot adapt to execution context: as we note in
-Section~\ref{sec:approach}, a series of calls to library routines can incur
-high memory costs.
+numerical algorithms, for example BLAS~\cite{Dongarra:1990fk} and
+LAPACK~\cite{LAPACK} for dense linear algebra.
+%, SPARSPAK~\cite{sparspak} for
+%sparse linear algebra, and BLITZ++~\cite{blitz} to support scientific
+%computing in C++, or GPULib by Tech-X, an open source library for
+%accelerating numerical computations on Graphics Processing
+%Units~\cite{gpulib}, to name just a few). Such libraries are difficult to
+%develop, and they generally cannot adapt to execution context: as we note in
+%Section~\ref{sec:approach}, a series of calls to library routines can incur
+%high memory costs.
 
@@
 generation . They include tools for basic dense linear algebra operations
-(ATLAS~\cite{WN147}, PhiPAC~\cite{bilmes97optimizing,phipacwww}), sparse
-linear algebra (OSKI~\cite{vuduc05}), FFTs~\cite{FFTW,Spiral}), stencil-based
-operations~\cite{kamil06}, and tensor computations (TCE~\cite{TCE}) among
-others.  While these libraries target a specific kernel, our approach aims 
+(ATLAS~\cite{WN147}, PhiPAC~\cite{bilmes97optimizing},
+%,phipacwww})
+sparse linear algebra (OSKI~\cite{vuduc05}), 
+%FFTs~\cite{FFTW,Spiral}), stencil-based
+
+%operations~\cite{kamil06}, and tensor computations (TCE~\cite{TCE}) among
+among others.  While these libraries target a specific kernel, our approach aims 
 at enabling the definition of arbitrary kernels involving dense matrix linear algebra.
-%It is often impossible to predict the performance of code precisely
-%on modern computer architectures. Thus, many of these specialized code
-%generators exploit search strategies to identify the best (or nearly best)
-%code for a particular choice of problem parameters and machine. Such
-%autotuning tools are not general but rather each is valuable in one specific
-%problem domain. 
+It is often impossible to predict the performance of code precisely
+on modern computer architectures. Thus, many of these specialized code
+generators exploit search strategies to identify the best (or nearly best)
+code for a particular choice of problem parameters and machine. Most existing
+autotuning tools are not general but rather focus on a specific domain or algorithm.
 %While not all applications can be written in terms of available libraries,
@@
 template-based syntax for defining transformations in a language-independent
-manner. Other recent research efforts whose goal, at least in part, is to
-enable optimizations of source code to be augmented with performance-related
-information include the X language~\cite{XLanguage} (a macro C-like language
-for annotating C code), the Broadway~\cite{broadway} compiler, and various
-meta-programming
-techniques~\cite{veldhuizen95,weise93,kiczales91,chiba95}. PLuTo~\cite{Pluto,uday08cc}
-is a source-to-source transformation tool for optimizing sequences of nested
-loops. PLuTo employs a polyhedral model of nested loops, where the dynamic
-instance (iteration) of each statement is viewed as an integer point in a
-well-defined space, called the statement's polyhedron. Combined with a
-characterization of data dependences, this representation allows the
-construction of mathematically correct complex loop transformations. The
-transformations target both improved cache locality and parallelism.
+manner (but currently only C++ is supported). 
+%Other recent research efforts whose goal, at least in part, is to
+%enable optimizations of source code to be augmented with performance-related
+%information include the X language~\cite{XLanguage} (a macro C-like language
+%for annotating C code), the Broadway~\cite{broadway} compiler, and various
+%meta-programming
+%~\cite{veldhuizen95,weise93,kiczales91,chiba95}
+%techniques (e.g., ~\cite{veldhuizen95,chiba95}). 
+PLuTo~\cite{Pluto}
+%uday08cc} 
+is a source-to-source transformation tool for
+optimizing sequences of nested loops. PLuTo employs a polyhedral model of
+nested loops, where the dynamic instance (iteration) of each statement is
+viewed as an integer point in a well-defined space, called the statement's
+polyhedron. Combined with a characterization of data dependences, this
+representation allows the construction of mathematically correct complex loop
+transformations. The transformations target both improved cache locality and
+parallelism.
 
@@
 calls to the Basic Linear Algebra Subprograms
-(BLAS)~\cite{Dongarra:1988uq,Lawson:1979kx,Dongarra:1990fk} and similar
+(BLAS)
+%~\cite{Dongarra:1988uq,Lawson:1979kx,Dongarra:1990fk} 
+and similar
 sparse matrix libraries~\cite{Saad:fr,George:1981uq}. Writing programs in
@@
 each routine call can have a profound effect on performance even when highly
-tuned implementations of the BLAS
-(e.g.~\cite{Bilmes:1997ye,Whaley:1998fk,IntelMath:oq,ESSL:kl,Goto:2006fk})
+tuned implementations of the BLAS (e.g.,~\cite{Goto:2006fk})
+%(e.g.~\cite{Bilmes:1997ye,Whaley:1998fk,IntelMath:oq,ESSL:kl,Goto:2006fk})
 are used.
@@
 operations than to make successive calls to separate BLAS routines (e.g., see
-~\cite{Blackford:2002vn,baker03lgmres,,gropp01,Vuduc:2003kl}.
+~\cite{gropp01,Vuduc:2003kl}.
 %~\cite{Ashby:uq,Blackford:2002vn,baker03blgmres,baker03lgmres,Dennis:2005tg,Howell:2008,gropp01,Vuduc:2003kl}.
@@
 To generate highly optimized code from a MATLAB prototype of the composed BLAS operation, we follow a three-step approach, illustrated in Figure~\ref{fig:process}.  To begin,
+we have developed a compiler that converts a MATLAB script into simple C code~\cite{Siek}.
+After generating the C code from the high-level MATLAB prototype, 
 %
@@
 \includegraphics[width=3in]{figures/process.png}
-\caption{Overview of the code generation and tuning process, optional steps are indicated with dashed arrows.}
+\caption{The code generation and tuning process.}
 \label{fig:process}
-\vspace{-.1in}
+\vspace{-.2in}
 \end{wrapfigure}
 %
-%~\cite{Siek}.  
-we have developed a compiler that converts a MATLAB script into simple C code.
-After generating the C code from the high-level MATLAB prototype, we (optionally) use the source-to-source automatic parallelization tool PLuTo\cite{Pluto} %developed at Ohio State University
-to optimize for coarse-grained parallelism and locality simultaneously.
-Using the results of the PLuTo analysis, we insert annotations into the C code, which are then processed by our extensible annotation system Orio to generate many tuned versions of the same operation using different optimization parameters.
-Orio then performs an empirical search to select the best among the multiple optimized code variants.
+we (optionally) use the source-to-source automatic parallelization tool PLuTo~\cite{Pluto} %developed at Ohio State University
+to optimize for coarse-grained parallelism and locality simultaneously. Using
+the results of the PLuTo analysis, we insert annotations into the C code,
+which are then processed by our extensible annotation system Orio to generate
+many tuned versions of the same operation using different optimization
+parameters. Orio then performs an empirical search to select the best among
+the multiple optimized code variants.
 
@@
 
-\vspace{-.2in}
+%\vspace{-.2in}
 \subsection{A MATLAB Compiler}
@@
 \end{wrapfigure}
-The MATLAB kernel specification is parsed into a high-level intermediate representation in the form of a dataflow graph.
-This dataflow graph is then iteratively processed until all of the implementation choices have been made.
-The compilation process consists of three phases: analysis, lowering, and optimization, which are together iterated until all of the implementation decisions have been made.
-The graph is then translated into C code.
-
-\subsubsection{The Dataflow Graph.}
-
-An example dataflow graph for the GEMVER kernel, defined below, is given in figure~\ref{fig:gemver-dataflow}.
+The MATLAB kernel specification is parsed into a high-level intermediate
+representation in the form of a dataflow graph.  This dataflow graph is then
+iteratively processed until all of the implementation choices have been made.
+The compilation process consists of three phases: analysis, refinement, and
+optimization, which are together iterated until all of the implementation
+decisions have been made.  The graph is then translated into C code.
+%\subsubsection{The Dataflow Graph.}
+%
+An example dataflow graph for the GEMVER kernel, defined below, is shown
+in Figure~\ref{fig:gemver-dataflow}.
+Each node represents a parameter of the kernel or an operation.  The arrows
+indicate the flow of data. At first the graph specifies what operations are
+to be performed but does not contain any implementation details. The symbol
+$\times$ in the depicted graph stands variously for outer product,
+matrix-vector multiplication, and scalar-vector multiplication, and it does
+not yet specify, for example, whether the outer products compute a row or a
+column at a time of the result matrix.
 \begin{figure}[ht]
@@
 
-Each node represents a parameter of the kernel or an operation.  The arrows 
-indicate the flow of data.
-At first the graph specifies what operations are to be performed but does not contain any implementation details.
-The symbol $\times$ in the depicted graph stands variously for outer product, matrix-vector multiplication, and scalar-vector multiplication, and it does not yet specify, for example, whether the outer products compute a row or a column at a time of the result matrix.
+Here we briefly describe the analysis, refinement, and optimization of the
+dataflow graph, which are discussed in more detail in~\cite{Siek}. During the
+analysis phase, all types of intermediate nodes are computed and
+assigned. The algorithm choice and storage format determination are computed
+simultaneously. For example, the $\times$ symbol in the graph in
+Figure~\ref{fig:gemver-dataflow} having the inputs $u_1$ and $v_1^T$ can be
+implemented by iterating over rows first, or over columns first. The choice
+depends on how the result is used downstream in the dataflow graph.  In this
+case, the result is added to the outer product of $u_2$ and $v_2^T$, so we
+still could choose either option as long as we make the same choice for both
+outer products. 
+%Going one more step downstream, there is an addition with
+%$A$, which was annotated in Figure~\ref{fig:gemver} to be column major. At
+%this point it is clear that the outer-products should be computed in
+%column-major order.
 
-\subsubsection{Analyze the Dataflow Graph.}
+The information on implementation possibilities for basic linear algebra
+operations are not hard-coded in the compiler; rather, these data are stored
+in a database, called the \emph{linear algebra database}.  This separation
+allows us to add new matrix formats, operations, and basic linear algebra
+algorithms without changes to the compiler algorithm.
 
-At the beginning of the analysis phase, each input and output node of the dataflow graph is labeled with its kind (scalar, vector, matrix, or unknown) and storage format (dense row-major, dense column-major, or unknown) according to the annotation given by the user.
-During analysis, the kind and storage format are computed for the intermediate nodes and algorithms are assigned to operation nodes.
-The choice of algorithm and the determination of kinds of storage formats are interrelated so they must be computed together.
-For example, the $\times$ symbol in the graph in Figure~\ref{fig:gemver-dataflow} having the inputs $u_1$ and $v_1^T$ can be implemented by iterating over rows first, or over columns first.
-The choice depends on how the result is used downstream in the dataflow graph.
-In this case, the result is added to the outer product of $u_2$ and $v_2^T$, so we still could choose either option as long as we make the same choice for both outer products.
-Going one more step downstream, there is an addition with $A$, which was annotated in Figure~\ref{fig:gemver} to be column major.
-At this point it is clear that the outer-products should be computed in column-major order.
+The analysis algorithm makes implementation choices using the
+most-con\-strained-first strategy (also known as minimum remaining
+values)~\cite{Russell:2003mz}.  The compiler chooses the node with the fewest
+matching implementations (in the linear algebra database) and assigns an
+algorithm name to the node.  If there is more than one match, the prototype
+compiler picks the first.  This process is repeated with all remaining nodes
+in the graph.
 
-We do not hard code this knowledge of basic linear algebra in the compiler algorithm itself, but instead use a data-driven approach and store information regarding how to implement basic linear algebra in a separate database, which we refer to as the \emph{linear algebra database}.
-This separation allows us to add new matrix formats, operations, and basic linear algebra algorithms without changes to the
-compiler algorithm.
+%Once
+%we add backtracking we will push each alternative onto the stack to be
+%explored later. Once the choice is made, the kind and storage format labels
+%for the operation node itself and for the input nodes are updated with the
+%labels specified in the linear algebra database for the chosen algorithm. The
+%algorithm then goes on to find the next node with the fewest matching
+%implementations and repeats.  In this phase, only the name of the algorithm
+%is assigned to the operation node.  The details of the implementation are not
+%specified until the refinement phase.
 
-The analysis algorithm makes implementation choices using the most-constrained first strategy (also called minimum remaining values)
-from the artificial intelligence literature~\cite{Russell:2003mz}.
-%This may need to be updated depending on how we decide to talk about the empirical search. -Jeremy 8/6/08 10:15 AM 
-The compiler chooses the node with the fewest matching implementations (in the linear algebra database) and assigns a matching implementation to the node.
-If there is more than one match, the prototype compiler picks the first.
-Once we add backtracking we will push each alternative onto the stack to be explored later.
-Once the choice is made, the kind and storage format labels for the operation node itself and for the input nodes are updated with the labels specified in the linear algebra database for the chosen algorithm.
-The algorithm then goes on to find the next node with the fewest matching implementations and repeats.
-In this phase, only the name of the algorithm is assigned to the operation node.
-The details of the implementation are not specified until the refinement phase.
+The refinement phase resolves the implementation for each operation node in
+the graph into a subgraph defining the details of the chosen algorithm.  Each
+subgraph is an abstract representation of the loop that implements the given
+operation that also contains an iteration strategy for traversing the
+elements of the matrix or vector.
 
-\subsubsection{Refine the Dataflow Graph.}
-\label{sec:refine}
-
-The refinement phase resolves the implementation for each operation node in the graph into a subgraph defining the details of the chosen algorithm.
-Each subgraph is an abstract representation of the loop that implements the given operation.
-In each case, the subgraph has an iteration strategy that specifies how to traverse the elements of the matrix or vector.
-
-Figure~\ref{fig:refine-mv-dot} shows two steps of refinement for a matrix-vector multiplication.
+%Figure~\ref{fig:refine-mv-dot} shows two steps of refinement for a matrix-vector multiplication.
 The first step expands the matrix-vector multiplication according to the mv-dot algorithm: $(Ax)(i) = A(i,:)x$.
@@
 %
-The second pass of refinement introduces another subgraph to implement the inner products.
+%The second pass of refinement introduces another subgraph to implement the inner products.
 The new subgraph iterates through each dense row, as indicated by the $j=1:n$ annotation.
@@
 
-\begin{figure}[hbtp]
-  \centering
-  \includegraphics[width=\textwidth]{figures/refine-mv-dot.png}
+%\begin{figure}[hbtp]
+%  \centering
+%  \includegraphics[width=\textwidth]{figures/refine-mv-dot.png}
 
-  \caption{The refinement of a matrix-vector product to a set of inner products and the refinement of inner products to scalar multiplications and additions.}
-  \label{fig:refine-mv-dot}
-\end{figure}
+%  \caption{The refinement of a matrix-vector product to a set of inner products and the refinement of inner products to scalar multiplications and additions.}
+%  \label{fig:refine-mv-dot}
+%\end{figure}
 
-\subsubsection{Optimize the Dataflow Graph.}
+%\subsubsection{Optimize the Dataflow Graph.}
 
-In the optimization step, the dataflow graph is optimized by applying conditional graph rewrite rules.
-One example rewrite rule is \emph{Merge Operand Sharing Subgraphs}: if two subgraphs share a common operand but do not depend on one another, and if the iteration strategies of the two subgraphs are compatible, then merge the subgraphs into a single subgraph.
-This rule is responsible for fusing the loops of the two matrix-vector products in the GEMVER kernel.
-Another example rule is \emph{Merge Dependent Subgraphs}: if one subgraph depends on another subgraph, and the iteration strategies of the two subgraphs are compatible, then merge them into a single subgraph.
-This rule, for example, is responsible for fusing the loops of the two outer products in the GEMVER kernel.
+In the optimization step, we apply conditional rewrite rules to optimize the
+dataflow graph, for example merging two subgraphs when they share a common
+operand.
+%but are not dependent on one another.
+%  One example rewrite rule is
+%\emph{Merge Operand Sharing Subgraphs}: if two subgraphs share a common
+%operand but do not depend on one another, and if the iteration strategies of
+%the two subgraphs are compatible, then merge the subgraphs into a single
+%subgraph.  
+This rule is responsible for fusing the loops of the two
+matrix-vector products in the GEMVER kernel.  
+%Another example rule is
+%\emph{Merge Dependent Subgraphs}: if one subgraph depends on another
+%subgraph, and the iteration strategies of the two subgraphs are compatible,
+%then merge them into a single subgraph.  This rule, for example, is
+%responsible for fusing the loops of the two outer products in the GEMVER
+%kernel.
 
-The choice of which optimizations to apply to various parts of the graph depends on the characteristics of the kernel, the algorithm, the matrix order and kind, and architectural features.
-The following is an instructive example: $y \gets A^T A (x + z + w)$.
-Assume that the compiler has already merged the two vector additions into one subgraph, call it $S_1$, and the two matrix products into another subgraph, call it $S_2$.
-The question then is whether it is profitable to apply the Merge Dependent Subgraphs rule one more time to merge $S_1$ and $S_2$.
-If the three vectors plus a row of $A$ do not fit in cache, then the merger will not be profitable because it will cause each row of $A$ to fall out of cache before the row gets used a second time.
-On the other hand, if all three vectors plus a row of $A$ do it in cache, then $S_1$ and $S_2$ should be merged.
-In general, we need a way to search through the many different combinations of optimization choices and choose a combination that minimizes memory traffic.
+%The choice of which optimizations to apply to various parts of the graph
+%depends on the characteristics of the kernel, the algorithm, the matrix order
+%and kind, and architectural features.
 
-\subsubsection{Translate the Dataflow Graph to C.}
+%The following is an instructive
+%example: $y \gets A^T A (x + z + w)$.  Assume that the compiler has already
+%merged the two vector additions into one subgraph, call it $S_1$, and the two
+%matrix products into another subgraph, call it $S_2$. The question then is
+%whether it is profitable to apply the Merge Dependent Subgraphs rule one more
+%time to merge $S_1$ and $S_2$. If the three vectors plus a row of $A$ do not
+%fit in cache, then the merger will not be profitable because it will cause
+%each row of $A$ to fall out of cache before the row gets used a second time.
+%On the other hand, if all three vectors plus a row of $A$ do it in cache,
+%then $S_1$ and $S_2$ should be merged.  In general, we need a way to search
+%through the many different combinations of optimization choices and choose a
+%combination that minimizes memory traffic.
 
-A graph that cannot be further refined has been reduced to a collection of subgraphs representing loops over scalar operations.
-The generator outputs a C loop for each subgraph.
-A topological sort is performed to determine the order in which the loops and scalar operations within the loops appear in the output.
+%\subsubsection{Translate the Dataflow Graph to C.}
+
+The final step performed by the MATLAB compiler when the graph cannot be
+refined further is the generation of C code. The generator outputs a C loop
+for each subgraph based on a topological sort of the graph.
+
 \subsection{Orio}
@@
 %~\cite{Norris:2007}
-Orio\cite{Norris:2007,ipdps09} is an empirical tuning tool that takes annotated C code as input,
-generates multiple transformed versions of the annotated code, and
-empirically evaluates the performance of the generated codes,
-ultimately choosing the best-performing version to use in production
+Orio\cite{Norris:2007,Hartono:IPDPS09} is an empirical tuning tool that takes
+annotated C code as input, generates multiple transformed versions of the
+annotated code, and empirically evaluates the performance of the generated
+codes, ultimately choosing the best-performing version to use in production
 runs.
@@
 \vspace{-.1in}
-\caption{Orio example: Annotated C source code (left) and tuning specifications for Blue Gene/P (right).}
+\caption{Orio example: Annotated C source code (left) and tuning specification excerpt for the Blue Gene/P (right).}
 \label{fig:orio-example}
@@
 
-\begin{minipage}[b]{.515\textwidth}
-\includegraphics[width=\textwidth]{figures/vadd_bgp.png}
+\begin{minipage}[b]{.5\textwidth}
+\includegraphics[height=2in,width=\textwidth]{figures/vadd_bgp.png}
 \end{minipage}
 &
-\begin{minipage}[b]{.515\textwidth}
-\includegraphics[width=\textwidth]{figures/atax.png}
+\begin{minipage}[b]{.5\textwidth}
+\includegraphics[height=2in,width=\textwidth]{figures/atax.png}
 \end{minipage}\\
@@
 
-\begin{minipage}[b]{.515\textwidth}
-\includegraphics[width=\textwidth]{figures/gemver.png}
+\begin{minipage}[b]{.5\textwidth}
+\includegraphics[height=2in,width=\textwidth]{figures/gemver.png}
 \end{minipage}
 &
-\begin{minipage}[b]{.515\textwidth}
-\includegraphics[width=\textwidth]{figures/bicgkernel.png}
+\begin{minipage}[b]{.5\textwidth}
+\includegraphics[height=2in,width=\textwidth]{figures/bicgkernel.png}
 \end{minipage}\\
@@
 }
-\caption{Performance numbers for different implementations of various composed of BLAS operations}
+\caption{Performance results for different implementations of several composed BLAS operations.}
 \label{fig:results}
@@
 
-We have described an approach to generating tuned linear algebra libraries from high-level annotated MATLAB code that involves a suite of tools to (1) translate the MATLAB code to C, (2) analyze the resulting loops and identify locatility and parallelism-enhancing optimizations using PLuTo, and (3) annotate the resulting C code with syntactic performance directives and use Orio to generate multiple optimized versions and empirically select the one with the best performance.
-Preliminary results from experiments with several composed BLAS operations show that in almost all cases, the optimized code generated by this suite of tools significantly outperforms the versions using tuned BLAS and aggressive compiler optimizations.
+We have described an approach to generating tuned linear algebra libraries
+from high-level annotated MATLAB code that involves a suite of tools to (1)
+translate the MATLAB code to C, (2) analyze the resulting loops and identify
+locatility and parallelism-enhancing optimizations using PLuTo, and (3)
+annotate the resulting C code with syntactic performance directives and use
+Orio to generate multiple optimized versions and empirically select the one
+with the best performance. Preliminary results from experiments with several
+composed BLAS operations show that the optimized code generated by this suite
+of tools significantly outperforms the versions using tuned BLAS and
+aggressive compiler optimizations.
 
-The positive initial results from our approach to generating tuned linear algebra routines motivate several future lines of investigation.
-At present there are a number of manual steps involved in obtaining a compiled tuned linear algebra library from the high-level extended MATLAB descriptions.
-We will work on interfacing our separately developed tools more closely and automating their interactions.
-This would enable application developers to easily produce custom linear algebra libraries tuned for specific architectures.
-We will also investigate extending the supported functionality of our MATLAB compiler beyond composed BLAS operations, thus benefiting a wider spectrum of applications.
+The positive initial results from our approach to generating tuned linear
+algebra routines motivate several future lines of investigation. At present
+there are a number of manual steps involved in obtaining a compiled tuned
+linear algebra library from the high-level extended MATLAB descriptions. We
+will work on the more close coupling of our separately developed tools and
+fully automating their interactions. This would enable application developers
+to easily produce custom linear algebra libraries tuned for specific
+architectures. We will also investigate extending the supported functionality
+of our MATLAB compiler beyond composed BLAS operations, thus benefiting a
+wider spectrum of applications.
 
-Orio is growing in a number of directions.
-While it automates much of the tuning process, a number of steps still require nontrivial manual input.
-Orio's modular design allows new transformations to be added easily, which presents many opportunities for expanding the existing set of optimizations with new ones implemented from scratch or through interfaces to other tools, as we have successfully done with PLuTo.
-We are also working on generating code that contains special cases for different input parameter values since in many cases different optimizations are optimal for different input sizes.
-We will also continue developing the parallel empirical search infrastructure; it is clear that more feedback between the MATLAB compiler and Orio will result in automation of the performance directive annotation process and a much more effective exploration of the search space.
-We plan to use a fast analytic cost function in the MATLAB compiler implementation to prune out regions of the search space that will not produce competitive implementations.
-The cost function is based on a model of the computer architecture, but does not necessarily account for all the details.
-Once the search space is narrowed, we will use the Orio empirical testing results with representative data sets to determine the exact performance of the alternatives and choose the best one.
+%Orio is growing in a number of directions. While it automates much of the
+%tuning process, a number of steps still require nontrivial manual
+%input. Orio's modular design allows new transformations to be added easily,
+%which presents many opportunities for expanding the existing set of
+%optimizations with new ones implemented from scratch or through interfaces to
+%other tools, as we have successfully done with PLuTo. We are also working on
+%generating code that contains special cases for different input parameter
+%values since in many cases different optimizations are optimal for different
+%input sizes.  We will also continue developing the parallel empirical search
+%infrastructure; it is clear that more feedback between the MATLAB compiler
+%and Orio will result in automation of the performance directive annotation
+%process and a much more effective exploration of the search space. We plan to
+%use a fast analytic cost function in the MATLAB compiler implementation to
+%prune out regions of the search space that will not produce competitive
+%implementations. The cost function is based on a model of the computer
+%architecture, but does not necessarily account for all the details. Once the
+%search space is narrowed, we will use the Orio empirical testing results with
+%representative data sets to determine the exact performance of the
+%alternatives and choose the best one.
 
-Finally, while Orio currently generates various loop transformations, it does not generate calls to existing library implementations.
-For this paper, we implemented the BLAS versions of all operations manually.
-Our past success with identifying Blue Gene/L intrinsics from expression graphs motivate further research on identifying BLAS and other established library operations from the expression graphs of the MATLAB code.
-\cite{Siek}
+%Finally, while Orio currently generates various loop transformations, it does
+%not generate calls to existing library implementations. For this paper, we
+%implemented the BLAS versions of all operations manually. Our past success
+%with identifying Blue Gene/L intrinsics from expression graphs motivate
+%further research on identifying BLAS and other established library operations
+%from the expression graphs of the MATLAB code.
+
 %\singlespacing

### 2008-12-14 src/module/ortil/transformator.py

@@ -327,7 +327,7 @@ class Transformator:
         # initialize all returned variables
         scan_stmts = []
         lbound_info_seq = []
-        new_int_vars = []
+        int_vars = []
 
         # generate the lower and upper values of each inter-tile loop
         val_table = {}
@@ -366,7 +366,7 @@ class Transformator:
             lb_name, ub_name = self.__getLoopBoundNames()
 
             # remember the new names
-            new_int_vars.extend([lb_name, ub_name])
+            int_vars.extend([lb_name, ub_name])
 
             # generate booleans to indicate the needs of prolog and epilog
             need_prolog, need_epilog = (len(lb_inames) > 0, len(ub_inames) > 0)
@@ -402,7 +402,7 @@ class Transformator:
             scan_stmts.append(ast.ExpStmt(a))
         
         # return all necessary information
-        return (scan_stmts, lbound_info_seq, new_int_vars)
+        return (scan_stmts, lbound_info_seq, int_vars)
 
     #----------------------------------------------
 
@@ -419,7 +419,7 @@ class Transformator:
         # initialize all returned variables
         scan_stmts = []
         lbound_info_seq = []
-        new_int_vars = []
+        int_vars = []
 
         # iterate over each statement to find loop bounds that are functions of outer loop iterators
         min_int = ast.NumLitExp(-2147483648, ast.NumLitExp.INT)
@@ -448,20 +448,36 @@ class Transformator:
                 lbound_info_seq.append(None)
                 continue
 
+            # check if this loop runs only once
+            is_one_time_loop = str(lb_exp) == str(ub_exp)
+
+            # generate booleans to indicate the needs of prolog, epilog, and main tiled loop
+            if is_one_time_loop:
+                need_main_tiled_loop = False
+                need_prolog = False
+                need_epilog = False
+            else:
+                need_main_tiled_loop = True
+                need_prolog = len(lb_inames) > 0
+                need_epilog = len(ub_inames) > 0
+
             # generate new variable names for both the new lower and upper loop bounds
-            lb_name, ub_name = self.__getLoopBoundNames()
-
-            # remember the new names
-            new_int_vars.extend([lb_name, ub_name])
-
-            # generate booleans to indicate the needs of prolog and epilog
-            need_prolog, need_epilog = (len(lb_inames) > 0, len(ub_inames) > 0)
+            if need_main_tiled_loop:
+                lb_name, ub_name = self.__getLoopBoundNames()
+                int_vars.extend([lb_name, ub_name])
+            else:
[…]

### 2008-12-14 src/module/ortil/ast.py

@@ -286,6 +286,11 @@ class ForStmt(Stmt):
         self.iter = iter      # may be null
         self.stmt = stmt
 
+        # this is a quick hack! this is a field with a string value, used to mark loops. 
+        # an instance of use of this field is to mark whether this loop nest iterates 
+        # full rectangular tiles.
+        self.label = ''
+
     def replicate(self):
         '''Replicate this abstract syntax tree node'''
         r_in = self.init
@@ -297,5 +302,7 @@ class ForStmt(Stmt):
             r_ts = r_ts.replicate()
         if r_it:
             r_it = r_it.replicate()
-        return ForStmt(r_in, r_ts, r_it, self.stmt.replicate(), self.line_no)
+        f = ForStmt(r_in, r_ts, r_it, self.stmt.replicate(), self.line_no)
+        f.label = self.label
+        return f
 

### 2008-12-14 src/module/ortil/transformator.py

@@ -110,9 +110,10 @@ class Transformator:
         # used for variable name generation
         self.counter = 1
 
-        # to turn on/off the static determination of loop bound values for scanning the full tiles
-        self.affine_lbound_exps = False
-
+        # booleans to switch on/off some (not all) of the complementary optimizations
+        self.affine_lbound_exps = True        # to use static loop-bound value determination
+        self.simplify_one_time_loop = True    # to use one-time loop simplification
+        
     #----------------------------------------------
 
     def __getIterName(self, iter_name, level):
@@ -362,18 +363,34 @@ class Transformator:
                 lbound_info_seq.append(None)
                 continue
 
+            # check if this loop runs only once
+            is_one_time_loop = str(lb_exp) == str(ub_exp)
+
+            # generate booleans to indicate the needs of prolog, epilog, and main tiled loop
+            if is_one_time_loop:
+                need_main_tiled_loop = False
+                need_prolog = False
+                need_epilog = False
+            else:
+                need_main_tiled_loop = True
+                need_prolog = len(lb_inames) > 0
+                need_epilog = len(ub_inames) > 0
+
             # generate new variable names for both the new lower and upper loop bounds
-            lb_name, ub_name = self.__getLoopBoundNames()
-
-            # remember the new names
-            int_vars.extend([lb_name, ub_name])
-
-            # generate booleans to indicate the needs of prolog and epilog
-            need_prolog, need_epilog = (len(lb_inames) > 0, len(ub_inames) > 0)
+            if need_main_tiled_loop:
+                lb_name, ub_name = self.__getLoopBoundNames()
+                int_vars.extend([lb_name, ub_name])
+            else:
+                lb_name = ''
+                ub_name = ''
 
             # append information about the new loop bounds
-            lbinfo = (lb_name, ub_name, need_prolog, need_epilog)
+            lbinfo = (lb_name, ub_name, need_prolog, need_epilog, need_main_tiled_loop)
             lbound_info_seq.append(lbinfo)
+
+            # skip generating loop-bound scanning code (if it's a one-time loop)
+            if not need_main_tiled_loop:
+                continue
 
             # determine the value of the new lower loop bound
             if str(lb_exp) in lb_exps_table:
@@ -597,6 +614,32 @@ class Transformator:
 
     #----------------------------------------------
 
+    def __labelFullCoreTiledLoop(self, stmt, outer_loop_inames):
[…]

### 2008-12-14 src/module/ortil/pprinter.py
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@389 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@ -132,6 +132,8 @@ class PrettyPrinter:
                     s += self.pprint(tnode.false_stmt, indent + extra_indent, extra_indent)
 
         elif isinstance(tnode, ast.ForStmt):
+            if tnode.label:
+                s += indent + '/* %s */\n' % tnode.label
             s += indent + 'for ('
             if tnode.init:
                 s += self.pprint(tnode.init, indent, extra_indent)

### 2008-12-14 src/module/ortil/ortil.py
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@390 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@ -32,20 +32,20 @@ class OrTil(module.module.Module):
         stmts = semant.SemanticAnalyzer(tiling_info).analyze(stmts)
 
         # perform loop-tiling transformation
-        t = transformator.Transformator(self.perf_params, tiling_info)
-        (stmts, new_int_vars) = t.transform(stmts)
+        t = transformator.Transformator(tiling_info)
+        (stmts, int_vars) = t.transform(stmts)
 
         # generate the declaration code for the newly declared integer variables
         code = ''
-        for i,iv in enumerate(new_int_vars):
+        for i,iv in enumerate(int_vars):
             if i%8 == 0:
                 code += '\n  int '
             code += iv
-            if i%8 == 7 or i == len(new_int_vars)-1:
+            if i%8 == 7 or i == len(int_vars)-1:
                 code += ';'
             else:
                 code += ','
-        if new_int_vars:
+        if int_vars:
             code += '\n\n'
 
         # generate the tiled code

### 2008-12-14 src/module/ortil/semant.py

@@ -13,12 +13,10 @@ class SemanticAnalyzer:
     def __init__(self, tiling_info):
         '''To instantiate a semantic analyzer'''
 
-        num_level, tiling_table = tiling_info
-        tiled_loop_inames = tiling_table.keys()
-        tile_size_table = tiling_table
+        num_level, iter_names = tiling_info
 
         self.num_level = num_level
-        self.tiled_loop_inames = tiled_loop_inames
+        self.iter_names = iter_names
         self.ast_util = ast_util.ASTUtil()
 
     #-----------------------------------------------------
@@ -105,7 +103,7 @@ class SemanticAnalyzer:
                 print ('error:OrTil: illegal loop nest where an inner loop has the same iterator ' +
                        'name as the outer loop')
                 sys.exit(1)
-            if id.name not in self.tiled_loop_inames:
+            if id.name not in self.iter_names:
                 print 'error:OrTil: missing tiled-loop iterator name: "%s"' % id.name
                 sys.exit(1)
             self.__checkStmt(stmt.stmt, oloop_inames + [id.name])

### 2008-12-14 src/module/ortil/transformator.py

@@ -94,17 +94,13 @@ class SimpleLoops:
 class Transformator:
     '''The code transformator that performs loop tiling'''
 
-    def __init__(self, perf_params, tiling_info):
+    def __init__(self, tiling_info):
         '''To instantiate a code transformator'''
 
-        num_level, tiling_table = tiling_info
-        tiled_loop_inames = tiling_table.keys()
-        tile_size_table = tiling_table
-        
-        self.perf_params = perf_params
+        num_level, iter_names = tiling_info
+
         self.num_level = num_level
-        self.tiled_loop_inames = tiled_loop_inames
-        self.tile_size_table = tile_size_table
+        self.iter_names = iter_names
         self.ast_util = ast_util.ASTUtil()
 
         # used for variable name generation
@@ -626,13 +622,18 @@ class Transformator:
         while True:
             if not isinstance(s, ast.ForStmt):
                 return
-            if s.label and s.label.startswith('full core tiles'):
+            if s.start_label and s.start_label.startswith('start full core tiles'):
                 return 
             id, lb_exp, ub_exp, st_exp, lbody = self.ast_util.getForLoopInfo(s)
             if id.name == outer_loop_inames[0]:
-                s.label = 'full core tiles:'
-                for i in outer_loop_inames:
-                    s.label += ' %s' % i
+                s.start_label = 'start full core tiles: '
+                s.end_label = 'end full core tiles: '
+                for i,iname in enumerate(outer_loop_inames):
+                    if i:
+                        s.start_label += ','
+                        s.end_label += ','
+                    s.start_label += '%s' % iname
+                    s.end_label += '%s' % iname
                 return
             if len(s.stmt.stmts) != 1:
                 return
@@ -938,8 +939,8 @@ class Transformator:
 
         # add new integer variable names used for both inter-tile and intra-tile loop iterators
         int_vars = []
-        int_vars.extend(self.tiled_loop_inames)
-        for iname in self.tiled_loop_inames:
+        int_vars.extend(self.iter_names)
+        for iname in self.iter_names:
             for level in range(1, self.num_level+1):
                 int_vars.append(self.__getIterName(iname, level))
 

### 2008-12-14 src/module/ortil/ann_parser.py
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@390 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@ -49,8 +49,8 @@ class AnnParser:
 
         # initialize the data structure to store all tiling information
         num_level = 1
-        tiling_table = {}
-        tiling_info = [num_level, tiling_table]
+        iter_names = []
+        tiling_info = [num_level, iter_names]
 
         # get all iterator names of the loops to be tiled
         m = re.match(__oparenth_re, text)
@@ -64,7 +64,6 @@ class AnnParser:
             sys.exit(1)
         itext = text[:m.end()-1]
         text = text[m.end():]
-        iter_names = [] 
         while True:
             if (not itext) or itext.isspace():
                 break
@@ -81,10 +80,6 @@ class AnnParser:
             m = re.match(__comma_re, itext)
             if m:
                 itext = itext[m.end():]
-
-        # insert all obtained iterator names into the tiling table
-        for i in iter_names:
-            tiling_table[i] = []
 
         # check if further parsing is needed
         if (not text) or text.isspace():
@@ -116,78 +111,10 @@ class AnnParser:
         # insert the obtained number of tiling levels into the tiling information
         tiling_info[0] = num_level
 
-        # check if further parsing is needed
-        if (not text) or text.isspace():
-            return tiling_info
-
-        # get a colon
-        m = re.match(__colon_re, text)
-        if not m:
-            print 'error:OrTil: annotation syntax error: "%s"' % orig_text
-            sys.exit(1)
-        text = text[m.end():]
-
-        # get the tile sizes
-        tile_size_sets = []
-        for tlevel in range(0,num_level):
-            m = re.match(__oparenth_re, text)
-            if not m:
-                print 'error:OrTil: annotation syntax error: "%s"' % orig_text
-                sys.exit(1)
-            text = text[m.end():]        
-            m = re.search(__cparenth_re, text)
-            if not m:
-                print 'error:OrTil: annotation syntax error: "%s"' % orig_text
-                sys.exit(1)
-            itext = text[:m.end()-1]
-            text = text[m.end():]
-            m = re.match(__comma_re, text)
-            if m:
-                text = text[m.end():]
-            tile_size_sets.append([])
-            for iname in iter_names:
-                m = re.match(__num_re, itext)
-                if not m:
-                    m = re.match(__var_re, itext)
-                if not m:
-                    print 'error:OrTil: annotation syntax error: "%s"' % orig_text
-                    sys.exit(1)
-                tsize = self.__evalExp(m.group(1))
[…]

### 2008-12-14 src/module/ortil/ast.py

@@ -289,7 +289,8 @@ class ForStmt(Stmt):
         # this is a quick hack! this is a field with a string value, used to mark loops. 
         # an instance of use of this field is to mark whether this loop nest iterates 
         # full rectangular tiles.
-        self.label = ''
+        self.start_label = ''
+        self.end_label = ''
 
     def replicate(self):
         '''Replicate this abstract syntax tree node'''
@@ -303,6 +304,7 @@ class ForStmt(Stmt):
         if r_it:
             r_it = r_it.replicate()
         f = ForStmt(r_in, r_ts, r_it, self.stmt.replicate(), self.line_no)
-        f.label = self.label
+        f.start_label = self.start_label
+        f.end_label = self.end_label
         return f
 

### 2008-12-14 src/module/ortil/pprinter.py

@@ -132,8 +132,8 @@ class PrettyPrinter:
                     s += self.pprint(tnode.false_stmt, indent + extra_indent, extra_indent)
 
         elif isinstance(tnode, ast.ForStmt):
-            if tnode.label:
-                s += indent + '/* %s */\n' % tnode.label
+            if tnode.start_label:
+                s += indent + '/* %s */\n' % tnode.start_label
             s += indent + 'for ('
             if tnode.init:
                 s += self.pprint(tnode.init, indent, extra_indent)
@@ -150,6 +150,8 @@ class PrettyPrinter:
             else:
                 s += '\n'
                 s += self.pprint(tnode.stmt, indent + extra_indent, extra_indent)
+            if tnode.end_label:
+                s += indent + '/* %s */\n' % tnode.end_label
 
         else:
             print 'internal error:OrTil: unrecognized type of AST: %s' % tnode.__class__.__name__

### 2008-12-14 testsuite/vadd/bgp/runjobs.py
vadd results on the BG/P

@@ -1,0 +1,120 @@
+#!/usr/bin/env python
+
+import os, sys, time, random, re
+
+def runExp(ename, flag, src_file, libs, numthreads):
+
+    print '-=-=-=-=-=-=-=-=-=-=-=-=-'
+    print ename
+    print '-=-=-=-=-=-=-=-=-=-=-=-=-'
+
+    CC = 'bgxlc_r -O3 -qstrict -qarch=450d -qtune=450 -qhot' 
+
+    REPS = [100000,10000,10000,1000,1000,500,500,100,50,20]
+    N = [10,100,1000,10000,50000,100000,500000,1000000,5000000,10000000]
+    T = [5,5,5,5,5,10,10,10,15,15]
+
+    # iterate over each problem size
+    jobs_queue = []
+    for reps, n, t in zip(REPS,N,T):
+
+        # generate a random integer
+        i = random.randint(1,100000)
+
+        # build the code
+        build_cmd = ('%s %s -DN=%s -DREPS=%s -o %s%s %s %s' %
+                     (CC, flag, n, reps, ename, i, src_file, libs))
+        print '********************************'
+        print build_cmd
+        print '********************************'
+        os.system(build_cmd)
+        
+        # submit a job
+        batch_cmd = ('qsub -n 1 -t %s -q short --env "OMP_NUM_THREADS=%s" ./%s%s' %
+                     (t, numthreads, ename, i))
+        print '********************************'
+        print batch_cmd
+        print '********************************'
+        f = os.popen(batch_cmd)
+        output = f.read()
+        f.close()
+            
+        # get the job ID
+        job_id = output.strip().split('\n')[-1]
+        print job_id
+
+        # insert job into the queue
+        jobs_queue.append([job_id, reps, n, t])
+
+    # iterate over each job stored in the queue
+    for job_id, reps, n, t in jobs_queue:
+        
+        # wait till the job is done
+        while 1:
+            print '.',
+            time.sleep(10)
+            status_cmd = 'qstat | grep %s | wc -l' % job_id
+            f = os.popen(status_cmd)
+            status = f.read().strip()
+            f.close()
+            if status == '0':
+                break
+        print ''
+
[…]

### 2008-12-14 testsuite/vadd/vadd.c

@@ -1,0 +1,20 @@
+
+void vadd(int n, double *x, double *w, double *y, double *z) {
+
+    register int i;
+
+/*@ begin Align (x1[],x2[],x3[],x4[],y[]) @*/
+/*@ begin Loop (
+  transform Unroll(ufactor=20, parallelize=True)
+    for (i=0; i<=n-1; i++)
+      x[i]=w[i]+y[i]+z[i];
+      ) @*/
+
+    for (i=0; i<=n-1; i++)
+        x[i]=w[i]+y[i]+z[i];
+
+/*@ end @*/
+/*@ end @*/
+
+}
+

### 2008-12-18 doc/papers/iccs09/paper.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@387 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb
Gail's edits to abstract
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@401 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@
-% ICCS'09 http://www.iccs-meeting.org/iccs2009/
+% ICCS'09 http://www.iccs-meeting.org/iccs2009/}
 % svn co https://svn.mcs.anl.gov/repos/performance/orio/doc/papers/iccs09
@@
 % first the title is needed
-\title{Generating Empirically Optimized Composed Matrix Kernles from MATLAB Prototypes}
+\title{Generating Empirically Optimized Composed Matrix Kernels from MATLAB Prototypes}
 
@@
 
-\author{Elizabeth Jessup\inst{1} 
-  \and Boyana Norris\inst{2}
-  \and Albert Hartono\inst{3}
-  \and Jeremy Siek\inst{1}}
+\author{Boyana Norris\thanks{The work of this author was supported by the Office of Advanced Scientific Computing Research, Office of Science, U.S.
+Dept.
+of Energy, under Contract DE-AC02-06CH11357.}\inst{1}
+  \and Albert Hartono\inst{2}
+  \and Elizabeth Jessup\thanks{The work of this author was supported by the
+  National Science Foundation under grants no. CCF-0430646 and CCF-830458.}\inst{3}
+  \and Jeremy Siek\inst{3}
+%
+\authorrunning{Norris, Hartono, Jessup, Siek}
+%
+\institute{Argonne National Laboratory \\
+  \email{norris@mcs.anl.gov}
+  \and Ohio State University \\
+  \email{hartonoa@cse.ohio-state.edu}
+  \and University of Colorado at Boulder \\
+  \email{\{elisabeth.jessup,jeremy.siek\}@colorado.edu}}}
 
-\authorrunning{Jessup, Norris, Hartono, Siek}
-
-\institute{University of Colorado at Boulder \\
-  \email{\{elisabeth.jessup,jeremy.siek\}@colorado.edu}
-  \and Argonne National Laboratory \\
-  \email{\{norris\}@mcs.anl.gov}
-  \and Ohio State University \\
-  \email{\{hartonoa\}@cse.ohio-state.edu}}
-
-\tocauthor{Elizabeth Jessup, Boyana Norris, Albert Hartono, and Jeremy Siek}
+\tocauthor{B. Norris, A. Hartono, E. Jessup, and J. Siek}
 \maketitle
@@
 \label{fig:compiler}
-\vspace{-.3in}
+\vspace{-.2in}
 \end{wrapfigure}
@@
 \label{fig:gemver-dataflow}
+\vspace{-.2in}
 \end{figure}
@@
 \begin{figure}[tb]
-\vspace{-.1in}
+\vspace{-.3in}
 \begin{center}
@@
 \label{fig:orio}
+\vspace{-.2in}
 \end{figure}
@@
 \vspace{-.1in} 
-\caption{Composed BLAS operations used in our experiments}
+\caption{Composed BLAS operations used in our experiments.}
 \label{tbl:blas-ops}
@@
 \centering 
-\vspace{-.1in} 
+%\vspace{-.1in} 
 \begin{tabular}{|c|c|c|c|} 
@@
 \hline 
+%& & & \\
 VADD & $\overrightarrow{w}$,$\overrightarrow{y}$,$\overrightarrow{z}$ & $\overrightarrow{x}$ &
-$\overrightarrow{x} = \overrightarrow{w} + \overrightarrow{y} + \overrightarrow{z}$ \\ 
+$\overrightarrow{x} = \overrightarrow{w} + \overrightarrow{y} + \overrightarrow{z}$ \\ [1ex]
 \hline 
+%& & & \\
 ATAX & $A$,$\overrightarrow{x}$ & $\overrightarrow{y}$ & 
-$\overrightarrow{y} = A' * (A * \overrightarrow{x})$ \\ 
+$\overrightarrow{y} = A' * (A * \overrightarrow{x})$ \\ [1ex]
 \hline 
- & ${\bf A}$,$a$,$b$, & ${\bf B}$, & ${\bf B} = {\bf A} + \overrightarrow{u_1} * \overrightarrow{v_1}' + \overrightarrow{u_2} * \overrightarrow{v_2}'$\\
-GEMVER & $\overrightarrow{u_1}$,$\overrightarrow{u_2}$,$\overrightarrow{v_1}$,$\overrightarrow{v_2}$, & $\overrightarrow{x}$,$\overrightarrow{w}$ & $\overrightarrow{x} = b * ({\bf B}' * \overrightarrow{y}) + \overrightarrow{z}$\\
-& $\overrightarrow{y}$,$\overrightarrow{z}$ & & $\overrightarrow{w} = a * ({\bf B} * \overrightarrow{x})$\\
+%& & & \\
+ & ${\bf A}$,$a$,$b$, &  & ${\bf B} = {\bf A} + \overrightarrow{u_1} * \overrightarrow{v_1}' + \overrightarrow{u_2} * \overrightarrow{v_2}'$\\
+GEMVER & $\overrightarrow{u_1}$,$\overrightarrow{u_2}$,$\overrightarrow{v_1}$,$\overrightarrow{v_2}$, & ${\bf B}$,  & $\overrightarrow{x} = b * ({\bf B}' * \overrightarrow{y}) + \overrightarrow{z}$\\
+& $\overrightarrow{y}$,$\overrightarrow{z}$ & $\overrightarrow{x}$,$\overrightarrow{w}$ & $\overrightarrow{w} = a * ({\bf B} * \overrightarrow{x})$\\ [1ex]
 \hline 
-BiCG Kernel & ${\bf A}$,$\overrightarrow{p}$,$\overrightarrow{r}$ & $\overrightarrow{q}$,$\overrightarrow{s}$ & $\overrightarrow{q} = {\bf A} * \overrightarrow{p}$ \\
-& & & $\overrightarrow{s} = {\bf A}' * \overrightarrow{r}$\\
+& & & $\overrightarrow{q} = {\bf A} * \overrightarrow{p}$ \\
+BiCG Kernel & ${\bf A}$,$\overrightarrow{p}$,$\overrightarrow{r}$ & $\overrightarrow{q}$,$\overrightarrow{s}$ & $\overrightarrow{s} = {\bf A}' * \overrightarrow{r}$\\ [1ex]
 \hline 
 \end{tabular} 
-\vspace{-.2in} 
+%\vspace{-.2in} 
 \end{table}
@@
 \begin{minipage}[b]{.5\textwidth}
-\includegraphics[height=2in,width=\textwidth]{figures/vadd_bgp.png}
+\includegraphics[width=\textwidth]{figures/vadd_bgp.png}
 \end{minipage}
@@
 \begin{minipage}[b]{.5\textwidth}
-\includegraphics[height=2in,width=\textwidth]{figures/atax.png}
+\includegraphics[width=\textwidth]{figures/atax.png}
 \end{minipage}\\
@@
 \begin{minipage}[b]{.5\textwidth}
-\includegraphics[height=2in,width=\textwidth]{figures/gemver.png}
+\includegraphics[width=\textwidth]{figures/gemver.png}
 \end{minipage}
@@
 \begin{minipage}[b]{.5\textwidth}
-\includegraphics[height=2in,width=\textwidth]{figures/bicgkernel.png}
+\includegraphics[width=\textwidth]{figures/bicgkernel.png}
 \end{minipage}\\
@@
 numbers for six code variants: the C code generated by the MATLAB
-compiler (''C from MATLAB''), three BLAS-based implementations that
+compiler (``C from MATLAB''), three BLAS-based implementations that
 use Intel MKL, ATLAS, and the default BLAS library on Ubuntu 8.04, and
@@
 
+%\newpage
+%\vfill
+%\begin{flushright}
+%\scriptsize
+%\framebox{\parbox{2.4in}{The submitted manuscript has been created by
+%UChicago Argonne, LLC, Operator of Argonne National Laboratory
+%("Argonne").  Argonne, a U.S. Department of Energy Office
+%of Science laboratory, is operated under Contract No.
+%DE-AC02-06CH11357.  The U.S. Government retains for itself, and
+%others acting on its behalf, a paid-up, nonexclusive, irrevocable
+%worldwide license in said article to reproduce, prepare derivative works,
+%distribute copies to the public, and perform publicly and display
+%publicly, by or on behalf of the Government.}}
+%\normalsize
+%\end{flushright}
+
 \end{document}

### 2008-12-18 doc/papers/iccs09/abstract.tex
fits in 10 pages, more editing needed, though
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@386 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb
Gail's edits to abstract
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@401 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@
 The development of optimized codes is time-consuming and requires extensive
-architecture, compiler, and language expertise; therefore, computational
-scientists are often forced to choose between investing too much time in
+architecture, compiler, and language expertise, therefore, computational
+scientists are often forced to choose between investing considerable time in
 tuning code or accepting lower performance.
@@
 for coarse-grained parallelism and locality simultaneously. Finally, we
-annotate the resulting code with performance tuning directives and use the
-empirical performance tuning system Orio to generate many tuned versions of
+annotate the resulting code with performance-tuning directives and use the
+empirical performance-tuning system Orio to generate many tuned versions of
 the same operation using different optimization techniques, such as loop
@@
 %based on source BLAS, ATLAS-optimized BLAS, and Intel MKL routines.
-\keywords{Code generation, empirical performance tuning}
+\keywords{MATLAB, code generation, empirical performance tuning}
 \end{abstract}

### 2008-12-18 doc/papers/iccs09/paper.tex
removed formatting hacks, cut down to 10 pages again.
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@402 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@
 % formatting cheats
-\renewcommand{\baselinestretch}{0.92}  
-\renewcommand\floatpagefraction{.9} 
-\renewcommand\topfraction{.9} 
-\renewcommand\bottomfraction{.9} 
-\renewcommand\textfraction{.1} 
-\setcounter{totalnumber}{50} 
-\setcounter{topnumber}{50} 
-\setcounter{bottomnumber}{50} 
-\setlength{\textwidth}{12.5cm} 
-\setlength{\textheight}{19.5cm}
+%\renewcommand{\baselinestretch}{0.92}  
+%\renewcommand\floatpagefraction{.9} 
+%\renewcommand\topfraction{.9} 
+%\renewcommand\bottomfraction{.9} 
+%\renewcommand\textfraction{.1} 
+%\setcounter{totalnumber}{50} 
+%\setcounter{topnumber}{50} 
+%\setcounter{bottomnumber}{50} 
+%\setlength{\textwidth}{12.5cm} 
+%\setlength{\textheight}{19.5cm}
 
@@
   \and University of Colorado at Boulder \\
-  \email{\{elisabeth.jessup,jeremy.siek\}@colorado.edu}}}
+  \email{\{elizabeth.jessup,jeremy.siek\}@colorado.edu}}}
 
@@
 The development of high-performance numerical codes is challenging because
-performance is determined by complex interactions between the algorithm, data
+performance is determined by complex interactions among the algorithm, data
 structure, programming language, compiler, and computer
@@
 must either invest substantial time in tuning their software or accept low
-performance. In either case, the productivity of the scientist degrades.
+performance. In either case, the productivity of the scientists degrades.
 
@@
 limited and varies from compiler to compiler, and (iv) apart from differences
-in execution time, it is difficult for users to evaluate the effectiveness of
+in execution time, it is difficult to evaluate the effectiveness of
 different compiler optimizations.
 
-The second path to performance optimization is to identify kernel
+When compilers alone cannot achieve the desired performance, 
+another path to performance optimization is to identify kernel
 routines that dominate the execution time of a wide variety of
@@
 %architectures of interest. 
-An example is the high-performance BLAS libraries produced by a combination
+An example is the high-performance Basic Linear Algebra Subprograms 
+(BLAS) libraries~\cite{BLAS} produced by a combination
 of hardware vendors, independent software vendors, and researchers.
 Developers who write their codes calling these routines can achieve high
-performance across all supported architectures. The efforts in
-high-performance library development aptly demonstrate, however, that the
-research into compiler technologies has not been fully successful.  If
-compilers could take the relatively simple code such as a dense matrix-matrix
-product and produce optimal code, this second approach would be unnecessary.
-Like compiler optimization, the kernel library-oriented approach has
-significant limitations.  Support for a variety of architectures is the most
-serious problem because the optimizations necessary to achieve near-peak
-performance are by nature not portable, requiring manual update or
-reimplementation for each new architecture.
+performance across all supported architectures, but are also subject to 
+the limitations of the library (e.g., portability and the types of operations available).
+%The efforts in
+%high-performance library development aptly demonstrate, however, that the
+%research into compiler technologies has not been fully successful.  If
+%compilers could take the relatively simple code such as a dense matrix-matrix
+%product and produce optimal code, this second approach would be unnecessary.
+%Like compiler optimization, the kernel library-oriented approach has
+%significant limitations.  Support for a variety of architectures is the most
+%serious problem because the optimizations necessary to achieve near-peak
+%performance are by nature not portable, requiring manual update or
+%reimplementation for each new architecture.
 
 This paper describes a combination of the two approaches designed to overcome
-some of their shortcomings. We take a multipronged approach
-toward the development of software infrastructure for \emph{generating}
-automatically tuned libraries for matrix algebra computations.  We briefly
-overview relevant prior and current research efforts in
-Section~\ref{sec:background}. In Section~\ref{sec:approach} we describe our
-MATLAB~\cite{matlab:webpage} to C compiler and the empirical performance tuning system Orio and its
-use in conjunction with the PLuTo tool suite to generate and empirically
-evaluate many tuned versions of the C code generated by the MATLAB compiler.
-In Section~\ref{sec:experiments}, we provide performance results showing that
-the code generated by using our system significantly outperforms not only the
-original simple C code but also code based on source BLAS, ATLAS-optimized
-BLAS, and Intel MKL routines.
+some of their shortcomings. We describe our initial efforts toward the
+development of software infrastructure for \emph{generating} automatically
+tuned libraries for matrix algebra computations.  In
+Section~\ref{sec:background} we briefly discuss relevant prior and current
+research efforts. In Section~\ref{sec:approach} we describe our
+MATLAB-to-C compiler and the empirical performance
+tuning system Orio and its use in conjunction with the PLuTo tool suite to
+generate and empirically evaluate many tuned versions of the C code generated
+by the MATLAB compiler.  In Section~\ref{sec:experiments} we provide
+performance results on two architectures. We conclude with a summary and
+discussion future work in Section~\ref{sec:conclusion}.
+
+%showing that the code generated by using our system
+%significantly outperforms not only the original simple C code but also code
+%based on source BLAS, ATLAS-optimized BLAS, and Intel MKL routines.
 %Text in intro repeats abstract some, but that might not be a big deal for a conf paper.
@@
 
-Existing optimizing MATLAB compilers, such as the MaJIC MATLAB compiler~\cite{MaJIC},
-include limited local optimizations for matrix expressions, but do not
-perform optimizations such as loop fusion across multiple operations as we do
-with the tools described in this paper. The telescoping languages
+Existing optimizing MATLAB~\cite{matlab:webpage} compilers, such as the MaJIC
+MATLAB compiler~\cite{MaJIC}, include limited local optimizations for matrix
+expressions, but do not perform optimizations such as loop fusion across
+multiple operations as we do with the tools described in this paper. The
+telescoping languages
 %~\cite{telescopingurl,teleoverview,Ken99} 
@@
 and similar
-sparse matrix libraries~\cite{Saad:fr,George:1981uq}. Writing programs in
+sparse matrix libraries~\cite{Saad:fr}.
+%,George:1981uq}. 
+Writing programs in
 this way promotes readability and maintainability but can be costly in terms
@@
 operations than to make successive calls to separate BLAS routines (e.g., see
-~\cite{gropp01,Vuduc:2003kl}.
+~\cite{gropp01}.
 %~\cite{Ashby:uq,Blackford:2002vn,baker03blgmres,baker03lgmres,Dennis:2005tg,Howell:2008,gropp01,Vuduc:2003kl}.
@@
 $q$, $r$, and $s$ are vectors, that comprise the computational bottlenecks of
-the BiConjugate Gradient Method~\cite{Barrett:1994kx,Saad:2003fk} and of the
+the BiConjugate Gradient Method~\cite{Saad:2003fk} and of the
 GEMVER kernel examined in Section~\ref{sec:matlab}.  These two operations can
@@
 
-To generate highly optimized code from a MATLAB prototype of the composed BLAS operation, we follow a three-step approach, illustrated in Figure~\ref{fig:process}.  To begin,
-we have developed a compiler that converts a MATLAB script into simple C code~\cite{Siek}.
-After generating the C code from the high-level MATLAB prototype, 
+To generate highly optimized code from a MATLAB prototype of the composed
+BLAS operation, we follow a three-step approach, illustrated in
+Figure~\ref{fig:process}.  
 %
@@
 %
-we (optionally) use the source-to-source automatic parallelization tool PLuTo~\cite{Pluto} %developed at Ohio State University
+To begin, we have developed a compiler that
+converts a MATLAB script into simple C code~\cite{Siek}.  After generating
+the C code from the high-level MATLAB prototype, we (optionally) use 
+the source-to-source automatic parallelization tool PLuTo~\cite{Pluto} %developed at Ohio State University
 to optimize for coarse-grained parallelism and locality simultaneously. Using
@@
 In this section we describe the design and operation of the MATLAB to C compiler we have developed~\cite{Siek}.    Figure~\ref{fig:compiler} gives an overview of the compilation process. 
+The MATLAB kernel specification is parsed into a high-level intermediate
+representation in the form of a dataflow graph, in which each node represents
 %
 \begin{wrapfigure}{r}{2in}
-\vspace{-.2in}
+%\vspace{-.2in}
 \centering
@@
 \end{wrapfigure}
-The MATLAB kernel specification is parsed into a high-level intermediate
-representation in the form of a dataflow graph.  This dataflow graph is then
-iteratively processed until all of the implementation choices have been made.
-The compilation process consists of three phases: analysis, refinement, and
-optimization, which are together iterated until all of the implementation
-decisions have been made.  The graph is then translated into C code.
+%
+a parameter (e.g., a scalar, matrix, or vector variable) of the kernel or an
+operation.  This dataflow graph is then iteratively processed until all of
+the implementation choices have been made.  The compilation process consists
+of three phases: analysis, refinement, and optimization, which are together
+iterated until all of the implementation decisions have been made.  The graph
+is then translated into C code.
+
 %\subsubsection{The Dataflow Graph.}
 %
-An example dataflow graph for the GEMVER kernel, defined below, is shown
-in Figure~\ref{fig:gemver-dataflow}.
-Each node represents a parameter of the kernel or an operation.  The arrows
-indicate the flow of data. At first the graph specifies what operations are
-to be performed but does not contain any implementation details. The symbol
-$\times$ in the depicted graph stands variously for outer product,
-matrix-vector multiplication, and scalar-vector multiplication, and it does
-not yet specify, for example, whether the outer products compute a row or a
-column at a time of the result matrix.
-\begin{figure}[ht]
-\begin{minipage}{.3\textwidth}
-\begin{align*}
-  A &\gets A + u_1 v_1^T + u_2 v_2^T \\[-0.5ex]
-  x &\gets \beta A^T y + z \\[-0.5ex]
-  w &\gets \alpha A x
-\end{align*} 
-\end{minipage}
-\begin{minipage}{.6\textwidth}
-\includegraphics[width=3in]{figures/gemver-dataflow.png}
-\end{minipage}
-\caption{Dataflow graph (right) for the GEMVER kernel (left).}
-\label{fig:gemver-dataflow}
-\vspace{-.2in}
-\end{figure}
+%An example dataflow graph for the GEMVER kernel, defined below, is shown
+%in Figure~\ref{fig:gemver-dataflow}.
+%Each node represents a parameter of the kernel or an operation.  The arrows
+%indicate the flow of data. At first the graph specifies what operations are
+%to be performed but does not contain any implementation details. The symbol
+%$\times$ in the depicted graph stands variously for outer product,
+%matrix-vector multiplication, and scalar-vector multiplication, and it does
+%not yet specify, for example, whether the outer products compute a row or a
+%column at a time of the result matrix.
+%\begin{figure}[ht]
+%\begin{minipage}{.3\textwidth}
+%\begin{align*}
+%  A &\gets A + u_1 v_1^T + u_2 v_2^T \\[-0.5ex]
+%  x &\gets \beta A^T y + z \\[-0.5ex]
+%  w &\gets \alpha A x
+%\end{align*} 
+%\end{minipage}
+%\begin{minipage}{.6\textwidth}
+%\includegraphics[width=3in]{figures/gemver-dataflow.png}
+%\end{minipage}
+%\caption{Dataflow graph (right) for the GEMVER kernel (left).}
+%\label{fig:gemver-dataflow}
+%\vspace{-.2in}
+%\end{figure}
 
@@
 
-%Once
-%we add backtracking we will push each alternative onto the stack to be
-%explored later. Once the choice is made, the kind and storage format labels
-%for the operation node itself and for the input nodes are updated with the
-%labels specified in the linear algebra database for the chosen algorithm. The
-%algorithm then goes on to find the next node with the fewest matching
-%implementations and repeats.  In this phase, only the name of the algorithm
-%is assigned to the operation node.  The details of the implementation are not
-%specified until the refinement phase.
-
 The refinement phase resolves the implementation for each operation node in
@@
 elements of the matrix or vector.
+In the optimization step, we apply conditional rewrite rules to optimize the
+dataflow graph, for example merging two subgraphs when they share a common
+operand. This rule is responsible for fusing the loops of the two
+matrix-vector products in the GEMVER kernel.  
+The final step performed by the MATLAB compiler when the graph cannot be
+refined further is the generation of C code. The generator outputs a C loop
+for each subgraph based on a topological sort of the graph.
 
@@
 %\subsubsection{Optimize the Dataflow Graph.}
-
-In the optimization step, we apply conditional rewrite rules to optimize the
-dataflow graph, for example merging two subgraphs when they share a common
-operand.
-%but are not dependent on one another.
-%  One example rewrite rule is
-%\emph{Merge Operand Sharing Subgraphs}: if two subgraphs share a common
-%operand but do not depend on one another, and if the iteration strategies of
-%the two subgraphs are compatible, then merge the subgraphs into a single
-%subgraph.  
-This rule is responsible for fusing the loops of the two
-matrix-vector products in the GEMVER kernel.  
-%Another example rule is
-%\emph{Merge Dependent Subgraphs}: if one subgraph depends on another
-%subgraph, and the iteration strategies of the two subgraphs are compatible,
-%then merge them into a single subgraph.  This rule, for example, is
-%responsible for fusing the loops of the two outer products in the GEMVER
-%kernel.
-
-%The choice of which optimizations to apply to various parts of the graph
-%depends on the characteristics of the kernel, the algorithm, the matrix order
-%and kind, and architectural features.
-
-%The following is an instructive
-%example: $y \gets A^T A (x + z + w)$.  Assume that the compiler has already
-%merged the two vector additions into one subgraph, call it $S_1$, and the two
-%matrix products into another subgraph, call it $S_2$. The question then is
-%whether it is profitable to apply the Merge Dependent Subgraphs rule one more
-%time to merge $S_1$ and $S_2$. If the three vectors plus a row of $A$ do not
-%fit in cache, then the merger will not be profitable because it will cause
-%each row of $A$ to fall out of cache before the row gets used a second time.
-%On the other hand, if all three vectors plus a row of $A$ do it in cache,
-%then $S_1$ and $S_2$ should be merged.  In general, we need a way to search
-%through the many different combinations of optimization choices and choose a
-%combination that minimizes memory traffic.
-
-%\subsubsection{Translate the Dataflow Graph to C.}
-
-The final step performed by the MATLAB compiler when the graph cannot be
-refined further is the generation of C code. The generator outputs a C loop
-for each subgraph based on a topological sort of the graph.
 
@@
 \begin{figure}[tb]
-\vspace{-.3in}
+%\vspace{-.3in}
 \begin{center}
@@
 \end{center}
-\vspace{-.25in}
+\vspace{-.2in}
 \caption{Overview of the Orio empirical tuning process.}
@@
 \end{figure}
+
+% --------------------------------------------------------------------------
 
@@
 \label{fig:orio-example}
-\vspace{-.2in}
+%\vspace{-.2in}
 \end{figure}
 
-Figure~\ref{fig:orio-example} shows an annotation example used by Orio
-to empirically optimize VADD operation on Blue Gene/P. The annotations
-contain performance hints that instruct Orio to perform memory
-alignment optimization, loop unrolling, and multicore parallelization
-(using OpenMP). In addition to these simple optimizations, Orio
-supports other transformations such as loop blocking, loop
-permutation, scalar replacement, array copy optimization, and other
-architecture-dependent optimizations (e.g., SIMDization for Intel and
-Blue Gene/P architectures). The right-hand side of
-Figure~\ref{fig:orio-example} shows separate tuning specifications
-used for building and running executable tests, including performance
-parameter values, execution environment details, input variable
-information, and the search algorithm. Orio also supports parallel
-search when parallel resources are available. In this example, the
-parallel Orio driver simultaneously executes 64 code variants in the
-same parallel job. The commands used by Orio to submit a parallel job
-and to query its status are also specified in the tuning
-specifications. At present, users must create the tuning
-specifications manually. When Orio is used in conjunction with
-compiler tools, such as the MATLAB compiler described in this paper,
-it should eventually be possible to automatically generate the tuning
+Figure~\ref{fig:orio-example} shows an annotation example used by Orio to
+empirically optimize VADD operation on Blue Gene/P. The annotations contain
+performance hints that instruct Orio to perform memory alignment
+optimization, loop unrolling, and multicore parallelization (using
+OpenMP). In addition to these simple optimizations, Orio supports other
+transformations such as loop blocking, loop permutation, scalar replacement,
+array copy optimization, and some architecture-dependent optimizations. The
+right-hand side of Figure~\ref{fig:orio-example} shows separate tuning
+specifications used for building and running executable tests, including
+performance parameter values, execution environment details, input variable
+information, and the search algorithm. Orio also supports parallel search
+when parallel resources are available. In this example, the parallel Orio
+driver simultaneously executes 64 code variants in the same parallel job. The
+commands used by Orio to submit a parallel job and to query its status are
+also specified in the tuning specifications. At present, users must create
+the tuning specifications manually. When Orio is used in conjunction with
+compiler tools, such as the MATLAB compiler described in this paper, it
+should eventually be possible to automatically generate the tuning
 specifications.
 
-\vspace{-0.1in}
+%\vspace{-0.1in}
 
@@
 
-Table~\ref{tbl:blas-ops} lists the composed BLAS operations used in
-our experiments, along with their input and output variables. Vectors
-are typeset in lowercase with an overhead arrow. Scalars and matrices
-are represented as lowercase and uppercase letters, respectively. A
-regular uppercase denotes a row matrix, whereas a bold uppercase
-symbolizes a column matrix. The extended MATLAB code that corresponds
-to each operation can be seen in the last column of
-Table~\ref{tbl:blas-ops}.
+Table~\ref{tbl:blas-ops} lists the composed BLAS operations used in our
+experiments, along with their input and output variables. Vectors are typeset
+in lowercase with an overhead arrow. Scalars and matrices are represented as
+lowercase and uppercase letters, respectively. A regular uppercase denotes a
+row matrix, whereas a bold uppercase symbolizes a column matrix. The extended
+MATLAB expression that corresponds to each operation can be seen in the last
+column of Table~\ref{tbl:blas-ops}.
 
 \begin{table}[htb] 
-\vspace{-.1in} 
+%\vspace{-.1in} 
 \caption{Composed BLAS operations used in our experiments.}
@@
 \hline 
-Name & Input & Output & C code from MATLAB compiler \\ 
+Name & Input & Output & Operation  \\ 
 \hline 
@@
 \end{tabular} 
-%\vspace{-.2in} 
+\vspace{-.1in} 
 \end{table}
@@
 }
+\vspace{-.1in}
 \caption{Performance results for different implementations of several composed BLAS operations.}
 \label{fig:results}
+\vspace{-.2in}
 \end{figure}
 
-The performance results of tuning the VADD operation are given in
-Figure~\ref{fig:results}(a). The ``Base'' label designates the simple
-loop C implementation generated by the MATLAB compiler, which was
-simply compiled with the options mentioned earlier and executed. We
-tested the performance of the IBM ESSL library and Goto BLAS for the
-two calls to \texttt{daxpy} used in the BLAS version of the code.
-Finally, we tuned the simple C loop version using Orio, with the
-performance annotations previously shown in
-Figure~\ref{fig:orio-example}. In this experiment, we measured the
-performance for both the sequential and parallel scenarios. We observe
-that even for a very simple operation, such as the vector addition
-above, the compiler alone is unable to obtain the same level of
-performance as the empirically tuned versions. Furthermore, as
-expected, the BLAS performance was worse than that of the single loop
-implementation, however, selecting the correct compiler options was
-crucial for achieving good performance of the base version.
+The performance results (in MFLOP/s) of tuning the VADD operation on the Blue
+Gene/P are given in Figure~\ref{fig:results}(a). The ``Base'' label
+designates the C implementation generated by the MATLAB compiler. We also
+tested the performance of an implementation that calls
+\texttt{daxpy} twice using available BLAS libraries.
+Finally, we tuned the simple C loop version using Orio, with the performance
+annotations previously shown in Figure~\ref{fig:orio-example}. In this
+experiment, we measured the performance for both the sequential and parallel
+scenarios. We observe that even for this simple operation, the compiler alone
+is unable to obtain the same level of performance as the empirically tuned
+versions. Furthermore, as expected, the BLAS implementation does not
+exploit locality and thus performed worse than the single loop version.
+%of the single loop implementation due to l
+% however, selecting the correct compiler
+%options was crucial for achieving good performance of the base version.
 
@@
 
-The performance results of ATAX are shown in
+The Xeon performance results (in seconds) of ATAX are shown in
 Figure~\ref{fig:results}(b). The Orio-tuned version that incorporates
@@
 
-Figure~\ref{fig:results}(c) exhibits the performance results of the
-GEMVER operation. Here we used the same PLuTo and Orio optimizations
-as for the ATAX example and, similarly, the parallel Orio version
-achieves the best performance, although in this case the sequential
-Orio version performs almost the same, suggesting that the compiler was
-not able to parallelize the code very effectively. For this operation,
-there are substantial performance differences between different BLAS
-versions, with the Intel MKL version achieving performance close to
-that of the simple compiler code.
+Figure~\ref{fig:results}(c) shows the performance (in seconds) of the GEMVER
+operation on the Xeon workstation. Here we used the same PLuTo and Orio
+optimizations as for the ATAX example and, similarly, the parallel Orio
+version achieved the best performance, although in this case the sequential
+Orio version performs almost the same, suggesting that the compiler was not
+able to parallelize the code very effectively. For this operation, there are
+substantial performance differences between different BLAS versions, with the
+Intel MKL version achieving performance close to that of the simple compiler
+code.
 
-The experimental results of BiCG Kernel operation are provided in
-Figure~\ref{fig:results}(d). For this operation, the PLuTo analysis
-did not result in performance improvement, thus we are showing the
-results obtained only through Orio transformations, which included
-vectorization, scalar replacement, and loop unroll/jam. Again the best
-performance was demonstrated by the parallel Orio version, while all
-the BLAS versions performed worse than the compiler-optimized C loop
-version.
+The performance (in seconds) for the BiCG Kernel operation is shown in
+Figure~\ref{fig:results}(d). For this operation, the PLuTo analysis did not
+result in performance improvement, thus we are showing the results obtained
+only through Orio transformations, which included vectorization, scalar
+replacement, and loop unroll/jam. Again the best performance was achieved
+by the parallel Orio version, while all the BLAS versions performed worse
+than the compiler-optimized C loop version.
 
 \section{Conclusions and Future Work}
+\label{sec:conclusions}
 
@@
 The positive initial results from our approach to generating tuned linear
-algebra routines motivate several future lines of investigation. At present
-there are a number of manual steps involved in obtaining a compiled tuned
-linear algebra library from the high-level extended MATLAB descriptions. We
-will work on the more close coupling of our separately developed tools and
-fully automating their interactions. This would enable application developers
-to easily produce custom linear algebra libraries tuned for specific
-architectures. We will also investigate extending the supported functionality
-of our MATLAB compiler beyond composed BLAS operations, thus benefiting a
-wider spectrum of applications.
+algebra routines motivate several future lines of investigation, including closer
+integration between the tools handling the different steps of the process and 
+more automation at each step.
+%At present
+%there are a number of manual steps involved in obtaining a compiled tuned
+%linear algebra library from the high-level extended MATLAB descriptions. We
+%will work on the more close coupling of our separately developed tools and
+%fully automating their interactions. This would enable application developers
+%to easily produce custom linear algebra libraries tuned for specific
+%architectures.
+
+% We will also investigate extending the supported functionality
+%of our MATLAB compiler beyond composed BLAS operations, thus benefiting a
+%wider spectrum of applications.
 
@@
 
+\nopagebreak
 \bibliographystyle{splncs}

### 2008-12-19 doc/papers/iccs09/paper.tex
final (I hope) edits
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@403 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@
 
-%showing that the code generated by using our system
-%significantly outperforms not only the original simple C code but also code
-%based on source BLAS, ATLAS-optimized BLAS, and Intel MKL routines.
-%Text in intro repeats abstract some, but that might not be a big deal for a conf paper.
-
 \section{Background}
@@
 MATLAB compiler~\cite{MaJIC}, include limited local optimizations for matrix
-expressions, but do not perform optimizations such as loop fusion across
+expressions but do not perform optimizations such as loop fusion across
 multiple operations as we do with the tools described in this paper. The
@@
 strength reduction, vectorization, and procedure specialization to optimize
-MATLAB scripts, but do not generate reusable optimized linear algebra
+MATLAB scripts but does not generate reusable optimized linear algebra
 routines as described in this paper.
@@
 of cache, and inserting prefetch instructions.  The pitfalls of this approach
-are well-understood~\cite{Goedecker01}.  It requires a significant amount of
+are well understood~\cite{Goedecker01}: It requires a significant amount of
 time and effort. Optimizing code for one particular platform may in fact make
@@
 understand or maintain.  An alternative is the use of tuned libraries of key
-numerical algorithms, for example BLAS~\cite{Dongarra:1990fk} and
+numerical algorithms, for example, BLAS~\cite{Dongarra:1990fk} and
 LAPACK~\cite{LAPACK} for dense linear algebra.
@@
 Specialized code generators circumvent the high costs of manual code
-generation . They include tools for basic dense linear algebra operations
-(ATLAS~\cite{WN147}, PhiPAC~\cite{bilmes97optimizing},
+generation. They include tools for basic dense linear algebra operations
+(ATLAS~\cite{WN147}, PhiPAC~\cite{bilmes97optimizing}),
 %,phipacwww})
-sparse linear algebra (OSKI~\cite{vuduc05}), 
+and sparse linear algebra (OSKI~\cite{vuduc05})
 %FFTs~\cite{FFTW,Spiral}), stencil-based
-
+%
 %operations~\cite{kamil06}, and tensor computations (TCE~\cite{TCE}) among
@@
 at enabling the definition of arbitrary kernels involving dense matrix linear algebra.
-It is often impossible to predict the performance of code precisely
+It is often impossible to predict precisely the performance of code
 on modern computer architectures. Thus, many of these specialized code
@@
 code for a particular choice of problem parameters and machine. Most existing
-autotuning tools are not general but rather focus on a specific domain or algorithm.
+autotuning tools are not general but focus on a specific domain or algorithm.
 %While not all applications can be written in terms of available libraries,
@@
 developed at Rice University, supports annotation-based loop fusion,
-unroll-and-jam, skewing and tiling.  A relatively new tool, POET~\cite{POET}
+unroll-and-jam, skewing, and tiling.  A relatively new tool, POET~\cite{POET},
 also supports a number of loop transformations. POET offers a complex
@@
 Codes based on matrix algebra are generally constructed as a sequence of
-calls to the Basic Linear Algebra Subprograms
-(BLAS)
+calls to the BLAS
 %~\cite{Dongarra:1988uq,Lawson:1979kx,Dongarra:1990fk} 
@@
 this way promotes readability and maintainability but can be costly in terms
-of memory efficiency. Specifically, the retrieval of a large order matrix at
-each routine call can have a profound effect on performance even when highly
+of memory efficiency. Specifically, the retrieval of a large-order matrix at
+each routine call can profoundly affect performance even when highly
 tuned implementations of the BLAS (e.g.,~\cite{Goto:2006fk})
@@
 
-The authors and other researchers have observed that it can be much more
-efficient to call a single specialized routine that performs multiple
-operations than to make successive calls to separate BLAS routines (e.g., see
-~\cite{gropp01}.
+A much more efficient approach is to call a single, specialized routine that
+performs multiple operations, rather than to make successive calls to separate BLAS
+routines (e.g., see ~\cite{gropp01}).
 %~\cite{Ashby:uq,Blackford:2002vn,baker03blgmres,baker03lgmres,Dennis:2005tg,Howell:2008,gropp01,Vuduc:2003kl}.
@@
 matrix-vector products $q = Ap$, $s = A^Tr$, where $A$ is a matrix and $p$,
-$q$, $r$, and $s$ are vectors, that comprise the computational bottlenecks of
-the BiConjugate Gradient Method~\cite{Saad:2003fk} and of the
+$q$, $r$, and $s$ are vectors, that represent the computational bottlenecks
+of the biconjugate gradient method (BiCG)~\cite{Saad:2003fk} and of the
 GEMVER kernel examined in Section~\ref{sec:matlab}.  These two operations can
-be implemented as a pair of calls to the BLAS routine {\tt GEMV} or they can
-be rewritten as a composed BLAS consisting of the doubly nested loop
-encompassing both matrix-vector products. In the former case, the matrix
-$A$ is accessed twice, while in the latter it is accessed only once. Our
-preliminary results in prior research and in this paper indicate that 
-loop fusion leads to a routine that delivers significantly better performance
-than does a pair of calls to the best optimized BLAS {\tt GEMV} routines for
-large matrix orders.  Composed routines are the focus of the work presented
-here, but much of what we discuss generalizes to a much broader array of
-computations.
+be implemented as a pair of calls to the BLAS routine {\tt GEMV}, or they can
+be rewritten as a composed BLAS consisting of a doubly nested loop
+encompassing both matrix-vector products. In the former case, the matrix $A$
+is accessed twice, whereas in the latter it is accessed only once. Our
+preliminary results in prior research and in the work reported in this paper
+indicate that loop fusion leads to a routine that delivers significantly
+better performance than does a pair of calls to the best optimized BLAS {\tt
+GEMV} routines for large matrix orders.  Composed routines are the focus of
+the work presented here, but much of what we discuss generalizes to a much
+broader array of computations.
 
@@
 \includegraphics[width=3in]{figures/process.png}
-\caption{The code generation and tuning process.}
+\caption{Code generation and tuning process.}
 \label{fig:process}
@@
 
-In the remainder of this section we describe each of the tools developed by the authors of this paper, namely the MATLAB to C compiler and the Orio empirical tuning tool.
+In the remainder of this section we describe each of the tools developed by
+the authors of this paper, namely, the MATLAB-to-C compiler~\cite{Siek} and the Orio
+empirical tuning tool~\cite{Norris:2007,Hartono:IPDPS09}.
 
@@
 \label{sec:matlab}
-%\input{matlab}
-In this section we describe the design and operation of the MATLAB to C compiler we have developed~\cite{Siek}.    Figure~\ref{fig:compiler} gives an overview of the compilation process. 
+
+Figure~\ref{fig:compiler} gives an overview of the MATLAB-to-C compilation process~\cite{Siek}. 
 The MATLAB kernel specification is parsed into a high-level intermediate
 representation in the form of a dataflow graph, in which each node represents
+a parameter (e.g., a scalar, matrix, or vector variable) of the kernel or an operation.
 %
@@
 \includegraphics[width=2in]{figures/compile.png}
-\caption{The MATLAB compiler.}
+\caption{MATLAB-to-C compiler.}
 \label{fig:compiler}
@@
 %
-a parameter (e.g., a scalar, matrix, or vector variable) of the kernel or an
-operation.  This dataflow graph is then iteratively processed until all of
+This dataflow graph is then iteratively processed until all of
 the implementation choices have been made.  The compilation process consists
-of three phases: analysis, refinement, and optimization, which are together
+of three phases -- analysis, refinement, and optimization -- that are together
 iterated until all of the implementation decisions have been made.  The graph
@@
 Here we briefly describe the analysis, refinement, and optimization of the
-dataflow graph, which are discussed in more detail in~\cite{Siek}. During the
-analysis phase, all types of intermediate nodes are computed and
+dataflow graph; these are discussed in more detail in~\cite{Siek}. 
+During the analysis phase, all types of intermediate nodes are computed and
 assigned. The algorithm choice and storage format determination are computed
-simultaneously. For example, the $\times$ symbol in the graph in
-Figure~\ref{fig:gemver-dataflow} having the inputs $u_1$ and $v_1^T$ can be
-implemented by iterating over rows first, or over columns first. The choice
-depends on how the result is used downstream in the dataflow graph.  In this
-case, the result is added to the outer product of $u_2$ and $v_2^T$, so we
-still could choose either option as long as we make the same choice for both
-outer products. 
+simultaneously.  Consider, for example, the GEMVER kernel computation:   
+$A \gets A + u_1 v_1^T + u_2 v_2^T;~
+  x \gets \beta A^T y + z;~
+  w \gets \alpha A x$.
+The multiplication of $u_1$ and $v_1^T$ can be implemented by iterating over 
+rows first or over columns first, depending on how the result is used
+downstream in the dataflow graph.  In this case, the result is added to the
+outer product of $u_2$ and $v_2^T$, so we still can choose either option as
+long as we make the same choice for both outer products.
 %Going one more step downstream, there is an addition with
@@
 The information on implementation possibilities for basic linear algebra
-operations are not hard-coded in the compiler; rather, these data are stored
+operations is not hard-coded in the compiler; rather, this data is stored
 in a database, called the \emph{linear algebra database}.  This separation
@@
 Orio. The input to Orio is C code containing semantic comments that
-describe both the computation, as well as various performance-tuning
+describe both the computation and various performance-tuning
 directives. Orio first extracts all annotation code regions by parsing
 the marked-up input code. Each annotated region is then passed to code
-transformator and code generator for potential optimizations. Then the
+transformation module and code generator for potential optimizations. Next the
 transformed C code with various incorporated optimizations
@@
 850 MHz PowerPC 450 processors with a dual floating-point unit and 2
-GB total memory per node, running a proprietary operating system. On
+GB total memory per node, running a proprietary operating system. On the
 Blue Gene/P, we used IBM XLC compiler (v9.0), with \texttt{-O3
@@
 %\vspace{-.1in} 
-\begin{tabular}{|c|c|c|c|} 
+\begin{tabular}{c c c c} 
 \hline 
@@
 \hline 
-\hline 
+%\hline 
 %& & & \\
@@
 $\overrightarrow{x} = \overrightarrow{w} + \overrightarrow{y} + \overrightarrow{z}$ \\ [1ex]
-\hline 
+%---------------------------------------------------------------------------------
+%\hline 
 %& & & \\
@@
 $\overrightarrow{y} = A' * (A * \overrightarrow{x})$ \\ [1ex]
-\hline 
+%---------------------------------------------------------------------------------
+%\hline 
 %& & & \\
- & ${\bf A}$,$a$,$b$, &  & ${\bf B} = {\bf A} + \overrightarrow{u_1} * \overrightarrow{v_1}' + \overrightarrow{u_2} * \overrightarrow{v_2}'$\\
-GEMVER & $\overrightarrow{u_1}$,$\overrightarrow{u_2}$,$\overrightarrow{v_1}$,$\overrightarrow{v_2}$, & ${\bf B}$,  & $\overrightarrow{x} = b * ({\bf B}' * \overrightarrow{y}) + \overrightarrow{z}$\\
-& $\overrightarrow{y}$,$\overrightarrow{z}$ & $\overrightarrow{x}$,$\overrightarrow{w}$ & $\overrightarrow{w} = a * ({\bf B} * \overrightarrow{x})$\\ [1ex]
-\hline 
-& & & $\overrightarrow{q} = {\bf A} * \overrightarrow{p}$ \\
-BiCG Kernel & ${\bf A}$,$\overrightarrow{p}$,$\overrightarrow{r}$ & $\overrightarrow{q}$,$\overrightarrow{s}$ & $\overrightarrow{s} = {\bf A}' * \overrightarrow{r}$\\ [1ex]
+GEMVER & ${\bf A}$,$a$,$b$, &  ${\bf B}$,  & ~~${\bf B} = {\bf A} + \overrightarrow{u_1} * \overrightarrow{v_1}' + \overrightarrow{u_2} * \overrightarrow{v_2}'$\\
+ & $\overrightarrow{u_1}$,$\overrightarrow{u_2}$,$\overrightarrow{v_1}$,$\overrightarrow{v_2}$, & $\overrightarrow{x}$,$\overrightarrow{w}$ & $\overrightarrow{x} = b * ({\bf B}' * \overrightarrow{y}) + \overrightarrow{z}$\\
+& $\overrightarrow{y}$,$\overrightarrow{z}$ &  & $\overrightarrow{w} = a * ({\bf B} * \overrightarrow{x})$\\ [1ex]
+%---------------------------------------------------------------------------------
+%\hline 
+BiCG Kernel & ${\bf A}$,$\overrightarrow{p}$,$\overrightarrow{r}$ & $\overrightarrow{q}$,$\overrightarrow{s}$ & $\overrightarrow{q} = {\bf A} * \overrightarrow{p}$ \\ 
+& & & $\overrightarrow{s} = {\bf A}' * \overrightarrow{r}$\\
+
 \hline 
@@
 \vspace{-.1in}
-\caption{Performance results for different implementations of several composed BLAS operations.}
+\caption{Performance results for several composed BLAS operations.}
 \label{fig:results}
-\vspace{-.2in}
+\vspace{-.1in}
 \end{figure}
@@
 tested the performance of an implementation that calls
-\texttt{daxpy} twice using available BLAS libraries.
+\texttt{DAXPY} twice using available BLAS libraries.
 Finally, we tuned the simple C loop version using Orio, with the performance
@@
 experiment, we measured the performance for both the sequential and parallel
-scenarios. We observe that even for this simple operation, the compiler alone
-is unable to obtain the same level of performance as the empirically tuned
-versions. Furthermore, as expected, the BLAS implementation does not
-exploit locality and thus performed worse than the single loop version.
+scenarios. Even for a very simple operation such as vector addition the
+compiler alone is unable to obtain the same level of performance as the
+empirically tuned versions. Furthermore, as expected, the BLAS implementation
+does not exploit locality and thus performed worse than the single-loop
+implementation.
 %of the single loop implementation due to l
@@
 operation on the Xeon workstation. Here we used the same PLuTo and Orio
-optimizations as for the ATAX example and, similarly, the parallel Orio
-version achieved the best performance, although in this case the sequential
-Orio version performs almost the same, suggesting that the compiler was not
-able to parallelize the code very effectively. For this operation, there are
-substantial performance differences between different BLAS versions, with the
-Intel MKL version achieving performance close to that of the simple compiler
-code.
+optimizations as for the ATAX example. Similarly, the parallel Orio version
+achieved the best performance, although in this case the sequential Orio
+version performs almost the same, suggesting that the compiler was not able
+to parallelize the code very effectively. For this operation, substantial
+performance differences exist between among the different BLAS versions, with
+the Intel MKL version achieving performance close to that of the simple
+compiler code.
 
-The performance (in seconds) for the BiCG Kernel operation is shown in
+The performance (in seconds) for the BiCG kernel operation is shown in
 Figure~\ref{fig:results}(d). For this operation, the PLuTo analysis did not
-result in performance improvement, thus we are showing the results obtained
+result in performance improvement. Thus we are showing the results obtained
 only through Orio transformations, which included vectorization, scalar
@@
 
-\section{Conclusions and Future Work}
-\label{sec:conclusions}
+\section{Conclusions}
+\label{sec:conclusion}
 
@@
 translate the MATLAB code to C, (2) analyze the resulting loops and identify
-locatility and parallelism-enhancing optimizations using PLuTo, and (3)
+locality and parallelism-enhancing optimizations using PLuTo, and (3)
 annotate the resulting C code with syntactic performance directives and use
@@
 
-%\vfill
-%\begin{flushright}
-%\scriptsize
-%\framebox{\parbox{2.4in}{The submitted manuscript has been created by
-%UChicago Argonne, LLC, Operator of Argonne National Laboratory
-%("Argonne").  Argonne, a U.S. Department of Energy Office
-%of Science laboratory, is operated under Contract No.
-%DE-AC02-06CH11357.  The U.S. Government retains for itself, and
-%others acting on its behalf, a paid-up, nonexclusive, irrevocable
-%worldwide license in said article to reproduce, prepare derivative works,
-%distribute copies to the public, and perform publicly and display
-%publicly, by or on behalf of the Government.}}
-%\normalsize
-%\end{flushright}
-
 %\newpage
@@
 \end{document}
-
-% Current bibtex warnings:
-This is BibTeX, Version 0.99c (Web2C 7.5.4)
-The top-level auxiliary file: paper.aux
-The style file: siam.bst
-Database file #1: paper.bib
-Warning--can't use both volume and number fields in TCE
-(There was 1 warning)
-
-% Put bibtex here only if you do not have access to the subversion repository, otherwise insert them into paper.bib

### 2009-02-10 src/tool/ZestyParser/Tokens.py
added source ZestyParser since previously included binary was not portable across different python versions.

@@ -1,0 +1,647 @@
+# ZestyParser 0.8.1 -- Parses in Python zestily
+# Copyright (C) 2006-2007 Adam Atlas
+# 
+# Permission is hereby granted, free of charge, to any person obtaining a copy
+# of this software and associated documentation files (the "Software"), to deal
+# in the Software without restriction, including without limitation the rights
+# to use, copy, modify, merge, publish, distribute, sublicense, and/or sell
+# copies of the Software, and to permit persons to whom the Software is
+# furnished to do so, subject to the following conditions:
+# 
+# The above copyright notice and this permission notice shall be included in
+# all copies or substantial portions of the Software.
+# 
+# THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
+# IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
+# FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT. IN NO EVENT SHALL THE
+# AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER
+# LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM,
+# OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN
+# THE SOFTWARE.
+
+'''
+@group Basic Tokens: Raw,RE,RawToken,Token,TakeToken
+@group Complex Tokens: CompositeToken,TokenSequence,TokenSeries
+@group Special Tokens: Defer,Default,Lookahead,Negative
+@group TokenSequence Flags: Omit,Skip,Only
+
+@version: 0.8.1
+@author: Adam Atlas
+@copyright: Copyright 2006-2007 Adam Atlas. Released under the MIT license (see LICENSE.txt).
+@contact: adam@atlas.st
+
+@var EmptyToken: A L{Default} instance initialized with the empty string.
+@var EOF: A token which matches (and returns C{None}) if the parser is at the end of its L{data <ZestyParser.data>} sequence.
+
[…]

### 2009-02-10 src/module/spmv/parser.py
-use ZestyParser (from source)

@@ -3,7 +3,7 @@
 #
 
 import re, sys
-from tool.zestyparser.ZestyParser import *
+from tool.ZestyParser import *
 
 #--------------------------------------------------------------------------------
 

### 2009-02-11 src/tool/ZestyParser/Tokens.py
- Removed deprecated use of 'as' which is now a keyword (in Python 2.6)

@@ -122,12 +122,10 @@ class AbstractToken (object):
     to = None
 
     #'as' is deprecated in favor of 'to' since it's becoming a reserved word
-    def __init__(self, desc, callback=None, to=None, as=None, name=None):
+    def __init__(self, desc, callback=None, to=None, name=None):
         self.desc = desc
         self.callback = callback
-        self.to = to or as
-        if as:
-            warnings.warn('`as` argument is deprecated; use `to`', DeprecationWarning, stacklevel=2)
+        self.to = to 
         self.name = name
     
     def __repr__(self):

### 2009-02-12 doc/papers/ipdps09/conclusions.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@427 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@
 Uday Bondhugula of Ohio State University for many productive discussions and
-his valuable help with PLuTo.
+his valuable help with Pluto.
 

### 2009-02-12 doc/papers/ipdps09/integration.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@427 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@
-\section{PLuTo-Orio Integration}
+\section{Pluto-Orio Integration}
 \label{sec:integration}
@@
 how Orio has been extended with an external transformation program, called
-PLuTo~\cite{uday08pldi}. This integration also demonstrates the easy
+Pluto~\cite{uday08pldi}. This integration also demonstrates the easy
 extensibility of Orio and the ability to leverage other source transformation
@@
 
-PLuTo is a source-to-source automatic transformation tool aimed at optimizing
+Pluto is a source-to-source automatic transformation tool aimed at optimizing
 a sequence of nested loops for data locality and coarse-grained parallelism
-simultaneously. PLuTo employs a polyhedral model of arbitrary loop nests,
+simultaneously. Pluto employs a polyhedral model of arbitrary loop nests,
 where the dynamic instance (iteration) of each statement is viewed as an
@@
 statement representation and a precise characterization of data dependences
-enable PLuTo to construct mathematically correct complex loop
-transformations. PLuTo's polyhedral-based transformations result in improved
+enable Pluto to construct mathematically correct complex loop
+transformations. Pluto's polyhedral-based transformations result in improved
 cache locality and loops parallelized for multicore architecture.
 
-Figure~\ref{fig:pluto-orio} outlines the overall structure of the PLuTo-Orio
+Figure~\ref{fig:pluto-orio} outlines the overall structure of the Pluto-Orio
 integrated system, which is implemented as a new optimization module in
@@
 parallelization, scalar replacement, and auto-vectorization. Using these
-parameter values, PLuTo then performs polyhedral transformations (e.g.,
+parameter values, Pluto then performs polyhedral transformations (e.g.,
 two-level tiling and OpenMP parallelization) on the input loop code. The
@@
 %information produced by gprof provides accurate locations (line numbers) of
-%the loop nests where the PLuTo-generated code spends most of its execution
+%the loop nests where the Pluto-generated code spends most of its execution
 %time. 
@@
 \vspace{-.2in}
-\caption{Integration of PLuTo into Orio's code transformation module.}
+\caption{Integration of Pluto into Orio's code transformation module.}
 \label{fig:pluto-orio}
@@
 
-PLuTo also performs syntactic loop unroll/jamming as a post-processing
+Pluto also performs syntactic loop unroll/jamming as a post-processing
 pass. The target loops, however, are limited only to innermost loops with a
@@
 generated after syntactic transformations is correct. Because of this, the
-tuning process using the integrated PLuTo-Orio system is currently
+tuning process using the integrated Pluto-Orio system is currently
 semi-automatic---user involvement is required to decide whether it is safe
@@
 %consistently have similar basic loop structure for different
-%PLuTo-generated code versions. Once all illegal syntactic
+%Pluto-generated code versions. Once all illegal syntactic
 %transformations are known by the user, they must be switched off so

### 2009-02-12 doc/papers/ipdps09/results.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@364 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@427 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@
 
-\subsection{Evaluation of the PLuTo-Orio Integrated System}
+\subsection{Evaluation of the Pluto-Orio Integrated System}
 
 This section discusses the performance evaluation of the integrated
-PLuTo-Orio system (discussed in Section~\ref{sec:integration}) on the
+Pluto-Orio system (discussed in Section~\ref{sec:integration}) on the
 multicore Intel Xeon by using a number of application kernels that are
 nontrivial to optimize and parallelize. We compare the performance of the
-code tuned by Orio with the base code and the PLuTo-generated codes. The
-PLuTo code was obtained by running the base code with
-PLuTo-0.3.0~\cite{pluto030} using \texttt{--tile} to employ loop tiling for
+code tuned by Orio with the base code and the Pluto-generated codes. The
+Pluto code was obtained by running the base code with
+Pluto-0.3.0~\cite{pluto030} using \texttt{--tile} to employ loop tiling for
 L1 cache, \texttt{--l2tile} to employ loop tiling for L2 cache, and
 \texttt{--unroll} to employ loop unrolling, and for parallel code generation,
-an additional \texttt{--parallel} option. We used PLuTo's default tile
+an additional \texttt{--parallel} option. We used Pluto's default tile
 sizes (L1: 32x32 or 32x32x32; L2: 256x256 or 256x256x256) and default
@@
 In the following sections, we refer to the icc-optimized base code as
-``ICC,'' the PLuTo-generated code with L1 tiling and
-unroll/jamming as ``PLuTo (L1 tiling),'' and the PLuTo-generated code with L1
-and L2 tilings and unroll/jamming as ``PLuTo (L1+L2 tiling).'' The code
-tuned by Orio is referred to as ``PLuTo+Orio.''
+``ICC,'' the Pluto-generated code with L1 tiling and
+unroll/jamming as ``Pluto (L1 tiling),'' and the Pluto-generated code with L1
+and L2 tilings and unroll/jamming as ``Pluto (L1+L2 tiling).'' The code
+tuned by Orio is referred to as ``Pluto+Orio.''
 
@@
 shown in Figure~\ref{fig:fdtd-2d-cookie-seq}. The base code optimized by icc
-alone performs better than PLuTo for small problem sizes since all input
-arrays fit in the L2 cache (insufficient computation to offset PLuTo's tiling
+alone performs better than Pluto for small problem sizes since all input
+arrays fit in the L2 cache (insufficient computation to offset Pluto's tiling
 overhead). As the array sizes increase, the lack of data reuse impairs the
-base code's performance, whereas the PLuTo performance remains about the
+base code's performance, whereas the Pluto performance remains about the
 same.
@@
 When the input arrays are small, Orio discovers that applying
-PLuTo's polyhedral transformations is not beneficial, and therefore,
+Pluto's polyhedral transformations is not beneficial, and therefore,
 it employs only its syntactic transformations on the original FDTD
-code. For large problem sizes, Orio exploits some of the PLuTo's code
+code. For large problem sizes, Orio exploits some of the Pluto's code
 transformations and enhances it further with its syntactic
 optimizations, resulting in performance consistently and significantly
-higher than both the base and PLuTo codes (up to 86\% over PLuTo).
+higher than both the base and Pluto codes (up to 86\% over Pluto).
 
 %By tuning the sequential code with Orio,
-%the attainable performance enhancement over PLuTo is up to 86\%.
+%the attainable performance enhancement over Pluto is up to 86\%.
 
@@
 for $tmax=500$ and $nx=ny=2000$. The results indicate that whereas icc is
-unable to auto-parallelize the code, PLuTo detects the existence of pipelined
+unable to auto-parallelize the code, Pluto detects the existence of pipelined
 parallelism and then successfully parallelizes the code. Optimizing and
-tuning by Orio additionally improves the PLuTo performance by up to
+tuning by Orio additionally improves the Pluto performance by up to
 78\%. Moreover, we observe that because of memory contention between the two
-quad-core Intel processors, the speedup of both the PLuTo and Orio codes
+quad-core Intel processors, the speedup of both the Pluto and Orio codes
 slightly decreases when the number of cores used is greater than four.
@@
 %%BN: commented out since this needs more explanation (no room in this version)
-%It is to be noted that collecting performance results of the PLuTo 
+%It is to be noted that collecting performance results of the Pluto 
 %code with L1 and L2 tilings was not possible because of the large code 
@@
 
-%% BN: comment for space
-\comment{
 \subsubsection{3-D Gauss-Seidel Successive Over-Relaxation Method}  
@@
 Figure~\ref{fig:seidel-cookie-seq} contains the sequential performance
-results for $T=500$, which show that applying PLuTo's polyhedral tiling on
+results for $T=500$, which show that applying Pluto's polyhedral tiling on
 the original code always delivers performance boosts, which range from 126\%
-to 142\%. When using PLuTo, performing two-level tiling (for both L1 and L2
+to 142\%. When using Pluto, performing two-level tiling (for both L1 and L2
 caches) yields 13\% lower performance than performing only L1 tiling. 
@@
 one-level tiling (for L1 cache only) is always performed. The sequential
-speedup obtained from tuning the PLuTo code with Orio is significant, ranging
+speedup obtained from tuning the Pluto code with Orio is significant, ranging
 from 51\% to 133\%.
@@
 Figure~\ref{fig:seidel-cookie-par}. Again we observe that icc failed to
-parallelize the code, whereas PLuTo was able to parallelize it and Orio
-improved the performance of the one-level tiled PLuTo code further by up to
-128\% (and up to 548\% for the two-level tiled PLuTo code). We observed that
+parallelize the code, whereas Pluto was able to parallelize it and Orio
+improved the performance of the one-level tiled Pluto code further by up to
+128\% (and up to 548\% for the two-level tiled Pluto code). We observed that
 tiling over both L1 and L2 can result in worse performance than L1-only
@@
 %Similar earlier trend also exists
-%in these performance results: PLuTo parallelizes the code through its
+%in these performance results: Pluto parallelizes the code through its
 %polyhedral transformations (i.e., skewing and tiling), while the icc
 %compiler alone does not parallelize the loop. Empirical tuning via
-%Orio creates up to 122\% speedups over the one-level tiled PLuTo code,
-%and up to 548\% speedups over the two-level tiled PLuTo code. Another
+%Orio creates up to 122\% speedups over the one-level tiled Pluto code,
+%and up to 548\% speedups over the two-level tiled Pluto code. Another
 %finding is that tiling for both L1 and L2 caches contributes to a poor
@@
 %processors, resulting in underutilization of the available resources.
-} % end comment
 
@@
 Figure~\ref{fig:lu-cookie-seq}. Similarly to the 2-D FDTD results, the icc-optimized code is more
-efficient than the PLuTo-generated codes when the input arrays are small and
+efficient than the Pluto-generated codes when the input arrays are small and
 fit in the L2 cache. Because of better data locality, however, the
-performance of the PLuTo-tiled codes is better than that of the base code as
+performance of the Pluto-tiled codes is better than that of the base code as
 the input arrays get larger. 
@@
 %by Orio: one for small problem sizes and one for large problem sizes. And
-Consequently, Orio employs PLuTo's polyhedral transformations only for large
+Consequently, Orio employs Pluto's polyhedral transformations only for large
 input sizes prior to applying its own syntactic transformations. 
@@
 %generating the best tuned code. 
-Compared to the two PLuTo codes, the
+Compared to the two Pluto codes, the
 Orio-tuned code yields performance improvements ranging between 26\% and 277\%.
@@
 obtained for $N=4000$. We observe that icc is not able to parallelize the
-code, whereas PLuTo achieves higher performance by exploiting multicore
-parallelism. Furthermore, both PLuTo-generated codes have comparable
+code, whereas Pluto achieves higher performance by exploiting multicore
+parallelism. Furthermore, both Pluto-generated codes have comparable
 performance, with slightly better scalability exhibited by the single-level tiled
-code. Orio further improved the performance of the PLuTo-generated versions
+code. Orio further improved the performance of the Pluto-generated versions
 by a factor of 51\% to 120\%.

### 2009-02-12 doc/papers/ipdps09/results-long.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@427 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@
 This section discusses the performance evaluations of the integrated
-PLuTo-Orio system (explained earlier in Section~\ref{sec:integration})
+Pluto-Orio system (explained earlier in Section~\ref{sec:integration})
 on the multicore Intel machine by using a number of application
@@
 the performance of the code tuned by Orio with the base code and the
-PLuTo-generated codes. The PLuTo code was obtained via running the
-base code with PLuTo-0.3.0~\cite{pluto030} using \texttt{--tile} to
+Pluto-generated codes. The Pluto code was obtained via running the
+base code with Pluto-0.3.0~\cite{pluto030} using \texttt{--tile} to
 employ loop tiling for L1 cache,
@@
 \texttt{--unroll} to employ loop unrolling. Generating a parallel
-version of PLuTo code needs an additional \texttt{--parallel}
-option. Note that we used PLuTo's default tile sizes and unroll
-factors to produce all PLuTo codes. By default, PLuTo always sets the
+version of Pluto code needs an additional \texttt{--parallel}
+option. Note that we used Pluto's default tile sizes and unroll
+factors to produce all Pluto codes. By default, Pluto always sets the
 L1 tile sizes to 32x32 (or 32x32x32) and L2 tile sizes to 256x256 (or
@@
 In the following sections, we refer to the ICC-optimized base code as
-'Compiler-optimized', the PLuTo-generated code with L1 tiling and
-unroll/jamming as 'PLuTo (L1 tiling)', and the PLuTo-generated code
-with L1 and L2 tilings and unroll/jamming as 'PLuTo (L1+L2
-tiling)'. Also, the code tuned by Orio is referred to as 'PLuTo+Orio'.
+'Compiler-optimized', the Pluto-generated code with L1 tiling and
+unroll/jamming as 'Pluto (L1 tiling)', and the Pluto-generated code
+with L1 and L2 tilings and unroll/jamming as 'Pluto (L1+L2
+tiling)'. Also, the code tuned by Orio is referred to as 'Pluto+Orio'.
 
@@
 $nx=ny$ is presented in Figure~\ref{fig:fdtd-2d-cookie-seq}. The base
-code optimized by ICC performs better than PLuTo for small problem
+code optimized by ICC performs better than Pluto for small problem
 sizes since all input arrays fit in the L2 cache, thus tiling
-performed by PLuTo will only introduce more overhead. As the array
+performed by Pluto will only introduce more overhead. As the array
 sizes become larger, the lack of data reuse impairs the base code's
-performance, whereas the PLuTo performance remains steady. Code tuning
+performance, whereas the Pluto performance remains steady. Code tuning
 using Orio generates two code variants: one for small problem sizes
 and another for large problem sizes. When the input array sizes are
-small, Orio discovers that applying PLuTo's polyhedral transformations
+small, Orio discovers that applying Pluto's polyhedral transformations
 is not beneficial, and therefore, it employs only its syntactic
 transformations on the original FDTD code. As to tuning for large
-problem size, Orio exploits some of the PLuTo's code transformations
+problem size, Orio exploits some of the Pluto's code transformations
 and enhances it further with its syntactic optimizations, resulting in
-performance constantly higher than both the base and PLuTo codes. By
+performance constantly higher than both the base and Pluto codes. By
 tuning the sequential code with Orio, the attainable performance
-enhancement over PLuTo is up to 86\%.
+enhancement over Pluto is up to 86\%.
 
@@
 obtained for $tmax=500$ and $nx=ny=2000$. The results indicate that
-while ICC is unable to auto-parallelize the code, PLuTo detects the
+while ICC is unable to auto-parallelize the code, Pluto detects the
 existence of pipelined parallelism and then parallelizes the
-code. Optimizing and tuning by Orio additionally improves the PLuTo
+code. Optimizing and tuning by Orio additionally improves the Pluto
 performance by up to 78\%. Moreover, it can be seen that due to memory
 contention between the two quad-core Intel processors, the performance
-speedup of both the PLuTo and Orio codes slightly decreases when the
+speedup of both the Pluto and Orio codes slightly decreases when the
 number of used cores is greater than four.
@@
 Figure~\ref{fig:seidel-cookie-seq} are when $T=500$. As the figure
-shows, applying PLuTo's polyhedral tiling on the original code always
+shows, applying Pluto's polyhedral tiling on the original code always
 delivers performance boosts, which range from 126\% to
 142\%. Performing two-level tiling (for both L1 and L2 caches) by
-PLuTo yields 13\% lower performance than only using L1 tiling; and
+Pluto yields 13\% lower performance than only using L1 tiling; and
 this is also reflected in the best sequential code found by Orio,
 where one-level tiling (for L1 cache only) is always performed. The
-sequential speedup obtained from tuning the PLuTo code with Orio is
+sequential speedup obtained from tuning the Pluto code with Orio is
 significant, ranging from 51\% to 133\%.
@@
 Figure~\ref{fig:seidel-cookie-par}. Similar earlier trend also exists
-in these performance results: PLuTo parallelizes the code through its
+in these performance results: Pluto parallelizes the code through its
 polyhedral transformations (i.e., skewing and tiling), while the ICC
 compiler alone does not parallelize the loop. Empirical tuning via
-Orio creates up to 122\% speedups over the one-level tiled PLuTo code,
-and up to 548\% speedups over the two-level tiled PLuTo code. Another
+Orio creates up to 122\% speedups over the one-level tiled Pluto code,
+and up to 548\% speedups over the two-level tiled Pluto code. Another
 finding is that tiling for both L1 and L2 caches contributes to a poor
@@
 in Figure~\ref{fig:lu-cookie-seq}. Like 2-D FDTD, the ICC-optimized
-code is more efficient than the PLuTo-generated codes when the input
+code is more efficient than the Pluto-generated codes when the input
 arrays are small and fit in the L2 cache. However, due to better data
-locality, the performance of the PLuTo-tiled codes exceeds the base
+locality, the performance of the Pluto-tiled codes exceeds the base
 code as the input arrays get larger. Again, there are two different
@@
 problem sizes. And only the code tuned for large input arrays employs
-PLuTo's polyhedral transformations, prior to applying Orio's syntactic
+Pluto's polyhedral transformations, prior to applying Orio's syntactic
 transformations. Unlike 3-D Gauss-Seidel, this time Orio chooses to
 apply both L1 and L2 tilings when generating the best tuned
-code. Compared to the two PLuTo codes, the Orio-tuned code yields
+code. Compared to the two Pluto codes, the Orio-tuned code yields
 performance speedups in the range of 26\% to 277\%.
@@
 were obtained for $N=4000$. From the results, we observe that ICC is
-not able to parallelize the code, whereas PLuTo achieves higher
+not able to parallelize the code, whereas Pluto achieves higher
 performance by exploitation of multicore parallelism. Furthermore,
-both PLuTo-generated codes have comparable performance numbers, with a
+both Pluto-generated codes have comparable performance numbers, with a
 slightly better scalability on the one-level tiled code. By the use of
-Orio tuning, the performances of the PLuTo base codes were improved by
+Orio tuning, the performances of the Pluto base codes were improved by
 a factor of 51\% to 120\%.

### 2009-02-12 doc/papers/ipdps09/results.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@428 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@
 
-\begin{figure}[thb] 
+\begin{figure*}%[t] 
 \begin{center} 
   \subfigure[Sequential (single core)]{ 
-  \includegraphics[width=.4\textwidth]{figures/axpy4_bgp/seq.eps}  
+  \includegraphics[width=.45\textwidth]{figures/axpy4_bgp/seq.eps}  
   \label{fig:axpy4-bgp-seq} 
@@
   \subfigure[Parallel (four cores)]{ 
-  \includegraphics[width=.4\textwidth]{figures/axpy4_bgp/par.eps}  
+  \includegraphics[width=.45\textwidth]{figures/axpy4_bgp/par.eps}  
   \label{fig:axpy4-bgp-par} 
@@
 \end{center}
-\vspace{-.2in} 
 \caption{Performance of AXPY-4 operations on the Blue Gene/P.} 
 \label{fig:axpy4-bgp-results} 
-\end{figure}
+\end{figure*}
 
@@
 \subsection{Sparse Matrix Computations}
+
+\begin{figure}%[tb]
+\begin{tabular}{c}
+\begin{minipage}[b]{.47\textwidth}
+\centering
+\includegraphics[width=1.0\textwidth]{figures/spmv.eps}  
+\end{minipage}
+\\
+{\footnotesize (a)}
+\\
+\begin{minipage}[b]{.3\textwidth}
+\scriptsize
+\begin{verbatim}
+
+for (i=0; i<num_rows; i++)
+  for (j=Aptr[i]; j<Aptr[i+1]; j++)
+    y[i] += Aval[j]*x[Aind[j]];
+
+\end{verbatim}
+\end{minipage}
+\\
+{\footnotesize (b)}
+\\
+\end{tabular}
+
+\caption{(a) Compressed sparse row (CSR) format (b) Basic implementation of CSR-based SpMV.}
+\label{fig:spmv}
+\end{figure}
 
@@
 reuse of $x$ and $y$. One of the most commonly used data structures for
-storing a sparse matrix is compressed sparse row (CSR) storage,
-%~\cite{vuduc-thesis} 
+storing a sparse matrix is compressed sparse row (CSR) storage
+~\cite{vuduc-thesis},
 illustrated in Figure~\ref{fig:spmv}(a). Elements in each row of $A$ are
@@
 shown in Figure~\ref{fig:spmv}(b).
-
-\begin{figure}[tb]
-\centering
-\vspace{-.2in}
-\begin{tabular}{cc}
-\begin{minipage}[b]{.5\textwidth}
-\centering
-\includegraphics[width=.8\textwidth]{figures/spmv.eps}  
-\end{minipage}
-&
-\begin{minipage}[b]{.5\textwidth}
-\scriptsize
-\begin{verbatim}
-for (i=0; i<num_rows; i++)
-  for (j=Aptr[i]; j<Aptr[i+1]; j++)
-    y[i] += Aval[j]*x[Aind[j]];
-
-\end{verbatim}
-\end{minipage}\\
-(a) & (b)\\
-\end{tabular}
-\vspace{-.2in}
-\caption{(a) Compressed sparse row (CSR) format (b) Basic implementation of CSR-based SpMV.}
-\label{fig:spmv}
-%\vspace{.2in}
-\end{figure}
 
@@
 %code for the many-thread cases comprises two distinct tuned codes to handle
-\begin{figure} [ht]
+
+\begin{figure*}%[ht]
 \begin{center} 
   \subfigure[SMP: $p=1$, $t/p=8$]{
-  \includegraphics[width=.3\textwidth]{figures/ex27_cookie/smp.eps}
+  \includegraphics[width=.315\textwidth]{figures/ex27_cookie/smp.eps}
   \label{fig:ex27-cookie-smp} } \subfigure[Dual: $p=4$, $t/p=2$]{
-  \includegraphics[width=.3\textwidth]{figures/ex27_cookie/dual.eps}
+  \includegraphics[width=.315\textwidth]{figures/ex27_cookie/dual.eps}
   \label{fig:ex27-cookie-dual} } \subfigure[VN: $p=8$, $t/p=1$]{
-  \includegraphics[width=.3\textwidth]{figures/ex27_cookie/vn.eps}
+  \includegraphics[width=.315\textwidth]{figures/ex27_cookie/vn.eps}
   \label{fig:ex27-cookie-vn} }
 \end{center}
-\vspace{-.2in} 
 \caption{Performance of inode SpMV on eight-core Intel machine; $p$ is the number of processes, and $t/p$ is the threads per process.} 
 \label{fig:ex27-cookie-results} 
-\end{figure}
+\end{figure*}
 
@@
 
-\begin{figure} [htb]
+\begin{figure*} %[htb]
 \begin{center} 
   \subfigure[1 node, SMP mode]{ 
-  \includegraphics[width=.3\textwidth]{figures/ex27_bgp/n1_smp.eps}  
+  \includegraphics[width=.315\textwidth]{figures/ex27_bgp/n1_smp.eps}  
   \label{fig:ex27-bgp-smp-n1} 
@@
   \subfigure[1 node, Dual mode]{ 
-  \includegraphics[width=.3\textwidth]{figures/ex27_bgp/n1_dual.eps}  
+  \includegraphics[width=.315\textwidth]{figures/ex27_bgp/n1_dual.eps}  
   \label{fig:ex27-bgp-dual-n1} 
@@
   \subfigure[1 node, VN mode]{ 
-  \includegraphics[width=.3\textwidth]{figures/ex27_bgp/n1_vn.eps}  
+  \includegraphics[width=.315\textwidth]{figures/ex27_bgp/n1_vn.eps}  
   \label{fig:ex27-bgp-vn-n1} 
@@
   \subfigure[8 nodes, SMP mode]{ 
-  \includegraphics[width=.3\textwidth]{figures/ex27_bgp/n8_smp.eps}  
+  \includegraphics[width=.315\textwidth]{figures/ex27_bgp/n8_smp.eps}  
   \label{fig:ex27-bgp-smp-n8} 
@@
   \subfigure[8 nodes, Dual mode]{ 
-  \includegraphics[width=.3\textwidth]{figures/ex27_bgp/n8_dual.eps}  
+  \includegraphics[width=.315\textwidth]{figures/ex27_bgp/n8_dual.eps}  
   \label{fig:ex27-bgp-dual-n8} 
@@
   \subfigure[8 nodes, VN mode]{ 
-  \includegraphics[width=.3\textwidth]{figures/ex27_bgp/n8_vn.eps}  
+  \includegraphics[width=.315\textwidth]{figures/ex27_bgp/n8_vn.eps}  
   \label{fig:ex27-bgp-vn-n8} 
@@
 %  \subfigure[32 nodes, SMP mode]{ 
-%  \includegraphics[width=.3\textwidth]{figures/ex27_bgp/n32_smp.eps}  
+%  \includegraphics[width=.315\textwidth]{figures/ex27_bgp/n32_smp.eps}  
 %  \label{fig:ex27-bgp-smp-n32} 
@@
 %  \subfigure[32 nodes, Dual mode]{ 
-%  \includegraphics[width=.3\textwidth]{figures/ex27_bgp/n32_dual.eps}  
+%  \includegraphics[width=.315\textwidth]{figures/ex27_bgp/n32_dual.eps}  
 %  \label{fig:ex27-bgp-dual-n32} 
@@
 %  \subfigure[32 nodes, VN mode]{ 
-%  \includegraphics[width=.3\textwidth]{figures/ex27_bgp/n32_vn.eps}  
+%  \includegraphics[width=.315\textwidth]{figures/ex27_bgp/n32_vn.eps}  
 %  \label{fig:ex27-bgp-vn-n32} 
@@
 \end{center}
-\vspace{-.2in} 
-\caption{Performance of inode SpMV on Blue Gene/P for 1, 8, and 32 nodes.} 
+\caption{Performance of inode SpMV on Blue Gene/P for 1 and 8 nodes.} 
 \label{fig:ex27-bgp-results1} 
-\end{figure}
+\end{figure*}
 
@@
   \subfigure[32x32, SMP mode]{ 
-  \includegraphics[width=.3\textwidth]{figures/ex27_bgp/s32_smp.eps}  
+  \includegraphics[width=.315\textwidth]{figures/ex27_bgp/s32_smp.eps}  
   \label{fig:ex27-bgp-smp-32x32} 
@@
   \subfigure[32x32, Dual mode]{ 
-  \includegraphics[width=.3\textwidth]{figures/ex27_bgp/s32_dual.eps}  
+  \includegraphics[width=.315\textwidth]{figures/ex27_bgp/s32_dual.eps}  
   \label{fig:ex27-bgp-dual-32x32} 
@@
   \subfigure[32x32, VN mode]{ 
-  \includegraphics[width=.3\textwidth]{figures/ex27_bgp/s32_vn.eps}  
+  \includegraphics[width=.315\textwidth]{figures/ex27_bgp/s32_vn.eps}  
   \label{fig:ex27-bgp-vn-32x32} 
@@
   \subfigure[64x64, SMP mode]{ 
-  \includegraphics[width=.3\textwidth]{figures/ex27_bgp/s64_smp.eps}  
+  \includegraphics[width=.315\textwidth]{figures/ex27_bgp/s64_smp.eps}  
   \label{fig:ex27-bgp-smp-64x64} 
@@
   \subfigure[64x64, Dual mode]{ 
-  \includegraphics[width=.3\textwidth]{figures/ex27_bgp/s64_dual.eps}  
+  \includegraphics[width=.315\textwidth]{figures/ex27_bgp/s64_dual.eps}  
   \label{fig:ex27-bgp-dual-64x64} 
@@
   \subfigure[64x64, VN mode]{ 
-  \includegraphics[width=.3\textwidth]{figures/ex27_bgp/s64_vn.eps}  
+  \includegraphics[width=.315\textwidth]{figures/ex27_bgp/s64_vn.eps}  
   \label{fig:ex27-bgp-vn-64x64} 
@@
   \subfigure[128x128, SMP mode]{ 
-  \includegraphics[width=.3\textwidth]{figures/ex27_bgp/s128_smp.eps}  
+  \includegraphics[width=.315\textwidth]{figures/ex27_bgp/s128_smp.eps}  
   \label{fig:ex27-bgp-smp-128x128} 
@@
   \subfigure[128x128, Dual mode]{ 
-  \includegraphics[width=.3\textwidth]{figures/ex27_bgp/s128_dual.eps}  
+  \includegraphics[width=.315\textwidth]{figures/ex27_bgp/s128_dual.eps}  
   \label{fig:ex27-bgp-dual-128x128} 
@@
   \subfigure[128x128, VN mode]{ 
-  \includegraphics[width=.3\textwidth]{figures/ex27_bgp/s128_vn.eps}  
+  \includegraphics[width=.315\textwidth]{figures/ex27_bgp/s128_vn.eps}  
   \label{fig:ex27-bgp-vn-128x128} 
@@
 \end{center}
-\vspace{-.2in}
 \caption{Performance of inode SpMV on Blue Gene/P for problem sizes 32x32, 64x64, and 128x128.} 
@@
 \subsubsection{2-D Finite-Difference Time-Domain Method for Computational Electromagnetics}  
-We consider the two-dimensional Finite Difference Time Domain (FDTD) algorithm, a
-popular method for solving the time-dependent Maxwell's equations in the
-context of computational 
+We consider the two-dimensional Finite Difference Time Domain (FDTD)
+algorithm, a popular method for solving the time-dependent Maxwell's
+equations in the context of computational electrodynamic problems. As
+shown in Figure~\ref{fig:fdtd-2d-code}, the 2-D FDTD method is
+implemented as an outer iteration over time containing four
+imperfectly nested loops.  The arrays $ex$ and $ey$ denote the
+electric field components, and the array $hz$ denotes the magnetic
+field.
+
 %
-\begin{wrapfigure}{r}{3in}
-\vspace{-0.2in}
+\begin{figure}
 \begin{center}
@@
 \end{center}
-\vspace{-0.2in}
 \caption{2-D FDTD code.} 
 \label{fig:fdtd-2d-code} 
-\vspace{-.1in}
-\end{wrapfigure}
+\end{figure}
 %
-electrodynamic problems. As shown in Figure~\ref{fig:fdtd-2d-code}, the 2-D
-FDTD method is implemented as an outer iteration over time containing four imperfectly
-nested loops.  The arrays $ex$ and $ey$ denote the electric field components,
-and the array $hz$ denotes the magnetic field.
 
@@
 
-\begin{figure}[htb] 
+\begin{figure*}%[htb] 
 \begin{center} 
   \subfigure[Sequential (T=500)]{ 
-  \includegraphics[width=.4\textwidth]{figures/fdtd-2d-cookie/seq.eps} 
+  \includegraphics[width=.45\textwidth]{figures/fdtd-2d-cookie/seq.eps} 
   \label{fig:fdtd-2d-cookie-seq} } \subfigure[Parallel (T=500, 
   N=2000)]{ 
-  \includegraphics[width=.4\textwidth]{figures/fdtd-2d-cookie/par.eps} 
+  \includegraphics[width=.45\textwidth]{figures/fdtd-2d-cookie/par.eps} 
   \label{fig:fdtd-2d-cookie-par} } 
 \end{center}
-\vspace{-.2in} 
 \caption{2-D FDTD performance on an eight-core Intel Xeon.} 
 \label{fig:fdtd-2d-cookie-results} 
-%\vspace{-.2in}
-\end{figure}
+\end{figure*}
+
+\begin{figure*}%[htb]
+\begin{center} 
+  \subfigure[Sequential (T=500)]{ 
+  \includegraphics[width=.45\textwidth]{figures/seidel-cookie/seq.eps} 
+  \label{fig:seidel-cookie-seq} } \subfigure[Parallel (T=500, 
+  N=4000)]{ 
+  \includegraphics[width=.45\textwidth]{figures/seidel-cookie/par.eps} 
+  \label{fig:seidel-cookie-par} } 
+\end{center}
+\caption{3-D Gauss-Seidel performance on eight-core Intel machine.} 
+\label{fig:seidel-cookie-results} 
+\end{figure*}
+
+\begin{figure*}%[htb]
+\begin{center} 
+  \subfigure[Sequential]{ 
+  \includegraphics[width=.45\textwidth]{figures/lu-cookie/seq.eps} 
+  \label{fig:lu-cookie-seq} } \subfigure[Parallel (N=4000)]{ 
+  \includegraphics[width=.45\textwidth]{figures/lu-cookie/par.eps} 
+  \label{fig:lu-cookie-par} } 
+\end{center} 
+\caption{LU Decomposition performance on eight-core Intel machine.} 
+\label{fig:lu-cookie-results} 
+\end{figure*}
 
@@
 \emph{successive displacement method}, indicating the dependence of the 
-iterations on the ordering. If the ordering is
+iterations on the ordering. If the ordering is changed, the components
+of the new iterations will change as well.
 %
 %\begin{figure}[ht]
-\begin{wrapfigure}{r}{3in}
-\vspace{-0.3in}
+\begin{figure}%[ht]
 \begin{center}
@@
 \end{center}
-\vspace{-0.2in}
 \caption{3-D Gauss-Seidel code} 
 \label{fig:seidel-code} 
-\vspace{-.1in}
-\end{wrapfigure}
+\end{figure}
 %
-changed, the components of the new iterations will change as well.
 
@@
 from 51\% to 133\%.
-
-\begin{figure} [htb]
-\begin{center} 
-  \subfigure[Sequential (T=500)]{ 
-  \includegraphics[width=.4\textwidth]{figures/seidel-cookie/seq.eps} 
-  \label{fig:seidel-cookie-seq} } \subfigure[Parallel (T=500, 
-  N=4000)]{ 
-  \includegraphics[width=.4\textwidth]{figures/seidel-cookie/par.eps} 
-  \label{fig:seidel-cookie-par} } 
-\end{center}
-\vspace{-.2in} 
-\caption{3-D Gauss-Seidel performance on eight-core Intel machine.} 
-\label{fig:seidel-cookie-results} 
-\vspace{-.1in}
-\end{figure}
 
@@
 
-LU factorization or decomposition is a numerical method for the solution of linear
-systems 
+LU factorization or decomposition is a numerical method for the
+solution of linear systems of equations; a simple implementation is
+shown in Figure~\ref{fig:lu-code}.
 %
-\begin{wrapfigure}{r}{3in}
+\begin{figure}%[ht]
 \begin{center}
-\vspace{-.2in}
 \begin{minipage}{2.8in} 
@@
 \end{center}
-\vspace{-0.2in}
 \caption{LU Decomposition code.} 
 \label{fig:lu-code} 
-\end{wrapfigure}
+\end{figure}
 %
-of equations; a simple implementation is shown in Figure~\ref{fig:lu-code}.
 
@@
 by a factor of 51\% to 120\%.
-
-\begin{figure}[htb]
-\begin{center} 
-  \subfigure[Sequential]{ 
-  \includegraphics[width=.4\textwidth]{figures/lu-cookie/seq.eps} 
-  \label{fig:lu-cookie-seq} } \subfigure[Parallel (N=4000)]{ 
-  \includegraphics[width=.4\textwidth]{figures/lu-cookie/par.eps} 
-  \label{fig:lu-cookie-par} } 
-\end{center} 
-\vspace{-.2in}
-\caption{LU Decomposition performance on eight-core Intel machine.} 
-\label{fig:lu-cookie-results} 
-\end{figure}

### 2009-02-12 doc/papers/ipdps09/motivation.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@428 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@
 
-The size and complexity of scientific computations are increasing at least as
-fast as the improvements in processor technology.  Programming such
-scientific applications is hard, and optimizing them for high performance is
-even harder.  This results in a potentially large gap between the achieved
-performance of applications and the peak available performance, with
-many applications achieving 10\% or less of the peak (e.g.,~
-%\cite{Oliker,
-\cite{conf/vecpar/CarterOS06}).  
-A greater concern is the inability of existing languages, compilers, and
-systems to deliver the available performance for the application through
-fully automated code optimizations.
+The size and complexity of scientific computations are increasing at
+least as fast as the improvements in processor technology.
+Programming such scientific applications is hard, and optimizing them
+for high performance is even harder.  This results in a potentially
+large gap between the achieved performance of applications and the
+peak available performance, with many applications achieving 10\% or
+less of the peak (e.g.,~\cite{Oliker,conf/vecpar/CarterOS06}).  A
+greater concern is the inability of existing languages, compilers, and
+systems to deliver the available performance for the application
+through fully automated code optimizations.
 
-Delivering performance without degrading productivity is crucial for the
-success of scientific computing. Scientific code developers generally attempt
-to improve performance by applying one or more of the following three
-approaches: manually optimizing code fragments, using tuned libraries for key
-numerical algorithms, and, less frequently, using compiler-based source
-transformation tools for loop-level optimizations. Manual tuning is
-time-consuming and impedes readability and performance portability. Tuned
-libraries often deliver great performance without requiring significant
-programming effort, but can only provide limited
-functionality. General-purpose source transformation tools for performance
-optimizations are few and have not yet gained popularity among computational
-scientists, at least in part because of poor portability and steep learning
-curves.
+Delivering performance without degrading productivity is crucial for
+the success of scientific computing. Scientific code developers
+generally attempt to improve performance by applying one or more of
+the following three approaches: manually optimizing code fragments,
+using tuned libraries for key numerical algorithms, and, less
+frequently, using compiler-based source transformation tools for
+loop-level optimizations. Manual tuning is time-consuming and impedes
+readability and performance portability. Tuned libraries often deliver
+great performance without requiring significant programming effort,
+but can only provide limited functionality. General-purpose source
+transformation tools for performance optimizations are few and have
+not yet gained popularity among computational scientists, at least in
+part because of poor portability and steep learning curves.

### 2009-02-12 doc/papers/ipdps09/paper.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@428 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@
-\documentclass[12pt]{article}
+\documentclass[conference]{IEEEtran}
+
+%\documentclass[12pt]{article}
 %\usepackage{latex8}
@@
 \usepackage{amsmath,amsthm}
-\usepackage[letterpaper,margin=1in]{geometry}
 
-\ifpdf
-\DeclareGraphicsExtensions{.pdf,.png,.jpg}
-\else
-\DeclareGraphicsExtensions{.eps,.ps}
-\fi
+%\usepackage[letterpaper,margin=1in]{geometry}
 
-\newenvironment{itemizer}{\begin{itemize}\setlength{\parsep}{0cm}\setlength{\itemsep}{-.3em}}{\end{itemize}}
+%\ifpdf
+%\DeclareGraphicsExtensions{.pdf,.png,.jpg}
+%\else
+%\DeclareGraphicsExtensions{.eps,.ps}
+%\fi
+
+%\newenvironment{itemizer}{\begin{itemize}\setlength{\parsep}{0cm}\setlength{\itemsep}{-.3em}}{\end{itemize}}
 
@@
 
-\title{\vspace{-.6in}Annotation-Based Empirical Performance Tuning\\ Using Orio}
+\begin{document}
+
+\title{Annotation-Based Empirical Performance Tuning Using Orio}
 
 \author{
-Albert Hartono$^1$, Boyana Norris$^2$, and P. Sadayappan$^1$ \\
-{\small $^1$Ohio State University, Dept. of Computer Science and Engineering, Columbus, OH 43210}\\
-{\small \texttt{\{hartonoa, saday\}@cse.ohio-state.edu}}\\
-{\small $^2$Argonne National Laboratory, Mathematics and Computer Science Division, Argonne, IL 60439}\\
-{\small \texttt{norris@mcs.anl.gov}}
+\IEEEauthorblockN{Albert Hartono} 
+\IEEEauthorblockA{\small Dept. of Computer Science and Engg.\\ 
+Ohio State University\\ 
+Columbus, Ohio 43210--1277\\ 
+Email: hartonoa@cse.ohio-state.edu} 
+\and 
+\IEEEauthorblockN{Boyana Norris} 
+\IEEEauthorblockA{\small Mathematics and Computer Science Division\\ 
+Argonne National Laboratory\\ 
+Argonne, Illinois 60439--4844\\
+Email: norris@mcs.anl.gov} 
+\and 
+\IEEEauthorblockN{P. Sadayappan} 
+\IEEEauthorblockA{\small Dept. of Computer Science and Engg.\\ 
+Ohio State University\\ 
+Columbus, Ohio 43210--1277\\ 
+Email: saday@cse.ohio-state.edu}
 }
-
-\begin{document}
 
@@
 \end{abstract}
+
+\IEEEpeerreviewmaketitle
 
@@
 
+\section*{Acknowledgment} 
+The authors would like to thank Uday Bondhugula of IBM Research for
+many productive discussions and his valuable help with Pluto. This
+work was supported in part by the U.S. Dept. of Energy under Contract
+DE-AC02-06CH11357.
+
 \small
-\bibliographystyle{abbrv} 
-\bibliography{paper}
+\bibliographystyle{IEEEtran} 
+\bibliography{IEEEabrv,paper}
 

### 2009-02-12 doc/papers/ipdps09/results.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@430 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@
 2.6.25 (x86-64). Each compute node of the Blue Gene/P is equipped with four
-850MHz IBM PowerPC 450 processors with a dual floating-point unit and 2GB
+850 MHz IBM PowerPC 450 processors with a dual floating-point unit and 2 GB
 total memory per node, private L1 (32 KB) and L2 (4 MB) caches, a shared L3
@@
 
-\caption{(a) Compressed sparse row (CSR) format (b) Basic implementation of CSR-based SpMV.}
+\caption{(a) Compressed sparse row (CSR) format; (b) Basic implementation of CSR-based SpMV.}
 \label{fig:spmv}
@@
 and $x$, $y$ are dense vectors. Each element of $A$ is used precisely once
-and element reuse is only possible for $x$ and $y$. Thus, to optimize SpMV,
+and element reuse is possible for $x$ and $y$. Thus, to optimize SpMV,
 one should use compact data structures for $A$ and try to maximize temporal
@@
 structure. The second inode contains three consecutive rows with identical
-nonzero structure, while the last inode only has a single row. Knowing the
+nonzero structure, and the last inode only has a single row. Knowing the
 properties of each inode at runtime enables maximum reuse of vector $x$ since
@@
 to automatically select the best version of the optimized inode SpMV. Only
-the outer loop which iterates over inodes was parallelized by using OpenMP.
+the outer loop that iterates over inodes was parallelized by using OpenMP.
 
@@
 directions, not the size of the sparse input matrix. The matrix dimension is
-based the grid size and the number of field components computed for each grid
-point, for example, an $8 \times 8$ problem involves sparse matrices of
+based on the grid size and the number of field components computed for each grid
+point; for example, an $8 \times 8$ problem involves sparse matrices of
 dimension 900 with 17,040 nonzero entries. In this experiment, we ran in
-three different multicore modes: SMP (one MPI process, eight
+three multicore modes: SMP (one MPI process, eight
 threads/process), dual (four MPI processes, two threads/process), and virtual
@@
 thread-level parallelism is not exploited; nevertheless, the Orio version
-still performs better than the other two alternatives.
+still performs better than the other two choices.
 %This validates the
@@
 When the input arrays are small, Orio discovers that applying
-Pluto's polyhedral transformations is not beneficial, and therefore,
+Pluto's polyhedral transformations is not beneficial, and therefore
 it employs only its syntactic transformations on the original FDTD
 code. For large problem sizes, Orio exploits some of the Pluto's code
-transformations and enhances it further with its syntactic
+transformations and enhances them further with its syntactic
 optimizations, resulting in performance consistently and significantly
@@
 
-Figure~\ref{fig:fdtd-2d-cookie-par} shows the multicore performance obtained
-for $tmax=500$ and $nx=ny=2000$. The results indicate that whereas icc is
-unable to auto-parallelize the code, Pluto detects the existence of pipelined
-parallelism and then successfully parallelizes the code. Optimizing and
-tuning by Orio additionally improves the Pluto performance by up to
-78\%. Moreover, we observe that because of memory contention between the two
-quad-core Intel processors, the speedup of both the Pluto and Orio codes
-slightly decreases when the number of cores used is greater than four.
+Figure~\ref{fig:fdtd-2d-cookie-par} shows the multicore performance
+obtained for $tmax=500$ and $nx=ny=2000$. The results indicate that
+whereas icc is unable to auto-parallelize the code, Pluto detects the
+existence of pipelined parallelism and then successfully parallelizes
+the code. Locality-exploiting optimizations by Orio additionally
+improve the Pluto performance by up to 78\%. Moreover, we observe
+that because of memory contention between the two quad-core Intel
+processors, the speedup of both the Pluto and Orio codes slightly
+decreases when the number of cores used is greater than four.
 
@@
 Figure~\ref{fig:seidel-code} shows the 3-D Gauss-Seidel computation, 
-which is sometimes is referred to as
+which is sometimes referred to as
 \emph{successive displacement method}, indicating the dependence of the 

### 2009-02-12 doc/papers/ipdps09/results.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@431 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@
 \leftarrow y_{i} + A_{i,j} \cdot x_{j}$, where $A$ is a sparse matrix,
-and $x$, $y$ are dense vectors. Each element of $A$ is used precisely once
-and element reuse is possible for $x$ and $y$. Thus, to optimize SpMV,
+and $x$, $y$ are dense vectors. Each element of $A$ is used precisely once,
+and element reuse is possible only for $x$ and $y$. Thus, to optimize SpMV,
 one should use compact data structures for $A$ and try to maximize temporal
@@
 structure. The second inode contains three consecutive rows with identical
-nonzero structure, and the last inode only has a single row. Knowing the
+nonzero structure, and the last inode has only a single row. Knowing the
 properties of each inode at runtime enables maximum reuse of vector $x$ since
@@
 node or VN (eight MPI processes, one thread/process). The code tuned by Orio
-consistently outperforms the alternatives. Furthermore, the performance of
+consistently outperforms the others. Furthermore, the performance of
 the hand-tuned code is almost equivalent to that of the icc-optimized simple
@@
 unroll factors (64 for 1-D unrolling and 8x8 for 2-D
-unroll/jamming). All codes were compiled with the Intel C Compiler
+unroll/jamming). All codes were compiled with the Intel C compiler
 using the \texttt{-O3} optimization flag to enable auto-vectorization
@@
 \subsubsection{2-D Finite-Difference Time-Domain Method for Computational Electromagnetics}  
-We consider the two-dimensional Finite Difference Time Domain (FDTD)
+We consider the two-dimensional finite difference time domain (FDTD)
 algorithm, a popular method for solving the time-dependent Maxwell's
@@
 code. For large problem sizes, Orio exploits some of the Pluto's code
-transformations and enhances them further with its syntactic
+transformations and enhances these further with its syntactic
 optimizations, resulting in performance consistently and significantly
@@
 obtained for $tmax=500$ and $nx=ny=2000$. The results indicate that
-whereas icc is unable to auto-parallelize the code, Pluto detects the
+whereas icc is unable to autoparallelize the code, Pluto detects the
 existence of pipelined parallelism and then successfully parallelizes
@@
 
-\subsubsection{3-D Gauss-Seidel Successive Over-Relaxation Method}  
+\subsubsection{3-D Gauss-Seidel Successive Overrelaxation Method}  
 Figure~\ref{fig:seidel-code} shows the 3-D Gauss-Seidel computation, 
@@
 \end{center}
-\caption{LU Decomposition code.} 
+\caption{LU decomposition code.} 
 \label{fig:lu-code} 

### 2009-02-12 doc/papers/ipdps09/integration.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@428 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@433 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@
+\begin{figure*}[bt]
+\begin{center}
+\includegraphics[width=0.6\textwidth]{figures/pluto-orio.eps}    
+\end{center}   
+\caption{Integration of Pluto into Orio's code transformation module.}
+\label{fig:pluto-orio}
+\end{figure*}
+
 \section{Pluto-Orio Integration}
@@
 
-Figure~\ref{fig:pluto-orio} outlines the overall structure of the Pluto-Orio
-integrated system, which is implemented as a new optimization module in
-Orio.
-%% BN: the detailed description below is good, but commented out for now for space.
-%Initially, the transformation process takes two kinds of input:
-%the loop code to be optimized and the optimization parameters code. The loop
-%code is embedded in the annotation body block, while the performance
-%parameters code is written in the module body block. A simple parser
-%implemented inside the new module parses the performance parameters code to
-%extract their values. 
-In addition to the loop code to be optimized, the module parses performance
-parameters, which include tile sizes, unroll factors, loop order
-(permutation), as well as several boolean values for triggering OpenMP
-parallelization, scalar replacement, and auto-vectorization. Using these
-parameter values, Pluto then performs polyhedral transformations (e.g.,
-two-level tiling and OpenMP parallelization) on the input loop code. The
-resulting generated code is subsequently passed to the widely-available
-profiling tool gprof~\cite{gprof} for hotspot detection. 
-%The statistical
-%information produced by gprof provides accurate locations (line numbers) of
-%the loop nests where the Pluto-generated code spends most of its execution
-%time. 
-The identified hotspot loop nests are then decorated with Orio's performance
-annotations for complementary syntactic transformations (e.g., loop
-permutation, loop unroll/jamming, scalar replacement, and explicit
-auto-vectorization). Finally, Orio optimizes the annotated hotspots as 
-described in Section~\ref{sec:implm}.
-
-\begin{figure}[tb]
-\begin{center}
-\includegraphics[width=0.6\textwidth]{figures/pluto-orio.eps}    
-\end{center}   
-\vspace{-.2in}
-\caption{Integration of Pluto into Orio's code transformation module.}
-\label{fig:pluto-orio}
-\end{figure}
+Figure~\ref{fig:pluto-orio} outlines the overall structure of the
+Pluto-Orio integrated system, which is implemented as a new
+optimization module in Orio. Initially, the transformation process
+takes two kinds of input: the loop code to be optimized and the
+optimization parameters code. The loop code is embedded in the
+annotation body block, while the performance parameters code is
+written in the module body block. A simple parser implemented inside
+the new module parses the performance parameters code to extract their
+values. In addition to the loop code to be optimized, the module
+parses performance parameters, which include tile sizes, unroll
+factors, loop order (permutation), as well as several boolean values
+for triggering OpenMP parallelization, scalar replacement, and
+auto-vectorization. Using these parameter values, Pluto then performs
+polyhedral transformations (e.g., two-level tiling and OpenMP
+parallelization) on the input loop code. The resulting generated code
+is subsequently passed to the widely-available profiling tool
+gprof~\cite{gprof} for hotspot detection. The statistical information
+produced by gprof provides accurate locations (line numbers) of the
+loop nests where the Pluto-generated code spends most of its execution
+time. The identified hotspot loop nests are then decorated with Orio's
+performance annotations for complementary syntactic transformations
+(e.g., loop permutation, loop unroll/jamming, scalar replacement, and
+explicit auto-vectorization). Finally, Orio optimizes the annotated
+hotspots as described in Section~\ref{sec:implm}.
 

### 2009-02-12 doc/papers/ipdps09/related.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@428 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@433 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@
 
-Ideally, a developer should only have to specify a few simple command-line
-options and then rely on the compiler to optimize the performance of an
-application on any architecture. Compilers alone, however, cannot fully
-satisfy the performance needs of scientific applications.  First,
-compilers must operate in a black-box fashion and at a very low level,
-limiting both the type and number of optimizations that can be done.  Second,
-static analysis of general-purpose languages, such as C, C++, and Fortran, is
-necessarily conservative, thereby precluding many possible optimizations.
-%Third, in the process of transforming a mathematical model into a computer
-%program, much potentially useful (for optimization purposes) information is
-%lost since it cannot be represented by the programming language.  
-Finally, extensive manual tuning of a code may prevent certain compiler optimizations
-and result in worse performance on new architectures, resulting in 
-loss of performance portability.
+Ideally, a developer should only have to specify a few simple
+command-line options and then rely on the compiler to optimize the
+performance of an application on any architecture. Compilers alone,
+however, cannot fully satisfy the performance needs of scientific
+applications.  First, compilers must operate in a black-box fashion
+and at a very low level, limiting both the type and number of
+optimizations that can be done.  Second, static analysis of
+general-purpose languages, such as C, C++, and Fortran, is necessarily
+conservative, thereby precluding many possible optimizations.  Third,
+in the process of transforming a mathematical model into a computer
+program, much potentially useful (for optimization purposes)
+information is lost since it cannot be represented by the programming
+language.  Finally, extensive manual tuning of a code may prevent
+certain compiler optimizations and result in worse performance on new
+architectures, resulting in loss of performance portability.
 
@@
 %% Libraries
+
 An alternative to manual or automated tuning of application codes is the use
@@
 \cite{BLAS,ESSL,Goto:2006fk}) and generating optimized code automatically
-for the given problem and machine parameters. 
-ATLAS~\cite{atlas_sc98}
-%,WN147} 
-for BLAS and LAPACK
-%~\cite{BLAS} and LAPACK~\cite{laug}
-routines, OSKI~\cite{OSKI} for sparse linear algebra,
-PHiPAC~\cite{bilmes97optimizing} for matrix-matrix products, and
-domain-specific libraries such as FFTW~\cite{frigo98} and
+for the given problem and machine
+parameters. ATLAS~\cite{atlas_sc98,WN147} for BLAS~\cite{BLAS} and
+some LAPACK~\cite{laug} routines, OSKI~\cite{OSKI} for sparse linear
+algebra, PHiPAC~\cite{bilmes97optimizing} for matrix-matrix products,
+and domain-specific libraries such as FFTW~\cite{frigo98} and
 SPIRAL~\cite{SPIRAL} are all examples of the latter approach. Most
-automatic tuning approaches perform empirical
-parameter searches on the target platform.  
-%FFTW uses a combination of static
-%models and empirical techniques to optimize FFTs. SPIRAL generates optimized
-%digital signal processing libraries by an extensive empirical search over
-%implementation variants.  GotoBLAS~\cite{Goto:2006fk,Goto:fk}, on the other
-%hand, achieves great performance results on several architectures by using
-%hand-tuned data structures and kernel operations.  
-These auto- or hand-tuned
-approaches can deliver performance that can be five times as fast as that
-produced by many optimizing compilers \cite{WN147}.  The library approach,
-however, is limited by the fact that optimizations are highly problem- and
+automatic tuning approaches perform empirical parameter searches on
+the target platform.  FFTW uses a combination of static models and
+empirical techniques to optimize FFTs. SPIRAL generates optimized
+digital signal processing libraries by an extensive empirical search
+over implementation variants.  GotoBLAS~\cite{Goto:2006fk,Goto:fk}, on
+the other hand, achieves great performance results on several
+architectures by using hand-tuned data structures and kernel
+operations.  These auto- or hand-tuned approaches can deliver
+performance that can be five times as fast as that produced by many
+optimizing compilers
+\cite{WN147}.  The library approach, however, is limited by the fact
+that optimizations are highly problem- and
 machine-dependent. Furthermore, at this time, the functionality of the
@@
 %% Other annotation-based source transformation approaches
-General-purpose tools for optimizing loop performance are also available.
-LoopTool~\cite{LoopTool} supports annotation-based loop fusion,
-unroll/jamming, skewing and tiling.  The Matrix Template Library
-\cite{Siek:1998ys} uses template metaprograms to tile at both the register
-and cache levels.  A new tool, POET~\cite{POET} also supports a number of
-loop transformations. 
-%POET offers a complex template-based syntax for
-%defining transformations in a language-independent manner. 
-Other research efforts whose goal, at least in part, is to enable
-optimizations of source code to be augmented with performance-related
-information include the X language~\cite{XLanguage} (a macro C-like language
-for annotating C code), the Broadway~\cite{broadway} compiler, and
-telescoping languages~\cite{telescopingurl}.
-%,teleoverview,Ken99}
-%, and various
-%meta-programming techniques~\cite{veldhuizen95,weise93,kiczales91,chiba95}.
 
-\comment{
+General-purpose tools for optimizing loop performance are also
+available.  LoopTool~\cite{LoopTool} supports annotation-based loop
+fusion, unroll/jamming, skewing and tiling.  The Matrix Template
+Library \cite{Siek:1998ys} uses template metaprograms to tile at both
+the register and cache levels.  A new tool, POET~\cite{POET} also
+supports a number of loop transformations. POET offers a complex
+template-based syntax for defining transformations in a
+language-independent manner. Other research efforts whose goal, at
+least in part, is to enable optimizations of source code to be
+augmented with performance-related information include the X
+language~\cite{XLanguage} (a macro C-like language for annotating C
+code), the Broadway~\cite{broadway} compiler, and telescoping
+languages~\cite{telescopingurl,teleoverview,Ken99}, and various
+meta-programming
+techniques~\cite{veldhuizen95,weise93,kiczales91,chiba95}.
+
 Emerging annotation-based tools are normally designed by compiler researchers
@@
 capabilities and performance.
-}

### 2009-02-12 doc/papers/ipdps09/results.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@433 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@
 
+\begin{figure}%[b]
+\begin{tabular}{c}
+\begin{minipage}[t]{.47\textwidth}
+\centering
+\vspace{0.2in}
+\includegraphics[width=1.0\textwidth]{figures/spmv.eps}  
+\end{minipage}
+\\
+{\footnotesize (a)}
+\\
+\begin{minipage}[t]{.3\textwidth}
+\scriptsize
+\begin{verbatim}
+
+for (i=0; i<num_rows; i++)
+  for (j=Aptr[i]; j<Aptr[i+1]; j++)
+    y[i] += Aval[j]*x[Aind[j]];
+
+\end{verbatim}
+\end{minipage}
+\\
+{\footnotesize (b)}
+\\
+\end{tabular}
+
+\caption{(a) Compressed sparse row (CSR) format; (b) Basic implementation of CSR-based SpMV.}
+\label{fig:spmv}
+\end{figure}
+
 \begin{figure*}%[t] 
@@
 
-\begin{figure}%[tb]
-\begin{tabular}{c}
-\begin{minipage}[b]{.47\textwidth}
-\centering
-\includegraphics[width=1.0\textwidth]{figures/spmv.eps}  
-\end{minipage}
-\\
-{\footnotesize (a)}
-\\
-\begin{minipage}[b]{.3\textwidth}
-\scriptsize
-\begin{verbatim}
-
-for (i=0; i<num_rows; i++)
-  for (j=Aptr[i]; j<Aptr[i+1]; j++)
-    y[i] += Aval[j]*x[Aind[j]];
-
-\end{verbatim}
-\end{minipage}
-\\
-{\footnotesize (b)}
-\\
-\end{tabular}
-
-\caption{(a) Compressed sparse row (CSR) format; (b) Basic implementation of CSR-based SpMV.}
-\label{fig:spmv}
-\end{figure}
-
 In this section we examine the effectiveness of Orio in optimizing key
@@
 
-The SpMV operation computes $\forall_{A_{i,j}} \neq 0:y_{i}
-\leftarrow y_{i} + A_{i,j} \cdot x_{j}$, where $A$ is a sparse matrix,
-and $x$, $y$ are dense vectors. Each element of $A$ is used precisely once,
-and element reuse is possible only for $x$ and $y$. Thus, to optimize SpMV,
-one should use compact data structures for $A$ and try to maximize temporal
-reuse of $x$ and $y$. One of the most commonly used data structures for
-storing a sparse matrix is compressed sparse row (CSR) storage
-~\cite{vuduc-thesis},
-illustrated in Figure~\ref{fig:spmv}(a). Elements in each row of $A$ are
-packed together in a dense array, \texttt{Aval}, and a corresponding array of
-integers \texttt{Aind} stores the column indices. The \texttt{Aptr} array
-designates where each sparse row begins in \texttt{Aval} and
-\texttt{Aind}. An implementation of SpMV using CSR storage is
-shown in Figure~\ref{fig:spmv}(b).
-
-PETSc's implementation of SpMV used in this experiment exploits matrix
-structure by using \textit{inodes}, which represent rows with identical
-nonzero structure.  PETSc detects inodes automatically in arbitrary sparse
-matrices. For example, the sparse matrix shown in Figure~\ref{fig:spmv}(a)
-contains three inodes. The first inode is of size two because it holds two
-adjacent rows (i.e., the first and the second rows) with the same nonzero
-structure. The second inode contains three consecutive rows with identical
-nonzero structure, and the last inode has only a single row. Knowing the
-properties of each inode at runtime enables maximum reuse of vector $x$ since
-multiple elements of $x$ can be loaded and used exactly once for all rows in
-the same inode structure. To accomplish this, one must use a
-register-blocking transformation. To optimize the inode-based SpMV routine,
-we incorporated a new transformation module inside Orio that implements
-various optimization strategies including register blocking, SIMDization,
-memory alignment optimization, loop-control optimization, accumulator
-expansion, and thread-level parallelization (with OpenMP). We then used Orio
-to automatically select the best version of the optimized inode SpMV. Only
-the outer loop that iterates over inodes was parallelized by using OpenMP.
-
-To evaluate the performance of the tuned SpMV routine, we conducted the
-experiment using a 2-D driven cavity flow simulation
-application~\cite{coff:kell:keye} (SNES ex27 in PETSc), on both the Xeon and
-Blue Gene/P. The ``Compiler-optimized'' label is used as a base case that
-represents the performance of the naive implementation of SpMV (shown in
-Figure~\ref{fig:spmv}(b)) optimized by the native compiler. We also tested
-the performance of the hand-tuned (by PETSc developers) SpMV code, which is
-included in PETSc releases.
-
-Figure~\ref{fig:ex27-cookie-results} shows the performance results on the
-Intel Xeon. The problem size is the number of grid points in the $x$ and $y$
-directions, not the size of the sparse input matrix. The matrix dimension is
-based on the grid size and the number of field components computed for each grid
-point; for example, an $8 \times 8$ problem involves sparse matrices of
-dimension 900 with 17,040 nonzero entries. In this experiment, we ran in
-three multicore modes: SMP (one MPI process, eight
-threads/process), dual (four MPI processes, two threads/process), and virtual
-node or VN (eight MPI processes, one thread/process). The code tuned by Orio
-consistently outperforms the others. Furthermore, the performance of
-the hand-tuned code is almost equivalent to that of the icc-optimized simple
-code. Performance differences between the Orio version and the hand-tuned
-version become more significant as more threads per process are employed to
-facilitate OpenMP parallelization. The SMP and VN results show that
-optimizations using loop-level parallelism (through OpenMP) achieve much
-better performance than using MPI coarse-grained parallelism.
-%Moreover, for the multithread cases, tuning using Orio determined that
-%the performance benefit of SIMDization is more significant than that of
-%thread-level parallelization only when the problem sizes are small. The
-%reason for this is due to the overhead introduced by OpenMP from coordinating
-%its threads across processors.
-%So, each optimized
-%code for the many-thread cases comprises two distinct tuned codes to handle
-
 \begin{figure*}%[ht]
@@
 \end{figure*}
-
-The Blue Gene/P results in Figure~\ref{fig:ex27-bgp-results1} again show
-%Blue Gene/P was done for a varied number of nodes (i.e., 1, 8, and 32)
-%and the same three different parallel modes as before (i.e., SMP,
-%dual, and VN). 
-that empirical optimization using Orio produces the best performance for
-all cases. 
-%outperforming both the hand-tuned and the xlc-optimized codes. 
-On this architecture, the performance gap between the xlc-optimized and the
-hand-tuned codes is now larger than in the Intel experiment. Similarly, by
-exploiting thread-level parallelism, the Orio-optimized code performs better
-than the hand-tuned version when the number of threads per process increases
-and the number of nodes decreases. For the single-thread cases (VN mode),
-thread-level parallelism is not exploited; nevertheless, the Orio version
-still performs better than the other two choices.
-%This validates the
-%effectiveness of Orio in empirically tuning sequential codes.
 
@@
 
+The SpMV operation computes $\forall_{A_{i,j}} \neq 0:y_{i}
+\leftarrow y_{i} + A_{i,j} \cdot x_{j}$, where $A$ is a sparse matrix,
+and $x$, $y$ are dense vectors. Each element of $A$ is used precisely once,
+and element reuse is possible only for $x$ and $y$. Thus, to optimize SpMV,
+one should use compact data structures for $A$ and try to maximize temporal
+reuse of $x$ and $y$. One of the most commonly used data structures for
+storing a sparse matrix is compressed sparse row (CSR) storage
+~\cite{vuduc-thesis},
+illustrated in Figure~\ref{fig:spmv}(a). Elements in each row of $A$ are
+packed together in a dense array, \texttt{Aval}, and a corresponding array of
+integers \texttt{Aind} stores the column indices. The \texttt{Aptr} array
+designates where each sparse row begins in \texttt{Aval} and
+\texttt{Aind}. An implementation of SpMV using CSR storage is
+shown in Figure~\ref{fig:spmv}(b).
+
+PETSc's implementation of SpMV used in this experiment exploits matrix
+structure by using \textit{inodes}, which represent rows with identical
+nonzero structure.  PETSc detects inodes automatically in arbitrary sparse
+matrices. For example, the sparse matrix shown in Figure~\ref{fig:spmv}(a)
+contains three inodes. The first inode is of size two because it holds two
+adjacent rows (i.e., the first and the second rows) with the same nonzero
+structure. The second inode contains three consecutive rows with identical
+nonzero structure, and the last inode has only a single row. Knowing the
+properties of each inode at runtime enables maximum reuse of vector $x$ since
+multiple elements of $x$ can be loaded and used exactly once for all rows in
+the same inode structure. To accomplish this, one must use a
+register-blocking transformation. To optimize the inode-based SpMV routine,
+we incorporated a new transformation module inside Orio that implements
+various optimization strategies including register blocking, SIMDization,
+memory alignment optimization, loop-control optimization, accumulator
+expansion, and thread-level parallelization (with OpenMP). We then used Orio
+to automatically select the best version of the optimized inode SpMV. Only
+the outer loop that iterates over inodes was parallelized by using OpenMP.
+
+To evaluate the performance of the tuned SpMV routine, we conducted the
+experiment using a 2-D driven cavity flow simulation
+application~\cite{coff:kell:keye} (SNES ex27 in PETSc), on both the Xeon and
+Blue Gene/P. The ``Compiler-optimized'' label is used as a base case that
+represents the performance of the naive implementation of SpMV (shown in
+Figure~\ref{fig:spmv}(b)) optimized by the native compiler. We also tested
+the performance of the hand-tuned (by PETSc developers) SpMV code, which is
+included in PETSc releases.
+
+Figure~\ref{fig:ex27-cookie-results} shows the performance results on the
+Intel Xeon. The problem size is the number of grid points in the $x$ and $y$
+directions, not the size of the sparse input matrix. The matrix dimension is
+based on the grid size and the number of field components computed for each grid
+point; for example, an $8 \times 8$ problem involves sparse matrices of
+dimension 900 with 17,040 nonzero entries. In this experiment, we ran in
+three multicore modes: SMP (one MPI process, eight
+threads/process), dual (four MPI processes, two threads/process), and virtual
+node or VN (eight MPI processes, one thread/process). The code tuned by Orio
+consistently outperforms the others. Furthermore, the performance of
+the hand-tuned code is almost equivalent to that of the icc-optimized simple
+code. Performance differences between the Orio version and the hand-tuned
+version become more significant as more threads per process are employed to
+facilitate OpenMP parallelization. The SMP and VN results show that
+optimizations using loop-level parallelism (through OpenMP) achieve much
+better performance than using MPI coarse-grained parallelism.
+%Moreover, for the multithread cases, tuning using Orio determined that
+%the performance benefit of SIMDization is more significant than that of
+%thread-level parallelization only when the problem sizes are small. The
+%reason for this is due to the overhead introduced by OpenMP from coordinating
+%its threads across processors.
+%So, each optimized
+%code for the many-thread cases comprises two distinct tuned codes to handle
+
+The Blue Gene/P results in Figure~\ref{fig:ex27-bgp-results1} again show
+%Blue Gene/P was done for a varied number of nodes (i.e., 1, 8, and 32)
+%and the same three different parallel modes as before (i.e., SMP,
+%dual, and VN). 
+that empirical optimization using Orio produces the best performance for
+all cases. 
+%outperforming both the hand-tuned and the xlc-optimized codes. 
+On this architecture, the performance gap between the xlc-optimized and the
+hand-tuned codes is now larger than in the Intel experiment. Similarly, by
+exploiting thread-level parallelism, the Orio-optimized code performs better
+than the hand-tuned version when the number of threads per process increases
+and the number of nodes decreases. For the single-thread cases (VN mode),
+thread-level parallelism is not exploited; nevertheless, the Orio version
+still performs better than the other two choices.
+%This validates the
+%effectiveness of Orio in empirically tuning sequential codes.
+
 \subsection{Evaluation of the Pluto-Orio Integrated System}
@@
 In the following sections, we refer to the icc-optimized base code as
-``ICC,'' the Pluto-generated code with L1 tiling and
+xs``ICC,'' the Pluto-generated code with L1 tiling and
 unroll/jamming as ``Pluto (L1 tiling),'' and the Pluto-generated code with L1
@@
 tuned by Orio is referred to as ``Pluto+Orio.''
-
-\subsubsection{2-D Finite-Difference Time-Domain Method for Computational Electromagnetics}  
-We consider the two-dimensional finite difference time domain (FDTD)
-algorithm, a popular method for solving the time-dependent Maxwell's
-equations in the context of computational electrodynamic problems. As
-shown in Figure~\ref{fig:fdtd-2d-code}, the 2-D FDTD method is
-implemented as an outer iteration over time containing four
-imperfectly nested loops.  The arrays $ex$ and $ey$ denote the
-electric field components, and the array $hz$ denotes the magnetic
-field.
-
-%
-\begin{figure}
-\begin{center}
-\begin{minipage}{3in} 
-\scriptsize
-\begin{verbatim} 
-for(t=0; t<tmax; t++) { 
-  for (j=0; j<ny; j++) ey[0][j] = t; 
-  for (i=1; i<nx; i++) 
-    for (j=0; j<ny; j++) 
-      ey[i][j] -= 0.5*(hz[i][j] - hz[i-1][j]); 
-  for (i=0; i<nx; i++) 
-    for (j=1; j<ny; j++) 
-      ex[i][j] -= 0.5*(hz[i][j] - hz[i][j-1]); 
-  for (i=0; i<nx; i++) 
-    for (j=0; j<ny; j++) 
-      hz[i][j] -= 0.7*(ex[i][j+1] - ex[i][j]
-                  + ey[i+1][j] - ey[i][j]); 
-} 
-\end{verbatim} 
-\end{minipage} 
-\end{center}
-\caption{2-D FDTD code.} 
-\label{fig:fdtd-2d-code} 
-\end{figure}
-%
-
-The performance of the sequential 2-D FDTD code for $tmax=500$ and $nx=ny$ is
-shown in Figure~\ref{fig:fdtd-2d-cookie-seq}. The base code optimized by icc
-alone performs better than Pluto for small problem sizes since all input
-arrays fit in the L2 cache (insufficient computation to offset Pluto's tiling
-overhead). As the array sizes increase, the lack of data reuse impairs the
-base code's performance, whereas the Pluto performance remains about the
-same.
-%We used Orio to generate two code variants: one for small problem sizes
-%and another for large problem sizes. 
-When the input arrays are small, Orio discovers that applying
-Pluto's polyhedral transformations is not beneficial, and therefore
-it employs only its syntactic transformations on the original FDTD
-code. For large problem sizes, Orio exploits some of the Pluto's code
-transformations and enhances these further with its syntactic
-optimizations, resulting in performance consistently and significantly
-higher than both the base and Pluto codes (up to 86\% over Pluto).
-
-%By tuning the sequential code with Orio,
-%the attainable performance enhancement over Pluto is up to 86\%.
-
-Figure~\ref{fig:fdtd-2d-cookie-par} shows the multicore performance
-obtained for $tmax=500$ and $nx=ny=2000$. The results indicate that
-whereas icc is unable to autoparallelize the code, Pluto detects the
-existence of pipelined parallelism and then successfully parallelizes
-the code. Locality-exploiting optimizations by Orio additionally
-improve the Pluto performance by up to 78\%. Moreover, we observe
-that because of memory contention between the two quad-core Intel
-processors, the speedup of both the Pluto and Orio codes slightly
-decreases when the number of cores used is greater than four.
 
@@
 
-%%BN: commented out since this needs more explanation (no room in this version)
-%It is to be noted that collecting performance results of the Pluto 
-%code with L1 and L2 tilings was not possible because of the large code 
-%size of the generated loop nest, which led to an impractically huge 
-%compile time.
+\subsubsection{2-D Finite-Difference Time-Domain Method for Computational Electromagnetics}  
+We consider the two-dimensional finite difference time domain (FDTD)
+algorithm, a popular method for solving the time-dependent Maxwell's
+equations in the context of computational electrodynamic problems. As
+shown in Figure~\ref{fig:fdtd-2d-code}, the 2-D FDTD method is
+implemented as an outer iteration over time containing four
+imperfectly nested loops.  The arrays $ex$ and $ey$ denote the
+electric field components, and the array $hz$ denotes the magnetic
+field.
 
-\subsubsection{3-D Gauss-Seidel Successive Overrelaxation Method}  
-Figure~\ref{fig:seidel-code} shows the 3-D Gauss-Seidel computation, 
-which is sometimes referred to as
-\emph{successive displacement method}, indicating the dependence of the 
-iterations on the ordering. If the ordering is changed, the components
-of the new iterations will change as well.
-%
-%\begin{figure}[ht]
+\begin{figure}
+\begin{center}
+\begin{minipage}{3in} 
+\scriptsize
+\begin{verbatim} 
+for(t=0; t<tmax; t++) { 
+  for (j=0; j<ny; j++) ey[0][j] = t; 
+  for (i=1; i<nx; i++) 
+    for (j=0; j<ny; j++) 
+      ey[i][j] -= 0.5*(hz[i][j] - hz[i-1][j]); 
+  for (i=0; i<nx; i++) 
+    for (j=1; j<ny; j++) 
+      ex[i][j] -= 0.5*(hz[i][j] - hz[i][j-1]); 
+  for (i=0; i<nx; i++) 
+    for (j=0; j<ny; j++) 
+      hz[i][j] -= 0.7*(ex[i][j+1] - ex[i][j]
+                  + ey[i+1][j] - ey[i][j]); 
+} 
+\end{verbatim} 
+\end{minipage} 
+\end{center}
+\caption{2-D FDTD code.} 
+\label{fig:fdtd-2d-code} 
+\end{figure}
+
 \begin{figure}%[ht]
@@
 \end{figure}
-%
+
+\begin{figure}%[ht]
+\begin{center}
+\begin{minipage}{2.8in} 
+\scriptsize
+\begin{verbatim} 
+for (k=0; k<=N-1; k++) { 
+  for (j=k+1; j<=N-1; j++) 
+    A[k][j] /= A[k][k]; 
+  for(i=k+1; i<=N-1; i++) 
+    for (j=k+1; j<=N-1; j++) 
+      A[i][j] -= A[i][k]*A[k][j]; 
+} 
+\end{verbatim} 
+\end{minipage} 
+\end{center}
+\caption{LU decomposition code.} 
+\label{fig:lu-code} 
+\end{figure}
+
+The performance of the sequential 2-D FDTD code for $tmax=500$ and $nx=ny$ is
+shown in Figure~\ref{fig:fdtd-2d-cookie-seq}. The base code optimized by icc
+alone performs better than Pluto for small problem sizes since all input
+arrays fit in the L2 cache (insufficient computation to offset Pluto's tiling
+overhead). As the array sizes increase, the lack of data reuse impairs the
+base code's performance, whereas the Pluto performance remains about the
+same.
+%We used Orio to generate two code variants: one for small problem sizes
+%and another for large problem sizes. 
+When the input arrays are small, Orio discovers that applying
+Pluto's polyhedral transformations is not beneficial, and therefore
+it employs only its syntactic transformations on the original FDTD
+code. For large problem sizes, Orio exploits some of the Pluto's code
+transformations and enhances these further with its syntactic
+optimizations, resulting in performance consistently and significantly
+higher than both the base and Pluto codes (up to 86\% over Pluto).
+
+%By tuning the sequential code with Orio,
+%the attainable performance enhancement over Pluto is up to 86\%.
+
+Figure~\ref{fig:fdtd-2d-cookie-par} shows the multicore performance
+obtained for $tmax=500$ and $nx=ny=2000$. The results indicate that
+whereas icc is unable to autoparallelize the code, Pluto detects the
+existence of pipelined parallelism and then successfully parallelizes
+the code. Locality-exploiting optimizations by Orio additionally
+improve the Pluto performance by up to 78\%. Moreover, we observe
+that because of memory contention between the two quad-core Intel
+processors, the speedup of both the Pluto and Orio codes slightly
+decreases when the number of cores used is greater than four.
+
+%%BN: commented out since this needs more explanation (no room in this version)
+%It is to be noted that collecting performance results of the Pluto 
+%code with L1 and L2 tilings was not possible because of the large code 
+%size of the generated loop nest, which led to an impractically huge 
+%compile time.
+
+\subsubsection{3-D Gauss-Seidel Successive Overrelaxation Method}  
+Figure~\ref{fig:seidel-code} shows the 3-D Gauss-Seidel computation, 
+which is sometimes referred to as
+\emph{successive displacement method}, indicating the dependence of the 
+iterations on the ordering. If the ordering is changed, the components
+of the new iterations will change as well.
 
@@
 shown in Figure~\ref{fig:lu-code}.
-%
-\begin{figure}%[ht]
-\begin{center}
-\begin{minipage}{2.8in} 
-\scriptsize
-\begin{verbatim} 
-for (k=0; k<=N-1; k++) { 
-  for (j=k+1; j<=N-1; j++) 
-    A[k][j] /= A[k][k]; 
-  for(i=k+1; i<=N-1; i++) 
-    for (j=k+1; j<=N-1; j++) 
-      A[i][j] -= A[i][k]*A[k][j]; 
-} 
-\end{verbatim} 
-\end{minipage} 
-\end{center}
-\caption{LU decomposition code.} 
-\label{fig:lu-code} 
-\end{figure}
-%
 

### 2009-02-12 doc/papers/iccs09/paper.tex
the final submitted version, really
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@405 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@435 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@
 MATLAB-to-C compiler and the empirical performance
-tuning system Orio and its use in conjunction with the PLuTo tool suite to
+tuning system Orio and its use in conjunction with the Pluto tool suite to
 generate and empirically evaluate many tuned versions of the C code generated
 by the MATLAB compiler.  In Section~\ref{sec:experiments} we provide
-performance results on two architectures. We conclude with a summary and
-discussion future work in Section~\ref{sec:conclusion}.
+performance results on two architectures. In Section~\ref{sec:conclusion} we conclude 
+with a brief summary.
 
@@
 %techniques (e.g., ~\cite{veldhuizen95,chiba95}). 
-PLuTo~\cite{Pluto}
+Pluto~\cite{Pluto}
 %uday08cc} 
 is a source-to-source transformation tool for
-optimizing sequences of nested loops. PLuTo employs a polyhedral model of
+optimizing sequences of nested loops. Pluto employs a polyhedral model of
 nested loops, where the dynamic instance (iteration) of each statement is
@@
 the C code from the high-level MATLAB prototype, we (optionally) use 
-the source-to-source automatic parallelization tool PLuTo~\cite{Pluto} %developed at Ohio State University
+the source-to-source automatic parallelization tool Pluto~\cite{Pluto} %developed at Ohio State University
 to optimize for coarse-grained parallelism and locality simultaneously. Using
-the results of the PLuTo analysis, we insert annotations into the C code,
+the results of the Pluto analysis, we insert annotations into the C code,
 which are then processed by our extensible annotation system Orio to generate
@@
 Figure~\ref{fig:results}(b). The Orio-tuned version that incorporates
-PLuTo-generated loop fusion optimizations and Orio parallelization
+Pluto-generated loop fusion optimizations and Orio parallelization
 directives achieves the best performance for most problem sizes,
@@
 Figure~\ref{fig:results}(c) shows the performance (in seconds) of the GEMVER
-operation on the Xeon workstation. Here we used the same PLuTo and Orio
+operation on the Xeon workstation. Here we used the same Pluto and Orio
 optimizations as for the ATAX example. Similarly, the parallel Orio version
@@
 The performance (in seconds) for the BiCG kernel operation is shown in
-Figure~\ref{fig:results}(d). For this operation, the PLuTo analysis did not
+Figure~\ref{fig:results}(d). For this operation, the Pluto analysis did not
 result in performance improvement. Thus we are showing the results obtained
@@
 translate the MATLAB code to C, (2) analyze the resulting loops and identify
-locality and parallelism-enhancing optimizations using PLuTo, and (3)
+locality and parallelism-enhancing optimizations using Pluto, and (3)
 annotate the resulting C code with syntactic performance directives and use
@@
 %optimizations with new ones implemented from scratch or through interfaces to
-%other tools, as we have successfully done with PLuTo. We are also working on
+%other tools, as we have successfully done with Pluto. We are also working on
 %generating code that contains special cases for different input parameter

### 2009-02-12 doc/papers/ipdps09/results.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@434 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@436 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@
 
+The SpMV operation computes $\forall_{A_{i,j}} \neq 0:y_{i}
+\leftarrow y_{i} + A_{i,j} \cdot x_{j}$, where $A$ is a sparse matrix,
+and $x$, $y$ are dense vectors. Each element of $A$ is used precisely once,
+and element reuse is possible only for $x$ and $y$. Thus, to optimize SpMV,
+one should use compact data structures for $A$ and try to maximize temporal
+reuse of $x$ and $y$. One of the most commonly used data structures for
+storing a sparse matrix is compressed sparse row (CSR) storage
+~\cite{vuduc-thesis},
+illustrated in Figure~\ref{fig:spmv}(a). Elements in each row of $A$ are
+packed together in a dense array, \texttt{Aval}, and a corresponding array of
+integers \texttt{Aind} stores the column indices. The \texttt{Aptr} array
+designates where each sparse row begins in \texttt{Aval} and
+\texttt{Aind}. An implementation of SpMV using CSR storage is
+shown in Figure~\ref{fig:spmv}(b).
+
+PETSc's implementation of SpMV used in this experiment exploits matrix
+structure by using \textit{inodes}, which represent rows with identical
+nonzero structure.  PETSc detects inodes automatically in arbitrary sparse
+matrices. For example, the sparse matrix shown in Figure~\ref{fig:spmv}(a)
+contains three inodes. The first inode is of size two because it holds two
+adjacent rows (i.e., the first and the second rows) with the same nonzero
+structure. The second inode contains three consecutive rows with identical
+nonzero structure, and the last inode has only a single row. Knowing the
+properties of each inode at runtime enables maximum reuse of vector $x$ since
+multiple elements of $x$ can be loaded and used exactly once for all rows in
+the same inode structure. To accomplish this, one must use a
+register-blocking transformation. To optimize the inode-based SpMV routine,
+we incorporated a new transformation module inside Orio that implements
+various optimization strategies including register blocking, SIMDization,
+memory alignment optimization, loop-control optimization, accumulator
+expansion, and thread-level parallelization (with OpenMP). We then used Orio
+to automatically select the best version of the optimized inode SpMV. Only
+the outer loop that iterates over inodes was parallelized by using OpenMP.
+
+To evaluate the performance of the tuned SpMV routine, we conducted the
+experiment using a 2-D driven cavity flow simulation
+application~\cite{coff:kell:keye} (SNES ex27 in PETSc), on both the Xeon and
+Blue Gene/P. The ``Compiler-optimized'' label is used as a base case that
+represents the performance of the naive implementation of SpMV (shown in
+Figure~\ref{fig:spmv}(b)) optimized by the native compiler. We also tested
+the performance of the hand-tuned (by PETSc developers) SpMV code, which is
+included in PETSc releases.
+
 \begin{figure*}%[ht]
@@
 
-The SpMV operation computes $\forall_{A_{i,j}} \neq 0:y_{i}
-\leftarrow y_{i} + A_{i,j} \cdot x_{j}$, where $A$ is a sparse matrix,
-and $x$, $y$ are dense vectors. Each element of $A$ is used precisely once,
-and element reuse is possible only for $x$ and $y$. Thus, to optimize SpMV,
-one should use compact data structures for $A$ and try to maximize temporal
-reuse of $x$ and $y$. One of the most commonly used data structures for
-storing a sparse matrix is compressed sparse row (CSR) storage
-~\cite{vuduc-thesis},
-illustrated in Figure~\ref{fig:spmv}(a). Elements in each row of $A$ are
-packed together in a dense array, \texttt{Aval}, and a corresponding array of
-integers \texttt{Aind} stores the column indices. The \texttt{Aptr} array
-designates where each sparse row begins in \texttt{Aval} and
-\texttt{Aind}. An implementation of SpMV using CSR storage is
-shown in Figure~\ref{fig:spmv}(b).
-
-PETSc's implementation of SpMV used in this experiment exploits matrix
-structure by using \textit{inodes}, which represent rows with identical
-nonzero structure.  PETSc detects inodes automatically in arbitrary sparse
-matrices. For example, the sparse matrix shown in Figure~\ref{fig:spmv}(a)
-contains three inodes. The first inode is of size two because it holds two
-adjacent rows (i.e., the first and the second rows) with the same nonzero
-structure. The second inode contains three consecutive rows with identical
-nonzero structure, and the last inode has only a single row. Knowing the
-properties of each inode at runtime enables maximum reuse of vector $x$ since
-multiple elements of $x$ can be loaded and used exactly once for all rows in
-the same inode structure. To accomplish this, one must use a
-register-blocking transformation. To optimize the inode-based SpMV routine,
-we incorporated a new transformation module inside Orio that implements
-various optimization strategies including register blocking, SIMDization,
-memory alignment optimization, loop-control optimization, accumulator
-expansion, and thread-level parallelization (with OpenMP). We then used Orio
-to automatically select the best version of the optimized inode SpMV. Only
-the outer loop that iterates over inodes was parallelized by using OpenMP.
-
-To evaluate the performance of the tuned SpMV routine, we conducted the
-experiment using a 2-D driven cavity flow simulation
-application~\cite{coff:kell:keye} (SNES ex27 in PETSc), on both the Xeon and
-Blue Gene/P. The ``Compiler-optimized'' label is used as a base case that
-represents the performance of the naive implementation of SpMV (shown in
-Figure~\ref{fig:spmv}(b)) optimized by the native compiler. We also tested
-the performance of the hand-tuned (by PETSc developers) SpMV code, which is
-included in PETSc releases.
-
 Figure~\ref{fig:ex27-cookie-results} shows the performance results on the
@@
 \end{center}
-\caption{3-D Gauss-Seidel performance on eight-core Intel machine.} 
+\caption{3-D Gauss-Seidel performance on eight-core Intel Xeon.} 
 \label{fig:seidel-cookie-results} 
@@
 \end{center} 
-\caption{LU Decomposition performance on eight-core Intel machine.} 
+\caption{LU decomposition performance on eight-core Intel Xeon.} 
 \label{fig:lu-cookie-results} 
@@
 obtained for $tmax=500$ and $nx=ny=2000$. The results indicate that
-whereas icc is unable to autoparallelize the code, Pluto detects the
+whereas icc is unable to auto-parallelize the code, Pluto detects the
 existence of pipelined parallelism and then successfully parallelizes

### 2009-02-12 doc/papers/ipdps09/related.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@437 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@
 
-An alternative to manual or automated tuning of application codes is the use
-of tuned libraries. The two basic approaches to supplying high-performance
-libraries include providing a library of hand-coded options (e.g.,
-\cite{BLAS,ESSL,Goto:2006fk}) and generating optimized code automatically
-for the given problem and machine
+An alternative to manual or automated tuning of application codes is
+the use of tuned libraries. The two basic approaches to supplying
+high-performance libraries include providing a library of hand-coded
+options (e.g., \cite{BLAS,ESSL,Goto:2006fk}) and generating optimized
+code automatically for the given problem and machine
 parameters. ATLAS~\cite{atlas_sc98,WN147} for BLAS~\cite{BLAS} and
@@
 approach at a time. Furthermore, at present, there is no good way for
-\emph{users} to learn about the tools available and compare their
+users to learn about the tools available and compare their
 capabilities and performance.

### 2009-02-13 doc/papers/iccs09/paper.tex
Final changes to address reviewer comments (mainly Reviewer 3)
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@441 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@
 BLAS operation, we follow a three-step approach, illustrated in
-Figure~\ref{fig:process}.  
+Figure~\ref{fig:process}.  This process is repeated when the MATLAB code changes
+or the code must be tuned on a new architecture.
+To begin, we have developed a compiler that
 %
@@
 \centering
-\includegraphics[width=3in]{figures/process.png}
+\includegraphics[width=3in]{figures/process.pdf}
+\vspace{-.15in}
 \caption{Code generation and tuning process.}
 \label{fig:process}
-\vspace{-.2in}
+\vspace{-.1in}
 \end{wrapfigure}
 %
-To begin, we have developed a compiler that
 converts a MATLAB script into simple C code~\cite{Siek}.  After generating
@@
 \centering
-\includegraphics[width=2in]{figures/compile.png}
+\includegraphics[width=2in]{figures/compile.pdf}
 \caption{MATLAB-to-C compiler.}
@@
 Orio. The input to Orio is C code containing semantic comments that
-describe both the computation and various performance-tuning
-directives. Orio first extracts all annotation code regions by parsing
-the marked-up input code. Each annotated region is then passed to code
-transformation module and code generator for potential optimizations. Next the
-transformed C code with various incorporated optimizations
-corresponding to the specified annotations is produced. The
-source-to-source transformation system of Orio generates an optimized
-code version for each distinct combination of performance parameter
-values. Each generated code variant is then executed and its
-performance cost evaluated. After iteratively testing all code
-variants, the best-performing code is selected as the final output of
-Orio. Because the search space of all possible optimized code variants
-can be exponentially large, the search engine implements a number of
-search heuristics (i.e., random, simplex, and simulated annealing) to
-effectively narrow the search for near-optimal performance. The tuning
-specifications, written by users in the form of annotations, contains
-information required for guiding the transformation and tuning
-process.
+describe both the computation (using a syntax that is more restricted
+than the original C) and various performance-tuning directives. Orio
+first extracts all annotation code regions by parsing the marked-up
+input code. Each annotated region is then passed to code
+transformation module and code generator for potential
+optimizations. Next the transformed C code with various incorporated
+optimizations corresponding to the specified annotations is
+produced. Orio generates an optimized code version for each distinct
+combination of performance parameter values. Each code
+variant is then executed and its performance measured. After
+iteratively testing all code variants, the best-performing code is
+selected as the final output of Orio. While each variant is
+computationally very cheap (a single code variant takes between a
+fraction of a second to a few seconds depending on the input sizes),
+the search space of all possible optimized code variants can be
+exponentially large. Therefore, the search engine implements a number
+of search heuristics (i.e., random, simplex, and simulated annealing)
+to effectively narrow the search for near-optimal performance and
+reduce the empirical search time. 
+%The tuning
+%specifications, written by users in the form of annotations, contains
+%information required for guiding the transformation and tuning
+%process.
 
@@
 \begin{center}
-\includegraphics[width=.65\textwidth]{figures/orio.png}
+\includegraphics[width=.65\textwidth]{figures/orio.pdf}
 \end{center}
@@
 \label{fig:orio}
-\vspace{-.2in}
+\vspace{-.15in}
 \end{figure}
@@
 when parallel resources are available. In this example, the parallel Orio
-driver simultaneously executes 64 code variants in the same parallel job. The
-commands used by Orio to submit a parallel job and to query its status are
-also specified in the tuning specifications. At present, users must create
-the tuning specifications manually. When Orio is used in conjunction with
-compiler tools, such as the MATLAB compiler described in this paper, it
-should eventually be possible to automatically generate the tuning
+driver simultaneously executes 64 code variants in the same parallel job. 
+%The
+%commands used by Orio to submit a parallel job and to query its status are
+%also specified in the tuning specifications. 
+At present, users must create the tuning specifications manually for
+each architecture. When Orio is used in conjunction with compiler
+tools, such as the MATLAB compiler described in this paper, it should
+eventually be possible to automatically generate the tuning
 specifications.
@@
 \begin{minipage}[b]{.5\textwidth}
-\includegraphics[width=\textwidth]{figures/vadd_bgp.png}
+\includegraphics[width=\textwidth]{figures/vadd_bgp_time2.pdf}
 \end{minipage}
@@
 \begin{minipage}[b]{.5\textwidth}
-\includegraphics[width=\textwidth]{figures/atax.png}
+\includegraphics[width=\textwidth]{figures/atax.pdf}
 \end{minipage}\\
@@
 \begin{minipage}[b]{.5\textwidth}
-\includegraphics[width=\textwidth]{figures/gemver.png}
+\includegraphics[width=\textwidth]{figures/gemver.pdf}
 \end{minipage}
@@
 \begin{minipage}[b]{.5\textwidth}
-\includegraphics[width=\textwidth]{figures/bicgkernel.png}
+\includegraphics[width=\textwidth]{figures/bicgkernel.pdf}
 \end{minipage}\\
@@
 experiment, we measured the performance for both the sequential and parallel
-scenarios. Even for a very simple operation such as vector addition the
+scenarios (indicated by (S) and (P) in the legend, respectively). Even for a very simple operation such as vector addition the
 compiler alone is unable to obtain the same level of performance as the
@@
 the sequential and parallel code variants tuned by Orio (``Orio
-(Seq.)'' and ``Orio (Par.)'', respectively).
+(S)'' and ``Orio (P)'', respectively).
 

### 2009-02-15 doc/papers/ipdps09/implementation.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@428 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@433 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@443 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@
+\begin{figure*}%[b]
+\begin{center}
+\includegraphics[width=0.6\textwidth]{figures/orio.eps}   
+\end{center}
+\caption{Overview of Orio's code generation and empirical tuning process.}  
+\label{fig:orio}
+\end{figure*}
+
+\begin{figure*}%[b]
+\begin{center} 
+\begin{tabular}{rrl} 
+$<$annotation-region$>$ & ::= & $<$leader-annotation$>$ $<$annotation-body$>$ $<$trailer-annotation$>$\\ 
+$<$leader-annotation$>$ & ::= & \texttt{/*@ begin} $<$module-name$>$ \texttt{(} $<$module-body$>$ \texttt{) @*/} \\
+$<$trailer-annotation$>$ & ::= & \texttt{/*@ end @*/} \\
+\end{tabular} 
+\end{center}  
+\caption{Annotation language grammar excerpt.}  
+\label{fig:ann-lang}
+\end{figure*}
+
 \section{Orio Design and Implementation}
@@
 extent possible.
-
-%\subsection{Overall Approach} 
-%\label{sec:overall}
-\begin{figure}[tb]
-\vspace{-.2in}
-\begin{center}  
-\includegraphics[width=0.6\textwidth]{figures/orio.eps}   
-\end{center}  
-\vspace{-.2in}
-\caption{Overview of Orio's code generation and empirical tuning process.}  
-\label{fig:orio}  
-\end{figure}
 
@@
 
-\begin{figure}[thb]
-\begin{center} 
-\begin{tabular}{rrl} 
-$<$annotation-region$>$ & ::= & $<$leader-annotation$>$ $<$annotation-body$>$ $<$trailer-annotation$>$\\ 
-$<$leader-annotation$>$ & ::= & \texttt{/*@ begin} $<$module-name$>$ \texttt{(} $<$module-body$>$ \texttt{) @*/} \\
-$<$trailer-annotation$>$ & ::= & \texttt{/*@ end @*/} \\
-\end{tabular} 
-\end{center}  
-\vspace{-.1in}
-\caption{Annotation language grammar excerpt.}  
-\label{fig:ann-lang}
-\end{figure}
-
 \subsection{Orio Input Example}
@@
 
-\begin{figure}[thb] 
+\begin{figure*}%[!t] 
 \centering 
@@
 \\
-%(a) Annotated AXPY-4 source code
-%& 
-%(b) Tuning specification for the Blue Gene/P\\
 \end{tabular}
@@
 \label{fig:orio-example}  
-%\vspace{-.2in}
-\end{figure}
+\end{figure*}
+
+\begin{figure*}%[!t] 
+\centering 
+\begin{tabular}{cc} 
+\begin{minipage}{.5\textwidth}  
+\scriptsize
+\begin{verbatim}  
+ /*@ begin Loop (
+ transform UnrollJam(ufactor=Ui)
+ for (i=0; i<=M-1; i++)
+   transform UnrollJam(ufactor=Uj)
+   for (j=0; j<=N-1; j++)
+     transform UnrollJam(ufactor=Uk)   
+     for (k=0; k<=O-1; k++)
+       A[i][j] += B[i][k]*C[k][j];
+ ) @*/
+ for (i=0; i<=M-1; i++)
+  for (j=0; j<=N-1; j++)
+   for (k=0; k<=O-1; k++)
+     A[i][j] += B[i][k]*C[k][j];
+ /*@ end @*/
+\end{verbatim}  
+\end{minipage}
+&
+\begin{minipage}{.5\textwidth}  
+\scriptsize
+\begin{verbatim}  
+ def performance_params {
+   param Ui[] = range(1,33);
+   param Uj[] = range(1,33);
+   param Uk[] = range(1,33);
+   constraint reg_capacity = Ui*Uj+Ui*Uk+Uk*Uj<=32;
+ }
+
+ def input_params {
+   param M[] = [10,50,100,500,1000];
+   param N[] = [10,50,100,500,1000];
+   param O[] = [10,50,100,500,1000];
+   constraint square_matrices = (M==N) and (N==O);
+ }
+
+\end{verbatim}  
+\end{minipage}
+\\
+\end{tabular}
+\caption{Example of specifying parameter constraints in Orio; annotated code for matrix-matrix multiplication (left) and constraint specification (right).}
+\label{fig:par-constraint}  
+\end{figure*}
 
@@
 order to find the most optimal available version. In the context of empirical
-optimization, code variants are alternative, semantically equivalent,
+optimization, code variants are alternative, semantically equivalent
 implementations of the same computation. Each implementation variant is
@@
 parameter values is to use an exhaustive search procedure that is guaranteed
-to determine the optimal code version. Normally the size of the search space
+to find the optimal code version. Normally the size of the search space
 is too large, making full coverage impractical.  Thus, in addition to
@@
 
-\begin{figure}[htp] 
-\centering 
-\begin{tabular}{cc} 
-\begin{minipage}{.4\textwidth}  
-\scriptsize
-\begin{verbatim}  
-/*@ begin Loop (
-transform UnrollJam(ufactor=Ui)
-for (i=0; i<=M-1; i++)
-  transform UnrollJam(ufactor=Uj)
-  for (j=0; j<=N-1; j++)
-    transform UnrollJam(ufactor=Uk)   
-    for (k=0; k<=O-1; k++)
-      A[i][j] += B[i][k]*C[k][j];
-) @*/
-for (i=0; i<=M-1; i++)
- for (j=0; j<=N-1; j++)
-  for (k=0; k<=O-1; k++)
-    A[i][j] += B[i][k]*C[k][j];
-/*@ end @*/
-\end{verbatim}  
-\end{minipage}
-&
-\begin{minipage}{.45\textwidth}  
-\scriptsize
-\begin{verbatim}  
-def performance_params {
-  param Ui[] = range(1,33);
-  param Uj[] = range(1,33);
-  param Uk[] = range(1,33);
-  constraint reg_capacity = Ui*Uj+Ui*Uk+Uk*Uj<=32;
-}
-
-def input_params {
-  param M[] = [10,50,100,500,1000];
-  param N[] = [10,50,100,500,1000];
-  param O[] = [10,50,100,500,1000];
-  constraint square_matrices = (M==N) and (N==O);
-}
-
-\end{verbatim}  
-\end{minipage}
-\\
-%(a) Annotated code of matrix-matrix multiplication
-%& 
-%(b) Orio's parameter specifications\\
-\end{tabular}
-\caption{Example of specifying parameter constraints in Orio; annotated code for matrix-matrix multiplication (left) and constraint specification (right).}
-\label{fig:par-constraint}  
-\end{figure}
-
-Orio also supports parallel search when parallel resources are available,
-e.g., when tuning on the Blue Gene/P. The parallel Orio driver concurrently
-executes multiple independent code variants in the same parallel job. After
-Orio submits a parallel job, each node in the target machine executes a
-distinct generated code variant to collect the code performance. The
-resulting performance results are collected and stored in a temporary file,
-which is later processed by Orio to determine the best performing variant.
-Figure~\ref{fig:orio-example} has an example of using parallel search by
-specifying the number of nodes to use (per job) with the \texttt{num\_procs}
-variable. The remaining fields used by the parallel search are the
-\texttt{batch\_command} and the \texttt{status\_command} fields, which specify how Orio
-should submit a parallel job and query its status, respectively.
+Orio also supports parallel search when parallel resources are
+available, e.g., when tuning on the Blue Gene/P. The parallel Orio
+driver concurrently executes multiple independent code variants in the
+same parallel job. After Orio submits a parallel job, each node in the
+target machine executes a distinct generated code variant to collect
+the code performance. The resulting performance results are collected
+and stored in a temporary file, which is later processed by Orio to
+determine the best performing variant.  Figure~\ref{fig:orio-example}
+has an example of using parallel search by specifying the number of
+nodes to use (per job) with the \texttt{num\_procs} variable. The
+remaining fields used by the parallel search are the
+\texttt{batch\_command} and the \texttt{status\_command} fields, which
+specify how Orio should submit a parallel job and query its status,
+respectively.

### 2009-02-15 doc/papers/ipdps09/integration.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@443 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@
 the new module parses the performance parameters code to extract their
-values. In addition to the loop code to be optimized, the module
-parses performance parameters, which include tile sizes, unroll
-factors, loop order (permutation), as well as several boolean values
-for triggering OpenMP parallelization, scalar replacement, and
-auto-vectorization. Using these parameter values, Pluto then performs
-polyhedral transformations (e.g., two-level tiling and OpenMP
-parallelization) on the input loop code. The resulting generated code
-is subsequently passed to the widely-available profiling tool
-gprof~\cite{gprof} for hotspot detection. The statistical information
-produced by gprof provides accurate locations (line numbers) of the
-loop nests where the Pluto-generated code spends most of its execution
-time. The identified hotspot loop nests are then decorated with Orio's
-performance annotations for complementary syntactic transformations
-(e.g., loop permutation, loop unroll/jamming, scalar replacement, and
-explicit auto-vectorization). Finally, Orio optimizes the annotated
-hotspots as described in Section~\ref{sec:implm}.
+values. The extracted values of the performance parameters include
+tile sizes, unroll factors, loop order (permutation), as well as
+several boolean values for triggering OpenMP parallelization, scalar
+replacement, and auto-vectorization. Using these parameter values,
+Pluto then performs polyhedral transformations (e.g., two-level tiling
+and OpenMP parallelization) on the input loop code. The resulting
+generated code is subsequently passed to the widely-available
+profiling tool gprof~\cite{gprof} for hotspot detection. The
+statistical information produced by gprof provides accurate locations
+(line numbers) of the loop nests where the Pluto-generated code spends
+most of its execution time. The identified hotspot loop nests are then
+decorated with Orio's performance annotations for complementary
+syntactic transformations (e.g., loop permutation, loop
+unroll/jamming, scalar replacement, and explicit
+auto-vectorization). Finally, Orio optimizes the annotated hotspots as
+described in Section~\ref{sec:implm}.
 

### 2009-02-15 doc/papers/ipdps09/results.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@437 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@439 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@443 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@
 \begin{tabular}{c}
-\begin{minipage}[t]{.47\textwidth}
+\begin{minipage}[t]{.43\textwidth}
 \centering
 \vspace{0.2in}
-\includegraphics[width=1.0\textwidth]{figures/spmv.eps}  
+~~~~~~~~\includegraphics[width=1.0\textwidth]{figures/spmv.eps}  
 \end{minipage}
 \\
-{\footnotesize (a)}
+(a)
 \\
@@
 \\
-{\footnotesize (b)}
+(b)
 \\
@@
 \end{center}
-\caption{Performance of AXPY-4 operations on the Blue Gene/P.} 
+\caption{Performance of AXPY-4 operations on Blue Gene/P.} 
 \label{fig:axpy4-bgp-results} 
@@
 and temporal localities, resulting in inferior memory performance.
+
 %The sequential BLAS-based implementation perform better than the
@@
 
-The SpMV operation computes $\forall_{A_{i,j}} \neq 0:y_{i}
+The SpMV operation computes $\forall{A_{i,j}} \neq 0:y_{i}
 \leftarrow y_{i} + A_{i,j} \cdot x_{j}$, where $A$ is a sparse matrix,
@@
 
-The Blue Gene/P results in Figure~\ref{fig:ex27-bgp-results1} again show
-%Blue Gene/P was done for a varied number of nodes (i.e., 1, 8, and 32)
-%and the same three different parallel modes as before (i.e., SMP,
-%dual, and VN). 
-that empirical optimization using Orio produces the best performance for
-all cases. 
-%outperforming both the hand-tuned and the xlc-optimized codes. 
-On this architecture, the performance gap between the xlc-optimized and the
-hand-tuned codes is now larger than in the Intel experiment. Similarly, by
-exploiting thread-level parallelism, the Orio-optimized code performs better
-than the hand-tuned version when the number of threads per process increases
-and the number of nodes decreases. For the single-thread cases (VN mode),
-thread-level parallelism is not exploited; nevertheless, the Orio version
-still performs better than the other two choices.
+The Blue Gene/P results in Figure~\ref{fig:ex27-bgp-results1} again
+show that empirical optimization using Orio produces the best
+performance for all cases. On this architecture, the performance gap
+between the xlc-optimized and the hand-tuned codes is now larger than
+in the Intel experiment. Similarly, by exploiting thread-level
+parallelism, the Orio-optimized code performs better than the
+hand-tuned version when the number of threads per process increases
+and the number of nodes decreases. For the single-thread cases (VN
+mode), thread-level parallelism is not exploited; nevertheless, the
+Orio version still performs better than the other two implementations.
+
 %This validates the
@@
 In the following sections, we refer to the icc-optimized base code as
-xs``ICC,'' the Pluto-generated code with L1 tiling and
+``ICC,'' the Pluto-generated code with L1 tiling and
 unroll/jamming as ``Pluto (L1 tiling),'' and the Pluto-generated code with L1
 and L2 tilings and unroll/jamming as ``Pluto (L1+L2 tiling).'' The code
-tuned by Orio is referred to as ``Pluto+Orio.''
+tuned by Orio is referred to as ``Pluto+Orio''.
 
@@
 \end{center}
-\caption{2-D FDTD performance on an eight-core Intel Xeon.} 
+\caption{2-D FDTD performance on eight-core Intel Xeon.} 
 \label{fig:fdtd-2d-cookie-results} 
@@
 Figure~\ref{fig:seidel-cookie-seq} contains the sequential performance
-results for $T=500$, which show that applying Pluto's polyhedral tiling on
-the original code always delivers performance boosts, which range from 126\%
-to 142\%. When using Pluto, performing two-level tiling (for both L1 and L2
-caches) yields 13\% lower performance than performing only L1 tiling. 
-This is also reflected in the best sequential code found by Orio, where
-one-level tiling (for L1 cache only) is always performed. The sequential
-speedup obtained from tuning the Pluto code with Orio is significant, ranging
-from 51\% to 133\%.
+results for $T=500$, which show that applying Pluto's polyhedral
+tiling on the original code always delivers performance boosts, which
+range from 126\% to 142\%. When using Pluto, performing two-level
+tiling (for both L1 and L2 caches) yields 13\% lower performance than
+performing only L1 tiling. This is also reflected in the best
+sequential code found by Orio, where one-level tiling (for L1 cache
+only) is always found to be better than two-level tiling. The
+sequential speedup obtained from tuning the Pluto code with Orio is
+significant, ranging from 51\% to 133\%.
 
@@
 tiling over both L1 and L2 can result in worse performance than L1-only
-tiling because the number of tiles can be smaller than the number of cores.
+tiling because the number of tiles can be less than the number of cores.
 %Similar earlier trend also exists

### 2009-02-15 doc/papers/ipdps09/results.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@444 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@
 The parallel performance results for $T=500$ and $N=4000$ are shown in
-Figure~\ref{fig:seidel-cookie-par}. Again we observe that icc failed to
-parallelize the code, whereas Pluto was able to parallelize it and Orio
-improved the performance of the one-level tiled Pluto code further by up to
-128\% (and up to 548\% for the two-level tiled Pluto code). We observed that
+Figure~\ref{fig:seidel-cookie-par}. Again we observe that icc fails to
+parallelize the code, whereas Pluto is able to parallelize it and Orio
+improves the performance of the one-level tiled Pluto code further by up to
+128\% (and up to 548\% for the two-level tiled Pluto code). We observe that
 tiling over both L1 and L2 can result in worse performance than L1-only
@@
 performance, with slightly better scalability exhibited by the single-level tiled
-code. Orio further improved the performance of the Pluto-generated versions
+code. Orio further improves the performance of the Pluto-generated versions
 by a factor of 51\% to 120\%.

### 2009-02-15 doc/papers/iccs09/paper.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@446 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@
 ATAX & $A$,$\overrightarrow{x}$ & $\overrightarrow{y}$ & 
-$\overrightarrow{y} = A' * (A * \overrightarrow{x})$ \\ [1ex]
+$\overrightarrow{y} = A^T * (A * \overrightarrow{x})$ \\ [1ex]
 %---------------------------------------------------------------------------------
@@
 %& & & \\
-GEMVER & ${\bf A}$,$a$,$b$, &  ${\bf B}$,  & ~~${\bf B} = {\bf A} + \overrightarrow{u_1} * \overrightarrow{v_1}' + \overrightarrow{u_2} * \overrightarrow{v_2}'$\\
- & $\overrightarrow{u_1}$,$\overrightarrow{u_2}$,$\overrightarrow{v_1}$,$\overrightarrow{v_2}$, & $\overrightarrow{x}$,$\overrightarrow{w}$ & $\overrightarrow{x} = b * ({\bf B}' * \overrightarrow{y}) + \overrightarrow{z}$\\
+GEMVER & ${\bf A}$,$a$,$b$, &  ${\bf B}$,  & ~~${\bf B} = {\bf A} + \overrightarrow{u_1} * \overrightarrow{v_1}^T + \overrightarrow{u_2} * \overrightarrow{v_2}^T$\\
+ & $\overrightarrow{u_1}$,$\overrightarrow{u_2}$,$\overrightarrow{v_1}$,$\overrightarrow{v_2}$, & $\overrightarrow{x}$,$\overrightarrow{w}$ & $\overrightarrow{x} = b * ({\bf B}^T * \overrightarrow{y}) + \overrightarrow{z}$\\
 & $\overrightarrow{y}$,$\overrightarrow{z}$ &  & $\overrightarrow{w} = a * ({\bf B} * \overrightarrow{x})$\\ [1ex]
@@
 BiCG Kernel & ${\bf A}$,$\overrightarrow{p}$,$\overrightarrow{r}$ & $\overrightarrow{q}$,$\overrightarrow{s}$ & $\overrightarrow{q} = {\bf A} * \overrightarrow{p}$ \\ 
-& & & $\overrightarrow{s} = {\bf A}' * \overrightarrow{r}$\\
+& & & $\overrightarrow{s} = {\bf A}^T * \overrightarrow{r}$\\
 
@@
 
-The performance results (in MFLOP/s) of tuning the VADD operation on the Blue
+The performance results of tuning the VADD operation on the Blue
 Gene/P are given in Figure~\ref{fig:results}(a). The ``Base'' label
@@
 
-The Xeon performance results (in seconds) of ATAX are shown in
+The Xeon performance results of ATAX are shown in
 Figure~\ref{fig:results}(b). The Orio-tuned version that incorporates
@@
 
-Figure~\ref{fig:results}(c) shows the performance (in seconds) of the GEMVER
+Figure~\ref{fig:results}(c) shows the performance of the GEMVER
 operation on the Xeon workstation. Here we used the same Pluto and Orio
@@
 
-The performance (in seconds) for the BiCG kernel operation is shown in
+The performance for the BiCG kernel operation is shown in
 Figure~\ref{fig:results}(d). For this operation, the Pluto analysis did not

### 2009-02-16 doc/papers/ipdps09/implementation.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@451 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@
 %
-Finally, Orio is usable in real scientific applications without
+Finally, Orio is usable with real scientific applications without
 requiring reimplementation. This ensures that the significant investments in
@@
 optimized code version is executed and its performance evaluated.  After
-iteratively evaluating all code variants, the best-performing code is picked
+iteratively evaluating all code variants, the best-performing code is selected
 as the final output of Orio. Because the search space of all possible
@@
 Orio annotations are embedded into the source code as structured C comments
-that always start with \texttt{/*@} and end with \texttt{@*/}. For example,
+that start with \texttt{/*@} and end with \texttt{@*/}. For example,
 the annotation \texttt{/*@ end @*/} is used to indicate the end of an
@@
 consists of three main parts: leader annotation, annotation body, and trailer
-annotation. The annotation body can be either empty or contain C code that
+annotation. The annotation body can either be empty or contain C code that
 possibly includes other nested annotation regions. A leader annotation
@@
 Current optimizations supported by Orio span different types of code
-transformations that are not provided by production compilers in some
+transformation that are not provided by production compilers in some
 cases. Available optimizations include simple loop unrolling, memory

### 2009-02-16 doc/papers/ipdps09/results.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@451 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@
 here we discuss a limited number of computational kernels; more performance
-results for different computations are available at the Orio project
+data for different computations is available at the Orio project
 website~\cite{OrioURL}.
@@
 \leftarrow y_{i} + A_{i,j} \cdot x_{j}$, where $A$ is a sparse matrix,
-and $x$, $y$ are dense vectors. Each element of $A$ is used precisely once,
+and $x$, $y$ are dense vectors. Each element of $A$ is used only once,
 and element reuse is possible only for $x$ and $y$. Thus, to optimize SpMV,
@@
 optimizations using loop-level parallelism (through OpenMP) achieve much
-better performance than using MPI coarse-grained parallelism.
+better performance than using coarse-grained MPI parallelism.
 %Moreover, for the multithread cases, tuning using Orio determined that

### 2009-02-18 src/module/align/codegen.py
fixes to align module in preparation for enabling fortran code generation
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@452 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@ -9,22 +9,22 @@ import sys
 class CodeGen:
     '''The code generator for the Blue Gene's memory alignment optimizer'''
 
-    def __init__(self, vars, annot_body_code, indent):
+    def __init__(self, vars, annot_body_code, indent, language='c'):
         '''To instantiate a code generator instance'''
 
         self.generator = None
         
-        if language.lower() in ['c','c++','cxx']:
-            self.generator = CodeGen_C()
-        elif language.lower() in ['f', 'f90', 'fortran']:
-            self.generator = CodeGen_F()
+        if not language.strip(): language='c'
+	else: language = language.strip().lower()
+
+        if language in ['c','c++','cxx']:
+            self.generator = CodeGen_C(vars, annot_body_code, indent)
+        elif language in ['f', 'f90', 'fortran']:
+            self.generator = CodeGen_F(vars, annot_body_code, indent)
         else:
             print 'Error: Unknown language specified for code generation: %s' % language
             sys.exit(1)
-            
-        self.vars = vars
-        self.annot_body_code = annot_body_code
-        self.indent = indent
+
         pass
         
     #------------------------------------------------------
@@ -40,8 +40,10 @@ class CodeGen_C(CodeGen):
 
     def __init__(self, vars, annot_body_code, indent):
         '''To instantiate a code generator instance'''
-        
-        codegen.CodeGen(self, vars, annot_body_code, indent)
+        self.vars = vars
+        self.annot_body_code = annot_body_code
+        self.indent = indent
+        self.language = 'c'
         pass
 
     #------------------------------------------------------

### 2013-11-03 orio/main/cmd_line_opts.py
Merge orio/main/cmd_line_opts.py

@@ -16,13 +16,9 @@ Usage: %s [options] <ifile>
 
 Options:
   -c, --pre-command=<string>     Command string with which to prefix the execution of the 
-<<<<<<< HEAD
                                  Orio-built code, e.g., tau_exec
   --post-command=<string>        Command string to run after each execution of Orio-built code,
                                  e.g., taudb_loadtrial
-=======
-                                 Orio-built code, e.g., tauex
->>>>>>> 6216c0a019c5afe9ed29f3c14c81efcc224f37ef
   -e, --erase-annot              remove annotations from the output
   -h, --help                     display this message
   -o <file>, --output=<file>     place the output in <file> (only valid when processing 

### 2013-11-03 doc/papers/iccs09/latex/typeinst.tex

%%%%%%%%%%%%%%%%%%%%%%% file typeinst.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is the LaTeX source for the instructions to authors using
% the LaTeX document class 'llncs.cls' for contributions to
% the Lecture Notes in Computer Sciences series.
% http://www.springer.com/lncs       Springer Heidelberg 2006/05/04
%
% It may be used as a template for your own input - copy it
% to a new file with a new name and use it as the basis
% for your article.
%
% NB: the document class 'llncs' has its own and detailed documentation, see
% ftp://ftp.springer.de/data/pubftp/pub/tex/latex/llncs/latex2e/llncsdoc.pdf
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[runningheads]{llncs}

\usepackage{amssymb}
\setcounter{tocdepth}{3}
\usepackage{graphicx}

\usepackage{url}
\urldef{\mailsa}\path|{alfred.hofmann,ursula.barth,ingrid.beyer,natalie.brecht,|
\urldef{\mailsb}\path|christine.guenther,frank.holzwarth,piamaria.karbach,|
\urldef{\mailsc}\path|anna.kramer,erika.siebert-cole,lncs}@springer.com|
\newcommand{\keywords}[1]{\par\addvspace\baselineskip
\noindent\keywordname\enspace\ignorespaces#1}

\begin{document}

\mainmatter  % start of an individual contribution

% first the title is needed
\title{Lecture Notes in Computer Science:\\Authors' Instructions
for the Preparation\\of Camera-Ready
Contributions\\to LNCS/LNAI/LNBI Proceedings}

% a short form should be given in case it is too long for the running head
\titlerunning{Lecture Notes in Computer Science: Authors' Instructions}

% the name(s) of the author(s) follow(s) next
%
% NB: Chinese authors should write their first names(s) in front of
% their surnames. This ensures that the names appear correctly in
% the running heads and the author index.
%
\author{Alfred Hofmann%
\thanks{Please note that the LNCS Editorial assumes that all authors have used
the western naming convention, with given names preceding surnames. This determines
the structure of the names in the running heads and the author index.}%
\and Ursula Barth\and Ingrid Beyer\and Natalie Brecht\and\\
Christine G\"{u}nther\and Frank Holzwarth\and Pia Maria Karbach \and\\
Anna Kramer\and Erika
Siebert-Cole}
%
\authorrunning{Lecture Notes in Computer Science: Authors' Instructions}
% (feature abused for this document to repeat the title also on left hand pages)

% the affiliations are given next; don't give your e-mail address
% unless you accept that it will be published
\institute{Springer-Verlag, Computer Science Editorial,\\
Tiergartenstr. 17, 69121 Heidelberg, Germany\\
\mailsa\\
\mailsb\\
\mailsc\\
\url{http://www.springer.com/lncs}}

%
% NB: a more complex sample for affiliations and the mapping to the
% corresponding authors can be found in the file "llncs.dem"
% (search for the string "\mainmatter" where a contribution starts).
% "llncs.dem" accompanies the document class "llncs.cls".
%

\toctitle{Lecture Notes in Computer Science}
\tocauthor{Authors' Instructions}
\maketitle

\begin{abstract}
The abstract should summarize the contents of the paper and should
contain at least 70 and at most 150 words. It should be written using the
\emph{abstract} environment.
\keywords{We would like to encourage you to list your keywords within
the abstract section}
\end{abstract}

\section{Introduction}

You are strongly encouraged to use \LaTeXe{} for the
preparation of your camera-ready manuscript together with the
corresponding Springer class file \verb+llncs.cls+. Only if you use
\LaTeXe{} can hyperlinks be generated in the online version
of your manuscript.

The \LaTeX{} source of this instruction file for \LaTeX{} users may be used as
a template. This is located in the ``authors'' subdirectory in
\url{ftp://ftp.springer.de/pub/tex/latex/llncs/latex2e/instruct/} and
entitled \texttt{typeinst.tex}. Kindly
send the final and checked source and PDF files of your paper to the
Contact Volume Editor. This is usually one of the organizers of the
conference. You should make sure that the \LaTeX{} and the PDF files are
identical and correct and that only one version of your paper is sent.
It is not possible to update files at a later stage. Please note that we
do not need the printed paper.

We would like to draw your attention to the fact that it is not possible
to modify a paper in any way, once it has been published. This applies 
to both the printed book and the online version of the publication. 
Every detail, including the order of the names of the authors, should 
be checked before the paper is sent to the Volume Editors.

\subsection{Checking the PDF File}

Kindly assure that the Contact Volume Editor is given the name and email
address of the contact author for your paper. The Contact Volume Editor
uses these details to compile a list for our production department at
SPS in India. Once the files have been worked upon, SPS sends a copy of
the final pdf of each paper to its contact author. The contact author is
asked to check through the final pdf to make sure that no errors have
crept in during the transfer or preparation of the files. This should
not be seen as an opportunity to update or copyedit the papers, which is
not possible due to time constraints. Only errors introduced during the
preparation of the files will be corrected.

This round of checking takes place about two weeks after the files have
been sent to the Editorial by the Contact Volume Editor, i.e., roughly
seven weeks before the start of the conference for conference
proceedings, or seven weeks before the volume leaves the printer's, for
post-proceedings. If SPS does not receive a reply from a particular
contact author, within the timeframe given, then it is presumed that the
author has found no errors in the paper. The tight publication schedule
of LNCS does not allow SPS to send reminders or search for alternative
email addresses on the Internet.

In some cases, it is the Contact Volume Editor that checks all the final
pdfs. In such cases, the authors are not involved in the checking phase.

\subsection{Additional Information Required by the Volume Editor}

If you have more than one surname, please make sure that the Volume Editor
knows how you are to be listed in the author index.

\subsection{Copyright Forms}

The copyright form may be downloaded from the ``For Authors"
(Information for LNCS Authors)
section of the LNCS Website: \texttt{www.springer.com/lncs}.
Please send your signed copyright
form to the Contact Volume Editor, either as a scanned pdf or by fax or
by courier. One author may sign on behalf of all of the other authors of a particular
paper. Digital signatures are acceptable.

\section{Paper Preparation}

Springer
provides you with a complete integrated \LaTeX{} document class (\texttt{llncs.cls})
for multi-author books such as those in the LNCS series.
Papers not complying with the LNCS style will be reformatted. This can
lead to an increase in the overall number of pages. We would therefore
urge you not to squash your paper.

Please always cancel any superfluous definitions that are
not actually used in your text. If you do not, these may conflict with
the definitions of the macro package, causing changes in the structure
of the text and leading to numerous mistakes in the proofs.

If you wonder what \LaTeX{} is and where it can be obtained, see the
``\textit{LaTeX project site}'' (\url{http://www.latex-project.org})
and especially the webpage ``\textit{How to get it}''
(\url{http://www.latex-project.org/ftp.html}) respectively.

When you use \LaTeX\ together with our document class file,
\texttt{llncs.cls},
your text is typeset automatically in Computer Modern Roman (CM) fonts.
Please do
\emph{not} change the preset fonts. If you have to use fonts other
than the preset fonts, kindly submit these with your files.

Please use the commands \verb+\label+ and \verb+\ref+ for
cross-references and the commands \verb+\bibitem+ and \verb+\cite+ for
references to the bibliography, to enable us to create hyperlinks at
these places.

For preparing your figures electronically and integrating them into
your source file we recommend using the standard \LaTeX{} \verb+graphics+ or
\verb+graphicx+ package. These provide the \verb+\includegraphics+ command.
In general, please refrain from using the \verb+\special+ command.

Remember to submit any further style files and
fonts you have used together with your source files.

\subsubsection{Headings.}

Headings should be capitalized
(i.e., nouns, verbs, and all other words
except articles, prepositions, and conjunctions should be set with an
initial capital) and should,
with the exception of the title, be aligned to the left.
Words joined by a hyphen are subject to a special rule. If the first
word can stand alone, the second word should be capitalized.

Here are some examples of headings: ``Criteria to Disprove
Context-Freeness of Collage Language", ``On Correcting the Intrusion of
Tracing Non-deterministic Programs by Software", ``A User-Friendly and
Extendable Data Distribution System", ``Multi-flip Networks:
Parallelizing GenSAT", ``Self-determinations of Man".

\subsubsection{Lemmas, Propositions, and Theorems.}

The numbers accorded to lemmas, propositions, and theorems, etc. should
appear in consecutive order, starting with Lemma 1, and not, for
example, with Lemma 11.

\subsection{Figures}

For \LaTeX\ users, we recommend using the \emph{graphics} or \emph{graphicx}
package and the \verb+\includegraphics+ command.

Please check that the lines in line drawings are not
interrupted and are of a constant width. Grids and details within the
figures must be clearly legible and may not be written one on top of
the other. Line drawings should have a resolution of at least 800 dpi
(preferably 1200 dpi). The lettering in figures should have a height of
2~mm (10-point type). Figures should be numbered and should have a
caption which should always be positioned \emph{under} the figures, in
contrast to the caption belonging to a table, which should always appear
\emph{above} the table; this is simply achieved as matter of sequence in
your source.

Please center the figures or your tabular material by using the \verb+\centering+
declaration. Short captions are centered by default between the margins
and typeset in 9-point type (Fig.~\ref{fig:example} shows an example).
The distance between text and figure is preset to be about 8~mm, the
distance between figure and caption about 6~mm.

To ensure that the reproduction of your illustrations is of a reasonable
quality, we advise against the use of shading. The contrast should be as
pronounced as possible.

If screenshots are necessary, please make sure that you are happy with
the print quality before you send the files.
\begin{figure}
\centering
\includegraphics[height=6.2cm]{eijkel2}
\caption{One kernel at $x_s$ (\emph{dotted kernel}) or two kernels at
$x_i$ and $x_j$ (\textit{left and right}) lead to the same summed estimate
at $x_s$. This shows a figure consisting of different types of
lines. Elements of the figure described in the caption should be set in
italics, in parentheses, as shown in this sample caption.}
\label{fig:example}
\end{figure}

Please define figures (and tables) as floating objects. Please avoid
using optional location parameters like ``\verb+[h]+" for ``here".

\paragraph{Remark 1.}

In the printed volumes, illustrations are generally black and white
(halftones), and only in exceptional cases, and if the author is
prepared to cover the extra cost for color reproduction, are colored
pictures accepted. Colored pictures are welcome in the electronic
version free of charge. If you send colored figures that are to be
printed in black and white, please make sure that they really are
legible in black and white. Some colors as well as the contrast of
converted colors show up very poorly when printed in black and white.

\subsection{Formulas}

Displayed equations or formulas are centered and set on a separate
line (with an extra line or halfline space above and below). Displayed
expressions should be numbered for reference. The numbers should be
consecutive within each section or within the contribution,
with numbers enclosed in parentheses and set on the right margin --
which is the default if you use the \emph{equation} environment, e.g.,
\begin{equation}
  \psi (u) = \int_{o}^{T} \left[\frac{1}{2}
  \left(\Lambda_{o}^{-1} u,u\right) + N^{\ast} (-u)\right] dt \;  .
\end{equation}

Equations should be punctuated in the same way as ordinary
text but with a small space before the end punctuation mark.

\subsection{Footnotes}

The superscript numeral used to refer to a footnote appears in the text
either directly after the word to be discussed or -- in relation to a
phrase or a sentence -- following the punctuation sign (comma,
semicolon, or period). Footnotes should appear at the bottom of
the
normal text area, with a line of about 2~cm set
immediately above them.\footnote{The footnote numeral is set flush left
and the text follows with the usual word spacing.}

\subsection{Program Code}

Program listings or program commands in the text are normally set in
typewriter font, e.g., CMTT10 or Courier.

\medskip

\noindent
{\it Example of a Computer Program}
\begin{verbatim}
program Inflation (Output)
  {Assuming annual inflation rates of 7%, 8%, and 10%,...
   years};
   const
     MaxYears = 10;
   var
     Year: 0..MaxYears;
     Factor1, Factor2, Factor3: Real;
   begin
     Year := 0;
     Factor1 := 1.0; Factor2 := 1.0; Factor3 := 1.0;
     WriteLn('Year  7% 8% 10%'); WriteLn;
     repeat
       Year := Year + 1;
       Factor1 := Factor1 * 1.07;
       Factor2 := Factor2 * 1.08;
       Factor3 := Factor3 * 1.10;
       WriteLn(Year:5,Factor1:7:3,Factor2:7:3,Factor3:7:3)
     until Year = MaxYears
end.
\end{verbatim}
%
\noindent
{\small (Example from Jensen K., Wirth N. (1991) Pascal user manual and
report. Springer, New York)}

\subsection{Citations}

For citations in the text please use
square brackets and consecutive numbers: \cite{jour}, \cite{lncschap},
\cite{proceeding1} -- provided automatically
by \LaTeX 's \verb|\cite| \dots\verb|\bibitem| mechanism.

\subsection{Page Numbering and Running Heads}

There is no need to include page numbers. If your paper title is too
long to serve as a running head, it will be shortened. Your suggestion
as to how to shorten it would be most welcome.

\section{LNCS Online}

The online version of the volume will be available in LNCS Online.
Members of institutes subscribing to the Lecture Notes in Computer
Science series have access to all the pdfs of all the online
publications. Non-subscribers can only read as far as the abstracts. If
they try to go beyond this point, they are automatically asked, whether
they would like to order the pdf, and are given instructions as to how
to do so.

Please note that, if your email address is given in your paper,
it will also be included in the meta data of the online version.

\section{BibTeX Entries}

The correct BibTeX entries for the Lecture Notes in Computer Science
volumes can be found at the following Website shortly after the
publication of the book:
\url{http://www.informatik.uni-trier.de/~ley/db/journals/lncs.html}

\subsubsection*{Acknowledgments.} The heading should be treated as a
subsubsection heading and should not be assigned a number.

\section{The References Section}\label{references}

In order to permit cross referencing within LNCS-Online, and eventually
between different publishers and their online databases, LNCS will,
from now on, be standardizing the format of the references. This new
feature will increase the visibility of publications and facilitate
academic research considerably. Please base your references on the
examples below. References that don't adhere to this style will be
reformatted by Springer. You should therefore check your references
thoroughly when you receive the final pdf of your paper.
The reference section must be complete. You may not omit references.
Instructions as to where to find a fuller version of the references are
not permissible.

We only accept references written using the latin alphabet. If the title 
of the book you are referring to is in Russian or Chinese, then please write 
(in Russian) or (in Chinese) at the end of the transcript or translation 
of the title.

The following section shows a sample reference list with entries for
journal articles \cite{jour}, an LNCS chapter \cite{lncschap}, a book
\cite{book}, proceedings without editors \cite{proceeding1} and
\cite{proceeding2}, as well as a URL \cite{url}.
Please note that proceedings published in LNCS are not cited with their
full titles, but with their acronyms!

\begin{thebibliography}{4}

\bibitem{jour} Smith, T.F., Waterman, M.S.: Identification of Common Molecular
Subsequences. J. Mol. Biol. 147, 195--197 (1981)

\bibitem{lncschap} May, P., Ehrlich, H.C., Steinke, T.: ZIB Structure Prediction Pipeline:
Composing a Complex Biological Workflow through Web Services. In: Nagel,
W.E., Walter, W.V., Lehner, W. (eds.) Euro-Par 2006. LNCS, vol. 4128,
pp. 1148--1158. Springer, Heidelberg (2006)

\bibitem{book} Foster, I., Kesselman, C.: The Grid: Blueprint for a New Computing
Infrastructure. Morgan Kaufmann, San Francisco (1999)

\bibitem{proceeding1} Czajkowski, K., Fitzgerald, S., Foster, I., Kesselman, C.: Grid
Information Services for Distributed Resource Sharing. In: 10th IEEE
International Symposium on High Performance Distributed Computing, pp.
181--184. IEEE Press, New York (2001)

\bibitem{proceeding2} Foster, I., Kesselman, C., Nick, J., Tuecke, S.: The Physiology of the
Grid: an Open Grid Services Architecture for Distributed Systems
Integration. Technical report, Global Grid Forum (2002)

\bibitem{url} National Center for Biotechnology Information, \url{http://www.ncbi.nlm.nih.gov}

\end{thebibliography}

\section*{Appendix: Springer-Author Discount}

LNCS authors are entitled to a 33.3\% discount off all Springer
publications. Before placing an order, the author should send an email
to \url{SDC.bookorder@springer.com}, giving full
details of his or her
Springer publication, to obtain a so-called token. This token is a
number, which must be entered when placing an order via the Internet, in
order to obtain the discount.

\section{Checklist of Items to be Sent to Volume Editors}
Here is a checklist of everything the volume editor requires from you:

\begin{itemize}
\settowidth{\leftmargin}{{\Large$\square$}}\advance\leftmargin\labelsep
\itemsep8pt\relax
\renewcommand\labelitemi{{\lower1.5pt\hbox{\Large$\square$}}}

\item The final \LaTeX{} source files
\item A final PDF file
\item A copyright form, signed by one author on behalf of all of the
authors of the paper.
\item A readme giving the name and email address of the
corresponding author.
\end{itemize}
\end{document}

### 2013-11-03 doc/papers/iccs09/paper.tex

@@
 BLAS operation, we follow a three-step approach, illustrated in
-Figure~\ref{fig:process}.  This process is repeated when the MATLAB code changes
+Fig.~\ref{fig:process}.  This process is repeated when the MATLAB code changes
 or the code must be tuned on a new architecture.
@@
 array copy optimization, and some architecture-dependent optimizations. The
-right-hand side of Figure~\ref{fig:orio-example} shows separate tuning
+right-hand side of Fig.~\ref{fig:orio-example} shows separate tuning
 specifications used for building and running executable tests, including
@@
 quad-core E5462 Xeon processors (8 cores total) running at 2.8 GHz
-(1600 MHz FSB) with 2 GB RAM, running Ubuntu 8.04. Intel C compiler
+(1600 MHz FSB) with 16 GB RAM, running Ubuntu 8.04. Intel C compiler
 (v10.1) was used with \texttt{-O3} option (and
@@
 The performance results of tuning the VADD operation on the Blue
-Gene/P are given in Figure~\ref{fig:results}(a). The ``Base'' label
+Gene/P are given in Fig.~\ref{fig:results}(a). The ``Base'' label
 designates the C implementation generated by the MATLAB compiler. We also
@@
 Finally, we tuned the simple C loop version using Orio, with the performance
-annotations previously shown in Figure~\ref{fig:orio-example}. In this
+annotations previously shown in Fig.~\ref{fig:orio-example}. In this
 experiment, we measured the performance for both the sequential and parallel
@@
 The Xeon performance results of ATAX are shown in
-Figure~\ref{fig:results}(b). The Orio-tuned version that incorporates
+Fig.~\ref{fig:results}(b). The Orio-tuned version that incorporates
 Pluto-generated loop fusion optimizations and Orio parallelization
@@
 The performance for the BiCG kernel operation is shown in
-Figure~\ref{fig:results}(d). For this operation, the Pluto analysis did not
+Fig.~\ref{fig:results}(d). For this operation, the Pluto analysis did not
 result in performance improvement. Thus we are showing the results obtained

### 2013-11-03 doc/papers/ipdps09/abstract.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@427 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@
-In many scientific applications, significant time is spent in tuning codes for a particular high-performance architecture.
-Tuning approaches range from the relatively nonintrusive (e.g., by using compiler options) to extensive code modifications that attempt to exploit specific architecture features.
-Intrusive techniques often result in code changes that are not easily reversible, which can negatively impact readability, maintainability, and performance on different architectures.
-We introduce an extensible annotation-based empirical tuning system called Orio, which is aimed at improving both performance and productivity by enabling software developers to insert annotations in the form of structured comments into their source code that trigger a number of low-level performance optimizations on a specified code fragment.
-To maximize the performance tuning opportunities, the annotation processing infrastructure is designed to support both architecture-independent and architecture-specific code optimizations.
-Given the annotated code as input, Orio generates many tuned versions of the same operation and empirically evaluates the alternatives to select the best performing version for production use.
-We have also enabled the use of the PLuTo automatic parallelization tool in conjunction with Orio to generate efficient OpenMP-based parallel code.
-We describe our experimental results involving a number of computational kernels, including dense array and sparse matrix operations.
+For many scientific applications, significant time is spent in tuning
+codes for a particular high-performance architecture. Tuning
+approaches range from the relatively nonintrusive (e.g., by using
+compiler options) to extensive code modifications that attempt to
+exploit specific architecture features. Intrusive techniques often
+result in code changes that are not easily reversible, and can
+negatively impact readability, maintainability, and performance on
+different architectures. We introduce an extensible annotation-based
+empirical tuning system called Orio that is aimed at improving both
+performance and productivity. It allows software developers to insert
+annotations in the form of structured comments into their source code
+to trigger a number of low-level performance optimizations on a
+specified code fragment.  To maximize the performance tuning
+opportunities, the annotation processing infrastructure is designed to
+support both architecture-independent and architecture-specific code
+optimizations. Given the annotated code as input, Orio generates many
+tuned versions of the same operation and empirically evaluates the
+alternatives to select the best performing version for production
+use. We have also enabled the use of the Pluto automatic
+parallelization tool in conjunction with Orio to generate efficient
+OpenMP-based parallel code. We describe our experimental results
+involving a number of computational kernels, including dense array and
+sparse matrix operations.

### 2013-11-05 javascripts/main.js
images

@@ -1,1 +1,25 @@
 console.log('This would be the main JS file.');
+
+// JavaScript functions to show and hide drop-down menus.
+// In SimpleNavBar.html we call ShowMenuDiv each time the mouse goes over 
+// either the menu title or the submenu itself, and call HideMenuDiv when the
+// mouse goes out of the menu title or the submenu iteslf (onMouseOut).
+
+function ShowItem (itemID) {
+  var x = document.getElementById(itemID);
+  if (x)
+    x.style.visibility = "visible";
+  return true;
+}
+
+function HideItem (itemID) { 
+  var x = document.getElementById(itemID);
+  if (x)
+     x.style.visibility = "hidden";
+  return true;
+}
+
+//    As noted in the css file, using x.style.visibility as
+//    seen below seemed to have better cross browser support than using 
+//    x.style.display="block" and x.style.display="none" to show and hide 
+//    the menu.

### 2013-11-05 doc/papers/ipdps09/integration.tex

@@
 annotation body block, while the performance parameters code is
-written in the module body block. A simple parser implemented inside
+written in the module body block. A parser implemented inside
 the new module parses the performance parameters code to extract their
@@
 
-Pluto also performs syntactic loop unroll/jamming as a post-processing
+Pluto also performs syntactic loop unroll/jam in a post-processing
 pass. The target loops, however, are limited only to innermost loops with a
 maximum depth of two (i.e., 1-D unrolling and 2-D unroll/jamming). So we
-choose to use the loop unroll/jamming transformation already available in
+choose to use the loop unroll/jam transformation already available in
 Orio, which is more flexible because it can be applied to loop nests of
@@
 
-At present Orio does not employ any data dependency analysis when performing
+At present Orio does not employ any data dependence analysis when performing
 syntactic transformations. Therefore, there is no guarantee that the code

### 2013-11-05 doc/papers/ipdps09/related.tex

@@
 over implementation variants.  GotoBLAS~\cite{Goto:2006fk,Goto:fk}, on
-the other hand, achieves great performance results on several
+the other hand, achieves near-peak performance on several
 architectures by using hand-tuned data structures and kernel
@@
 computational scientists. The complexity of existing annotation languages and lack
-of common syntaxes for transformations (e.g., loop unrolling) result
+of common syntax for transformations (e.g., loop unrolling) result
 in steep learning curves and the inability to take advantage of more than one
